---
title: "gent_clean_11.16.24"
author: "Joyce Hu"
date: "2024-11-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages (JH - check if any are unused after code cleaned)
```{r}
library(tidyverse)
library(tigris)
library(arrow)
library(here)
library(mgcv)
library(fixest)
library(lmtest)
library(terra)
library(exactextractr)
library(tidyterra)
library(sf)
library(mgcViz)

library(DHARMa)
library(qqplotr)
library(readxl) # for read_excel()
library(spdep) # for moran.test()
library(tidycensus) # for median income
library(distr) # for qqbounds()
library(spatialreg) # for spatial lag and error models
library(gridExtra) # for viewing plots side-by-side
library(qgam) 
library(dplyr)
```

Load and clean data
```{r}
# Check if 'data' directory exists; if not, create it
if (!dir.exists(here("data"))) {
  dir.create(here("data"))
}

# load gentrification score by tract
gentrification <- read.csv(here("data", "Data_NYC_Gentrification_2000_16.csv")) %>% 
  mutate(tractid = as.character(tractid)) %>% 
  dplyr::select(tractid, score_0.5)

# load gentrification score if using xlsx file:
# gentrification <- read_xlsx(here("Gentrification","Data_NYC_Gentrification_2000_16.xlsx")) %>%
#   mutate(tractid = as.character(tractid)) %>%
#   dplyr::select(tractid, score_0.5)

# pull NYC tract IDs/fips from gentrification index dataset for filtering
nyc_fips <- gentrification$tractid

# load NYC tracts shapefile
nyc_tracts <- tracts(state = "NY", county = c("Bronx", "Kings", "New York", "Queens", "Richmond"), year = 2010) %>%
  # clean lat and long for Conley SEs
  mutate(lat = as.numeric(INTPTLAT10), long = as.numeric(INTPTLON10), GEOID = GEOID10) %>% 
  # filter for FIPS codes included in gentrification index dataset
  filter(GEOID %in% nyc_fips)


#load NY tracts populatuion weighted centroid points from Census data
nyc_weighted_centroids <- read.table("https://www2.census.gov/geo/docs/reference/cenpop2010/tract/CenPop2010_Mean_TR36.txt", 
                         header = TRUE, sep = ",", stringsAsFactors = FALSE, 
                         colClasses = c("character", "character", "character", "numeric", "numeric", "numeric")) %>%
  mutate(FIPS = paste(STATEFP, COUNTYFP, TRACTCE, sep = "")) %>% 
  ## filter for only NYC tracts
  filter(FIPS %in% nyc_fips) %>%  
  rename(Longitude_weighted = LONGITUDE, 
         Latitude_weighted = LATITUDE)

# JH - 10/12/24 warning persists 10/13, 10/14 Warning: R metadata may have unsafe or invalid elements ℹ Type: "externalptr" options(arrow.unsafe_metadata = TRUE) #

# temperature data
if(!file.exists(here("data","summarized_daily_temp_preds.parquet"))){download.file("https://zenodo.org/records/10557980/files/summarized_daily_temp_preds.parquet?download=1", 
                                                                                   destfile = here("data","summarized_daily_temp_preds.parquet"))}

# temperature by tract
temp <- read_parquet(here("data","summarized_daily_temp_preds.parquet")) %>%
  filter(str_detect(date, "2016")) %>%
  mutate(temp_c = mean_temp_daily_cK/100-273.15) %>% 
  mutate(temp_f = temp_c*9/5+32) %>% 
  group_by(GEOID, date) %>%
  ungroup() %>% 
  group_by(GEOID) %>% 
  summarize(mean_temp = mean(mean_temp_daily_cK/100))

# ndvi by tract (need to have it downloaded and stored locally)
ndvi_2016 <- rast(here("data", "ndvi_c2_2016_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

ndvi_2000 <- rast(here("data", "ndvi_c2_2000_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

# PM2.5 data
# download data from URL
download_data <- function() {
  # set longer timeout
  options(timeout = 7200)
  if(!file.exists(here("data", "2016_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_pm25_daily_average.txt.gz", destfile = here("data", "2016_pm25_daily_average.txt.gz"))} #2016
  if(!file.exists(here("data", "2002_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_pm25_daily_average.txt.gz", destfile = here("data", "2002_pm25_daily_average.txt.gz"))}} #2002
download_data()

# read in PM2.5 daily average by census tract from EPA FAQSD - 2002
pm_tract_2002 <- read.delim(gzfile(here("data", "2002_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## Include only NYC tracts
  filter(FIPS %in% nyc_fips) %>%
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## Calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(pm_mean_2002 = mean(pm25_daily_average.ug.m3.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# read in PM2.5 daily average by census tract from EPA FAQSD - 2016
pm_tract_2016 <- read.delim(gzfile(here("data", "2016_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(pm_mean_2016 = mean(Prediction), 
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))
  

# combine 2002 and 2016 pm data for nyc_tracts
pm_tract <- pm_tract_2002 %>%
  inner_join(pm_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# NO2 CACES
no2_2016_caces <- read.csv(here("data", "2016_no2_caces.csv")) %>% 
  mutate(no2_2016 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_2000_caces <- read.csv(here("data", "2000_no2_caces.csv")) %>% 
  mutate(no2_2000 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_caces <- left_join(no2_2000_caces, no2_2016_caces, by = "fips") %>% 
  select(no2_2000, no2_2016, fips)

#O3 data download
options(timeout = 7200)
if(!file.exists(here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))} #2016
if(!file.exists(here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))} #2002

# read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2002
o3_tract_2002 <- read.delim(gzfile(here("data", "2002_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>% 
  ## include only NYC tracts
  filter(FIPS %in% nyc_fips) %>% 
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(o3_mean_2002 = mean(ozone_daily_8hour_maximum.ppb.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2016
o3_tract_2016 <- read.delim(gzfile(here("data", "2016_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>% 
  ## calculate annual average by tract using  population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(o3_mean_2016 = mean(Prediction),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))


o3_tract <- o3_tract_2002 %>%
  left_join(o3_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# merge all exposure data
all_data <- nyc_tracts %>% 
  # join gentrification index by tract
  right_join(gentrification, by = c("GEOID10" = "tractid")) %>%
  left_join(pm_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(pm_2016 = pm_mean_2016, pm_2002 = pm_mean_2002, pm_change = pm_2016-pm_2002) %>% 
  left_join(o3_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(o3_2016 = o3_mean_2016, o3_2002 = o3_mean_2002, o3_change = o3_2016-o3_2002) %>% 
  left_join(no2_caces, by = c("GEOID" = "fips")) %>% 
  mutate(no2_change = no2_2016-no2_2000) %>% 
  mutate(ndvi_2016 = ndvi_2016, ndvi_2000 = ndvi_2000, ndvi_change = ndvi_2016-ndvi_2000) %>% 
  left_join(temp, by = "GEOID")

#Find median household income variable in 2000 decennial census
all_vars_dec <- 
  load_variables(year = 2000, dataset = "sf3")

# models adjusting for median income in 2000 (note: 2000 decennial census is 1999 data)
income_2000 <- get_decennial(variables = "P053001", year = 2000,  sumfile = "sf3", state = "NY", 
                  county = c("Queens", "King", "New York", "Richmond", "Bronx"), geography = "tract") %>% 
  rename(med_income_2000 = value)
#results in 226 missing tracts in terms of 2010 tracts.

#Find median household income variable in 2010 ACS
income_2010 <- get_acs(variables = "B06011_001", year = 2010, state = "NY", 
                  county = c("Queens", "King", "New York", "Richmond", "Bronx"), geography = "tract") %>% 
  rename(med_income_2010 = estimate)

# Joining both income sets with data and imputing 2010 data where 2000 data is missing
all_data_inc <- all_data %>%
  left_join(income_2000, by = "GEOID") %>%
  left_join(income_2010, by = "GEOID") %>%
  mutate(med_income_2000_impute = coalesce(med_income_2000, med_income_2010)) %>% 
  # coerce GEOID to numeric then factor class to avoid error in MRF smooth
  mutate(GEOID = as.numeric(GEOID)) %>% 
  mutate(GEOID = factor(GEOID)) %>% 
 # coerce COUNTYFP10 to numeric then factor class to avoid error ?
  mutate(COUNTYFP10 = as.numeric(COUNTYFP10)) %>% 
  mutate(COUNTYFP10 = factor(COUNTYFP10))

#ranking median income for use in GAM models
all_data_inc$med_income_ranked <- rank(all_data_inc$med_income_2000_impute)

# #temporary df using med_income_2000, removing missing tracts
# all_data_inc_na <- na.omit(all_data_inc)

############# models adjusting for median income in 2000 and distance to downtown #########
# load nyc roads
downtown_st <- roads(state = "NY", county = "New York", year = 2011)
ggplot() + geom_sf(data = downtown_st)

# Define bounding box
bbox_coords <- matrix(c(
  -74.0187, 40.70158,  # Southwest
  -74.0187, 40.77279,  # Northwest
  -73.95953, 40.77279, # Northeast
  -73.95953, 40.70158, # Southeast
  -74.0187, 40.70158   # Closing the polygon
), ncol = 2, byrow = TRUE)

# convert bounding box to sf
bbox_polygon <- st_sfc(st_polygon(list(bbox_coords)), crs = 4326)

# Ensure roads are the same CRS and filter for streets within bbox
downtown_st <- st_transform(downtown_st, crs = st_crs(bbox_polygon))
filtered_streets <- downtown_st[st_within(downtown_st, bbox_polygon, sparse = FALSE), ]

# Plot filtered downtown streets and bounding box
ggplot() +
  geom_sf(data = filtered_streets) +
  geom_sf(data = bbox_polygon, fill = NA, color = 'red') +
  ggtitle("Downtown Streets within Bounding Box")

# Ensure census tracts are in the same CRS
nyc_tracts <- st_transform(nyc_tracts, crs = st_crs(bbox_polygon))

# Filter intersecting tracts within downtown
intersecting_tracts <- st_intersection(nyc_tracts, bbox_polygon)
combined_area <- st_union(intersecting_tracts)
#get downtown centroid
area_centroid <- st_centroid(combined_area)

ggplot() +
  geom_sf(data = intersecting_tracts, fill = 'lightblue', color = 'black') +
  geom_sf(data = area_centroid, color = 'red', size = 3) +
  ggtitle('Centroid of Downtown Census Tracts')

# Calculate distances and add to "all_data_inc" dataframe
distances <- st_distance(nyc_tracts, area_centroid)
all_data_inc$dt_dist <- as.numeric(distances)
```

flood code (Simplify)
```{r}
# Load necessary libraries
library(dplyr)
library(stringr)
library(here)

# 1. Read in the flood data
flood <- read.csv(here("data", "tracts_flood.csv"))

# 2. Ensure COUNTYFP10 is a character column in all_data_inc
all_data_inc <- all_data_inc %>%
  mutate(COUNTYFP10 = as.character(COUNTYFP10))

# 3. Prepare the borough lookup table (boro_lookup)
boro_lookup <- tibble(
  county_fips = c("61", "5", "47", "81", "85"),
  boro_code = c("1", "2", "3", "4", "5")
)

# 4. Check the unique values in COUNTYFP10 and county_fips
print("Unique COUNTYFP10 values in all_data_inc:")
print(unique(all_data_inc$COUNTYFP10))

print("Unique county_fips values in boro_lookup:")
print(unique(boro_lookup$county_fips))

# 5. Trim any leading/trailing spaces from COUNTYFP10 and county_fips
all_data_inc <- all_data_inc %>%
  mutate(COUNTYFP10 = str_trim(COUNTYFP10))

boro_lookup <- boro_lookup %>%
  mutate(county_fips = str_trim(county_fips))

# 6. Ensure both COUNTYFP10 and county_fips are character type
all_data_inc <- all_data_inc %>%
  mutate(COUNTYFP10 = as.character(COUNTYFP10))

boro_lookup <- boro_lookup %>%
  mutate(county_fips = as.character(county_fips))

# 7. Check column names before join to ensure we have the correct ones
print("Column names in all_data_inc before join:")
print(colnames(all_data_inc))

print("Column names in boro_lookup before join:")
print(colnames(boro_lookup))

# 8. Perform the join with the borough lookup table
all_data_inc <- all_data_inc %>%
  left_join(boro_lookup, by = c("COUNTYFP10" = "county_fips"))

# 9. Check if boro_code was successfully added
print("Column names in all_data_inc after join:")
print(colnames(all_data_inc))

# Check if boro_code exists
if ("boro_code" %in% colnames(all_data_inc)) {
  print("boro_code successfully added!")
} else {
  stop("boro_code not found after join!")
}

# 10. Create BoroCT2010 by combining boro_code and TRACTCE10
all_data_inc <- all_data_inc %>%
  mutate(BoroCT2010 = paste0(boro_code, TRACTCE10))

# 11. Convert BoroCT2010 columns to character type to ensure compatibility for joining
all_data_inc <- all_data_inc %>%
  mutate(BoroCT2010 = as.character(BoroCT2010))

flood <- flood %>%
  mutate(BoroCT2010 = as.character(BoroCT2010))

# 12. Join with flood data to get flood hazard information
all_data_inc <- all_data_inc %>%
  left_join(
    flood %>% 
      select(BoroCT2010, NTAName, ZONE_SUBTY) %>%
      rename(flood_hazard = ZONE_SUBTY),  # Rename ZONE_SUBTY to flood_hazard
    by = "BoroCT2010"  # Joining on BoroCT2010 column
  )

# 13. Final check of the structure to confirm the results
str(all_data_inc)

# Convert flood_hazard to a factor
all_data_inc <- all_data_inc %>%
  mutate(flood_hazard = factor(flood_hazard))



```



################################# try mrf_income first to test k and concurvity with county

Neighbor relationships for MRF smooth model
```{r}
# Establish neighbor relationships for MRF model
nb <- poly2nb(all_data_inc, row.names = all_data_inc$GEOID)
names(nb) <- attr(nb, "region.id") 
```

PM2.5 MRF smooth model
```{r}
mrf_income_pm <- gam(pm_change ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=400, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") #changing the k, whether or not income has a spline/log/nothing affects sig of score_0.5; k needs to be at least 15 for residuals etc to be okay (k=15 highest sig but maybe not best diagnostic plots); k needs to be at least ~400 for Moran's I to be insignificant

summary(mrf_income_pm)
gam.check(mrf_income_pm)
concurvity(mrf_income_pm, full = TRUE)

#####Calculating Moran's I to assess residual spatial autocorrelation

# First identify areas with no neighbors
no_neighbors <- which(card(nb) == 0)
if(length(no_neighbors) > 0) {
    print("Areas with no neighbors:")
    print(no_neighbors)
}

# Use zero.policy = TRUE (keeping all observations - isolated tracts given a weight of zero?)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get model residuals
residuals <- residuals(mrf_income_pm)

# Perform Moran's I test with zero.policy
moran_test <- moran.test(residuals, listw, zero.policy = TRUE)
print(moran_test)

# Create Moran's I plot

# Calculate lagged residuals with zero.policy
lag_residuals <- lag.listw(listw, residuals, zero.policy = TRUE)

# Create data frame for plotting
moran_df <- data.frame(
    residuals = residuals,
    lag_residuals = lag_residuals
)

# Moran scatter plot
ggplot(moran_df, aes(x = residuals, y = lag_residuals)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Residuals",
        y = "Spatially Lagged Residuals",
        title = "Moran's I Scatter Plot of GAM Residuals",
        subtitle = paste("Moran's I =", round(moran_test$estimate[1], 3),
                        "\np-value =", round(moran_test$p.value, 3))
    ) +
    theme_minimal()

```
O3 MRF smooth model
```{r}
mrf_income_o3 <- gam(o3_change ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=400, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") #when k is upped to 450, it takes too long to run

summary(mrf_income_o3)
gam.check(mrf_income_o3)
concurvity(mrf_income_o3, full = TRUE)

#####Calculating Moran's I to assess residual spatial autocorrelation

# First identify areas with no neighbors
no_neighbors <- which(card(nb) == 0)
if(length(no_neighbors) > 0) {
    print("Areas with no neighbors:")
    print(no_neighbors)
}

# Use zero.policy = TRUE (keeping all observations - isolated tracts given a weight of zero?)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get model residuals
residuals <- residuals(mrf_income_o3)

# Perform Moran's I test with zero.policy
moran_test <- moran.test(residuals, listw, zero.policy = TRUE)
print(moran_test)

# Create Moran's I plot

# Calculate lagged residuals with zero.policy
lag_residuals <- lag.listw(listw, residuals, zero.policy = TRUE)

# Create data frame for plotting
moran_df <- data.frame(
    residuals = residuals,
    lag_residuals = lag_residuals
)

# Moran scatter plot
ggplot(moran_df, aes(x = residuals, y = lag_residuals)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Residuals",
        y = "Spatially Lagged Residuals",
        title = "Moran's I Scatter Plot of GAM Residuals",
        subtitle = paste("Moran's I =", round(moran_test$estimate[1], 3),
                        "\np-value =", round(moran_test$p.value, 3))
    ) +
    theme_minimal()
```
NO2 MRF smooth model
```{r}
mrf_income_no2 <- gam(no2_change ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=400, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") #at k=5 is significant but residuals weird, when k increases, residuals and stuff still patterned-ish but better overall and no longer sig
summary(mrf_income_no2)
gam.check(mrf_income_no2)
concurvity(mrf_income_no2, full = TRUE)

#####Calculating Moran's I to assess residual spatial autocorrelation

# First identify areas with no neighbors
no_neighbors <- which(card(nb) == 0)
if(length(no_neighbors) > 0) {
    print("Areas with no neighbors:")
    print(no_neighbors)
}

# Use zero.policy = TRUE (keeping all observations - isolated tracts given a weight of zero?)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get model residuals
residuals <- residuals(mrf_income_no2)

# Perform Moran's I test with zero.policy
moran_test <- moran.test(residuals, listw, zero.policy = TRUE)
print(moran_test)

# Create Moran's I plot

# Calculate lagged residuals with zero.policy
lag_residuals <- lag.listw(listw, residuals, zero.policy = TRUE)

# Create data frame for plotting
moran_df <- data.frame(
    residuals = residuals,
    lag_residuals = lag_residuals
)

# Moran scatter plot
ggplot(moran_df, aes(x = residuals, y = lag_residuals)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Residuals",
        y = "Spatially Lagged Residuals",
        title = "Moran's I Scatter Plot of GAM Residuals",
        subtitle = paste("Moran's I =", round(moran_test$estimate[1], 3),
                        "\np-value =", round(moran_test$p.value, 3))
    ) +
    theme_minimal()
```

NDVI MRF smooth model
```{r}
mrf_income_ndvi <- gam(ndvi_change ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=300, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 
summary(mrf_income_ndvi)
gam.check(mrf_income_ndvi)
concurvity(mrf_income_ndvi, full = TRUE) #K NEEDS TO BE 300 FOR MORANS I TO BE INSIGNIFICANT

#####Calculating Moran's I to assess residual spatial autocorrelation

# First identify areas with no neighbors
no_neighbors <- which(card(nb) == 0)
if(length(no_neighbors) > 0) {
    print("Areas with no neighbors:")
    print(no_neighbors)
}

# Use zero.policy = TRUE (keeping all observations - isolated tracts given a weight of zero?)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get model residuals
residuals <- residuals(mrf_income_ndvi)

# Perform Moran's I test with zero.policy
moran_test <- moran.test(residuals, listw, zero.policy = TRUE)
print(moran_test)

# Create Moran's I plot

# Calculate lagged residuals with zero.policy
lag_residuals <- lag.listw(listw, residuals, zero.policy = TRUE)

# Create data frame for plotting
moran_df <- data.frame(
    residuals = residuals,
    lag_residuals = lag_residuals
)

# Moran scatter plot
ggplot(moran_df, aes(x = residuals, y = lag_residuals)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Residuals",
        y = "Spatially Lagged Residuals",
        title = "Moran's I Scatter Plot of GAM Residuals",
        subtitle = paste("Moran's I =", round(moran_test$estimate[1], 3),
                        "\np-value =", round(moran_test$p.value, 3))
    ) +
    theme_minimal()
```

2016 Mean Temp MRF smooth model
```{r}
mrf_income_temp <- gam(mean_temp ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=300, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML")  #terrible residual plots
summary(mrf_income_temp)
gam.check(mrf_income_temp)
concurvity(mrf_income_temp, full = TRUE)

#####Calculating Moran's I to assess residual spatial autocorrelation

# First identify areas with no neighbors
no_neighbors <- which(card(nb) == 0)
if(length(no_neighbors) > 0) {
    print("Areas with no neighbors:")
    print(no_neighbors)
}

# Use zero.policy = TRUE (keeping all observations - isolated tracts given a weight of zero?)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get model residuals
residuals <- residuals(mrf_income_ndvi)

# Perform Moran's I test with zero.policy
moran_test <- moran.test(residuals, listw, zero.policy = TRUE)
print(moran_test)

# Create Moran's I plot

# Calculate lagged residuals with zero.policy
lag_residuals <- lag.listw(listw, residuals, zero.policy = TRUE)

# Create data frame for plotting
moran_df <- data.frame(
    residuals = residuals,
    lag_residuals = lag_residuals
)

# Moran scatter plot
ggplot(moran_df, aes(x = residuals, y = lag_residuals)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Residuals",
        y = "Spatially Lagged Residuals",
        title = "Moran's I Scatter Plot of GAM Residuals",
        subtitle = paste("Moran's I =", round(moran_test$estimate[1], 3),
                        "\np-value =", round(moran_test$p.value, 3))
    ) +
    theme_minimal()
```
Creating composite z-score
```{r}
# Calculate z-scores for each variable
all_data_inc$pm_z <- scale(all_data_inc$pm_change)
all_data_inc$o3_z <- scale(all_data_inc$o3_change)
all_data_inc$no2_z <- scale(all_data_inc$no2_change)
all_data_inc$ndvi_z <- scale(all_data_inc$ndvi_change)
all_data_inc$temp_z <- scale(all_data_inc$mean_temp)

# Create composite z-score (average of all z-scores)
all_data_inc$composite_z <- rowMeans(
  cbind(
    all_data_inc$pm_z,
    all_data_inc$o3_z,
    all_data_inc$no2_z,
    all_data_inc$ndvi_z,
    all_data_inc$temp_z
  ),
  na.rm = TRUE
)

# Create new GAM model with composite z-score
mrf_income_composite <- gam(
  composite_z ~ score_0.5 + med_income_ranked + 
    s(GEOID, bs='mrf', k=100, xt=list(nb=nb)),
  family = scat,
  data = all_data_inc,
  method = "REML"
) #sig when k = 100 but Moran's I is 0.265, insig when k = 300, Moran's I is 0.074 but p is sig
 
summary(mrf_income_composite)
gam.check(mrf_income_composite)

#####Calculating Moran's I to assess residual spatial autocorrelation

# First identify areas with no neighbors
no_neighbors <- which(card(nb) == 0)
if(length(no_neighbors) > 0) {
    print("Areas with no neighbors:")
    print(no_neighbors)
}

# Use zero.policy = TRUE (keeping all observations - isolated tracts given a weight of zero?)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get model residuals
residuals <- residuals(mrf_income_composite)

# Perform Moran's I test with zero.policy
moran_test <- moran.test(residuals, listw, zero.policy = TRUE)
print(moran_test)

# Create Moran's I plot

# Calculate lagged residuals with zero.policy
lag_residuals <- lag.listw(listw, residuals, zero.policy = TRUE)

# Create data frame for plotting
moran_df <- data.frame(
    residuals = residuals,
    lag_residuals = lag_residuals
)

# Moran scatter plot
ggplot(moran_df, aes(x = residuals, y = lag_residuals)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Residuals",
        y = "Spatially Lagged Residuals",
        title = "Moran's I Scatter Plot of GAM Residuals",
        subtitle = paste("Moran's I =", round(moran_test$estimate[1], 3),
                        "\np-value =", round(moran_test$p.value, 3))
    ) +
    theme_minimal()

```

Testing k vs Moran's I for MRF composite model
```{r}
# Function to get number of unique locations
n_locations <- length(unique(all_data_inc$GEOID))

# Calculate minimum suggested k
k_min <- ceiling(n_locations^(1/3))  # cube root rule of thumb

# Define sequence of k values to test
k_values <- seq(k_min, 350, by = 50)

# Function to run k.check for multiple k values
test_k_values <- function(data, k_values) {
  results <- data.frame(
    k = numeric(),
    edf = numeric(),
    k_index = numeric(),
    p_value = numeric(),
    gcv = numeric()
  )
  
  for(k in k_values) {
    cat(sprintf("\nTesting k = %d\n", k))
    
    # Fit model with current k
    tryCatch({
      test_model <- gam(
        composite_z ~ score_0.5 + med_income_ranked + 
          s(GEOID, bs='mrf', k=k, xt=list(nb=nb)),
        family = scat,
        data = data,
        method = "REML"
      )
      
      # Extract model summary
      model_summary <- summary(test_model)
      
      # Get basic diagnostics that don't depend on gam.check
      new_row <- data.frame(
        k = k,
        edf = sum(model_summary$edf),
        k_index = NA,  # Will be filled if gam.check works
        p_value = NA,  # Will be filled if gam.check works
        gcv = test_model$gcv.ubre
      )
      
      # Try to get gam.check results
      tryCatch({
        check <- gam.check(test_model)
        new_row$k_index <- check$k.index
        new_row$p_value <- check$s.table[4]
      }, error = function(e) {
        cat(sprintf("Warning: Could not compute gam.check for k=%d\n", k))
      })
      
      results <- rbind(results, new_row)
      
    }, error = function(e) {
      cat(sprintf("Error fitting model for k=%d: %s\n", k, e$message))
    })
  }
  return(results)
}

# Print minimum suggested k
cat(sprintf("Minimum suggested k based on number of locations: %d\n", k_min))

# Test k values
k_results <- test_k_values(all_data_inc, k_values)

# Print results
print(k_results)

# Plot available results
par(mfrow=c(2,2))

# Only plot if we have results
if(nrow(k_results) > 0) {
  # P-value plot (if available)
  if(!all(is.na(k_results$p_value))) {
    plot(k_results$k, k_results$p_value, type="b", 
         main="P-value vs k", xlab="k", ylab="p-value")
  }
  
  # GCV plot
  plot(k_results$k, k_results$gcv, type="b", 
       main="GCV Score vs k", xlab="k", ylab="GCV")
  
  # EDF plot
  plot(k_results$k, k_results$edf, type="b", 
       main="EDF vs k", xlab="k", ylab="Effective Degrees of Freedom")
  
  # k-index plot (if available)
  if(!all(is.na(k_results$k_index))) {
    plot(k_results$k, k_results$k_index, type="b", 
         main="k-index vs k", xlab="k", ylab="k-index")
  }
}

# Function to get Moran's I for model residuals
check_spatial_autocorr <- function(model, nb) {
  residuals <- residuals(model)
  moran_test <- spdep::moran.test(residuals, 
                                 spdep::nb2listw(nb, zero.policy=TRUE),
                                 zero.policy=TRUE)
  return(c(
    moran_i = moran_test$estimate[1],
    p_value = moran_test$p.value
  ))
}

# First, let's check which areas have no neighbors
n_neighbors <- sapply(nb, length)
isolated_areas <- which(n_neighbors == 0)
if(length(isolated_areas) > 0) {
  cat("Areas with no neighbors:\n")
  print(isolated_areas)
}

# Function to get Moran's I for model residuals
check_spatial_autocorr <- function(model, nb) {
  residuals <- residuals(model)
  moran_test <- spdep::moran.test(residuals, 
                                 spdep::nb2listw(nb, zero.policy=TRUE),
                                 zero.policy=TRUE)
  return(moran_test)  # Return the full test object
}

# Calculate Moran's I for raw composite z-score
raw_moran <- spdep::moran.test(all_data_inc$composite_z, 
                              spdep::nb2listw(nb, zero.policy=TRUE),
                              zero.policy=TRUE)
cat(sprintf("\nRaw data Moran's I:\n"))
cat(sprintf("Moran's I statistic: %.3f\n", raw_moran$estimate["Moran I statistic"]))
cat(sprintf("p-value: %.3e\n", raw_moran$p.value))

# Test k values and check both model fit and spatial autocorrelation
test_k_values <- function(data, k_values, raw_moran) {
  results <- data.frame()
  
  # Store raw Moran's I value
  raw_moran_i <- raw_moran$estimate["Moran I statistic"]
  
  for(k in k_values) {
    cat(sprintf("\nTesting k = %d\n", k))
    
    # Fit model
    test_model <- gam(
      composite_z ~ score_0.5 + med_income_ranked + 
        s(GEOID, bs='mrf', k=k, xt=list(nb=nb)),
      family = scat,
      data = data,
      method = "REML"
    )
    
    # Get model summary
    model_summary <- summary(test_model)
    
    # Get Moran's I results
    moran_results <- check_spatial_autocorr(test_model, nb)
    residual_moran_i <- moran_results$estimate["Moran I statistic"]
    
    # Calculate reduction in Moran's I
    moran_reduction <- ((raw_moran_i - residual_moran_i) / abs(raw_moran_i)) * 100
    
    # Store results
    results <- rbind(results,
      data.frame(
        k = k,
        edf = sum(model_summary$edf),
        score = test_model$gcv.ubre,
        moran_i = residual_moran_i,
        moran_p = moran_results$p.value,
        moran_reduction = moran_reduction
      )
    )
    
    # Print current results
    cat(sprintf("EDF: %.1f\n", sum(model_summary$edf)))
    cat(sprintf("Score: %.3f\n", test_model$gcv.ubre))
    cat(sprintf("Residual Moran's I: %.3f (p=%.3e)\n", 
                residual_moran_i, moran_results$p.value))
    cat(sprintf("Reduction in Moran's I: %.1f%%\n", moran_reduction))
  }
  return(results)
}

# Test reasonable k values
k_values <- c(113, 163, 213, 263)
k_results <- test_k_values(all_data_inc, k_values, raw_moran)

# Print final comparison table
cat("\nSummary Results:\n")
print(data.frame(
  k = k_results$k,
  EDF = round(k_results$edf, 1),
  Score = round(k_results$score, 3),
  Morans_I = round(k_results$moran_i, 3),
  Moran_p = format(k_results$moran_p, scientific = TRUE),
  Reduction = sprintf("%.1f%%", k_results$moran_reduction)
))

# Plot results
par(mfrow=c(2,3))

# Score vs k
plot(k_results$k, k_results$score, type="b", 
     main="Score vs k", xlab="k", ylab="Score")

# EDF vs k
plot(k_results$k, k_results$edf, type="b", 
     main="EDF vs k", xlab="k", ylab="Effective Degrees of Freedom")

# Moran's I vs k
plot(k_results$k, k_results$moran_i, type="b", 
     main="Moran's I vs k", xlab="k", ylab="Moran's I")
abline(h=raw_moran$estimate["Moran I statistic"], col="red", lty=2)
text(k_values[1], raw_moran$estimate["Moran I statistic"], 
     "Raw Data Moran's I", pos=3, col="red")

# Moran's p-value vs k
plot(k_results$k, k_results$moran_p, type="b", 
     main="Moran's I p-value vs k", xlab="k", ylab="p-value")
abline(h=0.05, col="red", lty=2)

# Reduction in Moran's I
plot(k_results$k, k_results$moran_reduction, type="b",
     main="% Reduction in Moran's I",
     xlab="k", 
     ylab="Percent Reduction")
```
```{r}
hurricane <- st_read(here("data", "hurricane_evacuation_zones.shp"))

mrf_income_composite_residuals <- residuals(mrf_income_composite)

aov_result <- aov(mrf_income_composite_residuals ~ hurricane_, data = hurricane)


#probably need to separate out the hurricane polygons - give each tract a score - then do residual analysis
```

```{r}
```

```{r}
```

#########################################
########################################
code scraps for mrf smooth
```{r}
#testing ranking baseline exposure
all_data_inc$pm_baseline_rank <- rank(all_data_inc$pm_2002)

mrf_income_pm_county <- gam(pm_change ~ score_0.5 + med_income_ranked + s(COUNTYFP10, bs='re')+ s(GEOID, bs='mrf', k=15, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 
summary(mrf_income_pm_county)
gam.check(mrf_income_pm_county)

mrf_income_pm_baseline <- gam(pm_change ~ score_0.5 + med_income_ranked +pm_baseline_rank + s(GEOID, bs='mrf', k=15, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML")

mrf_income_pm_flood <- gam(pm_change ~ score_0.5 + med_income_ranked + flood_hazard + s(GEOID, bs='mrf', k=400, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") #K=400, NO LONGER HAS SPATIAL AUTOCORRELATION.

summary(mrf_income_pm_flood)
gam.check(mrf_income_pm_flood)
plot(mrf_income_pm_flood)

summary(mrf_income_pm_baseline) #after taking into account baseline rank, gent score not significant at all p = 0.72

mrf_income_pm_5 <- gam(pm_change ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=5, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 
summary(mrf_income_pm_5)
gam.check(mrf_income_pm_5)

mrf_income_pm_county_dt <- gam(pm_change ~ score_0.5 + med_income_ranked + s(COUNTYFP10, bs='re')+ s(dt_dist) + s(GEOID, bs='mrf', k=15, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 
summary(mrf_income_pm_county_dt)
gam.check(mrf_income_pm_county_dt)

income_pm <- gam(pm_change ~ score_0.5 + med_income_ranked + s(COUNTYFP10, bs='re')+ s(dt_dist), family = scat, data = all_data_inc, method = "REML") 

mrf_income_pm_county_dt_300 <- gam(pm_change ~ score_0.5 + med_income_ranked + s(COUNTYFP10, bs='re')+ s(dt_dist) + s(GEOID, bs='mrf', k=300, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 


#ozone
mrf_income_o3 <- gam(o3_change ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=15, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML")

mrf_income_o3_flood_hazard <- gam(o3_change ~ score_0.5 + med_income_ranked + flood_hazard + s(GEOID, bs='mrf', k=400, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML")


mrf_income_o3_county <- gam(o3_change ~ score_0.5 + s(COUNTYFP10, bs='re') + med_income_ranked + s(GEOID, bs='mrf', k=15, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML")


summary(mrf_income_o3_flood_hazard)
gam.check(mrf_income_o3_flood_hazard)


summary(mrf_income_o3_county)
gam.check(mrf_income_o3_county) #worse than without county
concurvity(mrf_income_o3_county, full = TRUE)

#testing ranking baseline exposure
all_data_inc$o3_baseline_rank <- rank(all_data_inc$o3_2002)
mrf_income_o3_baseline <- gam(o3_change ~ score_0.5 + med_income_ranked + o3_baseline_rank + s(GEOID, bs='mrf', k=15, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML")

summary(mrf_income_o3_baseline)#but with o3, after adding baseline, score_0.5 still sig
gam.check(mrf_income_o3_baseline) #curve is worse
concurvity(mrf_income_o3, full = TRUE)



all_data_inc$no2_baseline_rank <- rank(all_data_inc$no2_2000)

mrf_income_no2_baseline <- gam(no2_change ~ score_0.5 + med_income_ranked + no2_baseline_rank + s(GEOID, bs='mrf', k=5, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 

summary(mrf_income_no2_baseline)
gam.check(mrf_income_no2_baseline)

mrf_income_no2_15 <- gam(no2_change ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=15, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML")

summary(mrf_income_no2_15)
gam.check(mrf_income_no2_15)


#ndvi

all_data_inc$ndvi_baseline_rank <- rank(all_data_inc$ndvi_2000)
mrf_income_ndvi_baseline <- gam(ndvi_change ~ score_0.5 + med_income_ranked + ndvi_baseline_rank + s(GEOID, bs='mrf', k=5, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 
summary(mrf_income_ndvi_baseline)
gam.check(mrf_income_ndvi_baseline)

mrf_income_ndvi_15 <- gam(ndvi_change ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=15, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 
summary(mrf_income_ndvi_15)
gam.check(mrf_income_ndvi_15)


```

```{r}
# First identify areas with no neighbors
no_neighbors <- which(card(nb) == 0)
if(length(no_neighbors) > 0) {
    print("Areas with no neighbors:")
    print(no_neighbors)
}

# Method 1: Simply use zero.policy = TRUE (keeping all observations)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get model residuals
residuals <- residuals(mrf_income_no2)

# Perform Moran's I test with zero.policy
moran_test <- moran.test(residuals, listw, zero.policy = TRUE)
print(moran_test)

# Create Moran's I plot
library(ggplot2)

# Calculate lagged residuals with zero.policy
lag_residuals <- lag.listw(listw, residuals, zero.policy = TRUE)

# Create data frame for plotting
moran_df <- data.frame(
    residuals = residuals,
    lag_residuals = lag_residuals
)

# Remove NA values if any exist
moran_df <- na.omit(moran_df)

# Moran scatter plot
ggplot(moran_df, aes(x = residuals, y = lag_residuals)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Residuals",
        y = "Spatially Lagged Residuals",
        title = "Moran's I Scatter Plot of GAM Residuals",
        subtitle = paste("Moran's I =", round(moran_test$estimate[1], 3),
                        "\np-value =", round(moran_test$p.value, 3))
    ) +
    theme_minimal()
```

################################ANALYZE COOK DISTANCE AND RESIDUALS ETC FOR MRF MODELS
```{r}
# Calculate Cook's distance
cooks_d <- cooks.distance(mrf_income_pm)
plot(cooks_d, type = "h", main = "Cook's Distance")
abline(h = 4/length(cooks_d), col = "red")  # Rule of thumb threshold

# Identify potential influential points
influential <- which(cooks_d > 4/length(cooks_d))

influential_data <- all_data_inc[influential, ]
print(influential_data)  # Look at characteristics of these points

# Investigate why these points are influential:
# Plot their locations
plot(all_data_inc_na$long, all_data_inc_na$lat)
points(all_data_inc_na$long[influential], 
       all_data_inc_na$lat[influential], 
       col = "red", pch = 16)

```
```{r}
# Plot the smooth effect for GEOID
plot(mrf_income_pm, select = 1, scale = 0, main = "MRF Smooth for GEOID")
summary(nb)

```

################################

```{r}

library(purrr)
library(ggplot2)

# List of model names
model_names <- c(
  "mrf_income_pm", "mrf_income_pm_baseline", "mrf_income_pm_5",
  "mrf_income_o3", "mrf_income_o3_baseline",
  "mrf_income_no2", "mrf_income_no2_baseline", "mrf_income_no2_15",
  "mrf_income_ndvi", "mrf_income_ndvi_baseline", "mrf_income_ndvi_15",
  "mrf_income_temp"
)

# Function to calculate Cook's Distance, plot results, and identify influential points
process_model <- function(model_name) {
  
  # Retrieve the model from the global environment using get()
  model <- get(model_name)
  
  # Calculate Cook's Distance
  cooks_d <- cooks.distance(model)
  
  # Plot Cook's Distance with model name in title
  plot(cooks_d, type = "h", main = paste("Cook's Distance:", model_name))
  abline(h = 4/length(cooks_d), col = "red")  # Rule of thumb threshold
  
  # Identify influential points
  influential <- which(cooks_d > 4/length(cooks_d))
  
  # Retrieve the data for the influential points
  influential_data <- all_data_inc[influential, ]
  print(paste("Influential points for", model_name))
  print(influential_data)  # Look at characteristics of these points
  
  # Plot the locations of influential points on the map
  plot(all_data_inc_na$long, all_data_inc_na$lat,
       main = paste("Influential Points for", model_name),  # Add model name to the map plot title
       xlab = "Longitude", ylab = "Latitude")
  points(all_data_inc_na$long[influential], 
         all_data_inc_na$lat[influential], 
         col = "red", pch = 16)
  
  return(influential_data)
}

# Use purrr::map to apply the function to all models
influential_points_list <- map(model_names, process_model)

# If you want to inspect the influential points for each model, you can print the list
influential_points_list

```
trying SLM and SEM models again (no spatial autocorrelation, but linear so qqplots bad)
```{r}

#### QQ plots with confidence intervals
get_qq_plot <- function(model) {
  
  model_resid <- data.frame(residuals = model$residuals)
  
  plot <- ggplot(data = model_resid, mapping = aes(sample = residuals)) + 
    geom_qq_band(bandType = "ks", mapping = aes(fill = "KS"), alpha = 0.5) +
    geom_qq_band(bandType = "ts", mapping = aes(fill = "TS"), alpha = 0.5) +
    geom_qq_band(bandType = "pointwise", mapping = aes(fill = "Normal"), alpha = 0.5) +
    geom_qq_band(bandType = "boot", mapping = aes(fill = "Bootstrap"), alpha = 0.5) +
    stat_qq_band() +
    stat_qq_line() +
    stat_qq_point()
  
  return(plot)
}

############## SLM and SEM models

vars <- list("pm_change", "o3_change", "no2_change", "ndvi_change", "mean_temp")

# build linear models to test spatial error and spatial lag terms
pm_lm <- lm(pm_change ~ score_0.5 + med_income_ranked, data = all_data_inc)

# testing spatial error model
lm.RStests(pm_lm, ny.6, test="RSerr", zero.policy = T) # significant p value
# testing spatial lag model
lm.RStests(pm_lm, ny.6, test="RSerr", zero.policy = T) # significant p value

# fit spatial error models
get_sem <- function(var) {
  
  data_sem <- all_data_inc %>% 
    rename(dv = all_of(var))
  
  sem <- errorsarlm(dv ~ score_0.5 + med_income_ranked, data = data_sem, listw = ny.6, zero.policy = TRUE)
  
  return(sem)
}

sem_list <- lapply(vars, get_sem)

# fit spatial lag models
get_slm <- function(var) {
  
  data_slm <- all_data_inc %>% 
    rename(dv = all_of(var))
  
  slm <- lagsarlm(dv ~ score_0.5 + med_income_ranked, data = data_slm, listw = ny.6, zero.policy = TRUE)
  
  return(slm)
}

slm_list <- lapply(vars, get_slm)

# test residuals
moran_sem <- purrr::map(sem_list, ~ moran.test(.x$residuals, ny.6)) 
moran_slm <- purrr::map(slm_list, ~ moran.test(.x$residuals, ny.6))

# qq plots
qq_sem <- lapply(sem_list, get_qq_plot)
qq_slm <- lapply(slm_list, get_qq_plot)

grid.arrange(qq_sem[[1]], qq_sem[[2]], qq_sem[[3]], qq_sem[[4]], qq_sem[[5]])
grid.arrange(qq_slm[[1]], qq_slm[[2]], qq_slm[[3]], qq_slm[[4]], qq_slm[[5]])

# stratified SLM
get_slm_strat <- function(countyfp) {
  
  # filter data for county/borough
  data_borough <- all_data_inc %>% 
    filter(COUNTYFP %in% countyfp)
  
  # list all exposure variables for borough dataset
  exposures <- list(data_borough$pm_change, data_borough$o3_change, data_borough$no2_change,
                    data_borough$ndvi_change, data_borough$mean_temp)
  
  # identify neighbor relationships
  spdat.sens_borough <- as_Spatial(data_borough)
  ny.6_borough <- knearneigh(sp::coordinates(spdat.sens_borough), k=6)
  ny.6_borough <- knn2nb(ny.6_borough)
  ny.6_borough <- make.sym.nb(ny.6_borough)
  ny.6_borough <- nb2listw(ny.6_borough, style="W")
  
  get_slm <- function(exposure) {
    
    # assign exposure of interest as "var" variable
    data_borough <- data_borough %>% 
      mutate(var = exposure)
    
    slm <- lagsarlm(var ~ score_0.5 + med_income_ranked, data = data_borough, listw = ny.6_borough, zero.policy = TRUE)
    
    return(slm)
  }
  
  models <- lapply(exposures, get_slm)
  
  return(models)
}

countyfp <- list("005", "047", "061", "081", "085")

  # filter data for county/borough
  data_borough <- all_data_inc %>% 
    filter(COUNTYFP %in% countyfp)
  
  # list all exposure variables for borough dataset
  exposures <- list(data_borough$pm_change, data_borough$o3_change, data_borough$no2_change,
                    data_borough$ndvi_change, data_borough$mean_temp)
  
 

slm_strat <- lapply(countyfp, get_slm_strat) %>% 
  # flatten list of 25 models - PM2.5, O3, NO2, NDVI, temp, for 5 boroughs
  unlist(recursive = FALSE)

qq_slm_strat <- lapply(slm_strat, get_qq_plot)

# Bronx
grid.arrange(qq_slm_strat[[1]], qq_slm_strat[[2]], qq_slm_strat[[3]], qq_slm_strat[[4]], qq_slm_strat[[5]])
# Brooklyn
grid.arrange(qq_slm_strat[[6]], qq_slm_strat[[7]], qq_slm_strat[[8]], qq_slm_strat[[9]], qq_slm_strat[[10]])
# Manhattan
grid.arrange(qq_slm_strat[[11]], qq_slm_strat[[12]], qq_slm_strat[[13]], qq_slm_strat[[14]], qq_slm_strat[[15]])
# Queens
grid.arrange(qq_slm_strat[[16]], qq_slm_strat[[17]], qq_slm_strat[[18]], qq_slm_strat[[19]], qq_slm_strat[[20]])
# Staten Island
grid.arrange(qq_slm_strat[[21]], qq_slm_strat[[22]], qq_slm_strat[[23]], qq_slm_strat[[24]], qq_slm_strat[[25]])

# stratified SEM
get_sem_strat <- function(countyfp) {
  
  # filter data for county/borough
  data_borough <- all_data_inc %>% 
    filter(COUNTYFP %in% countyfp)
  
  # list all exposure variables for borough dataset
  exposures <- list(data_borough$pm_change, data_borough$o3_change, data_borough$no2_change,
                    data_borough$ndvi_change, data_borough$mean_temp)
  
  # identify neighbor relationships
  spdat.sens_borough <- as_Spatial(data_borough)
  ny.6_borough <- knearneigh(sp::coordinates(spdat.sens_borough), k=6)
  ny.6_borough <- knn2nb(ny.6_borough)
  ny.6_borough <- make.sym.nb(ny.6_borough)
  ny.6_borough <- nb2listw(ny.6_borough, style="W")
  
  get_sem <- function(exposure) {
    
    # assign exposure of interest as "var" variable
    data_borough <- data_borough %>% 
      mutate(var = exposure)
    
    sem <- errorsarlm(var ~ score_0.5 +  + dt_dist, data = data_borough, listw = ny.6_borough, zero.policy = TRUE)
    
    return(sem)
  }
  
  models <- lapply(exposures, get_sem)
  
  return(models)
}

sem_strat <- lapply(countyfp, get_sem_strat) %>% 
  # flatten list of 25 models - PM2.5, O3, NO2, NDVI, temp, for 5 boroughs
  un(recursive = FALSE)

qq_sem_strat <- lapply(sem_strat, get_qq_plot)

# Bronx
grid.arrange(qq_sem_strat[[1]], qq_sem_strat[[2]], qq_sem_strat[[3]], qq_sem_strat[[4]], qq_sem_strat[[5]])
# Brooklyn
grid.arrange(qq_sem_strat[[6]], qq_sem_strat[[7]], qq_sem_strat[[8]], qq_sem_strat[[9]], qq_sem_strat[[10]])
# Manhattan
grid.arrange(qq_sem_strat[[11]], qq_sem_strat[[12]], qq_sem_strat[[13]], qq_sem_strat[[14]], qq_sem_strat[[15]])
# Queens
grid.arrange(qq_sem_strat[[16]], qq_sem_strat[[17]], qq_sem_strat[[18]], qq_sem_strat[[19]], qq_sem_strat[[20]])
# Staten Island
grid.arrange(qq_sem_strat[[21]], qq_sem_strat[[22]], qq_sem_strat[[23]], qq_sem_strat[[24]], qq_sem_strat[[25]])
```

```{r}
# Spatial Error Model with shifted data
library(spatialreg)
library(MASS)

# Find minimum value and shift data to positive
min_o3 <- min(all_data_inc$o3_change)
shift_value <- abs(min_o3) + 1  # add 1 to ensure all positive values
o3_shifted <- all_data_inc$o3_change + shift_value

# Create model matrix for initial regression
X <- model.matrix(~ score_0.5 + med_income_ranked + flood_hazard, data = all_data_inc)

# Perform Box-Cox transformation on shifted data
bc <- boxcox(lm(o3_shifted ~ X - 1), 
             lambda = seq(-2, 2, by = 0.1),
             plotit = TRUE)
lambda <- bc$x[which.max(bc$y)]

# Transform the shifted dependent variable
if(abs(lambda) < 0.01) {
    o3_transformed <- log(o3_shifted)
    transform_type <- "log of shifted data"
} else {
    o3_transformed <- (o3_shifted^lambda - 1)/lambda
    transform_type <- paste("Box-Cox with lambda =", round(lambda, 3), "on shifted data")
}

# Fit spatial error model
spatial_error <- errorsarlm(o3_transformed ~ 
                           score_0.5 + 
                           med_income_ranked + flood_hazard,
                           data = all_data_inc,
                           listw = listw,
                           zero.policy = TRUE)

# Function to back-transform predictions
back_transform <- function(pred) {
    if(abs(lambda) < 0.01) {
        return(exp(pred) - shift_value)
    } else {
        return((pred * lambda + 1)^(1/lambda) - shift_value)
    }
}

# Check residuals
residuals_sem <- residuals(spatial_error)
listw <- nb2listw(nb, style="W", zero.policy=TRUE)
moran_sem <- moran.test(residuals_sem, listw, zero.policy=TRUE)
print("SEM Moran's I test:")
print(moran_sem)


# Print summaries
print("Data transformation details:")
print(paste("Minimum PM change value:", min_pm))
print(paste("Shift applied:", shift_value))
print(paste("Transformation type:", transform_type))
print("\nSpatial Error Model Summary:")
print(summary(spatial_error))

# Function for spatial error model diagnostics
spatial_diagnostics <- function(model, data) {
    # Extract residuals and fitted values safely
    resids <- residuals(model)  # Simplified residuals extraction
    fitted_vals <- fitted(model)
    
    par(mfrow=c(2,2))
    
    # 1. Residuals vs Fitted
    plot(fitted_vals, resids,
         xlab="Fitted values",
         ylab="Residuals",
         main="Residuals vs Fitted",
         type="n")  # Set up empty plot first
    points(fitted_vals, resids, pch=19)  # Add points
    abline(h=0, col="red", lty=2)
    lines(lowess(fitted_vals, resids), col="blue")
    
    # 2. Normal Q-Q plot (simplified)
    qqnorm(resids, main="Normal Q-Q Plot")  # Basic Q-Q plot
    qqline(resids, col="red")
    
    # 3. Scale-Location plot
    sqrt_abs_resids <- sqrt(abs(resids))
    plot(fitted_vals, sqrt_abs_resids,
         xlab="Fitted values",
         ylab=expression(sqrt("|Standardized residuals|")),
         main="Scale-Location",
         type="n")  # Set up empty plot first
    points(fitted_vals, sqrt_abs_resids, pch=19)  # Add points
    lines(lowess(fitted_vals, sqrt_abs_resids), col="blue")
    
    # 4. Residual distribution plot
    hist(resids, 
         main="Distribution of Residuals",
         xlab="Residuals",
         breaks=30,
         probability=TRUE)
    lines(density(resids), col="blue", lwd=2)
}

# Generate predictions first
spatial_preds <- back_transform(fitted(spatial_error))

# Now create all plots
# Spatial Error model diagnostics
par(mfrow=c(2,2))
spatial_diagnostics(spatial_error, all_data_inc)
mtext("Spatial Error Model Diagnostics", side=3, line=-2, outer=TRUE)


# Plot Spatial Error predictions
plot(all_data_inc$pm_change, spatial_preds,
     main="Spatial Error: Observed vs Predicted",
     xlab="Observed", ylab="Predicted",
     type="n")  # Set up empty plot first
points(all_data_inc$pm_change, spatial_preds, pch=19)  # Add points
abline(0,1, col="red")
lines(lowess(all_data_inc$pm_change, spatial_preds), col="blue")

# Reset plotting parameters
par(mfrow=c(1,1))

# Print some summary statistics of the predictions
cat("\nSummary of Spatial Error predictions:\n")
print(summary(spatial_preds))
```
###############################
######################LOOK AT QQ PLOTS FOR SEM MODELS (RN JUST THE NORMAL MODEL QQPLOT BEFORE SEM)
```{r}
# Function to perform spatial error analysis with Box-Cox transformation
spatial_error_analysis <- function(var_name, data, listw) {
    dependent_var <- data[[var_name]]
    
    # Find minimum value and shift data to positive
    min_val <- min(dependent_var)
    shift_value <- abs(min_val) + 1  # add 1 to ensure all positive values
    shifted_data <- dependent_var + shift_value
    
    # Create temporary data frame for Box-Cox
    temp_data <- data.frame(
        y = shifted_data,
        score_0.5 = data$score_0.5,
        med_income_ranked = data$med_income_ranked
    )
    
    # Perform Box-Cox transformation on shifted data
    bc <- boxcox(y ~ score_0.5 + med_income_ranked, 
                 data = temp_data,
                 lambda = seq(-2, 2, by = 0.1),
                 plotit = FALSE)
    lambda <- bc$x[which.max(bc$y)]
    
    # Transform the shifted dependent variable
    if(abs(lambda) < 0.01) {
        transformed_data <- log(shifted_data)
        transform_type <- "log of shifted data"
    } else {
        transformed_data <- (shifted_data^lambda - 1)/lambda
        transform_type <- paste("Box-Cox with lambda =", round(lambda, 3), "on shifted data")
    }
    
    # Create data frame with transformed response
    model_data <- data
    model_data$transformed_response <- transformed_data
    
    # Fit spatial error model
    spatial_error <- errorsarlm(transformed_response ~ score_0.5 + med_income_ranked,
                               data = model_data,
                               listw = listw,
                               zero.policy = TRUE)
    
    # Calculate residuals and perform Moran's I test
    residuals_sem <- residuals(spatial_error)
    moran_sem <- moran.test(residuals_sem, listw, zero.policy=TRUE)
    
    # Back-transform predictions
    if(abs(lambda) < 0.01) {
        predictions <- exp(fitted(spatial_error)) - shift_value
    } else {
        predictions <- (fitted(spatial_error) * lambda + 1)^(1/lambda) - shift_value
    }
    
    # Create diagnostic plots
    par(mfrow=c(2,2))
    
    # Residuals vs Fitted
    plot(fitted(spatial_error), residuals_sem,
         xlab="Fitted values", ylab="Residuals",
         main="Residuals vs Fitted")
    abline(h=0, col="red", lty=2)
    lines(lowess(fitted(spatial_error), residuals_sem), col="blue")
    
    # Normal Q-Q plot
    qqnorm(residuals_sem)
    qqline(residuals_sem, col="red")
    
    # Scale-Location plot
    sqrt_abs_resids <- sqrt(abs(residuals_sem))
    plot(fitted(spatial_error), sqrt_abs_resids,
         xlab="Fitted values",
         ylab=expression(sqrt("|Standardized residuals|")),
         main="Scale-Location")
    lines(lowess(fitted(spatial_error), sqrt_abs_resids), col="blue")
    
    # Residual distribution
    hist(residuals_sem, 
         main="Distribution of Residuals",
         xlab="Residuals",
         breaks=30,
         probability=TRUE)
    lines(density(residuals_sem), col="blue", lwd=2)
    
    # Results list
    results <- list(
        model = spatial_error,
        predictions = predictions,
        moran_test = moran_sem,
        transform_info = list(
            min_value = min_val,
            shift_value = shift_value,
            lambda = lambda,
            transform_type = transform_type
        )
    )
    
    return(results)
}

# Make sure required packages are loaded
library(spatialreg)
library(MASS)  # for boxcox function

# Run analysis for each variable
variables <- c("pm_change", "o3_change", "no2_change", "ndvi_change", "mean_temp")
results_list <- list()

for(var in variables) {
    cat("\n\n=== Analysis for", var, "===\n")
    
    # Run analysis
    results_list[[var]] <- spatial_error_analysis(
        var_name = var,
        data = all_data_inc,
        listw = listw
    )
    
    # Print summaries
    cat("\nTransformation details:\n")
    print(results_list[[var]]$transform_info)
    cat("\nModel Summary:\n")
    print(summary(results_list[[var]]$model))
    cat("\nMoran's I test for residuals:\n")
    print(results_list[[var]]$moran_test)
    
    # Add title to current plot
    mtext(paste("Diagnostic Plots for", var), side=3, line=-2, outer=TRUE)
    
    # Create observed vs predicted plot
    par(mfrow=c(1,1))
    plot(all_data_inc[[var]], results_list[[var]]$predictions,
         main=paste(var, ": Observed vs Predicted"),
         xlab="Observed", ylab="Predicted")
    abline(0,1, col="red")
    lines(lowess(all_data_inc[[var]], results_list[[var]]$predictions), col="blue")
}

# Reset plotting parameters
par(mfrow=c(1,1))
```

################################################
#################################################
Create model functions (NEW FOR MRF ONLY)
```{r}
# Create exposure list
exposure_list <- list(
  pm = list(
    name = "pm",
    baseline = "pm_2002",
    change = "pm_change"
  ),
  o3 = list(
    name = "o3",
    baseline = "o3_2002",
    change = "o3_change"
  ),
  no2 = list(
    name = "no2",
    baseline = "no2_2000",
    change = "no2_change"
  ),
  ndvi = list(
    name = "ndvi",
    baseline = "ndvi_2000",
    change = "ndvi_change"
  ),
  temp = list(
    name = "mean_temp",
    baseline = NULL,
    change = "mean_temp"  # No change variable for temp
  )
)

# Function to run GAM models
run_gam_models <- function(data, exposure_info) {
  # Extract exposure information
  exp_var <- exposure_info$change
  baseline_exposure <- exposure_info$baseline
  
  # Skip if data is missing the exposure variable - gives warning 
  if (!exp_var %in% names(data)) {
    warning(paste("Variable", exp_var, "not found in dataset"))
    return(NULL)
  }
  
  # Base components that will be in all models: gentrification score and county/borough
  base_formula <- paste(exp_var, "~ score_0.5 + s(COUNTYFP10, bs='re')")
  
  # Create model formulas
  formulas <- list(
    # Base model (score + county + spatial) for mrf smooth
    mrf_base = paste(base_formula, "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),
    
    # Models with income
    mrf_income = paste(base_formula, "+ rank(med_income_2000_impute) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),

    # Models with distance to downtown
    mrf_dist = paste(base_formula, "+ s(dt_dist) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  )
  
  # Add baseline models if baseline exposure level exists (all except temp)
  if (!is.null(baseline_exposure) && baseline_exposure %in% names(data)) {
    formulas$mrf_baseline <- paste(base_formula, "+", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    # Combinations with baseline
    formulas$mrf_income_baseline <- paste(base_formula, "+ rank(med_income_2000_impute) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    formulas$mrf_dist_baseline <- paste(base_formula, "+ s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    formulas$mrf_full <- paste(base_formula, "+ rank(med_income_2000_impute) + s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  }
  
  # Add income + distance combinations
  formulas$mrf_income_dist <- paste(base_formula, "+ rank(med_income_2000_impute) + s(dt_dist) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  
  # Convert formulas to formula objects
  formulas <- lapply(formulas, as.formula)
  
  # Run models
  models <- purrr::map(formulas, ~try(gam(., 
                                        family = scat, 
                                        data = data, 
                                        method = "REML"), 
                                    silent = TRUE))
  
  return(models)
}

# drop geometry from spatial data while keeping coordinates
data_prepared <- all_data_inc %>%
  mutate(
    COUNTYFP10 = as.factor(COUNTYFP10),
    GEOID = as.factor(GEOID)
  ) %>%
  st_drop_geometry()

# Construct neighbors list for MRF models
nb <- poly2nb(all_data_inc, row.names = all_data_inc$GEOID)
names(nb) <- attr(nb, "region.id") 

# JH 10/25/24 Warning: some observations have no neighbours; if this seems unexpected, try increasing the snap argument.Warning: neighbour object has 4 sub-graphs; if this sub-graph count seems unexpected, try increasing the snap argument.
```

Run models and check models
```{r}
# Run models for each exposure
all_models <- purrr::map(exposure_list, ~run_gam_models(data_prepared, .))

# Function to check which models succeeded/failed
check_models <- function(models) {
  model_status <- map(models, ~!inherits(., "try-error"))
  successful_models <- names(models)[unlist(model_status)]
  failed_models <- names(models)[!unlist(model_status)]
  
  return(list(
    successful = successful_models,
    failed = failed_models
  ))
}

# Check status of all models
model_status <- map(all_models, check_models)
model_status

```
```{r}
# Function to check and summarize all GAM models
check_and_summarize_models <- function(models) {
  # Apply gam.check and summary to each model
  purrr::map(models, function(model_list) {
    purrr::map(model_list, ~ {
      # Perform a check on the GAM model
      gam_check_result <- try(gam.check(.), silent = TRUE)
      
      # If gam.check runs without error, print it
      if (inherits(gam_check_result, "try-error")) {
        message("Error in gam.check for model: ", deparse(substitute(.)))
      } else {
        print(gam_check_result)
      }
      
      # Get the summary of the model
      summary_result <- try(summary(.), silent = TRUE)
      
      # If summary runs without error, print it
      if (inherits(summary_result, "try-error")) {
        message("Error in summary for model: ", deparse(substitute(.)))
      } else {
        print(summary_result)
      }
    })
  })
}

# Example usage:
check_and_summarize_models(all_models)
```





######################################################################
######################################################################
#####################################################################
2000 vs 2010 income checking
```{r}
library(tidyverse)
library(stats)

analyze_tract_incomes <- function(all_data, income_2000, income_2010) {
  # Merge data
  analysis_df <- all_data %>%
    left_join(income_2000 %>% select(GEOID, med_income_2000), by = "GEOID") %>%
    left_join(income_2010 %>% select(GEOID, med_income_2010), by = "GEOID")
  
  # Create summary statistics grouped by whether 2000 data exists
  summary_stats <- analysis_df %>%
    mutate(has_2000_data = !is.na(med_income_2000)) %>%
    group_by(has_2000_data) %>%
    summarise(
      n_tracts = dplyr::n(),
      mean_2010 = mean(med_income_2010, na.rm = TRUE),
      median_2010 = median(med_income_2010, na.rm = TRUE),
      sd_2010 = sd(med_income_2010, na.rm = TRUE),
      q25_2010 = quantile(med_income_2010, 0.25, na.rm = TRUE),
      q75_2010 = quantile(med_income_2010, 0.75, na.rm = TRUE)
    )
  
  # Perform Wilcoxon rank sum test (Mann-Whitney U test)
  test_result <- wilcox.test(
    med_income_2010 ~ !is.na(med_income_2000),
    data = analysis_df,
    alternative = "two.sided"
  )
  
  # Create visualization
  plot <- analysis_df %>%
    mutate(has_2000_data = if_else(!is.na(med_income_2000), 
                                  "Has 2000 Data", 
                                  "Missing 2000 Data")) %>%
    ggplot(aes(x = has_2000_data, y = med_income_2010)) +
    geom_boxplot() +
    theme_minimal() +
    labs(
      title = "2010 Income Distribution: Tracts With vs Without 2000 Data",
      x = "",
      y = "2010 Median Income"
    )
  
  # Calculate percent difference in means
  pct_diff <- with(summary_stats, 
                  (mean_2010[!has_2000_data] - mean_2010[has_2000_data]) / 
                    mean_2010[has_2000_data] * 100)
  
  # Return results
  list(
    summary_stats = summary_stats,
    wilcox_test = test_result,
    percent_difference = pct_diff,
    plot = plot
  )
}

# Use the function
results <- analyze_tract_incomes(all_data, income, income_2010)

# Print results
print("Summary Statistics:")
print(results$summary_stats)

print("\nWilcoxon Test Results:")
print(results$wilcox_test)

print(sprintf("\nPercent difference in means: %.2f%%", results$percent_difference))

# Display plot
print(results$plot)
print(results$plot)

#additional analysis based on average, variation, standard deviation- checking normality to see which statistical signifiancetest is appropriate

# Create analysis dataset
analysis_df <- all_data %>%
  left_join(income %>% select(GEOID, med_income_2000), by = "GEOID") %>%
  left_join(income_2010 %>% select(GEOID, med_income_2010), by = "GEOID") %>%
  mutate(has_2000_data = !is.na(med_income_2000))

# Calculate summary statistics
stats_summary <- analysis_df %>%
  group_by(has_2000_data) %>%
  summarise(
    n = dplyr::n(),
    mean_2010 = mean(med_income_2010, na.rm = TRUE),
    var_2010 = var(med_income_2010, na.rm = TRUE),
    sd_2010 = sd(med_income_2010, na.rm = TRUE)
  ) %>%
  mutate(across(where(is.numeric), ~round(., 2)))

# Test for normality using Shapiro-Wilk test
normality_test_with_2000 <- shapiro.test(analysis_df$med_income_2010[analysis_df$has_2000_data])
normality_test_without_2000 <- shapiro.test(analysis_df$med_income_2010[!analysis_df$has_2000_data])

# Perform F-test for equality of variances
var_test <- var.test(med_income_2010 ~ has_2000_data, data = analysis_df)

# Perform t-test (both equal and unequal variance versions)
t_test_equal_var <- t.test(med_income_2010 ~ has_2000_data, data = analysis_df, var.equal = TRUE)
t_test_unequal_var <- t.test(med_income_2010 ~ has_2000_data, data = analysis_df, var.equal = FALSE)

# Perform Wilcoxon rank sum test
wilcox_test <- wilcox.test(med_income_2010 ~ has_2000_data, data = analysis_df)

# Print results
cat("Summary Statistics:\n")
print(stats_summary)

cat("\nNormality Tests (Shapiro-Wilk):\n")
cat("Tracts with 2000 data: p =", normality_test_with_2000$p.value, "\n")
cat("Tracts without 2000 data: p =", normality_test_without_2000$p.value, "\n")

cat("\nF-test for Equal Variances:\n")
cat("F =", round(var_test$statistic, 3), ", p =", var_test$p.value, "\n")

cat("\nt-test (Equal Variances):\n")
cat("t =", round(t_test_equal_var$statistic, 3), ", p =", t_test_equal_var$p.value, "\n")

cat("\nt-test (Unequal Variances):\n")
cat("t =", round(t_test_unequal_var$statistic, 3), ", p =", t_test_unequal_var$p.value, "\n")

cat("\nWilcoxon Rank Sum Test:\n")
cat("W =", wilcox_test$statistic, ", p =", wilcox_test$p.value, "\n")
```

```{r}
# Construct neighbors list for MRF models
nb <- poly2nb(all_data_inc, row.names = all_data_inc$GEOID)
names(nb) <- attr(nb, "region.id") 
```


testing tp model with med_income_2000 for pm2.5
```{r}
tp_income_pm_na <- gam(pm_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + s(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2), method = 'REML', family = scat, data = all_data_inc_na)

# Try increasing k for the tensor product smooth of longitude and latitude (from 24 default)
tp_income_pm_na_2 <- gam(pm_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + s(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 100), method = 'REML', family = scat, data = all_data_inc_na) #when k=40 and k=100, p value still significant (= bad)

tp_income_pm_na_3 <- gam(pm_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 150), method = 'REML', family = scat, data = all_data_inc_na) 

tp_income_pm_na_4 <- gam(pm_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 200), method = 'REML', family = scat, data = all_data_inc_na) 

tp_income_pm_na_4 <- gam(pm_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 200), method = 'REML', family = scat, data = all_data_inc_na) 

tp_income_pm_na_5 <- gam(pm_change ~ score_0.5 + s(COUNTYFP10, bs = "re", k=10) + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 200), method = 'REML', family = scat, data = all_data_inc_na) 



#model diagnostics using gam.check
gam.check(tp_income_pm_na)
gam.check(tp_income_pm_na_2)
gam.check(tp_income_pm_na_3)
gam.check(tp_income_pm_na_4)
gam.check(tp_income_pm_na_5)


# First, let's get the fitted random effects properly
fitted_terms <- predict(tp_income_pm_na_5, type="terms")
county_term <- fitted_terms[,"s(COUNTYFP10)"]  # Extract county random effects

# Create summary by county
county_summary <- data.frame(
    COUNTYFP10 = all_data_inc_na$COUNTYFP10,
    effect = county_term
) %>%
    group_by(COUNTYFP10) %>%
    summarize(mean_effect = mean(effect)) %>%
    arrange(desc(abs(mean_effect)))

# View the most extreme counties
print(head(county_summary))

# Visualize
library(ggplot2)
ggplot(county_summary, aes(y=mean_effect, x=reorder(COUNTYFP10, mean_effect))) +
    geom_point() +
    theme_minimal() +
    theme(axis.text.x = element_blank()) +  # Hide county codes for clarity
    labs(x="Counties", y="Random Effect")

# Look at the data for the most extreme county
extreme_county <- county_summary$COUNTYFP10[1]
extreme_county_data <- all_data_inc_na %>%
    filter(COUNTYFP10 == extreme_county) %>%
    select(pm_change, score_0.5, med_income_2000, long, lat)

print("Data for most extreme county:")
print(summary(extreme_county_data)) #extreme county is 85 - Manhattan/NY county

print("Overall data summary:")
print(summary(all_data_inc_na[,c("pm_change", "score_0.5", "med_income_2000")]))
####

####

# Calculate Cook's distance
cooks_d <- cooks.distance(tp_income_pm_na)
plot(cooks_d, type = "h", main = "Cook's Distance")
abline(h = 4/length(cooks_d), col = "red")  # Rule of thumb threshold

# Identify potential influential points
influential <- which(cooks_d > 4/length(cooks_d))

influential_data <- all_data_inc_na[influential, ]
print(influential_data)  # Look at characteristics of these points

# Options for handling influential points:
# a. Fit model without them to compare
tp_income_pm_na_reduced <- gam(
  pm_change ~ score_0.5 + 
    s(COUNTYFP10, bs = "re") + 
    rank(med_income_2000) + 
    te(long, lat, d = 2, bs = 'gp', m = 2, k = 150),
  method = 'REML',
  family = scat,
  data = all_data_inc_na[-influential, ]
)

# Compare coefficients and predictions
summary(tp_income_pm_na)
summary(tp_income_pm_na_reduced)

# b. Investigate why these points are influential:
# Plot their locations
plot(all_data_inc_na$long, all_data_inc_na$lat)
points(all_data_inc_na$long[influential], 
       all_data_inc_na$lat[influential], 
       col = "red", pch = 16)


# Checking that scat is indeed better (comparing t distribution to gaussian)
model_scat <- gam(
  pm_change ~ score_0.5 + 
    s(COUNTYFP10, bs = "re") + 
    rank(med_income_2000) + 
    te(long, lat, d = 2, bs = 'gp', m = 2, k = 150),
  method = 'REML',
  family = scat,
  data = all_data_inc_na
)

model_gaussian <- gam(
  pm_change ~ score_0.5 + 
    s(COUNTYFP10, bs = "re") + 
    rank(med_income_2000) + 
    te(long, lat, d = 2, bs = 'gp', m = 2, k = 150),
  method = 'REML',
  family = gaussian,
  data = all_data_inc_na
)

# Compare QQ plots
par(mfrow=c(1,2))
qq.gam(model_scat, main="Scaled t")
qq.gam(model_gaussian, main="Gaussian")

# Compare predictions
pred_scat <- predict(model_scat)
pred_gaussian <- predict(model_gaussian)
plot(pred_scat, pred_gaussian, 
     xlab="Scaled t predictions", 
     ylab="Gaussian predictions")
abline(0,1, col="red")

# Compare AIC (if models have same scale)
AIC(model_scat, model_gaussian)
# Check if residual patterns differ
par(mfrow=c(2,2))
gam.check(tp_income_pm_gaussian)


```

testing tp models for o3
```{r}
tp_income_o3_na <- gam(o3_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + s(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2), method = 'REML', family = scat, data = all_data_inc_na)


#increasing k in tp
tp_income_o3_na_2 <- gam(o3_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 150), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_o3_na_3 <- gam(o3_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 250), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_o3_na_4 <- gam(o3_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 200), method = 'REML', family = scat, data = all_data_inc_na) # k still 0.02

tp_income_o3_na_5 <- gam(o3_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 225), method = 'REML', family = scat, data = all_data_inc_na) #k no longer significant


gam.check(tp_income_o3_na)
gam.check(tp_income_o3_na_2)
gam.check(tp_income_o3_na_3)
gam.check(tp_income_o3_na_4)
gam.check(tp_income_o3_na_5)




```

testing tp models for no2
```{r}
tp_income_no2_na <- gam(no2_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + s(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_no2_na_2 <- gam(no2_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k= 200), method = 'REML', family = scat, data = all_data_inc_na) #still significant

tp_income_no2_na_3 <- gam(no2_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k= 250), method = 'REML', family = scat, data = all_data_inc_na) #still significant
 
tp_income_no2_na_4 <- gam(no2_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k= 450), method = 'REML', family = scat, data = all_data_inc_na) #takes forever to run - not feasible

gam.check(tp_income_no2_na)
gam.check(tp_income_no2_na_2)
gam.check(tp_income_no2_na_3)
gam.check(tp_income_no2_na_4)


```
testing tp models for ndvi
```{r}
tp_income_ndvi_na <- gam(ndvi_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + s(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_ndvi_na_2 <- gam(ndvi_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 200), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_ndvi_na_3 <- gam(ndvi_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 300), method = 'REML', family = scat, data = all_data_inc_na) # still very significant

gam.check(tp_income_ndvi_na)
gam.check(tp_income_ndvi_na_2)
gam.check(tp_income_ndvi_na_3)

```
testing tp models for temp
```{r}
tp_income_temp_na <- gam(mean_temp ~ score_0.5 + s(COUNTYFP10, bs = "re") + s(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_temp_na_2<- gam(mean_temp ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 200), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_temp_na_3<- gam(mean_temp ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 300), method = 'REML', family = scat, data = all_data_inc_na) # still very significant


gam.check(tp_income_temp_na)
gam.check(tp_income_temp_na_2)
gam.check(tp_income_temp_na_3)
```

Testing residuals using tp model for now
```{r}
# Extract residuals for each model
resid_pm <- resid(tp_income_pm_na_5)
resid_o3 <- resid(tp_income_o3_na_5)
resid_no2 <- resid(tp_income_no2_na_3)
resid_ndvi <- resid(tp_income_ndvi_na_3)
resid_temp <- resid(tp_income_ndvi_na)


# Add residuals to the original dataset (ensure this corresponds to the right model)
all_data_inc_na$resid_pm <- resid_pm
all_data_inc_na$resid_o3 <- resid_o3
all_data_inc_na$resid_no2 <- resid_no2
all_data_inc_na$resid_ndvi <- resid_ndvi
all_data_inc_na$resid_temp <- resid_temp
```

Code for just PM to base function on
```{r}
# # Visualize residuals for PM model, for example
# ggplot(all_data_inc_na, aes(x = long, y = lat, color = resid_pm)) +
#   geom_point() +
#   scale_color_gradient2(midpoint = 0, low = "blue", mid = "white", high = "red") +
#   theme_minimal() +
#   labs(title = "Spatial Distribution of Residuals (PM)", color = "Residuals")
#   
# # Calculate the 10th and 90th percentiles of the residuals
# lower_threshold <- quantile(all_data_inc_na$resid_pm, 0.1)
# upper_threshold <- quantile(all_data_inc_na$resid_pm, 0.9)
# 
# # Create dichotomous variable based on thresholds
# all_data_inc_na$residual_gap <- ifelse(all_data_inc_na$resid_pm > upper_threshold, "High Residual", 
#                                         ifelse(all_data_inc_na$resid_pm < lower_threshold, "Low Residual", "Normal"))
# 
# # Visualize the result
# ggplot(all_data_inc_na, aes(x = long, y = lat, color = residual_gap)) +
#   geom_point() +
#   scale_color_manual(values = c("High Residual" = "red", "Low Residual" = "blue", "Normal" = "grey")) +
#   theme_minimal() +
#   labs(title = "Spatial Distribution of Residual Gaps (PM)", color = "Residual Gap")
# 
# 
# # Calculate the 10th and 90th percentiles of the residuals
# lower_threshold <- quantile(all_data_inc_na$resid_pm, 0.1)
# upper_threshold <- quantile(all_data_inc_na$resid_pm, 0.9)
# 
# # Create dichotomous variable based on thresholds
# all_data_inc_na$residual_gap <- ifelse(all_data_inc_na$resid_pm > upper_threshold, "High Residual", 
#                                         ifelse(all_data_inc_na$resid_pm < lower_threshold, "Low Residual", "Normal"))
# 
# # Visualize the result
# ggplot(all_data_inc_na, aes(x = long, y = lat, color = residual_gap)) +
#   geom_point() +
#   scale_color_manual(values = c("High Residual" = "red", "Low Residual" = "blue", "Normal" = "grey")) +
#   theme_minimal() +
#   labs(title = "Spatial Distribution of Residual Gaps (PM)", color = "Residual Gap")
# 
# 
# # Visualize gaps using the dichotomous color scale
# ggplot(all_data_inc_na, aes(x = long, y = lat, color = residual_gap)) +
#   geom_point() +
#   scale_color_manual(values = c("High Residual" = "red", "Low Residual" = "blue", "Normal" = "grey")) +
#   theme_minimal() +
#   labs(title = "Spatial Distribution of Residual Gaps (PM)", color = "Residual Gap")
# 
# # Summary statistics for residuals (e.g., PM)
# summary(all_data_inc_na$resid_pm)
# 
# # Identify the locations with extreme residuals
# high_residuals <- all_data_inc_na[all_data_inc_na$resid_pm > threshold, ]
# low_residuals <- all_data_inc_na[all_data_inc_na$resid_pm < -threshold, ]
# 
# # View the locations of high and low residuals
# head(high_residuals[, c("long", "lat", "resid_pm")])
# head(low_residuals[, c("long", "lat", "resid_pm")])


```
############################################
function to analyze residuals for all exposures based on tp model (based on high and low quantiles)
```{r}
analyze_spatial_residuals <- function(data, residual_col, lon_col = "long", lat_col = "lat", 
                                    lower_quantile = 0.1, upper_quantile = 0.9) {
  # Input validation
  if (!all(c(residual_col, lon_col, lat_col) %in% names(data))) {
    stop("One or more specified columns not found in data")
  }
  
  # Create analysis environment
  results <- list()
  
  # Basic summary statistics
  results$summary <- summary(data[[residual_col]])
  
  # Calculate thresholds
  results$thresholds <- list(
    lower = quantile(data[[residual_col]], lower_quantile, na.rm = TRUE),
    upper = quantile(data[[residual_col]], upper_quantile, na.rm = TRUE)
  )
  
  # Create categorical variable for residual gaps
  data$residual_category <- case_when(
    data[[residual_col]] > results$thresholds$upper ~ "High Residual",
    data[[residual_col]] < results$thresholds$lower ~ "Low Residual",
    TRUE ~ "Normal"
  )
  
  # Generate plots
  # 1. Continuous residual plot
  results$continuous_plot <- ggplot(data, aes(x = .data[[lon_col]], 
                                            y = .data[[lat_col]], 
                                            color = .data[[residual_col]])) +
    geom_point(alpha = 0.6) +
    scale_color_gradient2(midpoint = 0, 
                         low = "blue", 
                         mid = "white", 
                         high = "red") +
    theme_minimal() +
    labs(title = paste("Spatial Distribution of Residuals:", residual_col),
         color = "Residuals") +
    theme(legend.position = "right")
  
  # 2. Categorical residual plot
  results$categorical_plot <- ggplot(data, aes(x = .data[[lon_col]], 
                                             y = .data[[lat_col]], 
                                             color = residual_category)) +
    geom_point(alpha = 0.6) +
    scale_color_manual(values = c("High Residual" = "red", 
                                 "Low Residual" = "blue", 
                                 "Normal" = "grey")) +
    theme_minimal() +
    labs(title = paste("Spatial Distribution of Residual Gaps:", residual_col),
         color = "Residual Category") +
    theme(legend.position = "right")
  
  # Extract extreme locations
  results$extreme_locations <- list(
    high = data %>%
      filter(residual_category == "High Residual") %>%
      select(all_of(c(lon_col, lat_col, residual_col))) %>%
      arrange(desc(.data[[residual_col]])),
    
    low = data %>%
      filter(residual_category == "Low Residual") %>%
      select(all_of(c(lon_col, lat_col, residual_col))) %>%
      arrange(.data[[residual_col]])
  )
  
  # Calculate summary statistics by category
  results$category_stats <- data %>%
    group_by(residual_category) %>%
    summarise(
      count = dplyr::n(),
      mean_residual = mean(.data[[residual_col]], na.rm = TRUE),
      sd_residual = sd(.data[[residual_col]], na.rm = TRUE),
      min_residual = min(.data[[residual_col]], na.rm = TRUE),
      max_residual = max(.data[[residual_col]], na.rm = TRUE),
      .groups = "drop"
    )
  
  return(results)
}

# Example usage for each environmental variable
results_pm <- analyze_spatial_residuals(all_data_inc_na, "resid_pm")
results_o3 <- analyze_spatial_residuals(all_data_inc_na, "resid_o3")
results_no2 <- analyze_spatial_residuals(all_data_inc_na, "resid_no2")
results_ndvi <- analyze_spatial_residuals(all_data_inc_na, "resid_ndvi")
results_temp <- analyze_spatial_residuals(all_data_inc_na, "resid_temp")


# Display results for pm2.5
print(results_pm$summary)
print(results_pm$category_stats)
results_pm$continuous_plot
results_pm$categorical_plot
head(results_pm$extreme_locations$high)
head(results_pm$extreme_locations$low)

print(results_o3$summary)
print(results_o3$category_stats)
results_o3$continuous_plot
results_o3$categorical_plot
head(results_o3$extreme_locations$high)
head(results_o3$extreme_locations$low)

print(results_no2$summary)
print(results_no2$category_stats)
results_no2$continuous_plot
results_no2$categorical_plot
head(results_no2$extreme_locations$high)
head(results_no2$extreme_locations$low)

print(results_ndvi$summary)
print(results_ndvi$category_stats)
results_ndvi$continuous_plot
results_ndvi$categorical_plot
head(results_ndvi$extreme_locations$high)
head(results_ndvi$extreme_locations$low)

print(results_temp$summary)
print(results_temp$category_stats)
results_temp$continuous_plot
results_temp$categorical_plot
head(results_temp$extreme_locations$high)
head(results_temp$extreme_locations$low)
```
function to analyze residuals for all exposures based on tp model (based on high and low quantiles, narrower limits)
```{r}
analyze_spatial_residuals <- function(data, residual_col, lon_col = "long", lat_col = "lat", 
                                    lower_quantile = 0.05, upper_quantile = 0.95) {
  # Input validation
  if (!all(c(residual_col, lon_col, lat_col) %in% names(data))) {
    stop("One or more specified columns not found in data")
  }
  
  # Create analysis environment
  results <- list()
  
  # Basic summary statistics
  results$summary <- summary(data[[residual_col]])
  
  # Calculate thresholds
  results$thresholds <- list(
    lower = quantile(data[[residual_col]], lower_quantile, na.rm = TRUE),
    upper = quantile(data[[residual_col]], upper_quantile, na.rm = TRUE)
  )
  
  # Create categorical variable for residual gaps
  data$residual_category <- case_when(
    data[[residual_col]] > results$thresholds$upper ~ "High Residual",
    data[[residual_col]] < results$thresholds$lower ~ "Low Residual",
    TRUE ~ "Normal"
  )
  
  # Generate plots
  # 1. Continuous residual plot
  results$continuous_plot <- ggplot(data, aes(x = .data[[lon_col]], 
                                            y = .data[[lat_col]], 
                                            color = .data[[residual_col]])) +
    geom_point(alpha = 0.6) +
    scale_color_gradient2(midpoint = 0, 
                         low = "blue", 
                         mid = "white", 
                         high = "red") +
    theme_minimal() +
    labs(title = paste("Spatial Distribution of Residuals:", residual_col),
         color = "Residuals") +
    theme(legend.position = "right")
  
  # 2. Categorical residual plot
  results$categorical_plot <- ggplot(data, aes(x = .data[[lon_col]], 
                                             y = .data[[lat_col]], 
                                             color = residual_category)) +
    geom_point(alpha = 0.6) +
    scale_color_manual(values = c("High Residual" = "red", 
                                 "Low Residual" = "blue", 
                                 "Normal" = "grey")) +
    theme_minimal() +
    labs(title = paste("Spatial Distribution of Residual Gaps:", residual_col),
         color = "Residual Category") +
    theme(legend.position = "right")
  
  # Extract extreme locations
  results$extreme_locations <- list(
    high = data %>%
      filter(residual_category == "High Residual") %>%
      select(all_of(c(lon_col, lat_col, residual_col))) %>%
      arrange(desc(.data[[residual_col]])),
    
    low = data %>%
      filter(residual_category == "Low Residual") %>%
      select(all_of(c(lon_col, lat_col, residual_col))) %>%
      arrange(.data[[residual_col]])
  )
  
  # Calculate summary statistics by category
  results$category_stats <- data %>%
    group_by(residual_category) %>%
    summarise(
      count = dplyr::n(),
      mean_residual = mean(.data[[residual_col]], na.rm = TRUE),
      sd_residual = sd(.data[[residual_col]], na.rm = TRUE),
      min_residual = min(.data[[residual_col]], na.rm = TRUE),
      max_residual = max(.data[[residual_col]], na.rm = TRUE),
      .groups = "drop"
    )
  
  return(results)
}

# Example usage for each environmental variable
results_pm <- analyze_spatial_residuals(all_data_inc_na, "resid_pm")
results_o3 <- analyze_spatial_residuals(all_data_inc_na, "resid_o3")
results_no2 <- analyze_spatial_residuals(all_data_inc_na, "resid_no2")
results_ndvi <- analyze_spatial_residuals(all_data_inc_na, "resid_ndvi")
results_temp <- analyze_spatial_residuals(all_data_inc_na, "resid_temp")


# Display results for pm2.5
print(results_pm$summary)
print(results_pm$category_stats)
results_pm$continuous_plot
results_pm$categorical_plot
head(results_pm$extreme_locations$high)
head(results_pm$extreme_locations$low)

print(results_o3$summary)
print(results_o3$category_stats)
results_o3$continuous_plot
results_o3$categorical_plot
head(results_o3$extreme_locations$high)
head(results_o3$extreme_locations$low)

print(results_no2$summary)
print(results_no2$category_stats)
results_no2$continuous_plot
results_no2$categorical_plot
head(results_no2$extreme_locations$high)
head(results_no2$extreme_locations$low)

print(results_ndvi$summary)
print(results_ndvi$category_stats)
results_ndvi$continuous_plot
results_ndvi$categorical_plot
head(results_ndvi$extreme_locations$high)
head(results_ndvi$extreme_locations$low)

print(results_temp$summary)
print(results_temp$category_stats)
results_temp$continuous_plot
results_temp$categorical_plot
head(results_temp$extreme_locations$high)
head(results_temp$extreme_locations$low)
```
Analyze residual overlap between exposures
```{r}
analyze_residual_overlap <- function(data, residual_cols, lon_col = "long", lat_col = "lat", 
                                   lower_quantile = 0.1, upper_quantile = 0.9) {
  # Input validation
  if (!all(c(residual_cols, lon_col, lat_col) %in% names(data))) {
    stop("One or more specified columns not found in data")
  }
  
  # Create category columns for each residual
  results <- list()
  data_enhanced <- data
  
  # Categorize each residual
  for (col in residual_cols) {
    thresholds <- quantile(data[[col]], c(lower_quantile, upper_quantile), na.rm = TRUE)
    cat_col_name <- paste0(col, "_category")
    
    data_enhanced[[cat_col_name]] <- case_when(
      data[[col]] > thresholds[2] ~ "High",
      data[[col]] < thresholds[1] ~ "Low",
      TRUE ~ "Normal"
    )
  }
  
  # Count overlapping high/low/extreme areas
  data_enhanced$high_overlap_count <- rowSums(
    sapply(residual_cols, function(col) {
      data_enhanced[[paste0(col, "_category")]] == "High"
    })
  )
  
  data_enhanced$low_overlap_count <- rowSums(
    sapply(residual_cols, function(col) {
      data_enhanced[[paste0(col, "_category")]] == "Low"
    })
  )
  
  # Count either high or low (extreme) overlaps
  data_enhanced$extreme_overlap_count <- rowSums(
    sapply(residual_cols, function(col) {
      data_enhanced[[paste0(col, "_category")]] %in% c("High", "Low")
    })
  )
  
  # Create overlap visualizations
  # 1. High residual overlap plot
  results$high_overlap_plot <- ggplot(data_enhanced, 
         aes(x = .data[[lon_col]], y = .data[[lat_col]], 
             color = factor(high_overlap_count))) +
    geom_point(alpha = 0.6) +
    scale_color_viridis_d(option = "magma", direction = -1) +
    theme_minimal() +
    labs(title = "Overlap of High Residuals",
         color = "Number of\nOverlapping\nHigh Residuals") +
    theme(legend.position = "right")
  
  # 2. Low residual overlap plot
  results$low_overlap_plot <- ggplot(data_enhanced, 
         aes(x = .data[[lon_col]], y = .data[[lat_col]], 
             color = factor(low_overlap_count))) +
    geom_point(alpha = 0.6) +
    scale_color_viridis_d(option = "mako", direction = -1) +
    theme_minimal() +
    labs(title = "Overlap of Low Residuals",
         color = "Number of\nOverlapping\nLow Residuals") +
    theme(legend.position = "right")
  
  # 3. Extreme (either high or low) residual overlap plot
  results$extreme_overlap_plot <- ggplot(data_enhanced, 
         aes(x = .data[[lon_col]], y = .data[[lat_col]], 
             color = factor(extreme_overlap_count))) +
    geom_point(alpha = 0.6) +
    scale_color_viridis_d(option = "viridis", direction = -1) +
    theme_minimal() +
    labs(title = "Overlap of Extreme Residuals (High or Low)",
         color = "Number of\nOverlapping\nExtreme Residuals") +
    theme(legend.position = "right")
  
  # Create summary of overlaps
  results$overlap_summary <- data_enhanced %>%
    summarise(
      locations_with_no_extremes = sum(extreme_overlap_count == 0),
      locations_with_one_extreme = sum(extreme_overlap_count == 1),
      locations_with_multiple_extremes = sum(extreme_overlap_count > 1),
      max_extreme_overlap = max(extreme_overlap_count),
      locations_with_no_high = sum(high_overlap_count == 0),
      locations_with_one_high = sum(high_overlap_count == 1),
      locations_with_multiple_high = sum(high_overlap_count > 1),
      max_high_overlap = max(high_overlap_count),
      locations_with_no_low = sum(low_overlap_count == 0),
      locations_with_one_low = sum(low_overlap_count == 1),
      locations_with_multiple_low = sum(low_overlap_count > 1),
      max_low_overlap = max(low_overlap_count)
    )
  
  # Create detailed overlap patterns
  results$overlap_patterns <- data_enhanced %>%
    group_by(extreme_overlap_count) %>%
    summarise(
      count = dplyr::n(),
      pct_of_total = round(100 * dplyr::n() / nrow(data_enhanced), 1),
      .groups = "drop"
    ) %>%
    arrange(desc(count))
  
  # Add the enhanced dataset to results
  results$data <- data_enhanced
  
  return(results)
}

# Example usage
residual_columns <- c("resid_pm", "resid_o3", "resid_no2", "resid_ndvi")
overlap_results <- analyze_residual_overlap(all_data_inc_na, residual_columns)

# Display summary statistics
print("Summary of overlapping patterns:")
print(overlap_results$overlap_summary)

print("\nDistribution of extreme residual overlaps:")
print(overlap_results$overlap_patterns)

# Create a combined visualization using patchwork
library(patchwork)
combined_plot <- overlap_results$high_overlap_plot + 
  overlap_results$low_overlap_plot +
  overlap_results$extreme_overlap_plot +
  plot_layout(ncol = 3) +
  plot_annotation(
    title = "Spatial Patterns of Overlapping Residuals",
    subtitle = "Left: High Residuals, Center: Low Residuals, Right: Combined Extreme Residuals"
  )

print(combined_plot)

# Find locations with multiple extreme residuals
extreme_overlap_locations <- overlap_results$data %>%
  filter(extreme_overlap_count >= 2) %>%
  select(long, lat, extreme_overlap_count, 
         ends_with("_category")) %>%
  arrange(desc(extreme_overlap_count))

print("Locations with multiple extreme residuals:")
head(extreme_overlap_locations, 10)

# Create a summary table of which combinations of exposures tend to have extreme residuals together
exposure_combinations <- overlap_results$data %>%
  filter(extreme_overlap_count >= 2) %>%
  summarise(
    PM_O3 = sum((resid_pm_category != "Normal") & (resid_o3_category != "Normal")),
    PM_NO2 = sum((resid_pm_category != "Normal") & (resid_no2_category != "Normal")),
    PM_NDVI = sum((resid_pm_category != "Normal") & (resid_ndvi_category != "Normal")),
    O3_NO2 = sum((resid_o3_category != "Normal") & (resid_no2_category != "Normal")),
    O3_NDVI = sum((resid_o3_category != "Normal") & (resid_ndvi_category != "Normal")),
    NO2_NDVI = sum((resid_no2_category != "Normal") & (resid_ndvi_category != "Normal"))
  )

print("\nCommon pairs of exposures with extreme residuals:")
print(exposure_combinations)
```
More detailed summaries on where residuals overlap
```{r}
analyze_residual_overlap <- function(data, residual_cols, lon_col = "long", lat_col = "lat", 
                                   geoid_col = "geoid", county_col = "NAME", # 
                                   lower_quantile = 0.1, upper_quantile = 0.9) {
  # Input validation
  if (!all(c(residual_cols, lon_col, lat_col, geoid_col, county_col) %in% names(data))) {
    stop("One or more specified columns not found in data")
  }
  
  # Create category columns for each residual
  results <- list()
  data_enhanced <- data
  
  # Categorize each residual
  for (col in residual_cols) {
    thresholds <- quantile(data[[col]], c(lower_quantile, upper_quantile), na.rm = TRUE)
    cat_col_name <- paste0(col, "_category")
    
    data_enhanced[[cat_col_name]] <- case_when(
      data[[col]] > thresholds[2] ~ "High",
      data[[col]] < thresholds[1] ~ "Low",
      TRUE ~ "Normal"
    )
  }
  
  # Count overlapping high/low/extreme areas
  data_enhanced$high_overlap_count <- rowSums(
    sapply(residual_cols, function(col) {
      data_enhanced[[paste0(col, "_category")]] == "High"
    })
  )
  
  data_enhanced$low_overlap_count <- rowSums(
    sapply(residual_cols, function(col) {
      data_enhanced[[paste0(col, "_category")]] == "Low"
    })
  )
  
  # Count either high or low (extreme) overlaps
  data_enhanced$extreme_overlap_count <- rowSums(
    sapply(residual_cols, function(col) {
      data_enhanced[[paste0(col, "_category")]] %in% c("High", "Low")
    })
  )
  
  # Create overlap visualizations
  # 1. High residual overlap plot
  results$high_overlap_plot <- ggplot(data_enhanced, 
         aes(x = .data[[lon_col]], y = .data[[lat_col]], 
             color = factor(high_overlap_count))) +
    geom_point(alpha = 0.6) +
    scale_color_viridis_d(option = "magma", direction = -1) +
    theme_minimal() +
    labs(title = "Overlap of High Residuals",
         color = "Number of\nOverlapping\nHigh Residuals") +
    theme(legend.position = "right")
  
  # Create location summaries with geoids and county names
  results$location_summary <- data_enhanced %>%
    filter(extreme_overlap_count > 0) %>%
    select(
      !!sym(geoid_col),
      !!sym(county_col), # Added county name
      !!sym(lon_col),
      !!sym(lat_col),
      extreme_overlap_count,
      high_overlap_count,
      low_overlap_count,
      ends_with("_category")
    ) %>%
    arrange(desc(extreme_overlap_count))
  
  # Summary of overlap patterns by geoid and county
  results$overlap_patterns <- data_enhanced %>%
    group_by(extreme_overlap_count) %>%
    summarise(
      n_locations = dplyr::n(),
      geoids = list(get(geoid_col)),
      counties = list(unique(get(county_col))), # Added county names
      pct_of_total = round(100 * dplyr::n() / nrow(data_enhanced), 1),
      .groups = "drop"
    ) %>%
    arrange(desc(extreme_overlap_count))
  
  # Create detailed overlap combinations for high overlap areas
  results$overlap_combinations <- data_enhanced %>%
    filter(extreme_overlap_count >= 2) %>%
    group_by(!!sym(geoid_col)) %>%
    summarise(
      county = first(!!sym(county_col)), # Added county name
      extreme_overlap_count = first(extreme_overlap_count),
      combination = paste(names(which(sapply(residual_cols, function(col) {
        first(get(paste0(col, "_category"))) != "Normal"
      }))), collapse = ", "),
      categories = paste(sapply(residual_cols, function(col) {
        paste0(col, ": ", first(get(paste0(col, "_category"))))
      }), collapse = "; "),
      .groups = "drop"
    ) %>%
    arrange(desc(extreme_overlap_count))
  
  # Add the enhanced dataset to results
  results$data <- data_enhanced
  
  return(results)
}

overlap_results <- analyze_residual_overlap(
   data = all_data_inc_na,
   residual_cols = c("resid_pm", "resid_o3", "resid_no2", "resid_ndvi"),
   geoid_col = "GEOID",
   county_col = "NAME"
 )

# View locations with overlaps
print(overlap_results$location_summary)

# View overlap patterns
print(overlap_results$overlap_patterns)

# View detailed combinations
print(overlap_results$overlap_combinations)



```

```{r}
library(spdep)  # for spatial weights and Moran's I
library(dbscan) # for DBSCAN clustering
library(sf)     # for spatial operations
library(ggplot2)
library(dplyr)

analyze_residual_clusters <- function(data, residual_cols, lon_col = "long", lat_col = "lat") {
  results <- list()
  
  # Debug data structure
  cat("Data structure diagnostics:\n")
  cat("Class of input data:", class(data), "\n")
  cat("Dimension of input data:", dim(data), "\n")
  cat("Classes of coordinate columns:\n")
  print(sapply(data[c(lon_col, lat_col)], class))
  
  # Input validation
  if (!all(c(lon_col, lat_col) %in% names(data))) {
    stop("Longitude and latitude columns not found in data")
  }
  
  if (!all(residual_cols %in% names(data))) {
    stop("One or more residual columns not found in data")
  }
  
  # Function to safely convert potentially nested columns to numeric
  safe_to_numeric <- function(x) {
    if (is.list(x)) {
      # If it's a list column, try to unlist it first
      x <- unlist(x)
    }
    # Try to convert to character first, then to numeric
    as.numeric(as.character(x))
  }
  
  # Extract and validate coordinates
  coords <- tryCatch({
    # Create a clean data frame for coordinates
    coords_df <- data.frame(
      longitude = safe_to_numeric(data[[lon_col]]),
      latitude = safe_to_numeric(data[[lat_col]])
    )
    
    # Check for conversion success
    if (any(is.na(coords_df$longitude)) && !all(is.na(coords_df$longitude))) {
      warning(paste("Some longitude values could not be converted to numeric. First problematic value:",
                   data[[lon_col]][which(!is.na(data[[lon_col]]) & is.na(coords_df$longitude))[1]]))
    }
    if (any(is.na(coords_df$latitude)) && !all(is.na(coords_df$latitude))) {
      warning(paste("Some latitude values could not be converted to numeric. First problematic value:",
                   data[[lat_col]][which(!is.na(data[[lat_col]]) & is.na(coords_df$latitude))[1]]))
    }
    
    # Remove rows with NA coordinates
    valid_rows <- complete.cases(coords_df)
    if (sum(!valid_rows) > 0) {
      warning(paste(sum(!valid_rows), "rows removed due to invalid coordinates"))
    }
    coords_df <- coords_df[valid_rows, ]
    
    # Validate coordinate ranges
    invalid_lon <- coords_df$longitude < -180 | coords_df$longitude > 180
    invalid_lat <- coords_df$latitude < -90 | coords_df$latitude > 90
    
    if (any(invalid_lon)) {
      warning(paste("Found", sum(invalid_lon), "longitude values outside valid range (-180 to 180)"))
    }
    if (any(invalid_lat)) {
      warning(paste("Found", sum(invalid_lat), "latitude values outside valid range (-90 to 90)"))
    }
    
    # Remove invalid coordinates
    coords_df <- coords_df[!invalid_lon & !invalid_lat, ]
    
    if (nrow(coords_df) == 0) {
      stop("No valid coordinates remaining after cleaning")
    }
    
    as.matrix(coords_df)
  }, error = function(e) {
    stop(paste("Error processing coordinates:", e$message, "\nPlease check your coordinate data structure."))
  })
  
  # Function to analyze single residual with error handling
  analyze_single_residual <- function(residual_col) {
    tryCatch({
      # Clean residual data
      residual_values <- safe_to_numeric(data[[residual_col]])
      if (all(is.na(residual_values))) {
        stop("All values in residual column are NA")
      }
      
      # Match residuals with valid coordinates
      residual_values <- residual_values[valid_rows][!invalid_lon & !invalid_lat]
      
      # 1. Global Moran's I
      knn <- knearneigh(coords, k = min(5, nrow(coords) - 1))
      nb <- knn2nb(knn)
      weights <- nb2listw(nb, style = "W")
      
      moran_result <- moran.test(residual_values, weights)
      
      # 2. Local Moran's I (LISA)
      lisa_result <- localmoran(residual_values, weights)
      
      # 3. DBSCAN Clustering
      clustering_data <- cbind(scale(coords), scale(residual_values))
      eps <- 0.5 * sd(dist(clustering_data))
      dbscan_result <- dbscan(clustering_data, eps = eps, minPts = min(5, nrow(clustering_data)/10))
      
      # Store results
      results_df <- data.frame(
        lisa_pvalue = lisa_result[, 5],
        cluster = dbscan_result$cluster
      )
      
      return(list(
        moran = moran_result,
        lisa = lisa_result,
        dbscan = dbscan_result,
        results_df = results_df
      ))
    }, error = function(e) {
      warning(paste("Error analyzing", residual_col, ":", e$message))
      return(NULL)
    })
  }
  
  # Analyze each residual
  results$individual_analyses <- lapply(residual_cols, analyze_single_residual)
  names(results$individual_analyses) <- residual_cols
  
  # Create summary
  results$summary <- data.frame(
    residual = residual_cols,
    n_valid_points = sapply(results$individual_analyses, function(x) {
      if (!is.null(x)) nrow(x$results_df) else NA
    }),
    moran_i = sapply(results$individual_analyses, function(x) {
      if (!is.null(x)) x$moran$estimate[1] else NA
    }),
    moran_p = sapply(results$individual_analyses, function(x) {
      if (!is.null(x)) x$moran$p.value else NA
    })
  )
  
  return(results)
}

# Check the structure of your data
str(all_data_inc_na[c("long", "lat")])

# If needed, unnest your coordinate columns
if (is.list(all_data_inc_na$long)) {
  all_data_inc_na$long <- unlist(all_data_inc_na$long)
}
if (is.list(all_data_inc_na$lat)) {
  all_data_inc_na$lat <- unlist(all_data_inc_na$lat)
}

# Then run the analysis
residual_columns <- c("resid_pm", "resid_o3", "resid_no2", "resid_ndvi")
cluster_results <- analyze_residual_clusters(all_data_inc_na, residual_columns)

# Check the results
print(cluster_results$summary)
```

################################
Create model functions
```{r}
# Create exposure list
exposure_list <- list(
  pm = list(
    name = "pm",
    baseline = "pm_2002",
    change = "pm_change"
  ),
  o3 = list(
    name = "o3",
    baseline = "o3_2002",
    change = "o3_change"
  ),
  no2 = list(
    name = "no2",
    baseline = "no2_2000",
    change = "no2_change"
  ),
  ndvi = list(
    name = "ndvi",
    baseline = "ndvi_2000",
    change = "ndvi_change"
  ),
  temp = list(
    name = "mean_temp",
    baseline = NULL,
    change = "mean_temp"  # No change variable for temp
  )
)

# Function to run GAM models
run_gam_models <- function(data, exposure_info) {
  # Extract exposure information
  exp_var <- exposure_info$change
  baseline_exposure <- exposure_info$baseline
  
  # Skip if data is missing the exposure variable - gives warning 
  if (!exp_var %in% names(data)) {
    warning(paste("Variable", exp_var, "not found in dataset"))
    return(NULL)
  }
  
  # Base components that will be in all models: gentrification score and county/borough
  base_formula <- paste(exp_var, "~ score_0.5 + s(COUNTYFP10, bs='re')")
  
  # Create model formulas
  formulas <- list(
    # Base models (score + county + spatial) for tensor product and mrf smooth
    tp_base = paste(base_formula, "+ te(long, lat, d = 2, bs = 'gp', m = 2)"),
    mrf_base = paste(base_formula, "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),
    
    # Models with income
    tp_income = paste(base_formula, "+ s(med_income_2016) + te(long, lat, d = 2, bs = 'gp', m = 2)"),
    mrf_income = paste(base_formula, "+ s(med_income_2016) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),
    mrf_income_rank = paste(base_formula, "+ rank(med_income_2016) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),

    # Models with distance to downtown
    tp_dist = paste(base_formula, "+ s(dt_dist) + te(long, lat, d = 2, bs = 'gp', m = 2)"),
    mrf_dist = paste(base_formula, "+ s(dt_dist) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  )
  
  # Add baseline models if baseline exposure level exists (all except temp)
  if (!is.null(baseline_exposure) && baseline_exposure %in% names(data)) {
    formulas$tp_baseline <- paste(base_formula, "+", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_baseline <- paste(base_formula, "+", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    # Combinations with baseline
    formulas$tp_income_baseline <- paste(base_formula, "+ s(med_income_2016) +", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_income_baseline <- paste(base_formula, "+ s(med_income_2016) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    formulas$tp_dist_baseline <- paste(base_formula, "+ s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_dist_baseline <- paste(base_formula, "+ s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    formulas$tp_full <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_full <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  }
  
  # Add income + distance combinations
  formulas$tp_income_dist <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) + te(long, lat, d = 2, bs = 'gp', m = 2)")
  formulas$mrf_income_dist <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  
  # Convert formulas to formula objects
  formulas <- lapply(formulas, as.formula)
  
  # Run models
  models <- purrr::map(formulas, ~try(gam(., 
                                        family = scat, 
                                        data = data, 
                                        method = "REML"), 
                                    silent = TRUE))
  
  return(models)
}

# drop geometry from spatial data while keeping coordinates
data_prepared <- all_data_inc %>%
  mutate(
    COUNTYFP10 = as.factor(COUNTYFP10),
    GEOID = as.factor(GEOID)
  ) %>%
  st_drop_geometry()

# Construct neighbors list for MRF models
nb <- poly2nb(all_data_inc, row.names = all_data_inc$GEOID)
names(nb) <- attr(nb, "region.id") 

# JH 10/25/24 Warning: some observations have no neighbours; if this seems unexpected, try increasing the snap argument.Warning: neighbour object has 4 sub-graphs; if this sub-graph count seems unexpected, try increasing the snap argument.
```
Checking neighbor relationships for MRF models
```{r}
# Visualize the sub-graphs
plot(all_data_inc$geometry)
colors <- rainbow(length(nb))
plot(nb, st_coordinates(st_centroid(all_data_inc)), col=colors, add=TRUE)
# the sub-graphs match the islands.
# since we are adjusting by borough in the models, the 4 sub-graphs might be okay?? JH 10/26/24
```

Run models and check models
```{r}
# Run models for each exposure
all_models <- purrr::map(exposure_list, ~run_gam_models(data_prepared, .))

# Function to check which models succeeded/failed
check_models <- function(models) {
  model_status <- map(models, ~!inherits(., "try-error"))
  successful_models <- names(models)[unlist(model_status)]
  failed_models <- names(models)[!unlist(model_status)]
  
  return(list(
    successful = successful_models,
    failed = failed_models
  ))
}

# Check status of all models
model_status <- map(all_models, check_models)
model_status

```


Model diagnostics (11/8/24)
```{r}
# Extract the mrf_income models
mrf_income_models <- purrr::map(all_models, "mrf_income")

# Check the summary of all mrf_income models
purrr::walk(mrf_income_models, summary)

summary(mrf_income_models$pm)
plot(resid(mrf_income_models$pm))
plot(mrf_income_models$pm, pages = 1)

#gam check
gam.check(mrf_income_models$pm)
gam.check(mrf_income_models$o3)
gam.check(mrf_income_models$no2)
gam.check(mrf_income_models$ndvi)
gam.check(mrf_income_models$temp)
# p values all non sig fr checking which is good  for med income, but GEOID and county return NA p values?

#concurvity - high worst between s(COUNTYFP10) and s(GEOID) - 0.86 -problematic?
concurvity(mrf_income_models$pm, full = TRUE)
concurvity(mrf_income_models$pm, full = FALSE)
concurvity(mrf_income_models$o3, full = TRUE)
concurvity(mrf_income_models$o3, full = FALSE)
concurvity(mrf_income_models$no2, full = TRUE)
concurvity(mrf_income_models$no2, full = FALSE)
concurvity(mrf_income_models$ndvi, full = TRUE)
concurvity(mrf_income_models$ndvi, full = FALSE)

# Get the fitted values and residuals
fitted_values <- fitted(mrf_income_models$pm)
residuals <- resid(mrf_income_models$pm)

# Create a data frame for plotting
plot_data <- data.frame(fitted = fitted_values, residuals = residuals)

# Plot Residuals vs Fitted
ggplot(plot_data, aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted", x = "Fitted Values", y = "Residuals") +
  theme_minimal()



traceback()

```
# Plot the residuals for the "pm" mrf_income model


Create basic QQ plots for models
```{r}
# Function to create QQ plots for a specific exposure
create_exposure_qq_plots <- function(models_list, exposure_name) {
  exposure_models <- models_list[[exposure_name]]
  
  plots <- names(exposure_models) %>%
    purrr::map(function(model_name) {
      model <- exposure_models[[model_name]]
      
      # Skip if model is an error
      if (inherits(model, "try-error")) {
        return(NULL)
      }
      
      # Create QQ plot
      resid_df <- data.frame(
        residuals = residuals(model, type = "deviance"),
        fitted = fitted(model)
      )
      
      ggplot(resid_df, aes(sample = residuals)) +
        stat_qq() +
        stat_qq_line(color = "red") +
        labs(
          title = paste0(exposure_name, "\n", model_name),
          x = "Theoretical Quantiles",
          y = "Sample Quantiles"
        ) +
        theme_minimal() +
        theme(
          plot.title = element_text(size = 10, hjust = 0.5),
          axis.title = element_text(size = 8),
          axis.text = element_text(size = 7)
        )
    }) %>%
    purrr::compact() # Remove NULL values from the list
  
  # Calculate grid layout
  n_plots <- length(plots)
  n_cols <- min(3, n_plots)
  n_rows <- ceiling(n_plots / n_cols)
  
  # Combine plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = plots,
    ncol = n_cols,
    nrow = n_rows,
    top = grid::textGrob(paste("QQ Plots for", exposure_name), 
                        gp = grid::gpar(fontsize = 12))
  )
  
  return(combined_plot)
}

# Function to create QQ plots for a specific model type across exposures
create_model_type_qq_plots <- function(models_list, model_type) {
  # Convert the for loop to purrr::map
  plots <- names(models_list) %>%
    purrr::map(function(exposure_name) {
      exposure_models <- models_list[[exposure_name]]
      
      # Check if the model type exists for this exposure
      if (!model_type %in% names(exposure_models)) {
        return(NULL)
      }
      
      model <- exposure_models[[model_type]]
      
      # Skip if model is an error
      if (inherits(model, "try-error")) {
        return(NULL)
      }
      
      # Create QQ plot
      resid_df <- data.frame(
        residuals = residuals(model, type = "deviance"),
        fitted = fitted(model)
      )
      
      ggplot(resid_df, aes(sample = residuals)) +
        stat_qq() +
        stat_qq_line(color = "red") +
        labs(
          title = paste0(exposure_name, "\n", model_type),
          x = "Theoretical Quantiles",
          y = "Sample Quantiles"
        ) +
        theme_minimal() +
        theme(
          plot.title = element_text(size = 10, hjust = 0.5),
          axis.title = element_text(size = 8),
          axis.text = element_text(size = 7)
        )
    }) %>%
    purrr::compact() # Remove NULL values from the list
  
  # Calculate grid layout
  n_plots <- length(plots)
  n_cols <- min(3, n_plots)
  n_rows <- ceiling(n_plots / n_cols)
  
  # Combine plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = plots,
    ncol = n_cols,
    nrow = n_rows,
    top = grid::textGrob(paste("QQ Plots for", model_type), 
                        gp = grid::gpar(fontsize = 12))
  )
  
  return(combined_plot)
}

# Function to list available model types
list_model_types <- function(models_list) {
  unique(unlist(lapply(models_list, names)))
}

# Function to list available exposures
list_exposures <- function(models_list) {
  names(models_list)
}

```


Display basic QQ plots for models
```{r}
# List all model types
model_types <- list_model_types(all_models)

# Display all QQ plots
purrr::walk(model_types, ~create_model_type_qq_plots(all_models, .))

```

Create functions for QQ plots with confidence intervals
```{r}
# Function to create a single QQ plot with confidence bands
create_single_qq_plot <- function(model, title) {
  if (inherits(model, "try-error")) return(NULL)
  
  data.frame(
    residuals = residuals(model, type = "deviance"),
    fitted = fitted(model)
  ) %>%
    ggplot(aes(sample = residuals)) +
    geom_qq_band(bandType = "ks", mapping = aes(fill = "KS"), alpha = 0.5) +
    geom_qq_band(bandType = "ts", mapping = aes(fill = "TS"), alpha = 0.5) +
    geom_qq_band(bandType = "pointwise", mapping = aes(fill = "Normal"), alpha = 0.5) +
    geom_qq_band(bandType = "boot", mapping = aes(fill = "Bootstrap"), alpha = 0.5) +  
    stat_qq_line(color = "red") +
    stat_qq_point(size = 1) +
    labs(
      title = title,
      x = "Theoretical Quantiles",
      y = "Sample Quantiles"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 10, hjust = 0.5),
      axis.title = element_text(size = 8),
      axis.text = element_text(size = 7)
    )
}

# Function to arrange plots in a grid
arrange_plots <- function(plots, main_title) {
  n_plots <- length(plots)
  n_cols <- min(3, n_plots)  
  n_rows <- ceiling(n_plots / n_cols)
  
  gridExtra::grid.arrange(
    grobs = plots,
    ncol = n_cols,
    nrow = n_rows,
    top = grid::textGrob(
      main_title,
      gp = grid::gpar(fontsize = 12)  
    )
  )
}

# Main display function
display_qq_plots_by_model_type <- function(models_list) {
  # Get all unique model types
  model_types <- models_list %>%
    map(names) %>%
    reduce(union)
  
  # Create plots for each model type
  walk(model_types, function(model_type) {
    # Extract all models of this type across exposures
    plots <- models_list %>%
      imap(function(exposure_models, exposure_name) {
        if (!model_type %in% names(exposure_models)) return(NULL)
        
        create_single_qq_plot(
          exposure_models[[model_type]], 
          title = paste0(exposure_name, "\n", model_type)
        )
      }) %>%
      compact()
    
    # Arrange and display plots for this model type
    arrange_plots(
      plots,
      main_title = paste("QQ Plots for", model_type)
    )
  })
}

```

Display QQ plots with confidence bands for models
```{r}
# Display QQ plots
display_qq_plots_by_model_type(all_models)
```

Create comprehensive diagnostic plots (11/2 - need to change the scale of the plots for visibility)
```{r}
create_model_diagnostics_by_type <- function(models_list, model_type) {
  # Create diagnostic plots for each exposure
  diagnostic_plots <- models_list %>%
    imap(function(exposure_models, exposure_name) {
      # Skip if model type doesn't exist or is error
      if (!model_type %in% names(exposure_models) || 
          inherits(exposure_models[[model_type]], "try-error")) {
        return(NULL)
      }
      
      model <- exposure_models[[model_type]]
      
      # Create residuals dataframe
      resid_df <- tibble(
        residuals = residuals(model, type = "deviance"),
        fitted = fitted(model),
        scale_residuals = residuals(model, type = "scaled.pearson"),
        response = model$y
      )
      
      # Create the four diagnostic plots
      list(
        # 1. QQ plot
        qq = ggplot(resid_df, aes(sample = residuals)) +
          stat_qq() +
          stat_qq_line(color = "red") +
          labs(title = paste(exposure_name, "- Normal Q-Q Plot")) +
          theme_minimal() +
          theme(plot.title = element_text(size = 8)),
        
        # 2. Residuals vs Fitted
        resid = ggplot(resid_df, aes(x = fitted, y = residuals)) +
          geom_point(alpha = 0.5) +
          geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
          geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
          labs(title = paste(exposure_name, "- Residuals vs Fitted"),
               x = "Fitted values",
               y = "Deviance residuals") +
          theme_minimal() +
          theme(plot.title = element_text(size = 8)),
        
        # 3. Scale-location plot
        scale = ggplot(resid_df, 
                      aes(x = fitted, y = sqrt(abs(scale_residuals)))) +
          geom_point(alpha = 0.5) +
          geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
          labs(title = paste(exposure_name, "- Scale-Location"),
               x = "Fitted values",
               y = "√|Standardized residuals|") +
          theme_minimal() +
          theme(plot.title = element_text(size = 8)),
        
        # 4. Response vs Fitted
        resp = ggplot(resid_df, aes(x = fitted, y = response)) +
          geom_point(alpha = 0.5) +
          geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
          labs(title = paste(exposure_name, "- Response vs Fitted"),
               x = "Fitted values",
               y = "Response") +
          theme_minimal() +
          theme(plot.title = element_text(size = 8))
      )
    }) %>%
    compact() # Remove NULL entries
  
  # Stop if no valid models found
  if (length(diagnostic_plots) == 0) {
    stop("No valid models found for the specified model type")
  }
  
  # Flatten the nested list of plots
  all_plots <- diagnostic_plots %>%
    flatten()
  
  # Calculate grid dimensions
  n_exposures <- length(diagnostic_plots)
  n_cols <- 4  # One column for each type of diagnostic plot
  
  # Combine all plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = all_plots,
    ncol = n_cols,
    nrow = n_exposures,
    top = grid::textGrob(
      paste("Diagnostic Plots for Model Type:", model_type),
      gp = grid::gpar(fontsize = 12, fontface = "bold")
    )
  )
  
  return(combined_plot)
}

display_all_model_diagnostics <- function(models_list) {
  # Get unique model types
  model_types <- models_list %>%
    map(names) %>%
    flatten_chr() %>%
    unique()
  
  # Create and display all plots at once
  plots <- model_types %>%
    map(~{
      tryCatch({
        create_model_diagnostics_by_type(models_list, .x)
      }, error = function(e) {
        message(paste("Error creating plots for model type:", .x))
        message(e)
        return(NULL)
      })
    }) %>%
    compact()
  
  # Calculate grid layout for all model types
  n_types <- length(plots)
  grid_dims <- c(ceiling(sqrt(n_types)), floor(sqrt(n_types)))
  if (prod(grid_dims) < n_types) grid_dims[2] <- grid_dims[2] + 1
  
  # Display all plots in a single grid
  do.call(gridExtra::grid.arrange, 
          c(plots, 
            list(ncol = grid_dims[1],
                 nrow = grid_dims[2])))
}
display_all_model_diagnostics(all_models)

```
```{r}

create_model_diagnostics_by_type <- function(models_list, model_type) {
  # Create diagnostic plots for each exposure
  diagnostic_plots <- models_list %>%
    imap(function(exposure_models, exposure_name) {
      # Skip if model type doesn't exist or is error
      if (!model_type %in% names(exposure_models) || 
          inherits(exposure_models[[model_type]], "try-error")) {
        return(NULL)
      }
      
      model <- exposure_models[[model_type]]
      
      # Create residuals dataframe
      resid_df <- tibble(
        residuals = residuals(model, type = "deviance"),
        fitted = fitted(model),
        scale_residuals = residuals(model, type = "scaled.pearson"),
        response = model$y
      )
      
      # Create the four diagnostic plots
      list(
        # 1. QQ plot
        qq = ggplot(resid_df, aes(sample = residuals)) +
          stat_qq() +
          stat_qq_line(color = "red") +
          labs(title = paste(exposure_name, "- Normal Q-Q Plot")) +
          theme_minimal(base_size = 14) +
          theme(plot.title = element_text(size = 16, face = "bold"),
                axis.title = element_text(size = 12),
                axis.text = element_text(size = 12)),
        
        # 2. Residuals vs Fitted
        resid = ggplot(resid_df, aes(x = fitted, y = residuals)) +
          geom_point(alpha = 0.5) +
          geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
          geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
          labs(title = paste(exposure_name, "- Residuals vs Fitted"),
               x = "Fitted values",
               y = "Deviance residuals") +
          theme_minimal(base_size = 14) +
          theme(plot.title = element_text(size = 16, face = "bold"),
                axis.title = element_text(size = 12),
                axis.text = element_text(size = 12)),
        
        # 3. Scale-location plot
        scale = ggplot(resid_df, 
                      aes(x = fitted, y = sqrt(abs(scale_residuals)))) +
          geom_point(alpha = 0.5) +
          geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
          labs(title = paste(exposure_name, "- Scale-Location"),
               x = "Fitted values",
               y = "√|Standardized residuals|") +
          theme_minimal(base_size = 14) +
          theme(plot.title = element_text(size = 16, face = "bold"),
                axis.title = element_text(size = 12),
                axis.text = element_text(size = 12)),
        
        # 4. Response vs Fitted
        resp = ggplot(resid_df, aes(x = fitted, y = response)) +
          geom_point(alpha = 0.5) +
          geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
          labs(title = paste(exposure_name, "- Response vs Fitted"),
               x = "Fitted values",
               y = "Response") +
          theme_minimal(base_size = 14) +
          theme(plot.title = element_text(size = 16, face = "bold"),
                axis.title = element_text(size = 12),
                axis.text = element_text(size = 12))
      )
    }) %>%
    compact() # Remove NULL entries
  
  # Stop if no valid models found
  if (length(diagnostic_plots) == 0) {
    stop("No valid models found for the specified model type")
  }
  
  # Flatten the nested list of plots
  all_plots <- diagnostic_plots %>%
    flatten()
  
  # Calculate grid dimensions
  n_exposures <- length(diagnostic_plots)
  n_cols <- 4  # One column for each type of diagnostic plot
  
  # Combine all plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = all_plots,
    ncol = n_cols,
    nrow = n_exposures,
    top = grid::textGrob(
      paste("Diagnostic Plots for Model Type:", model_type),
      gp = grid::gpar(fontsize = 18, fontface = "bold")
    ),
    heights = rep(1, n_exposures),  # Adjust height of each row to ensure plots are not squished
    padding = unit(1, "cm")  # Add padding between plots
  )
  
  return(combined_plot)
}

display_all_model_diagnostics <- function(models_list) {
  # Get unique model types
  model_types <- models_list %>%
    map(names) %>%
    flatten_chr() %>%
    unique()
  
  # Create and display all plots at once
  plots <- model_types %>%
    map(~{
      tryCatch({
        create_model_diagnostics_by_type(models_list, .x)
      }, error = function(e) {
        message(paste("Error creating plots for model type:", .x))
        message(e)
        return(NULL)
      })
    }) %>%
    compact()
  
  # Calculate grid layout for all model types
  n_types <- length(plots)
  grid_dims <- c(ceiling(sqrt(n_types)), floor(sqrt(n_types)))
  if (prod(grid_dims) < n_types) grid_dims[2] <- grid_dims[2] + 1
  
  # Display all plots in a single grid
  do.call(gridExtra::grid.arrange, 
          c(plots, 
            list(ncol = grid_dims[1],
                 nrow = grid_dims[2],
                 top = grid::textGrob("All Model Diagnostics", 
                                      gp = grid::gpar(fontsize = 18, fontface = "bold")))))
}
display_all_model_diagnostics(all_models)


```

Create effects summary data frame and forest plot
```{r}
# Function to extract effect estimates for score_0.5
get_score_effects <- function(models_list, model_type = "mrf_income") {
  effects_list <- list()
  
  for (exposure_name in names(models_list)) {
    model <- models_list[[exposure_name]][[model_type]]
    
    if (!inherits(model, "try-error")) {
      # Get predicted values across range of scores
      score_seq <- seq(min(model$model$score_0.5), max(model$model$score_0.5), length.out = 100)
      pred_data <- data.frame(
        score_0.5 = score_seq,
        med_income_2016 = mean(model$model$med_income_2016),
        COUNTYFP10 = levels(model$model$COUNTYFP10)[1],
        GEOID = levels(model$model$GEOID)[1]
      )
      
      # Get predictions with standard errors
      preds <- predict(model, pred_data, type = "terms", se.fit = TRUE)
      
      # Extract effect for score_0.5 smooth term
      score_effect <- data.frame(
        exposure = exposure_name,
        estimate = diff(range(preds$fit[,"s(score_0.5)"])),
        se = diff(range(preds$se.fit[,"s(score_0.5)"])),
        p_value = summary(model)$s.table["s(score_0.5)", "p-value"]
      )
      
      effects_list[[exposure_name]] <- score_effect
    }
  }
  
  # Combine all effects into one dataframe
  effects_df <- do.call(rbind, effects_list)
  
  # Calculate confidence intervals
  effects_df$lower <- effects_df$estimate - 1.96 * effects_df$se
  effects_df$upper <- effects_df$estimate + 1.96 * effects_df$se
  
  return(effects_df)
}

# Get effects
effects_df <- get_score_effects(all_models)

# Create forest plot
library(ggplot2)

# Create exposure labels with significance indicators
effects_df$exposure_label <- paste0(
  toupper(effects_df$exposure), 
  ifelse(effects_df$p_value < 0.001, "***",
         ifelse(effects_df$p_value < 0.01, "**",
                ifelse(effects_df$p_value < 0.05, "*", "")))
)

# Create the forest plot
ggplot(effects_df, aes(y = exposure_label)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(x = estimate), size = 3) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  labs(
    title = "Estimated Effects of Gentrification Score on Environmental Exposures",
    subtitle = "MRF Income Model Results",
    x = "Effect Size (Range of Smooth Term)",
    y = "Exposure Type",
    caption = "* p < 0.05, ** p < 0.01, *** p < 0.001"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    plot.caption = element_text(size = 8, hjust = 0),
    panel.grid.minor = element_blank()
  )

# Print numerical results
print(effects_df[, c("exposure_label", "estimate", "se", "p_value", "lower", "upper")])
```


Scale mrf_income models by IQR and create scaled forest plot
```{r}
# Function to extract IQR-scaled effect estimates for score_0.5
get_score_effects_scaled <- function(models_list, model_type = "mrf_income") {
  effects_list <- list()
  
  for (exposure_name in names(models_list)) {
    model <- models_list[[exposure_name]][[model_type]]
    
    if (!inherits(model, "try-error")) {
      # Get model data
      model_data <- model$model
      
      # Calculate IQRs
      score_iqr <- IQR(model_data$score_0.5)
      exposure_var <- exposure_list[[exposure_name]]$change
      exposure_iqr <- IQR(model_data[[exposure_var]])
      
      # Get predicted values across range of scores
      score_seq <- seq(min(model_data$score_0.5), max(model_data$score_0.5), length.out = 100)
      pred_data <- data.frame(
        score_0.5 = score_seq,
        med_income_2016 = mean(model_data$med_income_2016),
        COUNTYFP10 = levels(model_data$COUNTYFP10)[1],
        GEOID = levels(model_data$GEOID)[1]
      )
      
      # Get predictions with standard errors
      preds <- predict(model, pred_data, type = "terms", se.fit = TRUE)
      
      # Extract and scale effect for score_0.5 smooth term
      raw_effect <- diff(range(preds$fit[,"s(score_0.5)"]))
      raw_se <- diff(range(preds$se.fit[,"s(score_0.5)"]))
      
      # Scale effects by IQRs
      scaled_effect <- (raw_effect * score_iqr) / exposure_iqr
      scaled_se <- (raw_se * score_iqr) / exposure_iqr
      
      score_effect <- data.frame(
        exposure = exposure_name,
        estimate = scaled_effect,
        se = scaled_se,
        p_value = summary(model)$s.table["s(score_0.5)", "p-value"],
        exposure_iqr = exposure_iqr,
        score_iqr = score_iqr
      )
      
      effects_list[[exposure_name]] <- score_effect
    }
  }
  
  # Combine all effects into one dataframe
  effects_df <- do.call(rbind, effects_list)
  
  # Calculate confidence intervals
  effects_df$lower <- effects_df$estimate - 1.96 * effects_df$se
  effects_df$upper <- effects_df$estimate + 1.96 * effects_df$se
  
  return(effects_df)
}

# Get scaled effects
effects_df_scaled <- get_score_effects_scaled(all_models)

# Create forest plot with scaled effects
effects_df_scaled$exposure_label <- paste0(
  toupper(effects_df_scaled$exposure), 
  ifelse(effects_df_scaled$p_value < 0.001, "***",
         ifelse(effects_df_scaled$p_value < 0.01, "**",
                ifelse(effects_df_scaled$p_value < 0.05, "*", "")))
)

ggplot(effects_df_scaled, aes(y = exposure_label)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(x = estimate), size = 3) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  labs(
    title = "IQR-Scaled Effects of Gentrification Score on Environmental Exposures",
    subtitle = "MRF Income Model Results",
    x = "Effect Size (Change in Exposure IQR per Score IQR)",
    y = "Exposure Type",
    caption = "* p < 0.05, ** p < 0.01, *** p < 0.001"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    plot.caption = element_text(size = 8, hjust = 0),
    panel.grid.minor = element_blank()
  )

```

Understanding direction and magnitude of exposure changes 
```{r}
# Function to analyze changes in exposures
analyze_exposure_changes <- function(data, exposure_list) {
  results <- list()
  
  for (exp_name in names(exposure_list)) {
    exp_info <- exposure_list[[exp_name]]
    change_var <- exp_info$change
    
    # Skip temperature which doesn't have baseline/change structure
    if (exp_name != "temp") {
      baseline_var <- exp_info$baseline
      
      # Calculate summary statistics
      change_stats <- data %>%
        summarise(
          n = nrow(data),
          mean_change = mean(.data[[change_var]], na.rm = TRUE),
          median_change = median(.data[[change_var]], na.rm = TRUE),
          sd_change = sd(.data[[change_var]], na.rm = TRUE),
          pct_positive = mean(.data[[change_var]] > 0, na.rm = TRUE) * 100,
          q25 = quantile(.data[[change_var]], 0.25, na.rm = TRUE),
          q75 = quantile(.data[[change_var]], 0.75, na.rm = TRUE)
        )
      
      # Create histogram
      hist_plot <- ggplot(data, aes(x = .data[[change_var]])) +
        geom_histogram(bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
        geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
        labs(
          title = paste("Distribution of Changes in", toupper(exp_name)),
          x = "Change",
          y = "Count"
        ) +
        theme_minimal()
      
      # Store results
      results[[exp_name]] <- list(
        stats = change_stats,
        plot = hist_plot
      )
    }
  }
  
  return(results)
}

# Run the analysis
change_analysis <- analyze_exposure_changes(data_prepared, exposure_list)

# Print summary statistics 
print(change_analysis)

# Create a combined plot of all histograms
plot_list <- lapply(names(change_analysis), function(exp_name) {
  change_analysis[[exp_name]]$plot
})

# Arrange all plots in a grid
library(gridExtra)
grid.arrange(grobs = plot_list, ncol = 2)

# Create boxplots for another view of the distributions
combined_changes <- data.frame(
  Exposure = character(),
  Change = numeric()
)

for (exp_name in names(exposure_list)) {
  if (exp_name != "temp") {
    change_var <- exposure_list[[exp_name]]$change
    tmp_data <- data.frame(
      Exposure = toupper(exp_name),
      Change = data_prepared[[change_var]]
    )
    combined_changes <- rbind(combined_changes, tmp_data)
  }
}

# Create boxplot
ggplot(combined_changes, aes(x = Exposure, y = Change)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of Changes Across All Exposures",
    x = "Exposure Type",
    y = "Change"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```