---
title: "gent_clean_11.25.24"
author: "Joyce Hu"
date: "2024-11-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages (JH - check if any are unused after code cleaned)
```{r}
library(tidyverse)
library(tigris)
library(arrow)
library(here)
library(mgcv)
library(fixest)
library(lmtest)
library(terra)
library(exactextractr)
library(tidyterra)
library(sf)
library(mgcViz)

library(DHARMa)
library(qqplotr)
library(readxl) # for read_excel()
library(spdep) # for moran.test()
library(tidycensus) # for median income
library(distr) # for qqbounds()
library(spatialreg) # for spatial lag and error models
library(gridExtra) # for viewing plots side-by-side
library(qgam) 
library(dplyr)
library(INLA)
library(ggregplot) #INLA compatible plotting library
library(ggfields) # for plotting spatial field
library(lattice) # for levelplot
```

Load and clean data
```{r}
# Check if 'data' directory exists; if not, create it
if (!dir.exists(here("data"))) {
  dir.create(here("data"))
}

# Load gentrification score by tract
gentrification <- read.csv(here("data", "Data_NYC_Gentrification_2000_16.csv")) %>% 
  mutate(tractid = as.character(tractid)) %>% 
  dplyr::select(tractid, score_0.5)

# Load gentrification score if using xlsx file:
# gentrification <- read_xlsx(here("Gentrification","Data_NYC_Gentrification_2000_16.xlsx")) %>%
#   mutate(tractid = as.character(tractid)) %>%
#   dplyr::select(tractid, score_0.5)

# pull NYC tract IDs/fips from gentrification index dataset for filtering
nyc_fips <- gentrification$tractid

# Load NYC tracts shapefile
nyc_tracts <- tracts(state = "NY", county = c("Bronx", "Kings", "New York", "Queens", "Richmond"), year = 2010) %>%
  # clean lat and long for Conley SEs
  mutate(lat = as.numeric(INTPTLAT10), long = as.numeric(INTPTLON10), GEOID = GEOID10) %>% 
  # filter for FIPS codes included in gentrification index dataset
  filter(GEOID %in% nyc_fips)


# Load NY tracts populatuion weighted centroid points from Census data
nyc_weighted_centroids <- read.table("https://www2.census.gov/geo/docs/reference/cenpop2010/tract/CenPop2010_Mean_TR36.txt", 
                         header = TRUE, sep = ",", stringsAsFactors = FALSE, 
                         colClasses = c("character", "character", "character", "numeric", "numeric", "numeric")) %>%
  mutate(FIPS = paste(STATEFP, COUNTYFP, TRACTCE, sep = "")) %>% 
  ## filter for only NYC tracts
  filter(FIPS %in% nyc_fips) %>%  
  rename(Longitude_weighted = LONGITUDE, 
         Latitude_weighted = LATITUDE)

# JH - 10/12/24 warning persists 10/13, 10/14 Warning: R metadata may have unsafe or invalid elements ℹ Type: "externalptr" options(arrow.unsafe_metadata = TRUE) #

# Temperature data
if(!file.exists(here("data","summarized_daily_temp_preds.parquet"))){download.file("https://zenodo.org/records/10557980/files/summarized_daily_temp_preds.parquet?download=1", 
                                                                                   destfile = here("data","summarized_daily_temp_preds.parquet"))}

# Temperature by tract
temp <- read_parquet(here("data","summarized_daily_temp_preds.parquet")) %>%
  filter(str_detect(date, "2016")) %>%
  mutate(temp_c = mean_temp_daily_cK/100-273.15) %>% 
  mutate(temp_f = temp_c*9/5+32) %>% 
  group_by(GEOID, date) %>%
  ungroup() %>% 
  group_by(GEOID) %>% 
  summarize(mean_temp = mean(mean_temp_daily_cK/100))

# NDVI by tract (need to have it downloaded and stored locally)
ndvi_2016 <- rast(here("data", "ndvi_c2_2016_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

ndvi_2000 <- rast(here("data", "ndvi_c2_2000_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

# PM2.5 data
# Download data from URL
download_data <- function() {
  # set longer timeout
  options(timeout = 7200)
  if(!file.exists(here("data", "2016_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_pm25_daily_average.txt.gz", destfile = here("data", "2016_pm25_daily_average.txt.gz"))} #2016
  if(!file.exists(here("data", "2002_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_pm25_daily_average.txt.gz", destfile = here("data", "2002_pm25_daily_average.txt.gz"))}} #2002
download_data()

# Read in PM2.5 daily average by census tract from EPA FAQSD - 2002
pm_tract_2002 <- read.delim(gzfile(here("data", "2002_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## Include only NYC tracts
  filter(FIPS %in% nyc_fips) %>%
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## Calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(pm_mean_2002 = mean(pm25_daily_average.ug.m3.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# Read in PM2.5 daily average by census tract from EPA FAQSD - 2016
pm_tract_2016 <- read.delim(gzfile(here("data", "2016_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(pm_mean_2016 = mean(Prediction), 
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))
  

# Combine 2002 and 2016 pm data for nyc_tracts
pm_tract <- pm_tract_2002 %>%
  inner_join(pm_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# NO2 CACES
no2_2016_caces <- read.csv(here("data", "2016_no2_caces.csv")) %>% 
  mutate(no2_2016 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_2000_caces <- read.csv(here("data", "2000_no2_caces.csv")) %>% 
  mutate(no2_2000 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_caces <- left_join(no2_2000_caces, no2_2016_caces, by = "fips") %>% 
  select(no2_2000, no2_2016, fips)

#O3 data download
options(timeout = 7200)
if(!file.exists(here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))} #2016
if(!file.exists(here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))} #2002

# Read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2002
o3_tract_2002 <- read.delim(gzfile(here("data", "2002_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>% 
  ## include only NYC tracts
  filter(FIPS %in% nyc_fips) %>% 
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(o3_mean_2002 = mean(ozone_daily_8hour_maximum.ppb.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# Read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2016
o3_tract_2016 <- read.delim(gzfile(here("data", "2016_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>% 
  ## calculate annual average by tract using  population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(o3_mean_2016 = mean(Prediction),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))


o3_tract <- o3_tract_2002 %>%
  left_join(o3_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# Merge all exposure data
all_data <- nyc_tracts %>% 
  # join gentrification index by tract
  right_join(gentrification, by = c("GEOID10" = "tractid")) %>%
  left_join(pm_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(pm_2016 = pm_mean_2016, pm_2002 = pm_mean_2002, pm_change = pm_2016-pm_2002) %>% 
  left_join(o3_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(o3_2016 = o3_mean_2016, o3_2002 = o3_mean_2002, o3_change = o3_2016-o3_2002) %>% 
  left_join(no2_caces, by = c("GEOID" = "fips")) %>% 
  mutate(no2_change = no2_2016-no2_2000) %>% 
  mutate(ndvi_2016 = ndvi_2016, ndvi_2000 = ndvi_2000, ndvi_change = ndvi_2016-ndvi_2000) %>% 
  left_join(temp, by = "GEOID")

############ Median income ###############################
# Find median household income variable in 2000 decennial census
all_vars_dec <- 
  load_variables(year = 2000, dataset = "sf3")

# Models adjusting for median income in 2000 (note: 2000 decennial census is 1999 data)
income_2000 <- get_decennial(variables = "P053001", year = 2000,  sumfile = "sf3", state = "NY", 
                  county = c("Queens", "King", "New York", "Richmond", "Bronx"), geography = "tract") %>% 
  rename(med_income_2000 = value)
#results in 226 missing tracts in terms of 2010 tracts.

# Find median household income variable in 2010 ACS
income_2010 <- get_acs(variables = "B06011_001", year = 2010, state = "NY", 
                  county = c("Queens", "King", "New York", "Richmond", "Bronx"), geography = "tract") %>% 
  rename(med_income_2010 = estimate)

# Joining both income sets with data and imputing 2010 data where 2000 data is missing
all_data_inc <- all_data %>%
  left_join(income_2000, by = "GEOID") %>%
  left_join(income_2010, by = "GEOID") %>%
  mutate(med_income_2000_impute = coalesce(med_income_2000, med_income_2010)) %>% 
  # coerce GEOID to numeric then factor class to avoid error in MRF smooth
  mutate(GEOID = as.numeric(GEOID)) %>% 
  mutate(GEOID = factor(GEOID)) %>% 
 # coerce COUNTYFP10 to numeric then factor class to avoid error ?
  mutate(COUNTYFP10 = as.numeric(COUNTYFP10)) %>% 
  mutate(COUNTYFP10 = factor(COUNTYFP10))

# Ranking median income for use in GAM models
all_data_inc$med_income_ranked <- rank(all_data_inc$med_income_2000_impute)

############# Distance to downtown #########
# Load nyc roads
downtown_st <- roads(state = "NY", county = "New York", year = 2011)
ggplot() + geom_sf(data = downtown_st)

# Define bounding box
bbox_coords <- matrix(c(
  -74.0187, 40.70158,  # Southwest
  -74.0187, 40.77279,  # Northwest
  -73.95953, 40.77279, # Northeast
  -73.95953, 40.70158, # Southeast
  -74.0187, 40.70158   # Closing the polygon
), ncol = 2, byrow = TRUE)

# Convert bounding box to sf
bbox_polygon <- st_sfc(st_polygon(list(bbox_coords)), crs = 4326)

# Ensure roads are the same CRS and filter for streets within bbox
downtown_st <- st_transform(downtown_st, crs = st_crs(bbox_polygon))
filtered_streets <- downtown_st[st_within(downtown_st, bbox_polygon, sparse = FALSE), ]

# Plot filtered downtown streets and bounding box
ggplot() +
  geom_sf(data = filtered_streets) +
  geom_sf(data = bbox_polygon, fill = NA, color = 'red') +
  ggtitle("Downtown Streets within Bounding Box")

# Ensure census tracts are in the same CRS
nyc_tracts <- st_transform(nyc_tracts, crs = st_crs(bbox_polygon))

# Filter intersecting tracts within downtown
intersecting_tracts <- st_intersection(nyc_tracts, bbox_polygon)
combined_area <- st_union(intersecting_tracts)
#get downtown centroid
area_centroid <- st_centroid(combined_area)

ggplot() +
  geom_sf(data = intersecting_tracts, fill = 'lightblue', color = 'black') +
  geom_sf(data = area_centroid, color = 'red', size = 3) +
  ggtitle('Centroid of Downtown Census Tracts')

# Calculate distances and add to "all_data_inc" dataframe
distances <- st_distance(nyc_tracts, area_centroid)
all_data_inc$dt_dist <- as.numeric(distances)

#all_data_inc includes gentrification score (score_0.5), all exposure data, median income, distance to downtown

```
INLA set up
```{r}
# # Set INLA repository
# options(repos = c(getOption("repos"),
#    INLA="https://inla.r-inla-download.org/R/stable"))
# 
# # Install INLA and dependencies (from CRAN)
# install.packages("INLA", dep = TRUE)

#need older version of INLA to be compatible with R version
 # remotes::install_version("INLA", version="22.05.07",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/testing"), dep=TRUE)

# Install an even older version of INLA for GLIBC compatibility

# remotes::install_version("INLA", 
#                          version="21.02.23", 
#                          repos=c(getOption("repos"),
#                                  INLA="https://inla.r-inla-download.org/R/stable"), 
#                          dep=TRUE)

```

INLA PM2.5 models without spatial aspect 
```{r}
######INLA PM2.5 MODELS #####

IM_pm_1  <- inla(pm_change ~ score_0.5 + med_income_ranked,
               family = "gaussian", 
               data = all_data_inc, verbose = TRUE) 

IM_pm_2 <- inla(pm_change ~ score_0.5 + med_income_ranked + f(COUNTYFP10, model = "iid"),
                family = "gaussian", 
                data = all_data_inc, verbose = TRUE)

IM_pm_3 <- inla(pm_change ~ score_0.5 + med_income_ranked + f(COUNTYFP10, model = "iid") + dt_dist, family = "gaussian", data = all_data_inc, verbose = TRUE)

IM_pm_4 <- inla(pm_change ~ score_0.5 + med_income_ranked + dt_dist, family = "gaussian", data = all_data_inc, verbose = TRUE)

IM_pm_5 <- inla(pm_change ~ score_0.5, family = "gaussian", data = all_data_inc, verbose = TRUE)

IM_pm_6 <- inla(pm_change ~ score_0.5 + f(COUNTYFP10, model = "iid"), family = "gaussian", data = all_data_inc, verbose = TRUE)



# plotting INLA 2.5 models
Efxplot(list(IM_pm_1, IM_pm_2, IM_pm_3, IM_pm_4, IM_pm_5, IM_pm_6))

#effect estimate for median income and distance to downtown all zero.
```
INLA - Mesh creation
```{r}
#### Create mesh using tract centroids as locations

# Create matrix of locations
Locations <- as.matrix(cbind(
  as.numeric(nyc_weighted_centroids$Longitude_weighted),
  as.numeric(nyc_weighted_centroids$Latitude_weighted)
))

# Create mesh options
# Various mesh option specifications 
MeshA <- inla.mesh.2d(jitter(Locations), max.edge = c(20, 40))
MeshB <- inla.mesh.2d(Locations, max.edge = c(20, 40))
MeshC <- inla.mesh.2d(Locations, max.edge = c(50, 100))
mesh_simple <- inla.mesh.create(Locations)

# Visualize mesh options
plot(MeshA)
points(Locations, col = "red", pch = 2)

plot(MeshB)
points(Locations, col = "red", pch = 2)

plot(MeshC)
points(Locations, col = "red", pch = 2)


# "Generally, the mesh function automatically creates a mesh like mesh A, where closer-together sampling locations produce smaller triangles. When exploring/setting up preliminary analyses, use a mesh like mesh B. For analyses to be reported in a paper, use a mesh like mesh C. Be careful of edges, and try to allow some space around your sampling area for INLA to estimate. The edge triangles can be bigger to reduce computing power."

# Set chosen mesh to use for analysis
Mesh <- MeshB
```

INLA - Creating the A Matrix and SPDE
```{r}
# Create A matrix
A <- inla.spde.make.A(Mesh, loc = Locations) 
# "the A matrix maps the Gaussian Markov Random Field (GMRF) from the mesh nodes to the n observation location. This is represented by a matrix with the number of observations as column and the number of nodes as rows."


# Create spatial structure (SPDE object)
spde <- inla.spde2.matern(Mesh, alpha=2) #they fix alpha as 2

# Create all the required indexes for the SPDE model, naming the spatial field as “spatial.field”, which facilitates the extraction of the estimated values of the spatial field after the fitting process.
iset <- inla.spde.make.index(name = "spatial.field", spde$n.spde)
```

INLA stack (PM2.5)
```{r}
# Create stack
stk <- inla.stack(
  data=list(y=all_data_inc$pm_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_ranked = all_data_inc$med_income_ranked,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                               dt_dist = all_data_inc$dt_dist)))


#"Two projector matrices are needed A = list(A,1): the projector matrix for the latent field and a matrix that is a one-to-one map of the covariate. The latter matrix can simply be a constant rather than a diagonal matrix."
```


INLA - model fitting (no priors) for PM2.5
```{r}
# Specify formula
formula1 <- y ~ -1 + Intercept + score_0.5 +
  med_income_ranked + 
  dt_dist +
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
  #rw1 and rw2 don't work for the covariates, not sure how else to smooth, so right now just linear?? 

# Specify model
model1<-inla(formula1, 
             data = inla.stack.data(stk, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk),compute = TRUE), 
             control.compute = list(cpo = TRUE))

```

INLA model results for PM2.5
```{r}
###  Model results

# Posterior distribution summary of the fixed effects
model1$summary.fix

# Posterior distribution summary of the hyper parameters (spatial field parameters)
model1$summary.hyperpar

# Rounded summaries
round(model1$summary.fixed,3) #med_income_ranked and dt_dist still come out out as all zeroes --> rounded...
round(model1$summary.hyperpar,3)

# "The posterior distribution of the hyperparameters θ1 and θ2 are difficult to interpret. We can transform them into more interpretable quantities such as the range, and the variance. The range is of particular interest. It provides the distance value (in the unit of the point coordinates) above which spatial dependencies become negligible."
model1.res<-inla.spde2.result(model1, 'spatial.field', spde, do.transf=TRUE) #do.tranf = TRUE makes sure that marginals are calculated in the same scale as the data.

model1.res$summary.log.range.nominal
model1.res$summary.log.variance.nominal

# Plot effect estimate and credible intervals
graphics.off()
Efxplot(list(model1))

# Plot for fixed parameters for insight into the effect of covariates
par(mfrow=c(2,2))
plot(model1$marginals.fixed[[1]], ty = "l", xlab = expression(beta[0]), ylab = "Density")
plot(model1$marginals.fixed[[2]], ty = "l", xlab = expression(beta[score_0.5]), ylab = "Density") 
plot(model1$marginals.fixed[[3]], ty = "l", xlab = expression(beta[med_income_ranked]), ylab = "Density") 
plot(model1$marginals.fixed[[4]], ty = "l", xlab = expression(beta[dt_dist]), ylab = "Density") 

# Plots for hyper parameters for insight into the effect of the structure of the spatial field
par(mfrow=c(1,3))
plot(model1.res$marginals.var[[1]], ty = "l", xlab = expression(sigma[randomfield]^2), ylab = "Density") 
plot(model1.res$marginals.kap[[1]], type = "l", xlab = expression(kappa), ylab = "Density")
plot(model1.res$marginals.range[[1]], type = "l", xlab = "range nominal", ylab = "Density")

dev.off()

# Plot spatial field 
gproj <- inla.mesh.projector(Mesh,  dims = c(300, 300))
g.mean <- inla.mesh.project(gproj, model1$summary.random$spatial.field$mean)
g.sd <- inla.mesh.project(gproj, model1$summary.random$spatial.field$sd)

grid.arrange(levelplot(g.mean, scales=list(draw=F), xlab='', ylab='', main='mean',col.regions = heat.colors(16)),
             levelplot(g.sd, scal=list(draw=F), xla='', yla='', main='sd' ,col.regions = heat.colors(16)),
             nrow=1)
# "take a look at the spatial field to understand the spatial structure. It can inform on potential omitted variable - how much of the variance in the dependent variable is not explained by the covariates." but honestly idk what i am really looking at

# Other model plots
plot(model1, single = TRUE) # look into how to interpret?

```

INLA - model selection of covariates for PM2.5
```{r}
### Model selection 
# Model selection with WAIC (other criteria can be used)

# Create a vector of fixed effect covariates
covariates <- c("med_income_ranked", "dt_dist")

for(i in 1:length(covariates)) {
  
  # Build formula with random effect for county
  f1 <- as.formula(paste0("y ~ -1 + Intercept + score_0.5 + f(spatial.field, model=spde) + f(COUNTYFP10, model='iid') + ",
                         paste0(covariates[1:i], collapse = " + ")))
  
  model1 <- inla(f1, 
                 data = inla.stack.data(stk, spde=spde),
                 family = 'gaussian',
                 control.predictor = list(A=inla.stack.A(stk), compute=TRUE),
                 control.compute = list(dic = TRUE, cpo=TRUE, waic = TRUE))
  
  # Store model selection results
  model_selection <- if(i==1) {
    rbind(c(model = paste(covariates[1:i]), waic = model1$waic$waic))
  } else {
    rbind(model_selection, 
          c(model = paste(covariates[1:i], collapse = " + "), 
            waic = model1$waic$waic))
  }
}

model_selection

```

INLA - setting priors in the model for PM2.5
```{r}
## trying to set priors using GAM model results
formula2 <- y ~ -1 + Intercept + score_0.5 + f(spatial.field, model=spde) + med_income_ranked + dt_dist 

# Using the estimates and standard errors to inform priors
model2 <- inla(formula2, 
               data = inla.stack.data(stk, spde=spde),
               family = 'gaussian', 
               control.fixed = list(
                 mean = list(Intercept = -5.954,
                           score_0.5 = -7.859e-4,
                           med_income_ranked = 1.002e-6,
                           dt_dist = 0),  # No prior info for dt_dist from GAM
                 prec = list(Intercept = 1/(8.170e-04)^2,
                           score_0.5 = 1/(5.793e-04)^2,
                           med_income_ranked = 1/(7.453e-07)^2,
                           dt_dist = 1)  # Using default precision for dt_dist
               ),
               control.predictor = list(A=inla.stack.A(stk), compute=TRUE),
               control.compute = list(dic = TRUE, waic = TRUE, config = TRUE),
               verbose = FALSE)

dev.off()
# Plot effect estimates and credible intervals
Efxplot(list(model2))

plot(model2, single = TRUE) #don't know what this plot is and don't know why output is different from model 1

# Compare model 1 (no priors) and model 2 (priors) results
model1$summary.fix
model1$summary.hyperpar

model2$summary.fix
model2$summary.hyperpar



```

INLA stack (NDVI)
```{r}
# Create stack
stk_ndvi <- inla.stack(
  data=list(y=all_data_inc$ndvi_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_ranked = all_data_inc$med_income_ranked,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                               dt_dist = all_data_inc$dt_dist)))


#"Two projector matrices are needed A = list(A,1): the projector matrix for the latent field and a matrix that is a one-to-one map of the covariate. The latter matrix can simply be a constant rather than a diagonal matrix."
```


INLA - model fitting (no priors) for 
```{r}
# Specify formula
formula1_ndvi <- y ~ -1 + Intercept + score_0.5 +
  med_income_ranked + 
  dt_dist +
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
  #rw1 and rw2 don't work for the covariates, not sure how else to smooth, so right now just linear?? 

# Specify model
model1_ndvi<-inla(formula1_ndvi, 
             data = inla.stack.data(stk_ndvi, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_ndvi),compute = TRUE), 
             control.compute = list(cpo = TRUE))

```

INLA model results for NDVI
```{r}
###  Model results
# Posterior distribution summary of the fixed effects
model1_ndvi$summary.fix

# Posterior distribution summary of the hyper parameters (spatial field parameters)
model1_ndvi$summary.hyperpar

# Rounded summaries
round(model1_ndvi$summary.fixed,3) #med_income_ranked and dt_dist still come out out as all zeroes --> rounded...
round(model1_ndvi$summary.hyperpar,3)

# "The posterior distribution of the hyperparameters θ1 and θ2 are difficult to interpret. We can transform them into more interpretable quantities such as the range, and the variance. The range is of particular interest. It provides the distance value (in the unit of the point coordinates) above which spatial dependencies become negligible."
model1_ndvi.res<-inla.spde2.result(model1_ndvi, 'spatial.field', spde, do.transf=TRUE) #do.tranf = TRUE makes sure that marginals are calculated in the same scale as the data.

model1_ndvi.res$summary.log.range.nominal
model1_ndvi.res$summary.log.variance.nominal

# Plot effect estimate and credible intervals
graphics.off()
Efxplot(list(model1_ndvi))

# Plot for fixed parameters for insight into the effect of covariates
par(mfrow=c(2,2))
plot(model1_ndvi$marginals.fixed[[1]], ty = "l", xlab = expression(beta[0]), ylab = "Density")
plot(model1_ndvi$marginals.fixed[[2]], ty = "l", xlab = expression(beta[score_0.5]), ylab = "Density") 
plot(model1_ndvi$marginals.fixed[[3]], ty = "l", xlab = expression(beta[med_income_ranked]), ylab = "Density") 
plot(model1_ndvi$marginals.fixed[[4]], ty = "l", xlab = expression(beta[dt_dist]), ylab = "Density") 

# Plots for hyper parameters for insight into the effect of the structure of the spatial field
par(mfrow=c(1,3))
plot(model1_ndvi.res$marginals.var[[1]], ty = "l", xlab = expression(sigma[randomfield]^2), ylab = "Density") 
plot(model1_ndvi.res$marginals.kap[[1]], type = "l", xlab = expression(kappa), ylab = "Density")
plot(model1_ndvi.res$marginals.range[[1]], type = "l", xlab = "range nominal", ylab = "Density")


# Plot spatial field 
gproj <- inla.mesh.projector(Mesh,  dims = c(300, 300))
g.mean <- inla.mesh.project(gproj, model1_ndvi$summary.random$spatial.field$mean)
g.sd <- inla.mesh.project(gproj, model1_ndvi$summary.random$spatial.field$sd)

grid.arrange(levelplot(g.mean, scales=list(draw=F), xlab='', ylab='', main='mean',col.regions = heat.colors(16)),
             levelplot(g.sd, scal=list(draw=F), xla='', yla='', main='sd' ,col.regions = heat.colors(16)),
             nrow=1)
# "take a look at the spatial field to understand the spatial structure. It can inform on potential omitted variable - how much of the variance in the dependent variable is not explained by the covariates." but honestly idk what i am really looking at

# Other model plots
plot(model1_ndvi, single = TRUE) # look into how to interpret?

```

INLA - model selection of covariates for NDVI
```{r}
### Model selection 
# Model selection with WAIC (other criteria can be used)

# Create a vector of fixed effect covariates
covariates <- c("med_income_ranked", "dt_dist")

for(i in 1:length(covariates)) {
  
  # Build formula with random effect for county
  f1 <- as.formula(paste0("y ~ -1 + Intercept + score_0.5 + f(spatial.field, model=spde) + f(COUNTYFP10, model='iid') + ",
                         paste0(covariates[1:i], collapse = " + ")))
  
  model1 <- inla(f1, 
                 data = inla.stack.data(stk_ndvi, spde=spde),
                 family = 'gaussian',
                 control.predictor = list(A=inla.stack.A(stk_ndvi), compute=TRUE),
                 control.compute = list(dic = TRUE, cpo=TRUE, waic = TRUE))
  
  # Store model selection results
  model_selection <- if(i==1) {
    rbind(c(model = paste(covariates[1:i]), waic = model1$waic$waic))
  } else {
    rbind(model_selection, 
          c(model = paste(covariates[1:i], collapse = " + "), 
            waic = model1$waic$waic))
  }
}

model_selection
```

INLA - setting priors in the model for NDVI
```{r}
## trying to set priors using GAM model results
formula2_ndvi <- y ~ -1 + Intercept + score_0.5 + f(spatial.field, model=spde) + med_income_ranked + dt_dist 

# Using the estimates and standard errors to inform priors
model2_ndvi <- inla(formula2_ndvi, 
               data = inla.stack.data(stk_ndvi, spde=spde),
               family = 'gaussian', 
               control.fixed = list(
                 mean = list(Intercept = -2.422e-02,
                           score_0.5 = -2.128e-04,
                           med_income_ranked = -2.896e-06,
                           dt_dist = 0),  # No prior info for dt_dist
                 prec = list(Intercept = 1/(1.297e-03)^2,
                           score_0.5 = 1/(8.675e-04)^2,
                           med_income_ranked = 1/(1.179e-06)^2,
                           dt_dist = 1)  # Using default precision for dt_dist
               ),
               control.predictor = list(A=inla.stack.A(stk), compute=TRUE),
               control.compute = list(dic = TRUE, waic = TRUE, config = TRUE),
               verbose = FALSE)


# Plot effect estimates and credible intervals
Efxplot(list(model2_ndvi)) # not statistically significant compared to without priors
Efxplot(list(model1_ndvi))

plot(model2_ndvi, single = TRUE) #don't know what this plot is and don't know why output is different from model 1

# Compare model 1 (no priors) and model 2 (priors) results
model1_ndvi$summary.fix
model1_ndvi$summary.hyperpar

model2_ndvi$summary.fix
model2_ndvi$summary.hyperpar

```

INLA model functions for all variables
```{r}
run_spatial_model <- function(response_var, data, spde, Mesh) {
  # Create stack
  stk <- inla.stack(
    data = list(y = data[[response_var]]),
    A = list(A, 1),
    effects = list(
      c(list(Intercept = 1), iset),
      list(
        score_0.5 = data$score_0.5,
        med_income_ranked = data$med_income_ranked,
        COUNTYFP10 = data$COUNTYFP10,
        dt_dist = data$dt_dist
      )
    )
  )
  
  # Formula
  formula <- y ~ -1 + Intercept + score_0.5 +
    med_income_ranked + 
    dt_dist +
    f(spatial.field, model = spde) +
    f(COUNTYFP10, model = "iid")
  
  # Fit model
  model <- inla(formula, 
                data = inla.stack.data(stk, spde = spde),
                family = 'gaussian', 
                control.predictor = list(A = inla.stack.A(stk), compute = TRUE), 
                control.compute = list(cpo = TRUE, dic = TRUE, waic = TRUE))
  
  # Create spatial field results
  model.res <- inla.spde2.result(model, 'spatial.field', spde, do.transf = TRUE)
  
  # Return model results
  return(list(
    model = model,
    fixed_effects = round(model$summary.fixed, 3),
    hyperpar = round(model$summary.hyperpar, 3),
    range = model.res$summary.log.range.nominal,
    variance = model.res$summary.log.variance.nominal,
    waic = model$waic$waic,
    response = response_var
  ))
} # waic doesn't mean much here bc not doing model selection, would need to do that separately

# Run models for all response variables
response_vars <- c("pm_change", "o3_change", "no2_change", "ndvi_change", "mean_temp")

results <- map(response_vars, ~run_spatial_model(
  response_var = .x,
  data = all_data_inc,
  spde = spde,
  Mesh = Mesh
)) %>%
  setNames(response_vars)

# Print summarized results
print_model_results <- function(results_list) {
  for(var in names(results_list)) {
    cat("\n\n=== Results for", var, "===\n")
    cat("\nFixed Effects:\n")
    print(results_list[[var]]$fixed_effects)
    cat("\nHyper Parameters:\n")
    print(results_list[[var]]$hyperpar)
    cat("\nRange:\n")
    print(results_list[[var]]$range)
    cat("\nVariance:\n")
    print(results_list[[var]]$variance)
    cat("\nWAIC:\n")
    print(results_list[[var]]$waic)
    cat("\n----------------------------------------\n")
  }
}

# Print results
print_model_results(results)


### Effect estimate and credible interval plots

# For PM2.5 change
graphics.off()
Efxplot(list(results$pm_change$model))

# For O3 change
graphics.off()
Efxplot(list(results$o3_change$model))

# For NO2 change
graphics.off()
Efxplot(list(results$no2_change$model))

# For NDVI change
graphics.off()
Efxplot(list(results$ndvi_change$model))

# For mean temperature
graphics.off()
Efxplot(list(results$mean_temp$model))
```

INLA - setting priors in the model for NO2
```{r}
# Create stack
stk_no2<- inla.stack(
  data=list(y=all_data_inc$no2_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_ranked = all_data_inc$med_income_ranked,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                               dt_dist = all_data_inc$dt_dist)))

formula2_no2 <- y ~ -1 + Intercept + score_0.5 +
  med_income_ranked + 
  dt_dist +
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect

model2_no2 <- inla(
  formula2_no2, 
  data = inla.stack.data(stk_no2, spde=spde), 
  family = 'gaussian',
  control.fixed = list(
    mean = list(
      Intercept = -14.54,
      score_0.5 = -0.01881,
      med_income_ranked = -0.0001338,
      dt_dist = 0  # Keeping as is since no prior info
    ),
    prec = list(
      Intercept = 1/(0.03376)^2,
      score_0.5 = 1/(0.02485)^2,
      med_income_ranked = 1/(0.00003092)^2,
      dt_dist = 1  # Keeping default precision
    )
  ),
  control.predictor = list(A=inla.stack.A(stk_no2), compute=TRUE),
  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE),
  verbose = FALSE
)

Efxplot(list(model2_no2))

```

Old INLA code 
```{r}
# # Making the A matrix (JH - A matrix is not the same as MeshA, this is using MeshB)
# 
# Matrix_A <- inla.spde.make.A(Mesh, loc = Locations) # Making A matrix
# Hosts.spde = inla.spde2.pcmatern(mesh = Mesh, prior.range = c(10, 0.5), prior.sigma = c(.5, .5)) # Making SPDE ##JH- NOT SURE ABOUT PRIOR RANGE THIS WAS FROM THE TUTORIAL
# 
# Matrix_A_no_prior <- inla.spde.make.A(Mesh, loc = Locations) # Making A matrix
# 
# 
# # use frequentist results to inform the priors for the inla - range of effect estimates from gam
# #using the priors as a starting point for simulations 
# #try a few different priors to see how stable results are in the posteriors --? pick a few different parameterizations of models and use them as the starting points.
# 
# w.Host <- inla.spde.make.index('w', n.spde = Hosts.spde$n.spde) # making the w
# 
# # Making the model matrix #### 
# 
# # make the model matrix using the final model selection formula without a response variable
# #trying with IM_PM_1 specifications
# X0 <- model.matrix(~ score_0.5 + med_income_ranked, family = "gaussian", data = all_data_inc) 
# # convert to a data frame. Eliminate the base level of the first categorical variable if applicable (you will manually specify an intercept below)  (JH - not applicable?)
# X <- as.data.frame(X0) 
# 
# head(X)
# 
# # Making the stack #### 
# #did not add county as random effect here. if want to, need to first update X to include it. column names in X need to match effect names when specifying the stack and the vector length of A.
# 
# N <- nrow(all_data_inc)
# 
# # this code used to work?? but now gets an error.
# # Stack <- inla.stack(
# #   data = list(y = all_data_inc$pm_change), # specify the response variable
# #   
# #   A = list(1, 1, Matrix_A), # Vector of Multiplication factors for random and fixed effects              
# #   
# #   effects = list(
# #     
# #     Intercept = rep(1, N), # specify the manual intercept!
# #     
# #     X = X, # attach the model matrix
# #     
# #     w = w.Host)) # attach the w 
# 
# #revised above code that runs as o 16.12.24
# Stack <- inla.stack(
#   data = list(y = all_data_inc$pm_change),
#   A = list(1, 1, Matrix_A_no_prior),
#   effects = list(
#     Intercept = rep(1, N),
#     X = X[, c("score_0.5", "med_income_ranked")],  # Exclude intercept column
#     spatial = w.Host$w
#   )
# )
# # adding stack to model
# f1 <- as.formula(paste0("y ~ -1 + Intercept + ", paste0(colnames(X), collapse = " + ")))
# 
# IM1 <- inla(f1, # Base model (no random effects)
#             family = "gaussian",
#             data = inla.stack.data(Stack),
#             control.compute = list(dic = TRUE),
#             control.predictor = list(A = inla.stack.A(Stack))
# )
# 
# Efxplot(list(IM1))
# summary(IM1)
# 
# ##look into the warnings.
# 
# #look into diagnostics e.g., convergence plots, measure of conversion  rhat - but don't rely on it - see if it's a thing in INLA

# ### Model fitting
# 
# #maybe not supposed to include score_0.5 here??
# #model0 not accounting for stack/covariates
# formula0<- y ~ -1 + Intercept + f(spatial.field, model=spde) 
# 
# 
# model0 <- inla(formula0,
#                data = list (y = all_data_inc$pm_change,
#                             Intercept = rep(1,spde$n.spde),
#                             spatial.field = 1:spde$n.spde),
#                control.predictor = list (A = A, compute = T),
#                control.compute = list (cpo = T))
# 
# round(model0$summary.fixed,3)
# round(model0$summary.hyperpar[1,],3)


# model0<-inla(formula0, #the formula
#              data= inla.stack.data(stk,spde=spde),  #the data stack
#              family= 'gaussian',   #which family the data comes from
#              control.predictor=list(A=inla.stack.A(stk),compute=TRUE),  #compute gives you the marginals of the linear predictor
#              control.compute = list(dic = TRUE, waic = TRUE, config = TRUE), #model diagnostics and config = TRUE gives you the GMRF
#              verbose = FALSE) #can include verbose=TRUE to see the log of the model runs


```


```{r}

```




Neighbor relationships for MRF smooth model
```{r}
# Establish neighbor relationships for MRF model
nb <- poly2nb(all_data_inc, row.names = all_data_inc$GEOID)
names(nb) <- attr(nb, "region.id") 
```

PM 2.5 MRF smooth model
```{r}
# MRF smooth model for pm 2.5
mrf_income_pm <- bam(pm_change ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=200, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 

mrf_income_pm_15 <- bam(pm_change ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=400, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 


#changing the k, whether or not income has a spline/log/nothing affects sig of score_0.5; k needs to be at least 15 for residuals etc to be okay (k=15 highest sig but maybe not best diagnostic plots); k needs to be at least ~400 for Moran's I to be insignificant

summary(mrf_income_pm)
gam.check(mrf_income_pm)
concurvity(mrf_income_pm, full = TRUE)

summary(mrf_income_pm_15)
gam.check(mrf_income_pm_15)

##### Calculating Moran's I to assess residual spatial autocorrelation #####

# First identify areas with no neighbors
no_neighbors <- which(card(nb) == 0)
if(length(no_neighbors) > 0) {
    print("Areas with no neighbors:")
    print(no_neighbors)
}

# Use zero.policy = TRUE (keeping all observations - isolated tracts given a weight of zero?)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get model residuals
residuals <- residuals(mrf_income_pm)

# Perform Moran's I test with zero.policy
moran_test <- moran.test(residuals, listw, zero.policy = TRUE)
print(moran_test)

### Create Moran's I plot ###

# Calculate lagged residuals with zero.policy
lag_residuals <- lag.listw(listw, residuals, zero.policy = TRUE)

# Create data frame for plotting
moran_df <- data.frame(
    residuals = residuals,
    lag_residuals = lag_residuals
)

# Moran scatter plot
ggplot(moran_df, aes(x = residuals, y = lag_residuals)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Residuals",
        y = "Spatially Lagged Residuals",
        title = "Moran's I Scatter Plot of GAM Residuals",
        subtitle = paste("Moran's I =", round(moran_test$estimate[1], 3),
                        "\np-value =", round(moran_test$p.value, 3))
    ) +
    theme_minimal()

```

O3 MRF smooth model
```{r}
#MRF smooth model for o3
mrf_income_o3 <- bam(o3_change ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=450, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 

summary(mrf_income_o3)
gam.check(mrf_income_o3)
concurvity(mrf_income_o3, full = TRUE)

#####Calculating Moran's I to assess residual spatial autocorrelation

# First identify areas with no neighbors
no_neighbors <- which(card(nb) == 0)
if(length(no_neighbors) > 0) {
    print("Areas with no neighbors:")
    print(no_neighbors)
}

# Use zero.policy = TRUE (keeping all observations - isolated tracts given a weight of zero?)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get model residuals
residuals <- residuals(mrf_income_o3)

# Perform Moran's I test with zero.policy
moran_test <- moran.test(residuals, listw, zero.policy = TRUE)
print(moran_test)

### Create Moran's I plot ###

# Calculate lagged residuals with zero.policy
lag_residuals <- lag.listw(listw, residuals, zero.policy = TRUE)

# Create data frame for plotting
moran_df <- data.frame(
    residuals = residuals,
    lag_residuals = lag_residuals
)

# Moran scatter plot
ggplot(moran_df, aes(x = residuals, y = lag_residuals)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Residuals",
        y = "Spatially Lagged Residuals",
        title = "Moran's I Scatter Plot of GAM Residuals",
        subtitle = paste("Moran's I =", round(moran_test$estimate[1], 3),
                        "\np-value =", round(moran_test$p.value, 3))
    ) +
    theme_minimal()
```


NO2 MRF smooth model
```{r}
#MRF smooth model for NO2
mrf_income_no2 <- bam(no2_change ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=300, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 

#at k=5 is significant but residuals weird, when k increases, residuals and stuff still patterned-ish but better overall and no longer sig

summary(mrf_income_no2)
gam.check(mrf_income_no2)
concurvity(mrf_income_no2, full = TRUE)

#####Calculating Moran's I to assess residual spatial autocorrelation

# First identify areas with no neighbors
no_neighbors <- which(card(nb) == 0)
if(length(no_neighbors) > 0) {
    print("Areas with no neighbors:")
    print(no_neighbors)
}

# Use zero.policy = TRUE (keeping all observations - isolated tracts given a weight of zero?)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get model residuals
residuals <- residuals(mrf_income_no2)

# Perform Moran's I test with zero.policy
moran_test <- moran.test(residuals, listw, zero.policy = TRUE)
print(moran_test)

### Create Moran's I plot ###

# Calculate lagged residuals with zero.policy
lag_residuals <- lag.listw(listw, residuals, zero.policy = TRUE)

# Create data frame for plotting
moran_df <- data.frame(
    residuals = residuals,
    lag_residuals = lag_residuals
)

# Moran scatter plot
ggplot(moran_df, aes(x = residuals, y = lag_residuals)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Residuals",
        y = "Spatially Lagged Residuals",
        title = "Moran's I Scatter Plot of GAM Residuals",
        subtitle = paste("Moran's I =", round(moran_test$estimate[1], 3),
                        "\np-value =", round(moran_test$p.value, 3))
    ) +
    theme_minimal()
```

NDVI MRF smooth model
```{r}
#MRF smooth model for NDVI
# Create clean dataset without 8 NA values (follow up)
all_data_inc_clean <- all_data_inc[is.finite(all_data_inc$ndvi_change), ]

# Create new neighborhood structure from cleaned data
nb_clean <- poly2nb(all_data_inc_clean)
names(nb_clean) <- all_data_inc_clean$GEOID

mrf_income_ndvi <- bam(ndvi_change ~ score_0.5 + med_income_ranked + 
                      s(GEOID, bs='mrf', k=300, xt=list(nb=nb_clean)), 
                      family = scat, data = all_data_inc_clean, method = "REML")


# 
# mrf_income_ndvi <- bam(ndvi_change ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=300, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 

summary(mrf_income_ndvi)
gam.check(mrf_income_ndvi)
concurvity(mrf_income_ndvi, full = TRUE) 

#K NEEDS TO BE 300 FOR MORANS I TO BE INSIGNIFICANT

#####Calculating Moran's I to assess residual spatial autocorrelation

# First identify areas with no neighbors
no_neighbors <- which(card(nb) == 0)
if(length(no_neighbors) > 0) {
    print("Areas with no neighbors:")
    print(no_neighbors)
}

# Use zero.policy = TRUE (keeping all observations - isolated tracts given a weight of zero?)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get model residuals
residuals <- residuals(mrf_income_ndvi)

# Perform Moran's I test with zero.policy
moran_test <- moran.test(residuals, listw, zero.policy = TRUE)
print(moran_test)

### Create Moran's I plot ###

# Calculate lagged residuals with zero.policy
lag_residuals <- lag.listw(listw, residuals, zero.policy = TRUE)

# Create data frame for plotting
moran_df <- data.frame(
    residuals = residuals,
    lag_residuals = lag_residuals
)

# Moran scatter plot
ggplot(moran_df, aes(x = residuals, y = lag_residuals)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Residuals",
        y = "Spatially Lagged Residuals",
        title = "Moran's I Scatter Plot of GAM Residuals",
        subtitle = paste("Moran's I =", round(moran_test$estimate[1], 3),
                        "\np-value =", round(moran_test$p.value, 3))
    ) +
    theme_minimal()
```

Mean Temp MRF smooth model
```{r}
#MRF smooth model for 2016 mean temp
mrf_income_temp <- bam(mean_temp ~ score_0.5 + med_income_ranked + s(GEOID, bs='mrf', k=400, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML")  
#terrible residual plots

summary(mrf_income_temp)
gam.check(mrf_income_temp)
concurvity(mrf_income_temp, full = TRUE)

#####Calculating Moran's I to assess residual spatial autocorrelation

# First identify areas with no neighbors
no_neighbors <- which(card(nb) == 0)
if(length(no_neighbors) > 0) {
    print("Areas with no neighbors:")
    print(no_neighbors)
}

# Use zero.policy = TRUE (keeping all observations - isolated tracts given a weight of zero?)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get model residuals
residuals <- residuals(mrf_income_ndvi)

# Perform Moran's I test with zero.policy
moran_test <- moran.test(residuals, listw, zero.policy = TRUE)
print(moran_test)

### Create Moran's I plot ###

# Calculate lagged residuals with zero.policy
lag_residuals <- lag.listw(listw, residuals, zero.policy = TRUE)

# Create data frame for plotting
moran_df <- data.frame(
    residuals = residuals,
    lag_residuals = lag_residuals
)

# Moran scatter plot
ggplot(moran_df, aes(x = residuals, y = lag_residuals)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Residuals",
        y = "Spatially Lagged Residuals",
        title = "Moran's I Scatter Plot of GAM Residuals",
        subtitle = paste("Moran's I =", round(moran_test$estimate[1], 3),
                        "\np-value =", round(moran_test$p.value, 3))
    ) +
    theme_minimal()
```

Creating composite z-score for environmental exposures
```{r}
# Calculate z-scores for each variable
all_data_inc$pm_z <- scale(all_data_inc$pm_change)
all_data_inc$o3_z <- scale(all_data_inc$o3_change)
all_data_inc$no2_z <- scale(all_data_inc$no2_change)
all_data_inc$ndvi_z <- scale(all_data_inc$ndvi_change)
all_data_inc$temp_z <- scale(all_data_inc$mean_temp)

# Create composite z-score (average of all z-scores)
all_data_inc$composite_z <- rowMeans(
  cbind(
    all_data_inc$pm_z,
    all_data_inc$o3_z,
    all_data_inc$no2_z,
    all_data_inc$ndvi_z,
    all_data_inc$temp_z
  ),
  na.rm = TRUE
)

summary(all_data_inc$composite_z)
hist(all_data_inc$composite_z)
shapiro.test(all_data_inc$composite_z)

qqnorm(all_data_inc$composite_z)

# Save composite_z to new data frame and export
# composite_z <- all_data_inc  %>% 
#   dplyr::select(GEOID, score_0.5, composite_z, med_income_2000_impute, med_income_ranked, lat, long)  
# composite_z <- as.data.frame(composite_z)
# 
# # Write only the non-geometry columns
# write.csv(composite_z[, 1:7], here("composite_z.csv"), 
#           row.names = FALSE,
#           quote = TRUE)  # Changed to TRUE to handle the GEOID factor
# 
# # Read it back 
# composite_z <- read.csv(here("composite_z.csv"), 
#          check.names = FALSE,
#          stringsAsFactors = FALSE)


```

GAM model using composite z-score
```{r}
# Create new GAM model with composite z-score
mrf_income_composite <- bam(
  composite_z ~ score_0.5 + med_income_ranked + 
    s(GEOID, bs='mrf', k=563, xt=list(nb=nb)),
  family = scat,
  data = all_data_inc,
  method = "REML"
) 

#sig when k = 100 but Moran's I is 0.265, insig when k = 300, Moran's I is 0.074 but p is sig
 
summary(mrf_income_composite)
gam.check(mrf_income_composite)

##### Calculating Moran's I to assess residual spatial autocorrelation ##### 

# First identify areas with no neighbors
no_neighbors <- which(card(nb) == 0)
if(length(no_neighbors) > 0) {
    print("Areas with no neighbors:")
    print(no_neighbors)
}

# Use zero.policy = TRUE (keeping all observations - isolated tracts given a weight of zero?)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get model residuals
residuals <- residuals(mrf_income_composite)

# Perform Moran's I test with zero.policy
moran_test <- moran.test(residuals, listw, zero.policy = TRUE)
print(moran_test)

### Create Moran's I plot ###

# Calculate lagged residuals with zero.policy
lag_residuals <- lag.listw(listw, residuals, zero.policy = TRUE)

# Create data frame for plotting
moran_df <- data.frame(
    residuals = residuals,
    lag_residuals = lag_residuals
)

# Moran scatter plot
ggplot(moran_df, aes(x = residuals, y = lag_residuals)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Residuals",
        y = "Spatially Lagged Residuals",
        title = "Moran's I Scatter Plot of GAM Residuals",
        subtitle = paste("Moran's I =", round(moran_test$estimate[1], 3),
                        "\np-value =", round(moran_test$p.value, 3))
    ) +
    theme_minimal()

```

Testing k vs Moran's I for MRF composite model
```{r}
# Function to get number of unique locations
n_locations <- length(unique(all_data_inc$GEOID))

# Calculate minimum suggested k
k_min <- ceiling(n_locations^(1/3))  # cube root rule of thumb

# Define sequence of k values to test
k_values <- seq(k_min, 500, by = 50)

# Function to run k.check for multiple k values
test_k_values <- function(data, k_values) {
  results <- data.frame(
    k = numeric(),
    edf = numeric(),
    k_index = numeric(),
    p_value = numeric(),
    gcv = numeric()
  )
  
  for(k in k_values) {
    cat(sprintf("\nTesting k = %d\n", k))
    
    # Fit model with current k
    tryCatch({
      test_model <- bam(
        composite_z ~ score_0.5 + med_income_ranked + 
          s(GEOID, bs='mrf', k=k, xt=list(nb=nb)),
        family = scat,
        data = data,
        method = "REML"
      )
      
      # Extract model summary
      model_summary <- summary(test_model)
      
      # Get basic diagnostics that don't depend on gam.check
      new_row <- data.frame(
        k = k,
        edf = sum(model_summary$edf),
        k_index = NA,  # Will be filled if gam.check works
        p_value = NA,  # Will be filled if gam.check works
        gcv = test_model$gcv.ubre
      )
      
      # Try to get gam.check results
      tryCatch({
        check <- gam.check(test_model)
        new_row$k_index <- check$k.index
        new_row$p_value <- check$s.table[4]
      }, error = function(e) {
        cat(sprintf("Warning: Could not compute gam.check for k=%d\n", k))
      })
      
      results <- rbind(results, new_row)
      
    }, error = function(e) {
      cat(sprintf("Error fitting model for k=%d: %s\n", k, e$message))
    })
  }
  return(results)
}

# Print minimum suggested k
cat(sprintf("Minimum suggested k based on number of locations: %d\n", k_min))

# Test k values
k_results <- test_k_values(all_data_inc, k_values)

# Print results
print(k_results)

# Plot available results
par(mfrow=c(2,2))

# Only plot if we have results
if(nrow(k_results) > 0) {
  # P-value plot (if available)
  if(!all(is.na(k_results$p_value))) {
    plot(k_results$k, k_results$p_value, type="b", 
         main="P-value vs k", xlab="k", ylab="p-value")
  }
  
  # GCV plot
  plot(k_results$k, k_results$gcv, type="b", 
       main="GCV Score vs k", xlab="k", ylab="GCV")
  
  # EDF plot
  plot(k_results$k, k_results$edf, type="b", 
       main="EDF vs k", xlab="k", ylab="Effective Degrees of Freedom")
  
  # k-index plot (if available)
  if(!all(is.na(k_results$k_index))) {
    plot(k_results$k, k_results$k_index, type="b", 
         main="k-index vs k", xlab="k", ylab="k-index")
  }
}

# Function to get Moran's I for model residuals
check_spatial_autocorr <- function(model, nb) {
  residuals <- residuals(model)
  moran_test <- spdep::moran.test(residuals, 
                                 spdep::nb2listw(nb, zero.policy=TRUE),
                                 zero.policy=TRUE)
  return(c(
    moran_i = moran_test$estimate[1],
    p_value = moran_test$p.value
  ))
}

# Check which areas have no neighbors
n_neighbors <- sapply(nb, length)
isolated_areas <- which(n_neighbors == 0)
if(length(isolated_areas) > 0) {
  cat("Areas with no neighbors:\n")
  print(isolated_areas)
}

# Function to get Moran's I for model residuals
check_spatial_autocorr <- function(model, nb) {
  residuals <- residuals(model)
  moran_test <- spdep::moran.test(residuals, 
                                 spdep::nb2listw(nb, zero.policy=TRUE),
                                 zero.policy=TRUE)
  return(moran_test)  # Return the full test object
}

# Calculate Moran's I for raw composite z-score
raw_moran <- spdep::moran.test(all_data_inc$composite_z, 
                              spdep::nb2listw(nb, zero.policy=TRUE),
                              zero.policy=TRUE)
cat(sprintf("\nRaw data Moran's I:\n"))
cat(sprintf("Moran's I statistic: %.3f\n", raw_moran$estimate["Moran I statistic"]))
cat(sprintf("p-value: %.3e\n", raw_moran$p.value))

# Test k values and check both model fit and spatial autocorrelation
test_k_values <- function(data, k_values, raw_moran) {
  results <- data.frame()
  
  # Store raw Moran's I value
  raw_moran_i <- raw_moran$estimate["Moran I statistic"]
  
  for(k in k_values) {
    cat(sprintf("\nTesting k = %d\n", k))
    
    # Fit model
    test_model <- gam(
      composite_z ~ score_0.5 + med_income_ranked + 
        s(GEOID, bs='mrf', k=k, xt=list(nb=nb)),
      family = scat,
      data = data,
      method = "REML"
    )
    
    # Get model summary
    model_summary <- summary(test_model)
    
    # Get Moran's I results
    moran_results <- check_spatial_autocorr(test_model, nb)
    residual_moran_i <- moran_results$estimate["Moran I statistic"]
    
    # Calculate reduction in Moran's I
    moran_reduction <- ((raw_moran_i - residual_moran_i) / abs(raw_moran_i)) * 100
    
    # Store results
    results <- rbind(results,
      data.frame(
        k = k,
        edf = sum(model_summary$edf),
        score = test_model$gcv.ubre,
        moran_i = residual_moran_i,
        moran_p = moran_results$p.value,
        moran_reduction = moran_reduction
      )
    )
    
    # Print current results
    cat(sprintf("EDF: %.1f\n", sum(model_summary$edf)))
    cat(sprintf("Score: %.3f\n", test_model$gcv.ubre))
    cat(sprintf("Residual Moran's I: %.3f (p=%.3e)\n", 
                residual_moran_i, moran_results$p.value))
    cat(sprintf("Reduction in Moran's I: %.1f%%\n", moran_reduction))
  }
  return(results)
}

# Test reasonable k values
k_values <- c(113, 163, 213, 263, 313, 363, 413, 463, 513, 563)
k_values_2 <- c(613, 663, 713, 763, 813)
k_results <- test_k_values(all_data_inc, k_values, raw_moran)
k_results_2 <- test_k_values(all_data_inc, k_values_2, raw_moran)


# Print final comparison table
cat("\nSummary Results:\n")
print(data.frame(
  k = k_results$k,
  EDF = round(k_results$edf, 1),
  Score = round(k_results$score, 3),
  Morans_I = round(k_results$moran_i, 3),
  Moran_p = format(k_results$moran_p, scientific = TRUE),
  Reduction = sprintf("%.1f%%", k_results$moran_reduction)
))

# Plot results
par(mfrow=c(2,3))

# Score vs k
plot(k_results$k, k_results$score, type="b", 
     main="Score vs k", xlab="k", ylab="Score")

# EDF vs k
plot(k_results$k, k_results$edf, type="b", 
     main="EDF vs k", xlab="k", ylab="Effective Degrees of Freedom")

# Moran's I vs k
plot(k_results$k, k_results$moran_i, type="b", 
     main="Moran's I vs k", xlab="k", ylab="Moran's I")
abline(h=raw_moran$estimate["Moran I statistic"], col="red", lty=2)
text(k_values[1], raw_moran$estimate["Moran I statistic"], 
     "Raw Data Moran's I", pos=3, col="red")

# Moran's p-value vs k
plot(k_results$k, k_results$moran_p, type="b", 
     main="Moran's I p-value vs k", xlab="k", ylab="p-value")
abline(h=0.05, col="red", lty=2)

# Reduction in Moran's I
plot(k_results$k, k_results$moran_reduction, type="b",
     main="% Reduction in Moran's I",
     xlab="k", 
     ylab="Percent Reduction")
```

Hurricane evacuation zones
```{r}
# Load hurricane evauation zones shapefile 
hurricane <- st_read(here("data", "hurricane_evacuation_zones.shp"))

# Extract residuals and prepare data
all_data_inc$mrf_residuals <- residuals(mrf_income_composite)

# Clean geometries
hurricane_clean <- hurricane %>%
  st_make_valid() %>%
  st_buffer(0) %>%
  st_transform(st_crs(all_data_inc))

all_data_inc_clean <- all_data_inc %>%
  st_make_valid() %>%
  st_buffer(0)

# Use centroids for spatial join
tract_centroids <- all_data_inc_clean %>%
  st_centroid() %>%
  select(GEOID, mrf_residuals)

# Spatial join with hurricane zones
combined_data <- st_join(
  tract_centroids,
  hurricane_clean %>% select(hurricane_),
  join = st_intersects
) %>%
  filter(!is.na(hurricane_))

# Statistical Analysis

# ANOVA test
anova_result <- aov(mrf_residuals ~ hurricane_, data = combined_data)
anova_summary <- summary(anova_result)

# Kruskal-Wallis test
kw_result <- kruskal.test(mrf_residuals ~ hurricane_, data = combined_data)

# Summary statistics by zone 
zone_summary <- combined_data %>%
  st_drop_geometry() %>%
  group_by(hurricane_) %>%
  summarise(
    count = length(mrf_residuals),
    mean_residual = mean(mrf_residuals, na.rm = TRUE),
    median_residual = median(mrf_residuals, na.rm = TRUE),
    sd_residual = sd(mrf_residuals, na.rm = TRUE),
    extreme_count = sum(abs(mrf_residuals) > 2*sd(mrf_residuals, na.rm = TRUE), na.rm = TRUE)
  ) %>%
  mutate(
    extreme_percentage = (extreme_count / count) * 100
  )

# Box plot
residual_plot <- ggplot(combined_data %>% st_drop_geometry(), 
                       aes(x = factor(hurricane_), y = mrf_residuals)) +
  geom_boxplot(outlier.alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "MRF Model Residuals by Hurricane Evacuation Zone",
    x = "Hurricane Zone",
    y = "Residuals"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Spatial visualization
viz_data <- all_data_inc_clean %>%
  left_join(
    combined_data %>% 
      st_drop_geometry() %>%
      select(GEOID, hurricane_),
    by = "GEOID"
  )

spatial_plot <- ggplot() +
  geom_sf(data = viz_data, aes(fill = hurricane_)) +
  scale_fill_viridis_d(na.value = "grey80") +
  theme_minimal() +
  labs(
    title = "Hurricane Evacuation Zones with MRF Model Residuals",
    fill = "Hurricane Zone"
  )

# Scatter plot 
scatter_plot <- ggplot(combined_data %>% st_drop_geometry(), 
                      aes(x = factor(hurricane_), y = mrf_residuals)) +
  geom_jitter(alpha = 0.5, width = 0.2) +
  stat_summary(fun = mean, geom = "point", color = "red", size = 3) +
  stat_summary(fun = mean, geom = "line", color = "red", group = 1) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  theme_minimal() +
  labs(
    title = "Residuals Distribution by Hurricane Zone",
    subtitle = "Red points and line show mean values",
    x = "Hurricane Zone",
    y = "Residuals"
  )


# Test for normality within each zone
normality_tests <- combined_data %>%
  st_drop_geometry() %>%
  group_by(hurricane_) %>%
  summarise(
    sample_size = length(mrf_residuals),
    shapiro_p_value = if(length(mrf_residuals) >= 3 && 
                        length(mrf_residuals) <= 5000) {
      shapiro.test(mrf_residuals)$p.value
    } else {
      NA
    }
  )

# Print results
cat("\nSummary Statistics by Hurricane Zone:\n")
print(zone_summary)

cat("\nANOVA Results:\n")
print(anova_summary)

cat("\nKruskal-Wallis Test Results:\n")
print(kw_result)

cat("\nNormality Test Results by Zone:\n")
print(normality_tests)

# Calculate effect size (Eta-squared)
ss_total <- sum((combined_data$mrf_residuals - mean(combined_data$mrf_residuals))^2)
eta_squared <- anova_summary[[1]]$"Sum Sq"[1] / ss_total

cat("\nEffect Size (Eta-squared):", round(eta_squared, 4), "\n")

# Display plots
print(residual_plot)
print(spatial_plot)
print(scatter_plot)

# Store all results in a list
results <- list(
  summary_stats = zone_summary,
  anova_results = anova_summary,
  kruskal_wallis = kw_result,
  normality_tests = normality_tests,
  effect_size = eta_squared
)
```

```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(rstatix)
library(dunn.test)
library(effsize)

# Perform pairwise t-tests with multiple comparison adjustment
pairwise_tests <- pairwise.t.test(
  combined_data$mrf_residuals,
  combined_data$hurricane_,
  p.adjust.method = "bonferroni"
)

pairwise_tests

# Perform Dunn test (non-parametric alternative)
dunn_test <- dunn.test(
  combined_data$mrf_residuals,
  combined_data$hurricane_,
  method = "bonferroni"
)

# Create pairwise comparison visualization using ggplot2
# Convert matrix to long format for plotting
p_value_matrix <- as.matrix(pairwise_tests$p.value)
p_value_df <- as.data.frame(p_value_matrix) %>%
  mutate(group1 = rownames(.)) %>%
  pivot_longer(
    cols = -group1,
    names_to = "group2",
    values_to = "p_value"
  ) %>%
  filter(!is.na(p_value))

# Create heatmap
ggplot(p_value_df, aes(x = group1, y = group2, fill = -log10(p_value))) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.3f", p_value)), size = 3) +
  scale_fill_gradient2(
    low = "white",
    high = "red",
    mid = "pink",
    midpoint = -log10(0.05),
    name = "-log10(p-value)"
  ) +
  theme_minimal() +
  labs(
    title = "Pairwise Comparisons of Hurricane Zone Residuals",
    subtitle = "Bonferroni-adjusted p-values",
    x = "Zone",
    y = "Zone"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#interpret???
```

```{r}
```

#########################################
########################################

ANALYZE COOK DISTANCE AND RESIDUALS ETC FOR MRF MODELS
```{r}
# Calculate Cook's distance
cooks_d <- cooks.distance(mrf_income_pm)
plot(cooks_d, type = "h", main = "Cook's Distance")
abline(h = 4/length(cooks_d), col = "red")  # Rule of thumb threshold

# Identify potential influential points
influential <- which(cooks_d > 4/length(cooks_d))

influential_data <- all_data_inc[influential, ]
print(influential_data)  # Look at characteristics of these points

# Investigate why these points are influential:
# Plot their locations
plot(all_data_inc_na$long, all_data_inc_na$lat)
points(all_data_inc_na$long[influential], 
       all_data_inc_na$lat[influential], 
       col = "red", pch = 16)

```

################################

```{r}

library(purrr)
library(ggplot2)

# List of model names
model_names <- c(
  "mrf_income_pm", "mrf_income_pm_baseline", "mrf_income_pm_5",
  "mrf_income_o3", "mrf_income_o3_baseline",
  "mrf_income_no2", "mrf_income_no2_baseline", "mrf_income_no2_15",
  "mrf_income_ndvi", "mrf_income_ndvi_baseline", "mrf_income_ndvi_15",
  "mrf_income_temp"
)

# Function to calculate Cook's Distance, plot results, and identify influential points
process_model <- function(model_name) {
  
  # Retrieve the model from the global environment using get()
  model <- get(model_name)
  
  # Calculate Cook's Distance
  cooks_d <- cooks.distance(model)
  
  # Plot Cook's Distance with model name in title
  plot(cooks_d, type = "h", main = paste("Cook's Distance:", model_name))
  abline(h = 4/length(cooks_d), col = "red")  # Rule of thumb threshold
  
  # Identify influential points
  influential <- which(cooks_d > 4/length(cooks_d))
  
  # Retrieve the data for the influential points
  influential_data <- all_data_inc[influential, ]
  print(paste("Influential points for", model_name))
  print(influential_data)  # Look at characteristics of these points
  
  # Plot the locations of influential points on the map
  plot(all_data_inc_na$long, all_data_inc_na$lat,
       main = paste("Influential Points for", model_name),  # Add model name to the map plot title
       xlab = "Longitude", ylab = "Latitude")
  points(all_data_inc_na$long[influential], 
         all_data_inc_na$lat[influential], 
         col = "red", pch = 16)
  
  return(influential_data)
}

# Use purrr::map to apply the function to all models
influential_points_list <- map(model_names, process_model)

# If you want to inspect the influential points for each model, you can print the list
influential_points_list

```
trying SLM and SEM models again (no spatial autocorrelation, but linear so qqplots bad)
```{r}

#### QQ plots with confidence intervals
get_qq_plot <- function(model) {
  
  model_resid <- data.frame(residuals = model$residuals)
  
  plot <- ggplot(data = model_resid, mapping = aes(sample = residuals)) + 
    geom_qq_band(bandType = "ks", mapping = aes(fill = "KS"), alpha = 0.5) +
    geom_qq_band(bandType = "ts", mapping = aes(fill = "TS"), alpha = 0.5) +
    geom_qq_band(bandType = "pointwise", mapping = aes(fill = "Normal"), alpha = 0.5) +
    geom_qq_band(bandType = "boot", mapping = aes(fill = "Bootstrap"), alpha = 0.5) +
    stat_qq_band() +
    stat_qq_line() +
    stat_qq_point()
  
  return(plot)
}

############## SLM and SEM models

vars <- list("pm_change", "o3_change", "no2_change", "ndvi_change", "mean_temp")

# build linear models to test spatial error and spatial lag terms
pm_lm <- lm(pm_change ~ score_0.5 + med_income_ranked, data = all_data_inc)

# testing spatial error model
lm.RStests(pm_lm, ny.6, test="RSerr", zero.policy = T) # significant p value
# testing spatial lag model
lm.RStests(pm_lm, ny.6, test="RSerr", zero.policy = T) # significant p value

# fit spatial error models
get_sem <- function(var) {
  
  data_sem <- all_data_inc %>% 
    rename(dv = all_of(var))
  
  sem <- errorsarlm(dv ~ score_0.5 + med_income_ranked, data = data_sem, listw = ny.6, zero.policy = TRUE)
  
  return(sem)
}

sem_list <- lapply(vars, get_sem)

# fit spatial lag models
get_slm <- function(var) {
  
  data_slm <- all_data_inc %>% 
    rename(dv = all_of(var))
  
  slm <- lagsarlm(dv ~ score_0.5 + med_income_ranked, data = data_slm, listw = ny.6, zero.policy = TRUE)
  
  return(slm)
}

slm_list <- lapply(vars, get_slm)

# test residuals
moran_sem <- purrr::map(sem_list, ~ moran.test(.x$residuals, ny.6)) 
moran_slm <- purrr::map(slm_list, ~ moran.test(.x$residuals, ny.6))

# qq plots
qq_sem <- lapply(sem_list, get_qq_plot)
qq_slm <- lapply(slm_list, get_qq_plot)

grid.arrange(qq_sem[[1]], qq_sem[[2]], qq_sem[[3]], qq_sem[[4]], qq_sem[[5]])
grid.arrange(qq_slm[[1]], qq_slm[[2]], qq_slm[[3]], qq_slm[[4]], qq_slm[[5]])

# stratified SLM
get_slm_strat <- function(countyfp) {
  
  # filter data for county/borough
  data_borough <- all_data_inc %>% 
    filter(COUNTYFP %in% countyfp)
  
  # list all exposure variables for borough dataset
  exposures <- list(data_borough$pm_change, data_borough$o3_change, data_borough$no2_change,
                    data_borough$ndvi_change, data_borough$mean_temp)
  
  # identify neighbor relationships
  spdat.sens_borough <- as_Spatial(data_borough)
  ny.6_borough <- knearneigh(sp::coordinates(spdat.sens_borough), k=6)
  ny.6_borough <- knn2nb(ny.6_borough)
  ny.6_borough <- make.sym.nb(ny.6_borough)
  ny.6_borough <- nb2listw(ny.6_borough, style="W")
  
  get_slm <- function(exposure) {
    
    # assign exposure of interest as "var" variable
    data_borough <- data_borough %>% 
      mutate(var = exposure)
    
    slm <- lagsarlm(var ~ score_0.5 + med_income_ranked, data = data_borough, listw = ny.6_borough, zero.policy = TRUE)
    
    return(slm)
  }
  
  models <- lapply(exposures, get_slm)
  
  return(models)
}

countyfp <- list("005", "047", "061", "081", "085")

  # filter data for county/borough
  data_borough <- all_data_inc %>% 
    filter(COUNTYFP %in% countyfp)
  
  # list all exposure variables for borough dataset
  exposures <- list(data_borough$pm_change, data_borough$o3_change, data_borough$no2_change,
                    data_borough$ndvi_change, data_borough$mean_temp)
  
 

slm_strat <- lapply(countyfp, get_slm_strat) %>% 
  # flatten list of 25 models - PM2.5, O3, NO2, NDVI, temp, for 5 boroughs
  unlist(recursive = FALSE)

qq_slm_strat <- lapply(slm_strat, get_qq_plot)

# Bronx
grid.arrange(qq_slm_strat[[1]], qq_slm_strat[[2]], qq_slm_strat[[3]], qq_slm_strat[[4]], qq_slm_strat[[5]])
# Brooklyn
grid.arrange(qq_slm_strat[[6]], qq_slm_strat[[7]], qq_slm_strat[[8]], qq_slm_strat[[9]], qq_slm_strat[[10]])
# Manhattan
grid.arrange(qq_slm_strat[[11]], qq_slm_strat[[12]], qq_slm_strat[[13]], qq_slm_strat[[14]], qq_slm_strat[[15]])
# Queens
grid.arrange(qq_slm_strat[[16]], qq_slm_strat[[17]], qq_slm_strat[[18]], qq_slm_strat[[19]], qq_slm_strat[[20]])
# Staten Island
grid.arrange(qq_slm_strat[[21]], qq_slm_strat[[22]], qq_slm_strat[[23]], qq_slm_strat[[24]], qq_slm_strat[[25]])

# stratified SEM
get_sem_strat <- function(countyfp) {
  
  # filter data for county/borough
  data_borough <- all_data_inc %>% 
    filter(COUNTYFP %in% countyfp)
  
  # list all exposure variables for borough dataset
  exposures <- list(data_borough$pm_change, data_borough$o3_change, data_borough$no2_change,
                    data_borough$ndvi_change, data_borough$mean_temp)
  
  # identify neighbor relationships
  spdat.sens_borough <- as_Spatial(data_borough)
  ny.6_borough <- knearneigh(sp::coordinates(spdat.sens_borough), k=6)
  ny.6_borough <- knn2nb(ny.6_borough)
  ny.6_borough <- make.sym.nb(ny.6_borough)
  ny.6_borough <- nb2listw(ny.6_borough, style="W")
  
  get_sem <- function(exposure) {
    
    # assign exposure of interest as "var" variable
    data_borough <- data_borough %>% 
      mutate(var = exposure)
    
    sem <- errorsarlm(var ~ score_0.5 +  + dt_dist, data = data_borough, listw = ny.6_borough, zero.policy = TRUE)
    
    return(sem)
  }
  
  models <- lapply(exposures, get_sem)
  
  return(models)
}

sem_strat <- lapply(countyfp, get_sem_strat) %>% 
  # flatten list of 25 models - PM2.5, O3, NO2, NDVI, temp, for 5 boroughs
  un(recursive = FALSE)

qq_sem_strat <- lapply(sem_strat, get_qq_plot)

# Bronx
grid.arrange(qq_sem_strat[[1]], qq_sem_strat[[2]], qq_sem_strat[[3]], qq_sem_strat[[4]], qq_sem_strat[[5]])
# Brooklyn
grid.arrange(qq_sem_strat[[6]], qq_sem_strat[[7]], qq_sem_strat[[8]], qq_sem_strat[[9]], qq_sem_strat[[10]])
# Manhattan
grid.arrange(qq_sem_strat[[11]], qq_sem_strat[[12]], qq_sem_strat[[13]], qq_sem_strat[[14]], qq_sem_strat[[15]])
# Queens
grid.arrange(qq_sem_strat[[16]], qq_sem_strat[[17]], qq_sem_strat[[18]], qq_sem_strat[[19]], qq_sem_strat[[20]])
# Staten Island
grid.arrange(qq_sem_strat[[21]], qq_sem_strat[[22]], qq_sem_strat[[23]], qq_sem_strat[[24]], qq_sem_strat[[25]])
```

```{r}
# Spatial Error Model with shifted data
library(spatialreg)
library(MASS)

# Find minimum value and shift data to positive
min_o3 <- min(all_data_inc$o3_change)
shift_value <- abs(min_o3) + 1  # add 1 to ensure all positive values
o3_shifted <- all_data_inc$o3_change + shift_value

# Create model matrix for initial regression
X <- model.matrix(~ score_0.5 + med_income_ranked + flood_hazard, data = all_data_inc)

# Perform Box-Cox transformation on shifted data
bc <- boxcox(lm(o3_shifted ~ X - 1), 
             lambda = seq(-2, 2, by = 0.1),
             plotit = TRUE)
lambda <- bc$x[which.max(bc$y)]

# Transform the shifted dependent variable
if(abs(lambda) < 0.01) {
    o3_transformed <- log(o3_shifted)
    transform_type <- "log of shifted data"
} else {
    o3_transformed <- (o3_shifted^lambda - 1)/lambda
    transform_type <- paste("Box-Cox with lambda =", round(lambda, 3), "on shifted data")
}

# Fit spatial error model
spatial_error <- errorsarlm(o3_transformed ~ 
                           score_0.5 + 
                           med_income_ranked + flood_hazard,
                           data = all_data_inc,
                           listw = listw,
                           zero.policy = TRUE)

# Function to back-transform predictions
back_transform <- function(pred) {
    if(abs(lambda) < 0.01) {
        return(exp(pred) - shift_value)
    } else {
        return((pred * lambda + 1)^(1/lambda) - shift_value)
    }
}

# Check residuals
residuals_sem <- residuals(spatial_error)
listw <- nb2listw(nb, style="W", zero.policy=TRUE)
moran_sem <- moran.test(residuals_sem, listw, zero.policy=TRUE)
print("SEM Moran's I test:")
print(moran_sem)


# Print summaries
print("Data transformation details:")
print(paste("Minimum PM change value:", min_pm))
print(paste("Shift applied:", shift_value))
print(paste("Transformation type:", transform_type))
print("\nSpatial Error Model Summary:")
print(summary(spatial_error))

# Function for spatial error model diagnostics
spatial_diagnostics <- function(model, data) {
    # Extract residuals and fitted values safely
    resids <- residuals(model)  # Simplified residuals extraction
    fitted_vals <- fitted(model)
    
    par(mfrow=c(2,2))
    
    # 1. Residuals vs Fitted
    plot(fitted_vals, resids,
         xlab="Fitted values",
         ylab="Residuals",
         main="Residuals vs Fitted",
         type="n")  # Set up empty plot first
    points(fitted_vals, resids, pch=19)  # Add points
    abline(h=0, col="red", lty=2)
    lines(lowess(fitted_vals, resids), col="blue")
    
    # 2. Normal Q-Q plot (simplified)
    qqnorm(resids, main="Normal Q-Q Plot")  # Basic Q-Q plot
    qqline(resids, col="red")
    
    # 3. Scale-Location plot
    sqrt_abs_resids <- sqrt(abs(resids))
    plot(fitted_vals, sqrt_abs_resids,
         xlab="Fitted values",
         ylab=expression(sqrt("|Standardized residuals|")),
         main="Scale-Location",
         type="n")  # Set up empty plot first
    points(fitted_vals, sqrt_abs_resids, pch=19)  # Add points
    lines(lowess(fitted_vals, sqrt_abs_resids), col="blue")
    
    # 4. Residual distribution plot
    hist(resids, 
         main="Distribution of Residuals",
         xlab="Residuals",
         breaks=30,
         probability=TRUE)
    lines(density(resids), col="blue", lwd=2)
}

# Generate predictions first
spatial_preds <- back_transform(fitted(spatial_error))

# Now create all plots
# Spatial Error model diagnostics
par(mfrow=c(2,2))
spatial_diagnostics(spatial_error, all_data_inc)
mtext("Spatial Error Model Diagnostics", side=3, line=-2, outer=TRUE)


# Plot Spatial Error predictions
plot(all_data_inc$pm_change, spatial_preds,
     main="Spatial Error: Observed vs Predicted",
     xlab="Observed", ylab="Predicted",
     type="n")  # Set up empty plot first
points(all_data_inc$pm_change, spatial_preds, pch=19)  # Add points
abline(0,1, col="red")
lines(lowess(all_data_inc$pm_change, spatial_preds), col="blue")

# Reset plotting parameters
par(mfrow=c(1,1))

# Print some summary statistics of the predictions
cat("\nSummary of Spatial Error predictions:\n")
print(summary(spatial_preds))
```
###############################
######################LOOK AT QQ PLOTS FOR SEM MODELS (RN JUST THE NORMAL MODEL QQPLOT BEFORE SEM)
```{r}
# Function to perform spatial error analysis with Box-Cox transformation
spatial_error_analysis <- function(var_name, data, listw) {
    dependent_var <- data[[var_name]]
    
    # Find minimum value and shift data to positive
    min_val <- min(dependent_var)
    shift_value <- abs(min_val) + 1  # add 1 to ensure all positive values
    shifted_data <- dependent_var + shift_value
    
    # Create temporary data frame for Box-Cox
    temp_data <- data.frame(
        y = shifted_data,
        score_0.5 = data$score_0.5,
        med_income_ranked = data$med_income_ranked
    )
    
    # Perform Box-Cox transformation on shifted data
    bc <- boxcox(y ~ score_0.5 + med_income_ranked, 
                 data = temp_data,
                 lambda = seq(-2, 2, by = 0.1),
                 plotit = FALSE)
    lambda <- bc$x[which.max(bc$y)]
    
    # Transform the shifted dependent variable
    if(abs(lambda) < 0.01) {
        transformed_data <- log(shifted_data)
        transform_type <- "log of shifted data"
    } else {
        transformed_data <- (shifted_data^lambda - 1)/lambda
        transform_type <- paste("Box-Cox with lambda =", round(lambda, 3), "on shifted data")
    }
    
    # Create data frame with transformed response
    model_data <- data
    model_data$transformed_response <- transformed_data
    
    # Fit spatial error model
    spatial_error <- errorsarlm(transformed_response ~ score_0.5 + med_income_ranked,
                               data = model_data,
                               listw = listw,
                               zero.policy = TRUE)
    
    # Calculate residuals and perform Moran's I test
    residuals_sem <- residuals(spatial_error)
    moran_sem <- moran.test(residuals_sem, listw, zero.policy=TRUE)
    
    # Back-transform predictions
    if(abs(lambda) < 0.01) {
        predictions <- exp(fitted(spatial_error)) - shift_value
    } else {
        predictions <- (fitted(spatial_error) * lambda + 1)^(1/lambda) - shift_value
    }
    
    # Create diagnostic plots
    par(mfrow=c(2,2))
    
    # Residuals vs Fitted
    plot(fitted(spatial_error), residuals_sem,
         xlab="Fitted values", ylab="Residuals",
         main="Residuals vs Fitted")
    abline(h=0, col="red", lty=2)
    lines(lowess(fitted(spatial_error), residuals_sem), col="blue")
    
    # Normal Q-Q plot
    qqnorm(residuals_sem)
    qqline(residuals_sem, col="red")
    
    # Scale-Location plot
    sqrt_abs_resids <- sqrt(abs(residuals_sem))
    plot(fitted(spatial_error), sqrt_abs_resids,
         xlab="Fitted values",
         ylab=expression(sqrt("|Standardized residuals|")),
         main="Scale-Location")
    lines(lowess(fitted(spatial_error), sqrt_abs_resids), col="blue")
    
    # Residual distribution
    hist(residuals_sem, 
         main="Distribution of Residuals",
         xlab="Residuals",
         breaks=30,
         probability=TRUE)
    lines(density(residuals_sem), col="blue", lwd=2)
    
    # Results list
    results <- list(
        model = spatial_error,
        predictions = predictions,
        moran_test = moran_sem,
        transform_info = list(
            min_value = min_val,
            shift_value = shift_value,
            lambda = lambda,
            transform_type = transform_type
        )
    )
    
    return(results)
}

# Make sure required packages are loaded
library(spatialreg)
library(MASS)  # for boxcox function

# Run analysis for each variable
variables <- c("pm_change", "o3_change", "no2_change", "ndvi_change", "mean_temp")
results_list <- list()

for(var in variables) {
    cat("\n\n=== Analysis for", var, "===\n")
    
    # Run analysis
    results_list[[var]] <- spatial_error_analysis(
        var_name = var,
        data = all_data_inc,
        listw = listw
    )
    
    # Print summaries
    cat("\nTransformation details:\n")
    print(results_list[[var]]$transform_info)
    cat("\nModel Summary:\n")
    print(summary(results_list[[var]]$model))
    cat("\nMoran's I test for residuals:\n")
    print(results_list[[var]]$moran_test)
    
    # Add title to current plot
    mtext(paste("Diagnostic Plots for", var), side=3, line=-2, outer=TRUE)
    
    # Create observed vs predicted plot
    par(mfrow=c(1,1))
    plot(all_data_inc[[var]], results_list[[var]]$predictions,
         main=paste(var, ": Observed vs Predicted"),
         xlab="Observed", ylab="Predicted")
    abline(0,1, col="red")
    lines(lowess(all_data_inc[[var]], results_list[[var]]$predictions), col="blue")
}

# Reset plotting parameters
par(mfrow=c(1,1))
```

################################################
#################################################
Create model functions (NEW FOR MRF ONLY)
```{r}
# Create exposure list
exposure_list <- list(
  pm = list(
    name = "pm",
    baseline = "pm_2002",
    change = "pm_change"
  ),
  o3 = list(
    name = "o3",
    baseline = "o3_2002",
    change = "o3_change"
  ),
  no2 = list(
    name = "no2",
    baseline = "no2_2000",
    change = "no2_change"
  ),
  ndvi = list(
    name = "ndvi",
    baseline = "ndvi_2000",
    change = "ndvi_change"
  ),
  temp = list(
    name = "mean_temp",
    baseline = NULL,
    change = "mean_temp"  # No change variable for temp
  )
)

# Function to run GAM models
run_gam_models <- function(data, exposure_info) {
  # Extract exposure information
  exp_var <- exposure_info$change
  baseline_exposure <- exposure_info$baseline
  
  # Skip if data is missing the exposure variable - gives warning 
  if (!exp_var %in% names(data)) {
    warning(paste("Variable", exp_var, "not found in dataset"))
    return(NULL)
  }
  
  # Base components that will be in all models: gentrification score and county/borough
  base_formula <- paste(exp_var, "~ score_0.5 + s(COUNTYFP10, bs='re')")
  
  # Create model formulas
  formulas <- list(
    # Base model (score + county + spatial) for mrf smooth
    mrf_base = paste(base_formula, "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),
    
    # Models with income
    mrf_income = paste(base_formula, "+ rank(med_income_2000_impute) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),

    # Models with distance to downtown
    mrf_dist = paste(base_formula, "+ s(dt_dist) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  )
  
  # Add baseline models if baseline exposure level exists (all except temp)
  if (!is.null(baseline_exposure) && baseline_exposure %in% names(data)) {
    formulas$mrf_baseline <- paste(base_formula, "+", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    # Combinations with baseline
    formulas$mrf_income_baseline <- paste(base_formula, "+ rank(med_income_2000_impute) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    formulas$mrf_dist_baseline <- paste(base_formula, "+ s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    formulas$mrf_full <- paste(base_formula, "+ rank(med_income_2000_impute) + s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  }
  
  # Add income + distance combinations
  formulas$mrf_income_dist <- paste(base_formula, "+ rank(med_income_2000_impute) + s(dt_dist) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  
  # Convert formulas to formula objects
  formulas <- lapply(formulas, as.formula)
  
  # Run models
  models <- purrr::map(formulas, ~try(gam(., 
                                        family = scat, 
                                        data = data, 
                                        method = "REML"), 
                                    silent = TRUE))
  
  return(models)
}

# drop geometry from spatial data while keeping coordinates
data_prepared <- all_data_inc %>%
  mutate(
    COUNTYFP10 = as.factor(COUNTYFP10),
    GEOID = as.factor(GEOID)
  ) %>%
  st_drop_geometry()

# Construct neighbors list for MRF models
nb <- poly2nb(all_data_inc, row.names = all_data_inc$GEOID)
names(nb) <- attr(nb, "region.id") 

# JH 10/25/24 Warning: some observations have no neighbours; if this seems unexpected, try increasing the snap argument.Warning: neighbour object has 4 sub-graphs; if this sub-graph count seems unexpected, try increasing the snap argument.
```

Run models and check models
```{r}
# Run models for each exposure
all_models <- purrr::map(exposure_list, ~run_gam_models(data_prepared, .))

# Function to check which models succeeded/failed
check_models <- function(models) {
  model_status <- map(models, ~!inherits(., "try-error"))
  successful_models <- names(models)[unlist(model_status)]
  failed_models <- names(models)[!unlist(model_status)]
  
  return(list(
    successful = successful_models,
    failed = failed_models
  ))
}

# Check status of all models
model_status <- map(all_models, check_models)
model_status

```
```{r}
# Function to check and summarize all GAM models
check_and_summarize_models <- function(models) {
  # Apply gam.check and summary to each model
  purrr::map(models, function(model_list) {
    purrr::map(model_list, ~ {
      # Perform a check on the GAM model
      gam_check_result <- try(gam.check(.), silent = TRUE)
      
      # If gam.check runs without error, print it
      if (inherits(gam_check_result, "try-error")) {
        message("Error in gam.check for model: ", deparse(substitute(.)))
      } else {
        print(gam_check_result)
      }
      
      # Get the summary of the model
      summary_result <- try(summary(.), silent = TRUE)
      
      # If summary runs without error, print it
      if (inherits(summary_result, "try-error")) {
        message("Error in summary for model: ", deparse(substitute(.)))
      } else {
        print(summary_result)
      }
    })
  })
}

# Example usage:
check_and_summarize_models(all_models)
```





######################################################################
######################################################################
#####################################################################
2000 vs 2010 income checking
```{r}
library(tidyverse)
library(stats)

analyze_tract_incomes <- function(all_data, income_2000, income_2010) {
  # Merge data
  analysis_df <- all_data %>%
    left_join(income_2000 %>% select(GEOID, med_income_2000), by = "GEOID") %>%
    left_join(income_2010 %>% select(GEOID, med_income_2010), by = "GEOID")
  
  # Create summary statistics grouped by whether 2000 data exists
  summary_stats <- analysis_df %>%
    mutate(has_2000_data = !is.na(med_income_2000)) %>%
    group_by(has_2000_data) %>%
    summarise(
      n_tracts = dplyr::n(),
      mean_2010 = mean(med_income_2010, na.rm = TRUE),
      median_2010 = median(med_income_2010, na.rm = TRUE),
      sd_2010 = sd(med_income_2010, na.rm = TRUE),
      q25_2010 = quantile(med_income_2010, 0.25, na.rm = TRUE),
      q75_2010 = quantile(med_income_2010, 0.75, na.rm = TRUE)
    )
  
  # Perform Wilcoxon rank sum test (Mann-Whitney U test)
  test_result <- wilcox.test(
    med_income_2010 ~ !is.na(med_income_2000),
    data = analysis_df,
    alternative = "two.sided"
  )
  
  # Create visualization
  plot <- analysis_df %>%
    mutate(has_2000_data = if_else(!is.na(med_income_2000), 
                                  "Has 2000 Data", 
                                  "Missing 2000 Data")) %>%
    ggplot(aes(x = has_2000_data, y = med_income_2010)) +
    geom_boxplot() +
    theme_minimal() +
    labs(
      title = "2010 Income Distribution: Tracts With vs Without 2000 Data",
      x = "",
      y = "2010 Median Income"
    )
  
  # Calculate percent difference in means
  pct_diff <- with(summary_stats, 
                  (mean_2010[!has_2000_data] - mean_2010[has_2000_data]) / 
                    mean_2010[has_2000_data] * 100)
  
  # Return results
  list(
    summary_stats = summary_stats,
    wilcox_test = test_result,
    percent_difference = pct_diff,
    plot = plot
  )
}

# Use the function
results <- analyze_tract_incomes(all_data, income, income_2010)

# Print results
print("Summary Statistics:")
print(results$summary_stats)

print("\nWilcoxon Test Results:")
print(results$wilcox_test)

print(sprintf("\nPercent difference in means: %.2f%%", results$percent_difference))

# Display plot
print(results$plot)
print(results$plot)

#additional analysis based on average, variation, standard deviation- checking normality to see which statistical signifiancetest is appropriate

# Create analysis dataset
analysis_df <- all_data %>%
  left_join(income %>% select(GEOID, med_income_2000), by = "GEOID") %>%
  left_join(income_2010 %>% select(GEOID, med_income_2010), by = "GEOID") %>%
  mutate(has_2000_data = !is.na(med_income_2000))

# Calculate summary statistics
stats_summary <- analysis_df %>%
  group_by(has_2000_data) %>%
  summarise(
    n = dplyr::n(),
    mean_2010 = mean(med_income_2010, na.rm = TRUE),
    var_2010 = var(med_income_2010, na.rm = TRUE),
    sd_2010 = sd(med_income_2010, na.rm = TRUE)
  ) %>%
  mutate(across(where(is.numeric), ~round(., 2)))

# Test for normality using Shapiro-Wilk test
normality_test_with_2000 <- shapiro.test(analysis_df$med_income_2010[analysis_df$has_2000_data])
normality_test_without_2000 <- shapiro.test(analysis_df$med_income_2010[!analysis_df$has_2000_data])

# Perform F-test for equality of variances
var_test <- var.test(med_income_2010 ~ has_2000_data, data = analysis_df)

# Perform t-test (both equal and unequal variance versions)
t_test_equal_var <- t.test(med_income_2010 ~ has_2000_data, data = analysis_df, var.equal = TRUE)
t_test_unequal_var <- t.test(med_income_2010 ~ has_2000_data, data = analysis_df, var.equal = FALSE)

# Perform Wilcoxon rank sum test
wilcox_test <- wilcox.test(med_income_2010 ~ has_2000_data, data = analysis_df)

# Print results
cat("Summary Statistics:\n")
print(stats_summary)

cat("\nNormality Tests (Shapiro-Wilk):\n")
cat("Tracts with 2000 data: p =", normality_test_with_2000$p.value, "\n")
cat("Tracts without 2000 data: p =", normality_test_without_2000$p.value, "\n")

cat("\nF-test for Equal Variances:\n")
cat("F =", round(var_test$statistic, 3), ", p =", var_test$p.value, "\n")

cat("\nt-test (Equal Variances):\n")
cat("t =", round(t_test_equal_var$statistic, 3), ", p =", t_test_equal_var$p.value, "\n")

cat("\nt-test (Unequal Variances):\n")
cat("t =", round(t_test_unequal_var$statistic, 3), ", p =", t_test_unequal_var$p.value, "\n")

cat("\nWilcoxon Rank Sum Test:\n")
cat("W =", wilcox_test$statistic, ", p =", wilcox_test$p.value, "\n")
```

```{r}
# Construct neighbors list for MRF models
nb <- poly2nb(all_data_inc, row.names = all_data_inc$GEOID)
names(nb) <- attr(nb, "region.id") 
```


testing tp model with med_income_2000 for pm2.5
```{r}
tp_income_pm_na <- gam(pm_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + s(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2), method = 'REML', family = scat, data = all_data_inc_na)

# Try increasing k for the tensor product smooth of longitude and latitude (from 24 default)
tp_income_pm_na_2 <- gam(pm_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + s(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 100), method = 'REML', family = scat, data = all_data_inc_na) #when k=40 and k=100, p value still significant (= bad)

tp_income_pm_na_3 <- gam(pm_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 150), method = 'REML', family = scat, data = all_data_inc_na) 

tp_income_pm_na_4 <- gam(pm_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 200), method = 'REML', family = scat, data = all_data_inc_na) 

tp_income_pm_na_4 <- gam(pm_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 200), method = 'REML', family = scat, data = all_data_inc_na) 

tp_income_pm_na_5 <- gam(pm_change ~ score_0.5 + s(COUNTYFP10, bs = "re", k=10) + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 200), method = 'REML', family = scat, data = all_data_inc_na) 



#model diagnostics using gam.check
gam.check(tp_income_pm_na)
gam.check(tp_income_pm_na_2)
gam.check(tp_income_pm_na_3)
gam.check(tp_income_pm_na_4)
gam.check(tp_income_pm_na_5)


# First, let's get the fitted random effects properly
fitted_terms <- predict(tp_income_pm_na_5, type="terms")
county_term <- fitted_terms[,"s(COUNTYFP10)"]  # Extract county random effects

# Create summary by county
county_summary <- data.frame(
    COUNTYFP10 = all_data_inc_na$COUNTYFP10,
    effect = county_term
) %>%
    group_by(COUNTYFP10) %>%
    summarize(mean_effect = mean(effect)) %>%
    arrange(desc(abs(mean_effect)))

# View the most extreme counties
print(head(county_summary))

# Visualize
library(ggplot2)
ggplot(county_summary, aes(y=mean_effect, x=reorder(COUNTYFP10, mean_effect))) +
    geom_point() +
    theme_minimal() +
    theme(axis.text.x = element_blank()) +  # Hide county codes for clarity
    labs(x="Counties", y="Random Effect")

# Look at the data for the most extreme county
extreme_county <- county_summary$COUNTYFP10[1]
extreme_county_data <- all_data_inc_na %>%
    filter(COUNTYFP10 == extreme_county) %>%
    select(pm_change, score_0.5, med_income_2000, long, lat)

print("Data for most extreme county:")
print(summary(extreme_county_data)) #extreme county is 85 - Manhattan/NY county

print("Overall data summary:")
print(summary(all_data_inc_na[,c("pm_change", "score_0.5", "med_income_2000")]))
####

####

# Calculate Cook's distance
cooks_d <- cooks.distance(tp_income_pm_na)
plot(cooks_d, type = "h", main = "Cook's Distance")
abline(h = 4/length(cooks_d), col = "red")  # Rule of thumb threshold

# Identify potential influential points
influential <- which(cooks_d > 4/length(cooks_d))

influential_data <- all_data_inc_na[influential, ]
print(influential_data)  # Look at characteristics of these points

# Options for handling influential points:
# a. Fit model without them to compare
tp_income_pm_na_reduced <- gam(
  pm_change ~ score_0.5 + 
    s(COUNTYFP10, bs = "re") + 
    rank(med_income_2000) + 
    te(long, lat, d = 2, bs = 'gp', m = 2, k = 150),
  method = 'REML',
  family = scat,
  data = all_data_inc_na[-influential, ]
)

# Compare coefficients and predictions
summary(tp_income_pm_na)
summary(tp_income_pm_na_reduced)

# b. Investigate why these points are influential:
# Plot their locations
plot(all_data_inc_na$long, all_data_inc_na$lat)
points(all_data_inc_na$long[influential], 
       all_data_inc_na$lat[influential], 
       col = "red", pch = 16)


# Checking that scat is indeed better (comparing t distribution to gaussian)
model_scat <- gam(
  pm_change ~ score_0.5 + 
    s(COUNTYFP10, bs = "re") + 
    rank(med_income_2000) + 
    te(long, lat, d = 2, bs = 'gp', m = 2, k = 150),
  method = 'REML',
  family = scat,
  data = all_data_inc_na
)

model_gaussian <- gam(
  pm_change ~ score_0.5 + 
    s(COUNTYFP10, bs = "re") + 
    rank(med_income_2000) + 
    te(long, lat, d = 2, bs = 'gp', m = 2, k = 150),
  method = 'REML',
  family = gaussian,
  data = all_data_inc_na
)

# Compare QQ plots
par(mfrow=c(1,2))
qq.gam(model_scat, main="Scaled t")
qq.gam(model_gaussian, main="Gaussian")

# Compare predictions
pred_scat <- predict(model_scat)
pred_gaussian <- predict(model_gaussian)
plot(pred_scat, pred_gaussian, 
     xlab="Scaled t predictions", 
     ylab="Gaussian predictions")
abline(0,1, col="red")

# Compare AIC (if models have same scale)
AIC(model_scat, model_gaussian)
# Check if residual patterns differ
par(mfrow=c(2,2))
gam.check(tp_income_pm_gaussian)


```

testing tp models for o3
```{r}
tp_income_o3_na <- gam(o3_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + s(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2), method = 'REML', family = scat, data = all_data_inc_na)


#increasing k in tp
tp_income_o3_na_2 <- gam(o3_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 150), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_o3_na_3 <- gam(o3_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 250), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_o3_na_4 <- gam(o3_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 200), method = 'REML', family = scat, data = all_data_inc_na) # k still 0.02

tp_income_o3_na_5 <- gam(o3_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 225), method = 'REML', family = scat, data = all_data_inc_na) #k no longer significant


gam.check(tp_income_o3_na)
gam.check(tp_income_o3_na_2)
gam.check(tp_income_o3_na_3)
gam.check(tp_income_o3_na_4)
gam.check(tp_income_o3_na_5)




```

testing tp models for no2
```{r}
tp_income_no2_na <- gam(no2_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + s(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_no2_na_2 <- gam(no2_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k= 200), method = 'REML', family = scat, data = all_data_inc_na) #still significant

tp_income_no2_na_3 <- gam(no2_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k= 250), method = 'REML', family = scat, data = all_data_inc_na) #still significant
 
tp_income_no2_na_4 <- gam(no2_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k= 450), method = 'REML', family = scat, data = all_data_inc_na) #takes forever to run - not feasible

gam.check(tp_income_no2_na)
gam.check(tp_income_no2_na_2)
gam.check(tp_income_no2_na_3)
gam.check(tp_income_no2_na_4)


```
testing tp models for ndvi
```{r}
tp_income_ndvi_na <- gam(ndvi_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + s(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_ndvi_na_2 <- gam(ndvi_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 200), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_ndvi_na_3 <- gam(ndvi_change ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 300), method = 'REML', family = scat, data = all_data_inc_na) # still very significant

gam.check(tp_income_ndvi_na)
gam.check(tp_income_ndvi_na_2)
gam.check(tp_income_ndvi_na_3)

```
testing tp models for temp
```{r}
tp_income_temp_na <- gam(mean_temp ~ score_0.5 + s(COUNTYFP10, bs = "re") + s(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_temp_na_2<- gam(mean_temp ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 200), method = 'REML', family = scat, data = all_data_inc_na)

tp_income_temp_na_3<- gam(mean_temp ~ score_0.5 + s(COUNTYFP10, bs = "re") + rank(med_income_2000) + te(long, lat, d = 2, bs = 'gp', m = 2, k = 300), method = 'REML', family = scat, data = all_data_inc_na) # still very significant


gam.check(tp_income_temp_na)
gam.check(tp_income_temp_na_2)
gam.check(tp_income_temp_na_3)
```

Testing residuals using tp model for now
```{r}
# Extract residuals for each model
resid_pm <- resid(tp_income_pm_na_5)
resid_o3 <- resid(tp_income_o3_na_5)
resid_no2 <- resid(tp_income_no2_na_3)
resid_ndvi <- resid(tp_income_ndvi_na_3)
resid_temp <- resid(tp_income_ndvi_na)


# Add residuals to the original dataset (ensure this corresponds to the right model)
all_data_inc_na$resid_pm <- resid_pm
all_data_inc_na$resid_o3 <- resid_o3
all_data_inc_na$resid_no2 <- resid_no2
all_data_inc_na$resid_ndvi <- resid_ndvi
all_data_inc_na$resid_temp <- resid_temp
```

Code for just PM to base function on
```{r}
# # Visualize residuals for PM model, for example
# ggplot(all_data_inc_na, aes(x = long, y = lat, color = resid_pm)) +
#   geom_point() +
#   scale_color_gradient2(midpoint = 0, low = "blue", mid = "white", high = "red") +
#   theme_minimal() +
#   labs(title = "Spatial Distribution of Residuals (PM)", color = "Residuals")
#   
# # Calculate the 10th and 90th percentiles of the residuals
# lower_threshold <- quantile(all_data_inc_na$resid_pm, 0.1)
# upper_threshold <- quantile(all_data_inc_na$resid_pm, 0.9)
# 
# # Create dichotomous variable based on thresholds
# all_data_inc_na$residual_gap <- ifelse(all_data_inc_na$resid_pm > upper_threshold, "High Residual", 
#                                         ifelse(all_data_inc_na$resid_pm < lower_threshold, "Low Residual", "Normal"))
# 
# # Visualize the result
# ggplot(all_data_inc_na, aes(x = long, y = lat, color = residual_gap)) +
#   geom_point() +
#   scale_color_manual(values = c("High Residual" = "red", "Low Residual" = "blue", "Normal" = "grey")) +
#   theme_minimal() +
#   labs(title = "Spatial Distribution of Residual Gaps (PM)", color = "Residual Gap")
# 
# 
# # Calculate the 10th and 90th percentiles of the residuals
# lower_threshold <- quantile(all_data_inc_na$resid_pm, 0.1)
# upper_threshold <- quantile(all_data_inc_na$resid_pm, 0.9)
# 
# # Create dichotomous variable based on thresholds
# all_data_inc_na$residual_gap <- ifelse(all_data_inc_na$resid_pm > upper_threshold, "High Residual", 
#                                         ifelse(all_data_inc_na$resid_pm < lower_threshold, "Low Residual", "Normal"))
# 
# # Visualize the result
# ggplot(all_data_inc_na, aes(x = long, y = lat, color = residual_gap)) +
#   geom_point() +
#   scale_color_manual(values = c("High Residual" = "red", "Low Residual" = "blue", "Normal" = "grey")) +
#   theme_minimal() +
#   labs(title = "Spatial Distribution of Residual Gaps (PM)", color = "Residual Gap")
# 
# 
# # Visualize gaps using the dichotomous color scale
# ggplot(all_data_inc_na, aes(x = long, y = lat, color = residual_gap)) +
#   geom_point() +
#   scale_color_manual(values = c("High Residual" = "red", "Low Residual" = "blue", "Normal" = "grey")) +
#   theme_minimal() +
#   labs(title = "Spatial Distribution of Residual Gaps (PM)", color = "Residual Gap")
# 
# # Summary statistics for residuals (e.g., PM)
# summary(all_data_inc_na$resid_pm)
# 
# # Identify the locations with extreme residuals
# high_residuals <- all_data_inc_na[all_data_inc_na$resid_pm > threshold, ]
# low_residuals <- all_data_inc_na[all_data_inc_na$resid_pm < -threshold, ]
# 
# # View the locations of high and low residuals
# head(high_residuals[, c("long", "lat", "resid_pm")])
# head(low_residuals[, c("long", "lat", "resid_pm")])


```
############################################
function to analyze residuals for all exposures based on tp model (based on high and low quantiles)
```{r}
analyze_spatial_residuals <- function(data, residual_col, lon_col = "long", lat_col = "lat", 
                                    lower_quantile = 0.1, upper_quantile = 0.9) {
  # Input validation
  if (!all(c(residual_col, lon_col, lat_col) %in% names(data))) {
    stop("One or more specified columns not found in data")
  }
  
  # Create analysis environment
  results <- list()
  
  # Basic summary statistics
  results$summary <- summary(data[[residual_col]])
  
  # Calculate thresholds
  results$thresholds <- list(
    lower = quantile(data[[residual_col]], lower_quantile, na.rm = TRUE),
    upper = quantile(data[[residual_col]], upper_quantile, na.rm = TRUE)
  )
  
  # Create categorical variable for residual gaps
  data$residual_category <- case_when(
    data[[residual_col]] > results$thresholds$upper ~ "High Residual",
    data[[residual_col]] < results$thresholds$lower ~ "Low Residual",
    TRUE ~ "Normal"
  )
  
  # Generate plots
  # 1. Continuous residual plot
  results$continuous_plot <- ggplot(data, aes(x = .data[[lon_col]], 
                                            y = .data[[lat_col]], 
                                            color = .data[[residual_col]])) +
    geom_point(alpha = 0.6) +
    scale_color_gradient2(midpoint = 0, 
                         low = "blue", 
                         mid = "white", 
                         high = "red") +
    theme_minimal() +
    labs(title = paste("Spatial Distribution of Residuals:", residual_col),
         color = "Residuals") +
    theme(legend.position = "right")
  
  # 2. Categorical residual plot
  results$categorical_plot <- ggplot(data, aes(x = .data[[lon_col]], 
                                             y = .data[[lat_col]], 
                                             color = residual_category)) +
    geom_point(alpha = 0.6) +
    scale_color_manual(values = c("High Residual" = "red", 
                                 "Low Residual" = "blue", 
                                 "Normal" = "grey")) +
    theme_minimal() +
    labs(title = paste("Spatial Distribution of Residual Gaps:", residual_col),
         color = "Residual Category") +
    theme(legend.position = "right")
  
  # Extract extreme locations
  results$extreme_locations <- list(
    high = data %>%
      filter(residual_category == "High Residual") %>%
      select(all_of(c(lon_col, lat_col, residual_col))) %>%
      arrange(desc(.data[[residual_col]])),
    
    low = data %>%
      filter(residual_category == "Low Residual") %>%
      select(all_of(c(lon_col, lat_col, residual_col))) %>%
      arrange(.data[[residual_col]])
  )
  
  # Calculate summary statistics by category
  results$category_stats <- data %>%
    group_by(residual_category) %>%
    summarise(
      count = dplyr::n(),
      mean_residual = mean(.data[[residual_col]], na.rm = TRUE),
      sd_residual = sd(.data[[residual_col]], na.rm = TRUE),
      min_residual = min(.data[[residual_col]], na.rm = TRUE),
      max_residual = max(.data[[residual_col]], na.rm = TRUE),
      .groups = "drop"
    )
  
  return(results)
}

# Example usage for each environmental variable
results_pm <- analyze_spatial_residuals(all_data_inc_na, "resid_pm")
results_o3 <- analyze_spatial_residuals(all_data_inc_na, "resid_o3")
results_no2 <- analyze_spatial_residuals(all_data_inc_na, "resid_no2")
results_ndvi <- analyze_spatial_residuals(all_data_inc_na, "resid_ndvi")
results_temp <- analyze_spatial_residuals(all_data_inc_na, "resid_temp")


# Display results for pm2.5
print(results_pm$summary)
print(results_pm$category_stats)
results_pm$continuous_plot
results_pm$categorical_plot
head(results_pm$extreme_locations$high)
head(results_pm$extreme_locations$low)

print(results_o3$summary)
print(results_o3$category_stats)
results_o3$continuous_plot
results_o3$categorical_plot
head(results_o3$extreme_locations$high)
head(results_o3$extreme_locations$low)

print(results_no2$summary)
print(results_no2$category_stats)
results_no2$continuous_plot
results_no2$categorical_plot
head(results_no2$extreme_locations$high)
head(results_no2$extreme_locations$low)

print(results_ndvi$summary)
print(results_ndvi$category_stats)
results_ndvi$continuous_plot
results_ndvi$categorical_plot
head(results_ndvi$extreme_locations$high)
head(results_ndvi$extreme_locations$low)

print(results_temp$summary)
print(results_temp$category_stats)
results_temp$continuous_plot
results_temp$categorical_plot
head(results_temp$extreme_locations$high)
head(results_temp$extreme_locations$low)
```
function to analyze residuals for all exposures based on tp model (based on high and low quantiles, narrower limits)
```{r}
analyze_spatial_residuals <- function(data, residual_col, lon_col = "long", lat_col = "lat", 
                                    lower_quantile = 0.05, upper_quantile = 0.95) {
  # Input validation
  if (!all(c(residual_col, lon_col, lat_col) %in% names(data))) {
    stop("One or more specified columns not found in data")
  }
  
  # Create analysis environment
  results <- list()
  
  # Basic summary statistics
  results$summary <- summary(data[[residual_col]])
  
  # Calculate thresholds
  results$thresholds <- list(
    lower = quantile(data[[residual_col]], lower_quantile, na.rm = TRUE),
    upper = quantile(data[[residual_col]], upper_quantile, na.rm = TRUE)
  )
  
  # Create categorical variable for residual gaps
  data$residual_category <- case_when(
    data[[residual_col]] > results$thresholds$upper ~ "High Residual",
    data[[residual_col]] < results$thresholds$lower ~ "Low Residual",
    TRUE ~ "Normal"
  )
  
  # Generate plots
  # 1. Continuous residual plot
  results$continuous_plot <- ggplot(data, aes(x = .data[[lon_col]], 
                                            y = .data[[lat_col]], 
                                            color = .data[[residual_col]])) +
    geom_point(alpha = 0.6) +
    scale_color_gradient2(midpoint = 0, 
                         low = "blue", 
                         mid = "white", 
                         high = "red") +
    theme_minimal() +
    labs(title = paste("Spatial Distribution of Residuals:", residual_col),
         color = "Residuals") +
    theme(legend.position = "right")
  
  # 2. Categorical residual plot
  results$categorical_plot <- ggplot(data, aes(x = .data[[lon_col]], 
                                             y = .data[[lat_col]], 
                                             color = residual_category)) +
    geom_point(alpha = 0.6) +
    scale_color_manual(values = c("High Residual" = "red", 
                                 "Low Residual" = "blue", 
                                 "Normal" = "grey")) +
    theme_minimal() +
    labs(title = paste("Spatial Distribution of Residual Gaps:", residual_col),
         color = "Residual Category") +
    theme(legend.position = "right")
  
  # Extract extreme locations
  results$extreme_locations <- list(
    high = data %>%
      filter(residual_category == "High Residual") %>%
      select(all_of(c(lon_col, lat_col, residual_col))) %>%
      arrange(desc(.data[[residual_col]])),
    
    low = data %>%
      filter(residual_category == "Low Residual") %>%
      select(all_of(c(lon_col, lat_col, residual_col))) %>%
      arrange(.data[[residual_col]])
  )
  
  # Calculate summary statistics by category
  results$category_stats <- data %>%
    group_by(residual_category) %>%
    summarise(
      count = dplyr::n(),
      mean_residual = mean(.data[[residual_col]], na.rm = TRUE),
      sd_residual = sd(.data[[residual_col]], na.rm = TRUE),
      min_residual = min(.data[[residual_col]], na.rm = TRUE),
      max_residual = max(.data[[residual_col]], na.rm = TRUE),
      .groups = "drop"
    )
  
  return(results)
}

# Example usage for each environmental variable
results_pm <- analyze_spatial_residuals(all_data_inc_na, "resid_pm")
results_o3 <- analyze_spatial_residuals(all_data_inc_na, "resid_o3")
results_no2 <- analyze_spatial_residuals(all_data_inc_na, "resid_no2")
results_ndvi <- analyze_spatial_residuals(all_data_inc_na, "resid_ndvi")
results_temp <- analyze_spatial_residuals(all_data_inc_na, "resid_temp")


# Display results for pm2.5
print(results_pm$summary)
print(results_pm$category_stats)
results_pm$continuous_plot
results_pm$categorical_plot
head(results_pm$extreme_locations$high)
head(results_pm$extreme_locations$low)

print(results_o3$summary)
print(results_o3$category_stats)
results_o3$continuous_plot
results_o3$categorical_plot
head(results_o3$extreme_locations$high)
head(results_o3$extreme_locations$low)

print(results_no2$summary)
print(results_no2$category_stats)
results_no2$continuous_plot
results_no2$categorical_plot
head(results_no2$extreme_locations$high)
head(results_no2$extreme_locations$low)

print(results_ndvi$summary)
print(results_ndvi$category_stats)
results_ndvi$continuous_plot
results_ndvi$categorical_plot
head(results_ndvi$extreme_locations$high)
head(results_ndvi$extreme_locations$low)

print(results_temp$summary)
print(results_temp$category_stats)
results_temp$continuous_plot
results_temp$categorical_plot
head(results_temp$extreme_locations$high)
head(results_temp$extreme_locations$low)
```
Analyze residual overlap between exposures
```{r}
analyze_residual_overlap <- function(data, residual_cols, lon_col = "long", lat_col = "lat", 
                                   lower_quantile = 0.1, upper_quantile = 0.9) {
  # Input validation
  if (!all(c(residual_cols, lon_col, lat_col) %in% names(data))) {
    stop("One or more specified columns not found in data")
  }
  
  # Create category columns for each residual
  results <- list()
  data_enhanced <- data
  
  # Categorize each residual
  for (col in residual_cols) {
    thresholds <- quantile(data[[col]], c(lower_quantile, upper_quantile), na.rm = TRUE)
    cat_col_name <- paste0(col, "_category")
    
    data_enhanced[[cat_col_name]] <- case_when(
      data[[col]] > thresholds[2] ~ "High",
      data[[col]] < thresholds[1] ~ "Low",
      TRUE ~ "Normal"
    )
  }
  
  # Count overlapping high/low/extreme areas
  data_enhanced$high_overlap_count <- rowSums(
    sapply(residual_cols, function(col) {
      data_enhanced[[paste0(col, "_category")]] == "High"
    })
  )
  
  data_enhanced$low_overlap_count <- rowSums(
    sapply(residual_cols, function(col) {
      data_enhanced[[paste0(col, "_category")]] == "Low"
    })
  )
  
  # Count either high or low (extreme) overlaps
  data_enhanced$extreme_overlap_count <- rowSums(
    sapply(residual_cols, function(col) {
      data_enhanced[[paste0(col, "_category")]] %in% c("High", "Low")
    })
  )
  
  # Create overlap visualizations
  # 1. High residual overlap plot
  results$high_overlap_plot <- ggplot(data_enhanced, 
         aes(x = .data[[lon_col]], y = .data[[lat_col]], 
             color = factor(high_overlap_count))) +
    geom_point(alpha = 0.6) +
    scale_color_viridis_d(option = "magma", direction = -1) +
    theme_minimal() +
    labs(title = "Overlap of High Residuals",
         color = "Number of\nOverlapping\nHigh Residuals") +
    theme(legend.position = "right")
  
  # 2. Low residual overlap plot
  results$low_overlap_plot <- ggplot(data_enhanced, 
         aes(x = .data[[lon_col]], y = .data[[lat_col]], 
             color = factor(low_overlap_count))) +
    geom_point(alpha = 0.6) +
    scale_color_viridis_d(option = "mako", direction = -1) +
    theme_minimal() +
    labs(title = "Overlap of Low Residuals",
         color = "Number of\nOverlapping\nLow Residuals") +
    theme(legend.position = "right")
  
  # 3. Extreme (either high or low) residual overlap plot
  results$extreme_overlap_plot <- ggplot(data_enhanced, 
         aes(x = .data[[lon_col]], y = .data[[lat_col]], 
             color = factor(extreme_overlap_count))) +
    geom_point(alpha = 0.6) +
    scale_color_viridis_d(option = "viridis", direction = -1) +
    theme_minimal() +
    labs(title = "Overlap of Extreme Residuals (High or Low)",
         color = "Number of\nOverlapping\nExtreme Residuals") +
    theme(legend.position = "right")
  
  # Create summary of overlaps
  results$overlap_summary <- data_enhanced %>%
    summarise(
      locations_with_no_extremes = sum(extreme_overlap_count == 0),
      locations_with_one_extreme = sum(extreme_overlap_count == 1),
      locations_with_multiple_extremes = sum(extreme_overlap_count > 1),
      max_extreme_overlap = max(extreme_overlap_count),
      locations_with_no_high = sum(high_overlap_count == 0),
      locations_with_one_high = sum(high_overlap_count == 1),
      locations_with_multiple_high = sum(high_overlap_count > 1),
      max_high_overlap = max(high_overlap_count),
      locations_with_no_low = sum(low_overlap_count == 0),
      locations_with_one_low = sum(low_overlap_count == 1),
      locations_with_multiple_low = sum(low_overlap_count > 1),
      max_low_overlap = max(low_overlap_count)
    )
  
  # Create detailed overlap patterns
  results$overlap_patterns <- data_enhanced %>%
    group_by(extreme_overlap_count) %>%
    summarise(
      count = dplyr::n(),
      pct_of_total = round(100 * dplyr::n() / nrow(data_enhanced), 1),
      .groups = "drop"
    ) %>%
    arrange(desc(count))
  
  # Add the enhanced dataset to results
  results$data <- data_enhanced
  
  return(results)
}

# Example usage
residual_columns <- c("resid_pm", "resid_o3", "resid_no2", "resid_ndvi")
overlap_results <- analyze_residual_overlap(all_data_inc_na, residual_columns)

# Display summary statistics
print("Summary of overlapping patterns:")
print(overlap_results$overlap_summary)

print("\nDistribution of extreme residual overlaps:")
print(overlap_results$overlap_patterns)

# Create a combined visualization using patchwork
library(patchwork)
combined_plot <- overlap_results$high_overlap_plot + 
  overlap_results$low_overlap_plot +
  overlap_results$extreme_overlap_plot +
  plot_layout(ncol = 3) +
  plot_annotation(
    title = "Spatial Patterns of Overlapping Residuals",
    subtitle = "Left: High Residuals, Center: Low Residuals, Right: Combined Extreme Residuals"
  )

print(combined_plot)

# Find locations with multiple extreme residuals
extreme_overlap_locations <- overlap_results$data %>%
  filter(extreme_overlap_count >= 2) %>%
  select(long, lat, extreme_overlap_count, 
         ends_with("_category")) %>%
  arrange(desc(extreme_overlap_count))

print("Locations with multiple extreme residuals:")
head(extreme_overlap_locations, 10)

# Create a summary table of which combinations of exposures tend to have extreme residuals together
exposure_combinations <- overlap_results$data %>%
  filter(extreme_overlap_count >= 2) %>%
  summarise(
    PM_O3 = sum((resid_pm_category != "Normal") & (resid_o3_category != "Normal")),
    PM_NO2 = sum((resid_pm_category != "Normal") & (resid_no2_category != "Normal")),
    PM_NDVI = sum((resid_pm_category != "Normal") & (resid_ndvi_category != "Normal")),
    O3_NO2 = sum((resid_o3_category != "Normal") & (resid_no2_category != "Normal")),
    O3_NDVI = sum((resid_o3_category != "Normal") & (resid_ndvi_category != "Normal")),
    NO2_NDVI = sum((resid_no2_category != "Normal") & (resid_ndvi_category != "Normal"))
  )

print("\nCommon pairs of exposures with extreme residuals:")
print(exposure_combinations)
```
More detailed summaries on where residuals overlap
```{r}
analyze_residual_overlap <- function(data, residual_cols, lon_col = "long", lat_col = "lat", 
                                   geoid_col = "geoid", county_col = "NAME", # 
                                   lower_quantile = 0.1, upper_quantile = 0.9) {
  # Input validation
  if (!all(c(residual_cols, lon_col, lat_col, geoid_col, county_col) %in% names(data))) {
    stop("One or more specified columns not found in data")
  }
  
  # Create category columns for each residual
  results <- list()
  data_enhanced <- data
  
  # Categorize each residual
  for (col in residual_cols) {
    thresholds <- quantile(data[[col]], c(lower_quantile, upper_quantile), na.rm = TRUE)
    cat_col_name <- paste0(col, "_category")
    
    data_enhanced[[cat_col_name]] <- case_when(
      data[[col]] > thresholds[2] ~ "High",
      data[[col]] < thresholds[1] ~ "Low",
      TRUE ~ "Normal"
    )
  }
  
  # Count overlapping high/low/extreme areas
  data_enhanced$high_overlap_count <- rowSums(
    sapply(residual_cols, function(col) {
      data_enhanced[[paste0(col, "_category")]] == "High"
    })
  )
  
  data_enhanced$low_overlap_count <- rowSums(
    sapply(residual_cols, function(col) {
      data_enhanced[[paste0(col, "_category")]] == "Low"
    })
  )
  
  # Count either high or low (extreme) overlaps
  data_enhanced$extreme_overlap_count <- rowSums(
    sapply(residual_cols, function(col) {
      data_enhanced[[paste0(col, "_category")]] %in% c("High", "Low")
    })
  )
  
  # Create overlap visualizations
  # 1. High residual overlap plot
  results$high_overlap_plot <- ggplot(data_enhanced, 
         aes(x = .data[[lon_col]], y = .data[[lat_col]], 
             color = factor(high_overlap_count))) +
    geom_point(alpha = 0.6) +
    scale_color_viridis_d(option = "magma", direction = -1) +
    theme_minimal() +
    labs(title = "Overlap of High Residuals",
         color = "Number of\nOverlapping\nHigh Residuals") +
    theme(legend.position = "right")
  
  # Create location summaries with geoids and county names
  results$location_summary <- data_enhanced %>%
    filter(extreme_overlap_count > 0) %>%
    select(
      !!sym(geoid_col),
      !!sym(county_col), # Added county name
      !!sym(lon_col),
      !!sym(lat_col),
      extreme_overlap_count,
      high_overlap_count,
      low_overlap_count,
      ends_with("_category")
    ) %>%
    arrange(desc(extreme_overlap_count))
  
  # Summary of overlap patterns by geoid and county
  results$overlap_patterns <- data_enhanced %>%
    group_by(extreme_overlap_count) %>%
    summarise(
      n_locations = dplyr::n(),
      geoids = list(get(geoid_col)),
      counties = list(unique(get(county_col))), # Added county names
      pct_of_total = round(100 * dplyr::n() / nrow(data_enhanced), 1),
      .groups = "drop"
    ) %>%
    arrange(desc(extreme_overlap_count))
  
  # Create detailed overlap combinations for high overlap areas
  results$overlap_combinations <- data_enhanced %>%
    filter(extreme_overlap_count >= 2) %>%
    group_by(!!sym(geoid_col)) %>%
    summarise(
      county = first(!!sym(county_col)), # Added county name
      extreme_overlap_count = first(extreme_overlap_count),
      combination = paste(names(which(sapply(residual_cols, function(col) {
        first(get(paste0(col, "_category"))) != "Normal"
      }))), collapse = ", "),
      categories = paste(sapply(residual_cols, function(col) {
        paste0(col, ": ", first(get(paste0(col, "_category"))))
      }), collapse = "; "),
      .groups = "drop"
    ) %>%
    arrange(desc(extreme_overlap_count))
  
  # Add the enhanced dataset to results
  results$data <- data_enhanced
  
  return(results)
}

overlap_results <- analyze_residual_overlap(
   data = all_data_inc_na,
   residual_cols = c("resid_pm", "resid_o3", "resid_no2", "resid_ndvi"),
   geoid_col = "GEOID",
   county_col = "NAME"
 )

# View locations with overlaps
print(overlap_results$location_summary)

# View overlap patterns
print(overlap_results$overlap_patterns)

# View detailed combinations
print(overlap_results$overlap_combinations)



```

```{r}
library(spdep)  # for spatial weights and Moran's I
library(dbscan) # for DBSCAN clustering
library(sf)     # for spatial operations
library(ggplot2)
library(dplyr)

analyze_residual_clusters <- function(data, residual_cols, lon_col = "long", lat_col = "lat") {
  results <- list()
  
  # Debug data structure
  cat("Data structure diagnostics:\n")
  cat("Class of input data:", class(data), "\n")
  cat("Dimension of input data:", dim(data), "\n")
  cat("Classes of coordinate columns:\n")
  print(sapply(data[c(lon_col, lat_col)], class))
  
  # Input validation
  if (!all(c(lon_col, lat_col) %in% names(data))) {
    stop("Longitude and latitude columns not found in data")
  }
  
  if (!all(residual_cols %in% names(data))) {
    stop("One or more residual columns not found in data")
  }
  
  # Function to safely convert potentially nested columns to numeric
  safe_to_numeric <- function(x) {
    if (is.list(x)) {
      # If it's a list column, try to unlist it first
      x <- unlist(x)
    }
    # Try to convert to character first, then to numeric
    as.numeric(as.character(x))
  }
  
  # Extract and validate coordinates
  coords <- tryCatch({
    # Create a clean data frame for coordinates
    coords_df <- data.frame(
      longitude = safe_to_numeric(data[[lon_col]]),
      latitude = safe_to_numeric(data[[lat_col]])
    )
    
    # Check for conversion success
    if (any(is.na(coords_df$longitude)) && !all(is.na(coords_df$longitude))) {
      warning(paste("Some longitude values could not be converted to numeric. First problematic value:",
                   data[[lon_col]][which(!is.na(data[[lon_col]]) & is.na(coords_df$longitude))[1]]))
    }
    if (any(is.na(coords_df$latitude)) && !all(is.na(coords_df$latitude))) {
      warning(paste("Some latitude values could not be converted to numeric. First problematic value:",
                   data[[lat_col]][which(!is.na(data[[lat_col]]) & is.na(coords_df$latitude))[1]]))
    }
    
    # Remove rows with NA coordinates
    valid_rows <- complete.cases(coords_df)
    if (sum(!valid_rows) > 0) {
      warning(paste(sum(!valid_rows), "rows removed due to invalid coordinates"))
    }
    coords_df <- coords_df[valid_rows, ]
    
    # Validate coordinate ranges
    invalid_lon <- coords_df$longitude < -180 | coords_df$longitude > 180
    invalid_lat <- coords_df$latitude < -90 | coords_df$latitude > 90
    
    if (any(invalid_lon)) {
      warning(paste("Found", sum(invalid_lon), "longitude values outside valid range (-180 to 180)"))
    }
    if (any(invalid_lat)) {
      warning(paste("Found", sum(invalid_lat), "latitude values outside valid range (-90 to 90)"))
    }
    
    # Remove invalid coordinates
    coords_df <- coords_df[!invalid_lon & !invalid_lat, ]
    
    if (nrow(coords_df) == 0) {
      stop("No valid coordinates remaining after cleaning")
    }
    
    as.matrix(coords_df)
  }, error = function(e) {
    stop(paste("Error processing coordinates:", e$message, "\nPlease check your coordinate data structure."))
  })
  
  # Function to analyze single residual with error handling
  analyze_single_residual <- function(residual_col) {
    tryCatch({
      # Clean residual data
      residual_values <- safe_to_numeric(data[[residual_col]])
      if (all(is.na(residual_values))) {
        stop("All values in residual column are NA")
      }
      
      # Match residuals with valid coordinates
      residual_values <- residual_values[valid_rows][!invalid_lon & !invalid_lat]
      
      # 1. Global Moran's I
      knn <- knearneigh(coords, k = min(5, nrow(coords) - 1))
      nb <- knn2nb(knn)
      weights <- nb2listw(nb, style = "W")
      
      moran_result <- moran.test(residual_values, weights)
      
      # 2. Local Moran's I (LISA)
      lisa_result <- localmoran(residual_values, weights)
      
      # 3. DBSCAN Clustering
      clustering_data <- cbind(scale(coords), scale(residual_values))
      eps <- 0.5 * sd(dist(clustering_data))
      dbscan_result <- dbscan(clustering_data, eps = eps, minPts = min(5, nrow(clustering_data)/10))
      
      # Store results
      results_df <- data.frame(
        lisa_pvalue = lisa_result[, 5],
        cluster = dbscan_result$cluster
      )
      
      return(list(
        moran = moran_result,
        lisa = lisa_result,
        dbscan = dbscan_result,
        results_df = results_df
      ))
    }, error = function(e) {
      warning(paste("Error analyzing", residual_col, ":", e$message))
      return(NULL)
    })
  }
  
  # Analyze each residual
  results$individual_analyses <- lapply(residual_cols, analyze_single_residual)
  names(results$individual_analyses) <- residual_cols
  
  # Create summary
  results$summary <- data.frame(
    residual = residual_cols,
    n_valid_points = sapply(results$individual_analyses, function(x) {
      if (!is.null(x)) nrow(x$results_df) else NA
    }),
    moran_i = sapply(results$individual_analyses, function(x) {
      if (!is.null(x)) x$moran$estimate[1] else NA
    }),
    moran_p = sapply(results$individual_analyses, function(x) {
      if (!is.null(x)) x$moran$p.value else NA
    })
  )
  
  return(results)
}

# Check the structure of your data
str(all_data_inc_na[c("long", "lat")])

# If needed, unnest your coordinate columns
if (is.list(all_data_inc_na$long)) {
  all_data_inc_na$long <- unlist(all_data_inc_na$long)
}
if (is.list(all_data_inc_na$lat)) {
  all_data_inc_na$lat <- unlist(all_data_inc_na$lat)
}

# Then run the analysis
residual_columns <- c("resid_pm", "resid_o3", "resid_no2", "resid_ndvi")
cluster_results <- analyze_residual_clusters(all_data_inc_na, residual_columns)

# Check the results
print(cluster_results$summary)
```

################################
Create model functions
```{r}
# Create exposure list
exposure_list <- list(
  pm = list(
    name = "pm",
    baseline = "pm_2002",
    change = "pm_change"
  ),
  o3 = list(
    name = "o3",
    baseline = "o3_2002",
    change = "o3_change"
  ),
  no2 = list(
    name = "no2",
    baseline = "no2_2000",
    change = "no2_change"
  ),
  ndvi = list(
    name = "ndvi",
    baseline = "ndvi_2000",
    change = "ndvi_change"
  ),
  temp = list(
    name = "mean_temp",
    baseline = NULL,
    change = "mean_temp"  # No change variable for temp
  )
)

# Function to run GAM models
run_gam_models <- function(data, exposure_info) {
  # Extract exposure information
  exp_var <- exposure_info$change
  baseline_exposure <- exposure_info$baseline
  
  # Skip if data is missing the exposure variable - gives warning 
  if (!exp_var %in% names(data)) {
    warning(paste("Variable", exp_var, "not found in dataset"))
    return(NULL)
  }
  
  # Base components that will be in all models: gentrification score and county/borough
  base_formula <- paste(exp_var, "~ score_0.5 + s(COUNTYFP10, bs='re')")
  
  # Create model formulas
  formulas <- list(
    # Base models (score + county + spatial) for tensor product and mrf smooth
    tp_base = paste(base_formula, "+ te(long, lat, d = 2, bs = 'gp', m = 2)"),
    mrf_base = paste(base_formula, "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),
    
    # Models with income
    tp_income = paste(base_formula, "+ s(med_income_2016) + te(long, lat, d = 2, bs = 'gp', m = 2)"),
    mrf_income = paste(base_formula, "+ s(med_income_2016) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),
    mrf_income_rank = paste(base_formula, "+ rank(med_income_2016) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),

    # Models with distance to downtown
    tp_dist = paste(base_formula, "+ s(dt_dist) + te(long, lat, d = 2, bs = 'gp', m = 2)"),
    mrf_dist = paste(base_formula, "+ s(dt_dist) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  )
  
  # Add baseline models if baseline exposure level exists (all except temp)
  if (!is.null(baseline_exposure) && baseline_exposure %in% names(data)) {
    formulas$tp_baseline <- paste(base_formula, "+", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_baseline <- paste(base_formula, "+", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    # Combinations with baseline
    formulas$tp_income_baseline <- paste(base_formula, "+ s(med_income_2016) +", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_income_baseline <- paste(base_formula, "+ s(med_income_2016) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    formulas$tp_dist_baseline <- paste(base_formula, "+ s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_dist_baseline <- paste(base_formula, "+ s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    formulas$tp_full <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_full <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  }
  
  # Add income + distance combinations
  formulas$tp_income_dist <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) + te(long, lat, d = 2, bs = 'gp', m = 2)")
  formulas$mrf_income_dist <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  
  # Convert formulas to formula objects
  formulas <- lapply(formulas, as.formula)
  
  # Run models
  models <- purrr::map(formulas, ~try(gam(., 
                                        family = scat, 
                                        data = data, 
                                        method = "REML"), 
                                    silent = TRUE))
  
  return(models)
}

# drop geometry from spatial data while keeping coordinates
data_prepared <- all_data_inc %>%
  mutate(
    COUNTYFP10 = as.factor(COUNTYFP10),
    GEOID = as.factor(GEOID)
  ) %>%
  st_drop_geometry()

# Construct neighbors list for MRF models
nb <- poly2nb(all_data_inc, row.names = all_data_inc$GEOID)
names(nb) <- attr(nb, "region.id") 

# JH 10/25/24 Warning: some observations have no neighbours; if this seems unexpected, try increasing the snap argument.Warning: neighbour object has 4 sub-graphs; if this sub-graph count seems unexpected, try increasing the snap argument.
```
Checking neighbor relationships for MRF models
```{r}
# Visualize the sub-graphs
plot(all_data_inc$geometry)
colors <- rainbow(length(nb))
plot(nb, st_coordinates(st_centroid(all_data_inc)), col=colors, add=TRUE)
# the sub-graphs match the islands.
# since we are adjusting by borough in the models, the 4 sub-graphs might be okay?? JH 10/26/24
```

Run models and check models
```{r}
# Run models for each exposure
all_models <- purrr::map(exposure_list, ~run_gam_models(data_prepared, .))

# Function to check which models succeeded/failed
check_models <- function(models) {
  model_status <- map(models, ~!inherits(., "try-error"))
  successful_models <- names(models)[unlist(model_status)]
  failed_models <- names(models)[!unlist(model_status)]
  
  return(list(
    successful = successful_models,
    failed = failed_models
  ))
}

# Check status of all models
model_status <- map(all_models, check_models)
model_status

```


Model diagnostics (11/8/24)
```{r}
# Extract the mrf_income models
mrf_income_models <- purrr::map(all_models, "mrf_income")

# Check the summary of all mrf_income models
purrr::walk(mrf_income_models, summary)

summary(mrf_income_models$pm)
plot(resid(mrf_income_models$pm))
plot(mrf_income_models$pm, pages = 1)

#gam check
gam.check(mrf_income_models$pm)
gam.check(mrf_income_models$o3)
gam.check(mrf_income_models$no2)
gam.check(mrf_income_models$ndvi)
gam.check(mrf_income_models$temp)
# p values all non sig fr checking which is good  for med income, but GEOID and county return NA p values?

#concurvity - high worst between s(COUNTYFP10) and s(GEOID) - 0.86 -problematic?
concurvity(mrf_income_models$pm, full = TRUE)
concurvity(mrf_income_models$pm, full = FALSE)
concurvity(mrf_income_models$o3, full = TRUE)
concurvity(mrf_income_models$o3, full = FALSE)
concurvity(mrf_income_models$no2, full = TRUE)
concurvity(mrf_income_models$no2, full = FALSE)
concurvity(mrf_income_models$ndvi, full = TRUE)
concurvity(mrf_income_models$ndvi, full = FALSE)

# Get the fitted values and residuals
fitted_values <- fitted(mrf_income_models$pm)
residuals <- resid(mrf_income_models$pm)

# Create a data frame for plotting
plot_data <- data.frame(fitted = fitted_values, residuals = residuals)

# Plot Residuals vs Fitted
ggplot(plot_data, aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted", x = "Fitted Values", y = "Residuals") +
  theme_minimal()



traceback()

```
# Plot the residuals for the "pm" mrf_income model


Create basic QQ plots for models
```{r}
# Function to create QQ plots for a specific exposure
create_exposure_qq_plots <- function(models_list, exposure_name) {
  exposure_models <- models_list[[exposure_name]]
  
  plots <- names(exposure_models) %>%
    purrr::map(function(model_name) {
      model <- exposure_models[[model_name]]
      
      # Skip if model is an error
      if (inherits(model, "try-error")) {
        return(NULL)
      }
      
      # Create QQ plot
      resid_df <- data.frame(
        residuals = residuals(model, type = "deviance"),
        fitted = fitted(model)
      )
      
      ggplot(resid_df, aes(sample = residuals)) +
        stat_qq() +
        stat_qq_line(color = "red") +
        labs(
          title = paste0(exposure_name, "\n", model_name),
          x = "Theoretical Quantiles",
          y = "Sample Quantiles"
        ) +
        theme_minimal() +
        theme(
          plot.title = element_text(size = 10, hjust = 0.5),
          axis.title = element_text(size = 8),
          axis.text = element_text(size = 7)
        )
    }) %>%
    purrr::compact() # Remove NULL values from the list
  
  # Calculate grid layout
  n_plots <- length(plots)
  n_cols <- min(3, n_plots)
  n_rows <- ceiling(n_plots / n_cols)
  
  # Combine plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = plots,
    ncol = n_cols,
    nrow = n_rows,
    top = grid::textGrob(paste("QQ Plots for", exposure_name), 
                        gp = grid::gpar(fontsize = 12))
  )
  
  return(combined_plot)
}

# Function to create QQ plots for a specific model type across exposures
create_model_type_qq_plots <- function(models_list, model_type) {
  # Convert the for loop to purrr::map
  plots <- names(models_list) %>%
    purrr::map(function(exposure_name) {
      exposure_models <- models_list[[exposure_name]]
      
      # Check if the model type exists for this exposure
      if (!model_type %in% names(exposure_models)) {
        return(NULL)
      }
      
      model <- exposure_models[[model_type]]
      
      # Skip if model is an error
      if (inherits(model, "try-error")) {
        return(NULL)
      }
      
      # Create QQ plot
      resid_df <- data.frame(
        residuals = residuals(model, type = "deviance"),
        fitted = fitted(model)
      )
      
      ggplot(resid_df, aes(sample = residuals)) +
        stat_qq() +
        stat_qq_line(color = "red") +
        labs(
          title = paste0(exposure_name, "\n", model_type),
          x = "Theoretical Quantiles",
          y = "Sample Quantiles"
        ) +
        theme_minimal() +
        theme(
          plot.title = element_text(size = 10, hjust = 0.5),
          axis.title = element_text(size = 8),
          axis.text = element_text(size = 7)
        )
    }) %>%
    purrr::compact() # Remove NULL values from the list
  
  # Calculate grid layout
  n_plots <- length(plots)
  n_cols <- min(3, n_plots)
  n_rows <- ceiling(n_plots / n_cols)
  
  # Combine plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = plots,
    ncol = n_cols,
    nrow = n_rows,
    top = grid::textGrob(paste("QQ Plots for", model_type), 
                        gp = grid::gpar(fontsize = 12))
  )
  
  return(combined_plot)
}

# Function to list available model types
list_model_types <- function(models_list) {
  unique(unlist(lapply(models_list, names)))
}

# Function to list available exposures
list_exposures <- function(models_list) {
  names(models_list)
}

```


Display basic QQ plots for models
```{r}
# List all model types
model_types <- list_model_types(all_models)

# Display all QQ plots
purrr::walk(model_types, ~create_model_type_qq_plots(all_models, .))

```

Create functions for QQ plots with confidence intervals
```{r}
# Function to create a single QQ plot with confidence bands
create_single_qq_plot <- function(model, title) {
  if (inherits(model, "try-error")) return(NULL)
  
  data.frame(
    residuals = residuals(model, type = "deviance"),
    fitted = fitted(model)
  ) %>%
    ggplot(aes(sample = residuals)) +
    geom_qq_band(bandType = "ks", mapping = aes(fill = "KS"), alpha = 0.5) +
    geom_qq_band(bandType = "ts", mapping = aes(fill = "TS"), alpha = 0.5) +
    geom_qq_band(bandType = "pointwise", mapping = aes(fill = "Normal"), alpha = 0.5) +
    geom_qq_band(bandType = "boot", mapping = aes(fill = "Bootstrap"), alpha = 0.5) +  
    stat_qq_line(color = "red") +
    stat_qq_point(size = 1) +
    labs(
      title = title,
      x = "Theoretical Quantiles",
      y = "Sample Quantiles"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 10, hjust = 0.5),
      axis.title = element_text(size = 8),
      axis.text = element_text(size = 7)
    )
}

# Function to arrange plots in a grid
arrange_plots <- function(plots, main_title) {
  n_plots <- length(plots)
  n_cols <- min(3, n_plots)  
  n_rows <- ceiling(n_plots / n_cols)
  
  gridExtra::grid.arrange(
    grobs = plots,
    ncol = n_cols,
    nrow = n_rows,
    top = grid::textGrob(
      main_title,
      gp = grid::gpar(fontsize = 12)  
    )
  )
}

# Main display function
display_qq_plots_by_model_type <- function(models_list) {
  # Get all unique model types
  model_types <- models_list %>%
    map(names) %>%
    reduce(union)
  
  # Create plots for each model type
  walk(model_types, function(model_type) {
    # Extract all models of this type across exposures
    plots <- models_list %>%
      imap(function(exposure_models, exposure_name) {
        if (!model_type %in% names(exposure_models)) return(NULL)
        
        create_single_qq_plot(
          exposure_models[[model_type]], 
          title = paste0(exposure_name, "\n", model_type)
        )
      }) %>%
      compact()
    
    # Arrange and display plots for this model type
    arrange_plots(
      plots,
      main_title = paste("QQ Plots for", model_type)
    )
  })
}

```

Display QQ plots with confidence bands for models
```{r}
# Display QQ plots
display_qq_plots_by_model_type(all_models)
```

Create comprehensive diagnostic plots (11/2 - need to change the scale of the plots for visibility)
```{r}
create_model_diagnostics_by_type <- function(models_list, model_type) {
  # Create diagnostic plots for each exposure
  diagnostic_plots <- models_list %>%
    imap(function(exposure_models, exposure_name) {
      # Skip if model type doesn't exist or is error
      if (!model_type %in% names(exposure_models) || 
          inherits(exposure_models[[model_type]], "try-error")) {
        return(NULL)
      }
      
      model <- exposure_models[[model_type]]
      
      # Create residuals dataframe
      resid_df <- tibble(
        residuals = residuals(model, type = "deviance"),
        fitted = fitted(model),
        scale_residuals = residuals(model, type = "scaled.pearson"),
        response = model$y
      )
      
      # Create the four diagnostic plots
      list(
        # 1. QQ plot
        qq = ggplot(resid_df, aes(sample = residuals)) +
          stat_qq() +
          stat_qq_line(color = "red") +
          labs(title = paste(exposure_name, "- Normal Q-Q Plot")) +
          theme_minimal() +
          theme(plot.title = element_text(size = 8)),
        
        # 2. Residuals vs Fitted
        resid = ggplot(resid_df, aes(x = fitted, y = residuals)) +
          geom_point(alpha = 0.5) +
          geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
          geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
          labs(title = paste(exposure_name, "- Residuals vs Fitted"),
               x = "Fitted values",
               y = "Deviance residuals") +
          theme_minimal() +
          theme(plot.title = element_text(size = 8)),
        
        # 3. Scale-location plot
        scale = ggplot(resid_df, 
                      aes(x = fitted, y = sqrt(abs(scale_residuals)))) +
          geom_point(alpha = 0.5) +
          geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
          labs(title = paste(exposure_name, "- Scale-Location"),
               x = "Fitted values",
               y = "√|Standardized residuals|") +
          theme_minimal() +
          theme(plot.title = element_text(size = 8)),
        
        # 4. Response vs Fitted
        resp = ggplot(resid_df, aes(x = fitted, y = response)) +
          geom_point(alpha = 0.5) +
          geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
          labs(title = paste(exposure_name, "- Response vs Fitted"),
               x = "Fitted values",
               y = "Response") +
          theme_minimal() +
          theme(plot.title = element_text(size = 8))
      )
    }) %>%
    compact() # Remove NULL entries
  
  # Stop if no valid models found
  if (length(diagnostic_plots) == 0) {
    stop("No valid models found for the specified model type")
  }
  
  # Flatten the nested list of plots
  all_plots <- diagnostic_plots %>%
    flatten()
  
  # Calculate grid dimensions
  n_exposures <- length(diagnostic_plots)
  n_cols <- 4  # One column for each type of diagnostic plot
  
  # Combine all plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = all_plots,
    ncol = n_cols,
    nrow = n_exposures,
    top = grid::textGrob(
      paste("Diagnostic Plots for Model Type:", model_type),
      gp = grid::gpar(fontsize = 12, fontface = "bold")
    )
  )
  
  return(combined_plot)
}

display_all_model_diagnostics <- function(models_list) {
  # Get unique model types
  model_types <- models_list %>%
    map(names) %>%
    flatten_chr() %>%
    unique()
  
  # Create and display all plots at once
  plots <- model_types %>%
    map(~{
      tryCatch({
        create_model_diagnostics_by_type(models_list, .x)
      }, error = function(e) {
        message(paste("Error creating plots for model type:", .x))
        message(e)
        return(NULL)
      })
    }) %>%
    compact()
  
  # Calculate grid layout for all model types
  n_types <- length(plots)
  grid_dims <- c(ceiling(sqrt(n_types)), floor(sqrt(n_types)))
  if (prod(grid_dims) < n_types) grid_dims[2] <- grid_dims[2] + 1
  
  # Display all plots in a single grid
  do.call(gridExtra::grid.arrange, 
          c(plots, 
            list(ncol = grid_dims[1],
                 nrow = grid_dims[2])))
}
display_all_model_diagnostics(all_models)

```
```{r}

create_model_diagnostics_by_type <- function(models_list, model_type) {
  # Create diagnostic plots for each exposure
  diagnostic_plots <- models_list %>%
    imap(function(exposure_models, exposure_name) {
      # Skip if model type doesn't exist or is error
      if (!model_type %in% names(exposure_models) || 
          inherits(exposure_models[[model_type]], "try-error")) {
        return(NULL)
      }
      
      model <- exposure_models[[model_type]]
      
      # Create residuals dataframe
      resid_df <- tibble(
        residuals = residuals(model, type = "deviance"),
        fitted = fitted(model),
        scale_residuals = residuals(model, type = "scaled.pearson"),
        response = model$y
      )
      
      # Create the four diagnostic plots
      list(
        # 1. QQ plot
        qq = ggplot(resid_df, aes(sample = residuals)) +
          stat_qq() +
          stat_qq_line(color = "red") +
          labs(title = paste(exposure_name, "- Normal Q-Q Plot")) +
          theme_minimal(base_size = 14) +
          theme(plot.title = element_text(size = 16, face = "bold"),
                axis.title = element_text(size = 12),
                axis.text = element_text(size = 12)),
        
        # 2. Residuals vs Fitted
        resid = ggplot(resid_df, aes(x = fitted, y = residuals)) +
          geom_point(alpha = 0.5) +
          geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
          geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
          labs(title = paste(exposure_name, "- Residuals vs Fitted"),
               x = "Fitted values",
               y = "Deviance residuals") +
          theme_minimal(base_size = 14) +
          theme(plot.title = element_text(size = 16, face = "bold"),
                axis.title = element_text(size = 12),
                axis.text = element_text(size = 12)),
        
        # 3. Scale-location plot
        scale = ggplot(resid_df, 
                      aes(x = fitted, y = sqrt(abs(scale_residuals)))) +
          geom_point(alpha = 0.5) +
          geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
          labs(title = paste(exposure_name, "- Scale-Location"),
               x = "Fitted values",
               y = "√|Standardized residuals|") +
          theme_minimal(base_size = 14) +
          theme(plot.title = element_text(size = 16, face = "bold"),
                axis.title = element_text(size = 12),
                axis.text = element_text(size = 12)),
        
        # 4. Response vs Fitted
        resp = ggplot(resid_df, aes(x = fitted, y = response)) +
          geom_point(alpha = 0.5) +
          geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
          labs(title = paste(exposure_name, "- Response vs Fitted"),
               x = "Fitted values",
               y = "Response") +
          theme_minimal(base_size = 14) +
          theme(plot.title = element_text(size = 16, face = "bold"),
                axis.title = element_text(size = 12),
                axis.text = element_text(size = 12))
      )
    }) %>%
    compact() # Remove NULL entries
  
  # Stop if no valid models found
  if (length(diagnostic_plots) == 0) {
    stop("No valid models found for the specified model type")
  }
  
  # Flatten the nested list of plots
  all_plots <- diagnostic_plots %>%
    flatten()
  
  # Calculate grid dimensions
  n_exposures <- length(diagnostic_plots)
  n_cols <- 4  # One column for each type of diagnostic plot
  
  # Combine all plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = all_plots,
    ncol = n_cols,
    nrow = n_exposures,
    top = grid::textGrob(
      paste("Diagnostic Plots for Model Type:", model_type),
      gp = grid::gpar(fontsize = 18, fontface = "bold")
    ),
    heights = rep(1, n_exposures),  # Adjust height of each row to ensure plots are not squished
    padding = unit(1, "cm")  # Add padding between plots
  )
  
  return(combined_plot)
}

display_all_model_diagnostics <- function(models_list) {
  # Get unique model types
  model_types <- models_list %>%
    map(names) %>%
    flatten_chr() %>%
    unique()
  
  # Create and display all plots at once
  plots <- model_types %>%
    map(~{
      tryCatch({
        create_model_diagnostics_by_type(models_list, .x)
      }, error = function(e) {
        message(paste("Error creating plots for model type:", .x))
        message(e)
        return(NULL)
      })
    }) %>%
    compact()
  
  # Calculate grid layout for all model types
  n_types <- length(plots)
  grid_dims <- c(ceiling(sqrt(n_types)), floor(sqrt(n_types)))
  if (prod(grid_dims) < n_types) grid_dims[2] <- grid_dims[2] + 1
  
  # Display all plots in a single grid
  do.call(gridExtra::grid.arrange, 
          c(plots, 
            list(ncol = grid_dims[1],
                 nrow = grid_dims[2],
                 top = grid::textGrob("All Model Diagnostics", 
                                      gp = grid::gpar(fontsize = 18, fontface = "bold")))))
}
display_all_model_diagnostics(all_models)


```

Create effects summary data frame and forest plot
```{r}
# Function to extract effect estimates for score_0.5
get_score_effects <- function(models_list, model_type = "mrf_income") {
  effects_list <- list()
  
  for (exposure_name in names(models_list)) {
    model <- models_list[[exposure_name]][[model_type]]
    
    if (!inherits(model, "try-error")) {
      # Get predicted values across range of scores
      score_seq <- seq(min(model$model$score_0.5), max(model$model$score_0.5), length.out = 100)
      pred_data <- data.frame(
        score_0.5 = score_seq,
        med_income_2016 = mean(model$model$med_income_2016),
        COUNTYFP10 = levels(model$model$COUNTYFP10)[1],
        GEOID = levels(model$model$GEOID)[1]
      )
      
      # Get predictions with standard errors
      preds <- predict(model, pred_data, type = "terms", se.fit = TRUE)
      
      # Extract effect for score_0.5 smooth term
      score_effect <- data.frame(
        exposure = exposure_name,
        estimate = diff(range(preds$fit[,"s(score_0.5)"])),
        se = diff(range(preds$se.fit[,"s(score_0.5)"])),
        p_value = summary(model)$s.table["s(score_0.5)", "p-value"]
      )
      
      effects_list[[exposure_name]] <- score_effect
    }
  }
  
  # Combine all effects into one dataframe
  effects_df <- do.call(rbind, effects_list)
  
  # Calculate confidence intervals
  effects_df$lower <- effects_df$estimate - 1.96 * effects_df$se
  effects_df$upper <- effects_df$estimate + 1.96 * effects_df$se
  
  return(effects_df)
}

# Get effects
effects_df <- get_score_effects(all_models)

# Create forest plot
library(ggplot2)

# Create exposure labels with significance indicators
effects_df$exposure_label <- paste0(
  toupper(effects_df$exposure), 
  ifelse(effects_df$p_value < 0.001, "***",
         ifelse(effects_df$p_value < 0.01, "**",
                ifelse(effects_df$p_value < 0.05, "*", "")))
)

# Create the forest plot
ggplot(effects_df, aes(y = exposure_label)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(x = estimate), size = 3) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  labs(
    title = "Estimated Effects of Gentrification Score on Environmental Exposures",
    subtitle = "MRF Income Model Results",
    x = "Effect Size (Range of Smooth Term)",
    y = "Exposure Type",
    caption = "* p < 0.05, ** p < 0.01, *** p < 0.001"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    plot.caption = element_text(size = 8, hjust = 0),
    panel.grid.minor = element_blank()
  )

# Print numerical results
print(effects_df[, c("exposure_label", "estimate", "se", "p_value", "lower", "upper")])
```


Scale mrf_income models by IQR and create scaled forest plot
```{r}
# Function to extract IQR-scaled effect estimates for score_0.5
get_score_effects_scaled <- function(models_list, model_type = "mrf_income") {
  effects_list <- list()
  
  for (exposure_name in names(models_list)) {
    model <- models_list[[exposure_name]][[model_type]]
    
    if (!inherits(model, "try-error")) {
      # Get model data
      model_data <- model$model
      
      # Calculate IQRs
      score_iqr <- IQR(model_data$score_0.5)
      exposure_var <- exposure_list[[exposure_name]]$change
      exposure_iqr <- IQR(model_data[[exposure_var]])
      
      # Get predicted values across range of scores
      score_seq <- seq(min(model_data$score_0.5), max(model_data$score_0.5), length.out = 100)
      pred_data <- data.frame(
        score_0.5 = score_seq,
        med_income_2016 = mean(model_data$med_income_2016),
        COUNTYFP10 = levels(model_data$COUNTYFP10)[1],
        GEOID = levels(model_data$GEOID)[1]
      )
      
      # Get predictions with standard errors
      preds <- predict(model, pred_data, type = "terms", se.fit = TRUE)
      
      # Extract and scale effect for score_0.5 smooth term
      raw_effect <- diff(range(preds$fit[,"s(score_0.5)"]))
      raw_se <- diff(range(preds$se.fit[,"s(score_0.5)"]))
      
      # Scale effects by IQRs
      scaled_effect <- (raw_effect * score_iqr) / exposure_iqr
      scaled_se <- (raw_se * score_iqr) / exposure_iqr
      
      score_effect <- data.frame(
        exposure = exposure_name,
        estimate = scaled_effect,
        se = scaled_se,
        p_value = summary(model)$s.table["s(score_0.5)", "p-value"],
        exposure_iqr = exposure_iqr,
        score_iqr = score_iqr
      )
      
      effects_list[[exposure_name]] <- score_effect
    }
  }
  
  # Combine all effects into one dataframe
  effects_df <- do.call(rbind, effects_list)
  
  # Calculate confidence intervals
  effects_df$lower <- effects_df$estimate - 1.96 * effects_df$se
  effects_df$upper <- effects_df$estimate + 1.96 * effects_df$se
  
  return(effects_df)
}

# Get scaled effects
effects_df_scaled <- get_score_effects_scaled(all_models)

# Create forest plot with scaled effects
effects_df_scaled$exposure_label <- paste0(
  toupper(effects_df_scaled$exposure), 
  ifelse(effects_df_scaled$p_value < 0.001, "***",
         ifelse(effects_df_scaled$p_value < 0.01, "**",
                ifelse(effects_df_scaled$p_value < 0.05, "*", "")))
)

ggplot(effects_df_scaled, aes(y = exposure_label)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(x = estimate), size = 3) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  labs(
    title = "IQR-Scaled Effects of Gentrification Score on Environmental Exposures",
    subtitle = "MRF Income Model Results",
    x = "Effect Size (Change in Exposure IQR per Score IQR)",
    y = "Exposure Type",
    caption = "* p < 0.05, ** p < 0.01, *** p < 0.001"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    plot.caption = element_text(size = 8, hjust = 0),
    panel.grid.minor = element_blank()
  )

```

Understanding direction and magnitude of exposure changes 
```{r}
# Function to analyze changes in exposures
analyze_exposure_changes <- function(data, exposure_list) {
  results <- list()
  
  for (exp_name in names(exposure_list)) {
    exp_info <- exposure_list[[exp_name]]
    change_var <- exp_info$change
    
    # Skip temperature which doesn't have baseline/change structure
    if (exp_name != "temp") {
      baseline_var <- exp_info$baseline
      
      # Calculate summary statistics
      change_stats <- data %>%
        summarise(
          n = nrow(data),
          mean_change = mean(.data[[change_var]], na.rm = TRUE),
          median_change = median(.data[[change_var]], na.rm = TRUE),
          sd_change = sd(.data[[change_var]], na.rm = TRUE),
          pct_positive = mean(.data[[change_var]] > 0, na.rm = TRUE) * 100,
          q25 = quantile(.data[[change_var]], 0.25, na.rm = TRUE),
          q75 = quantile(.data[[change_var]], 0.75, na.rm = TRUE)
        )
      
      # Create histogram
      hist_plot <- ggplot(data, aes(x = .data[[change_var]])) +
        geom_histogram(bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
        geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
        labs(
          title = paste("Distribution of Changes in", toupper(exp_name)),
          x = "Change",
          y = "Count"
        ) +
        theme_minimal()
      
      # Store results
      results[[exp_name]] <- list(
        stats = change_stats,
        plot = hist_plot
      )
    }
  }
  
  return(results)
}

# Run the analysis
change_analysis <- analyze_exposure_changes(data_prepared, exposure_list)

# Print summary statistics 
print(change_analysis)

# Create a combined plot of all histograms
plot_list <- lapply(names(change_analysis), function(exp_name) {
  change_analysis[[exp_name]]$plot
})

# Arrange all plots in a grid
library(gridExtra)
grid.arrange(grobs = plot_list, ncol = 2)

# Create boxplots for another view of the distributions
combined_changes <- data.frame(
  Exposure = character(),
  Change = numeric()
)

for (exp_name in names(exposure_list)) {
  if (exp_name != "temp") {
    change_var <- exposure_list[[exp_name]]$change
    tmp_data <- data.frame(
      Exposure = toupper(exp_name),
      Change = data_prepared[[change_var]]
    )
    combined_changes <- rbind(combined_changes, tmp_data)
  }
}

# Create boxplot
ggplot(combined_changes, aes(x = Exposure, y = Change)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of Changes Across All Exposures",
    x = "Exposure Type",
    y = "Change"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```