---
title: "gent_clean_10.25.24"
author: "Joyce Hu"
date: "2024-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages (JH - check if any are unused after code cleaned)
```{r}
library(tidyverse)
library(tigris)
library(arrow)
library(here)
library(mgcv)
library(fixest)
library(lmtest)
library(terra)
library(exactextractr)
library(tidyterra)
library(sf)
library(mgcViz)

library(DHARMa)
library(qqplotr)
library(readxl) # for read_excel()
library(spdep) # for moran.test()
library(tidycensus) # for median income
library(distr) # for qqbounds()
library(spatialreg) # for spatial lag and error models
library(gridExtra) # for viewing plots side-by-side
library(qgam) 
library(dplyr)
```

Load and clean data
```{r}
# Check if 'data' directory exists; if not, create it
if (!dir.exists(here("data"))) {
  dir.create(here("data"))
}

# load gentrification score by tract
gentrification <- read.csv(here("data", "Data_NYC_Gentrification_2000_16.csv")) %>% 
  mutate(tractid = as.character(tractid)) %>% 
  dplyr::select(tractid, score_0.5)

# load gentrification score if using xlsx file:
# gentrification <- read_xlsx(here("Gentrification","Data_NYC_Gentrification_2000_16.xlsx")) %>%
#   mutate(tractid = as.character(tractid)) %>%
#   dplyr::select(tractid, score_0.5)

# pull NYC tract IDs/fips from gentrification index dataset for filtering
nyc_fips <- gentrification$tractid

# load NYC tracts shapefile
nyc_tracts <- tracts(state = "NY", county = c("Bronx", "Kings", "New York", "Queens", "Richmond"), year = 2010) %>%
  # clean lat and long for Conley SEs
  mutate(lat = as.numeric(INTPTLAT10), long = as.numeric(INTPTLON10), GEOID = GEOID10) %>% 
  # filter for FIPS codes included in gentrification index dataset
  filter(GEOID %in% nyc_fips)


#load NY tracts populatuion weighted centroid points from Census data
nyc_weighted_centroids <- read.table("https://www2.census.gov/geo/docs/reference/cenpop2010/tract/CenPop2010_Mean_TR36.txt", 
                         header = TRUE, sep = ",", stringsAsFactors = FALSE, 
                         colClasses = c("character", "character", "character", "numeric", "numeric", "numeric")) %>%
  mutate(FIPS = paste(STATEFP, COUNTYFP, TRACTCE, sep = "")) %>% 
  ## filter for only NYC tracts
  filter(FIPS %in% nyc_fips) %>%  
  rename(Longitude_weighted = LONGITUDE, 
         Latitude_weighted = LATITUDE)

# JH - 10/12/24 warning persists 10/13, 10/14 Warning: R metadata may have unsafe or invalid elements ℹ Type: "externalptr" options(arrow.unsafe_metadata = TRUE) #

# temperature data
if(!file.exists(here("data","summarized_daily_temp_preds.parquet"))){download.file("https://zenodo.org/records/10557980/files/summarized_daily_temp_preds.parquet?download=1", 
                                                                                   destfile = here("data","summarized_daily_temp_preds.parquet"))}

# temperature by tract
temp <- read_parquet(here("data","summarized_daily_temp_preds.parquet")) %>%
  filter(str_detect(date, "2016")) %>%
  mutate(temp_c = mean_temp_daily_cK/100-273.15) %>% 
  mutate(temp_f = temp_c*9/5+32) %>% 
  group_by(GEOID, date) %>%
  ungroup() %>% 
  group_by(GEOID) %>% 
  summarize(mean_temp = mean(mean_temp_daily_cK/100))

# ndvi by tract (need to have it downloaded and stored locally)
ndvi_2016 <- rast(here("data", "ndvi_c2_2016_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

ndvi_2000 <- rast(here("data", "ndvi_c2_2000_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

# PM2.5 data
# download data from URL
download_data <- function() {
  # set longer timeout
  options(timeout = 7200)
  if(!file.exists(here("data", "2016_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_pm25_daily_average.txt.gz", destfile = here("data", "2016_pm25_daily_average.txt.gz"))} #2016
  if(!file.exists(here("data", "2002_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_pm25_daily_average.txt.gz", destfile = here("data", "2002_pm25_daily_average.txt.gz"))}} #2002
download_data()

# read in PM2.5 daily average by census tract from EPA FAQSD - 2002
pm_tract_2002 <- read.delim(gzfile(here("data", "2002_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## Include only NYC tracts
  filter(FIPS %in% nyc_fips) %>%
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## Calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(pm_mean_2002 = mean(pm25_daily_average.ug.m3.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# read in PM2.5 daily average by census tract from EPA FAQSD - 2016
pm_tract_2016 <- read.delim(gzfile(here("data", "2016_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(pm_mean_2016 = mean(Prediction), 
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))
  

# combine 2002 and 2016 pm data for nyc_tracts
pm_tract <- pm_tract_2002 %>%
  inner_join(pm_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# NO2 CACES
no2_2016_caces <- read.csv(here("data", "2016_no2_caces.csv")) %>% 
  mutate(no2_2016 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_2000_caces <- read.csv(here("data", "2000_no2_caces.csv")) %>% 
  mutate(no2_2000 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_caces <- left_join(no2_2000_caces, no2_2016_caces, by = "fips") %>% 
  select(no2_2000, no2_2016, fips)

#O3 data download
options(timeout = 7200)
if(!file.exists(here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))} #2016
if(!file.exists(here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))} #2002

# read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2002
o3_tract_2002 <- read.delim(gzfile(here("data", "2002_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>% 
  ## include only NYC tracts
  filter(FIPS %in% nyc_fips) %>% 
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(o3_mean_2002 = mean(ozone_daily_8hour_maximum.ppb.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2016
o3_tract_2016 <- read.delim(gzfile(here("data", "2016_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>% 
  ## calculate annual average by tract using  population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(o3_mean_2016 = mean(Prediction),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))


o3_tract <- o3_tract_2002 %>%
  left_join(o3_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# merge all exposure data
all_data <- nyc_tracts %>% 
  # join gentrification index by tract
  right_join(gentrification, by = c("GEOID10" = "tractid")) %>%
  left_join(pm_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(pm_2016 = pm_mean_2016, pm_2002 = pm_mean_2002, pm_change = pm_2016-pm_2002) %>% 
  left_join(o3_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(o3_2016 = o3_mean_2016, o3_2002 = o3_mean_2002, o3_change = o3_2016-o3_2002) %>% 
  left_join(no2_caces, by = c("GEOID" = "fips")) %>% 
  mutate(no2_change = no2_2016-no2_2000) %>% 
  mutate(ndvi_2016 = ndvi_2016, ndvi_2000 = ndvi_2000, ndvi_change = ndvi_2016-ndvi_2000) %>% 
  left_join(temp, by = "GEOID")

# models adjusting for median income in 2016
income <- get_acs(variables = "B06011_001", year = 2016, state = "NY", 
                  county = c("Queens", "King", "New York", "Richmond", "Bronx"), geography = "tract") %>% 
  rename(med_income_2016 = estimate)

all_data_inc <- all_data %>% 
  left_join(income, by = "GEOID") %>% 
  # coerce GEOID to numeric then factor class to avoid error in MRF smooth
  mutate(GEOID = as.numeric(GEOID)) %>% 
  mutate(GEOID = factor(GEOID))

############# models adjusting for median income in 2016 and distance to downtown #########
# load nyc roads
downtown_st <- roads(state = "NY", county = "New York", year = 2011)
ggplot() + geom_sf(data = downtown_st)

# Define bounding box
bbox_coords <- matrix(c(
  -74.0187, 40.70158,  # Southwest
  -74.0187, 40.77279,  # Northwest
  -73.95953, 40.77279, # Northeast
  -73.95953, 40.70158, # Southeast
  -74.0187, 40.70158   # Closing the polygon
), ncol = 2, byrow = TRUE)

# convert bounding box to sf
bbox_polygon <- st_sfc(st_polygon(list(bbox_coords)), crs = 4326)

# Ensure roads are the same CRS and filter for streets within bbox
downtown_st <- st_transform(downtown_st, crs = st_crs(bbox_polygon))
filtered_streets <- downtown_st[st_within(downtown_st, bbox_polygon, sparse = FALSE), ]

# Plot filtered downtown streets and bounding box
ggplot() +
  geom_sf(data = filtered_streets) +
  geom_sf(data = bbox_polygon, fill = NA, color = 'red') +
  ggtitle("Downtown Streets within Bounding Box")

# Ensure census tracts are in the same CRS
nyc_tracts <- st_transform(nyc_tracts, crs = st_crs(bbox_polygon))

# Filter intersecting tracts within downtown
intersecting_tracts <- st_intersection(nyc_tracts, bbox_polygon)
combined_area <- st_union(intersecting_tracts)
#get downtown centroid
area_centroid <- st_centroid(combined_area)

ggplot() +
  geom_sf(data = intersecting_tracts, fill = 'lightblue', color = 'black') +
  geom_sf(data = area_centroid, color = 'red', size = 3) +
  ggtitle('Centroid of Downtown Census Tracts')

# Calculate distances and add to "all_data_inc" dataframe
distances <- st_distance(nyc_tracts, area_centroid)
all_data_inc$dt_dist <- as.numeric(distances)
```

Create model functions
```{r}
# Create exposure list
exposure_list <- list(
  pm = list(
    name = "pm",
    baseline = "pm_2002",
    change = "pm_change"
  ),
  o3 = list(
    name = "o3",
    baseline = "o3_2002",
    change = "o3_change"
  ),
  no2 = list(
    name = "no2",
    baseline = "no2_2000",
    change = "no2_change"
  ),
  ndvi = list(
    name = "ndvi",
    baseline = "ndvi_2000",
    change = "ndvi_change"
  ),
  temp = list(
    name = "mean_temp",
    baseline = NULL,
    change = "mean_temp"  # No change variable for temp
  )
)

# Function to run GAM models
run_gam_models <- function(data, exposure_info) {
  # Extract exposure information
  exp_var <- exposure_info$change
  baseline_exposure <- exposure_info$baseline
  
  # Skip if data is missing the exposure variable - gives warning 
  if (!exp_var %in% names(data)) {
    warning(paste("Variable", exp_var, "not found in dataset"))
    return(NULL)
  }
  
  # Base components that will be in all models: gentrification score and county/borough
  base_formula <- paste(exp_var, "~ s(score_0.5) + s(COUNTYFP10, bs='re')")
  
  # Create model formulas
  formulas <- list(
    # Base models (score + county + spatial) for tensor product and mrf smooth
    tp_base = paste(base_formula, "+ te(long, lat, d = 2, bs = 'gp', m = 2)"),
    mrf_base = paste(base_formula, "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),
    
    # Models with income
    tp_income = paste(base_formula, "+ s(med_income_2016) + te(long, lat, d = 2, bs = 'gp', m = 2)"),
    mrf_income = paste(base_formula, "+ s(med_income_2016) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),
    
    # Models with distance to downtown
    tp_dist = paste(base_formula, "+ s(dt_dist) + te(long, lat, d = 2, bs = 'gp', m = 2)"),
    mrf_dist = paste(base_formula, "+ s(dt_dist) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  )
  
  # Add baseline models if baseline exposure level exists (all except temp)
  if (!is.null(baseline_exposure) && baseline_exposure %in% names(data)) {
    formulas$tp_baseline <- paste(base_formula, "+", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_baseline <- paste(base_formula, "+", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    # Combinations with baseline
    formulas$tp_income_baseline <- paste(base_formula, "+ s(med_income_2016) +", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_income_baseline <- paste(base_formula, "+ s(med_income_2016) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    formulas$tp_dist_baseline <- paste(base_formula, "+ s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_dist_baseline <- paste(base_formula, "+ s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    formulas$tp_full <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_full <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  }
  
  # Add income + distance combinations
  formulas$tp_income_dist <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) + te(long, lat, d = 2, bs = 'gp', m = 2)")
  formulas$mrf_income_dist <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  
  # Convert formulas to formula objects
  formulas <- lapply(formulas, as.formula)
  
  # Run models
  models <- purrr::map(formulas, ~try(gam(., 
                                        family = scat, 
                                        data = data, 
                                        method = "REML"), 
                                    silent = TRUE))
  
  return(models)
}

# drop geometry from spatial data while keeping coordinates
data_prepared <- all_data_inc %>%
  mutate(
    COUNTYFP10 = as.factor(COUNTYFP10),
    GEOID = as.factor(GEOID)
  ) %>%
  st_drop_geometry()

# Construct neighbors list for MRF models
nb <- poly2nb(all_data_inc, row.names = all_data_inc$GEOID)
names(nb) <- attr(nb, "region.id") 

# JH 10/25/24 Warning: some observations have no neighbours; if this seems unexpected, try increasing the snap argument.Warning: neighbour object has 4 sub-graphs; if this sub-graph count seems unexpected, try increasing the snap argument.
```
Checking neighbor relationships for MRF models
```{r}
# Visualize the sub-graphs
plot(all_data_inc$geometry)
colors <- rainbow(length(nb))
plot(nb, st_coordinates(st_centroid(all_data_inc)), col=colors, add=TRUE)
# the sub-graphs match the islands.
# since we are adjusting by borough in the models, the 4 sub-graphs might be okay?? JH 10/26/24
```

Run models and check models
```{r}
# Run models for each exposure
all_models <- purrr::map(exposure_list, ~run_gam_models(data_prepared, .))

# Function to check which models succeeded/failed
check_models <- function(models) {
  model_status <- map(models, ~!inherits(., "try-error"))
  successful_models <- names(models)[unlist(model_status)]
  failed_models <- names(models)[!unlist(model_status)]
  
  return(list(
    successful = successful_models,
    failed = failed_models
  ))
}

# Check status of all models
model_status <- map(all_models, check_models)
model_status
```

Create basic QQ plots for models
```{r}
# Function to create QQ plots for a specific exposure
create_exposure_qq_plots <- function(models_list, exposure_name) {
  exposure_models <- models_list[[exposure_name]]
  
  plots <- names(exposure_models) %>%
    purrr::map(function(model_name) {
      model <- exposure_models[[model_name]]
      
      # Skip if model is an error
      if (inherits(model, "try-error")) {
        return(NULL)
      }
      
      # Create QQ plot
      resid_df <- data.frame(
        residuals = residuals(model, type = "deviance"),
        fitted = fitted(model)
      )
      
      ggplot(resid_df, aes(sample = residuals)) +
        stat_qq() +
        stat_qq_line(color = "red") +
        labs(
          title = paste0(exposure_name, "\n", model_name),
          x = "Theoretical Quantiles",
          y = "Sample Quantiles"
        ) +
        theme_minimal() +
        theme(
          plot.title = element_text(size = 10, hjust = 0.5),
          axis.title = element_text(size = 8),
          axis.text = element_text(size = 7)
        )
    }) %>%
    purrr::compact() # Remove NULL values from the list
  
  # Calculate grid layout
  n_plots <- length(plots)
  n_cols <- min(3, n_plots)
  n_rows <- ceiling(n_plots / n_cols)
  
  # Combine plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = plots,
    ncol = n_cols,
    nrow = n_rows,
    top = grid::textGrob(paste("QQ Plots for", exposure_name), 
                        gp = grid::gpar(fontsize = 12))
  )
  
  return(combined_plot)
}

# Function to create QQ plots for a specific model type across exposures
create_model_type_qq_plots <- function(models_list, model_type) {
  # Convert the for loop to purrr::map
  plots <- names(models_list) %>%
    purrr::map(function(exposure_name) {
      exposure_models <- models_list[[exposure_name]]
      
      # Check if the model type exists for this exposure
      if (!model_type %in% names(exposure_models)) {
        return(NULL)
      }
      
      model <- exposure_models[[model_type]]
      
      # Skip if model is an error
      if (inherits(model, "try-error")) {
        return(NULL)
      }
      
      # Create QQ plot
      resid_df <- data.frame(
        residuals = residuals(model, type = "deviance"),
        fitted = fitted(model)
      )
      
      ggplot(resid_df, aes(sample = residuals)) +
        stat_qq() +
        stat_qq_line(color = "red") +
        labs(
          title = paste0(exposure_name, "\n", model_type),
          x = "Theoretical Quantiles",
          y = "Sample Quantiles"
        ) +
        theme_minimal() +
        theme(
          plot.title = element_text(size = 10, hjust = 0.5),
          axis.title = element_text(size = 8),
          axis.text = element_text(size = 7)
        )
    }) %>%
    purrr::compact() # Remove NULL values from the list
  
  # Calculate grid layout
  n_plots <- length(plots)
  n_cols <- min(3, n_plots)
  n_rows <- ceiling(n_plots / n_cols)
  
  # Combine plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = plots,
    ncol = n_cols,
    nrow = n_rows,
    top = grid::textGrob(paste("QQ Plots for", model_type), 
                        gp = grid::gpar(fontsize = 12))
  )
  
  return(combined_plot)
}

# Function to list available model types
list_model_types <- function(models_list) {
  unique(unlist(lapply(models_list, names)))
}

# Function to list available exposures
list_exposures <- function(models_list) {
  names(models_list)
}

```


Display basic QQ plots for models
```{r}
# List all model types
model_types <- list_model_types(all_models)

# Display all QQ plots
purrr::walk(model_types, ~create_model_type_qq_plots(all_models, .))

```

Create functions for QQ plots with confidence intervals
```{r}
# Function to create a single QQ plot with confidence bands
create_single_qq_plot <- function(model, title) {
  if (inherits(model, "try-error")) return(NULL)
  
  data.frame(
    residuals = residuals(model, type = "deviance"),
    fitted = fitted(model)
  ) %>%
    ggplot(aes(sample = residuals)) +
    geom_qq_band(bandType = "ks", mapping = aes(fill = "KS"), alpha = 0.5) +
    geom_qq_band(bandType = "ts", mapping = aes(fill = "TS"), alpha = 0.5) +
    geom_qq_band(bandType = "pointwise", mapping = aes(fill = "Normal"), alpha = 0.5) +
    geom_qq_band(bandType = "boot", mapping = aes(fill = "Bootstrap"), alpha = 0.5) +  
    stat_qq_line(color = "red") +
    stat_qq_point(size = 1) +
    labs(
      title = title,
      x = "Theoretical Quantiles",
      y = "Sample Quantiles"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 10, hjust = 0.5),
      axis.title = element_text(size = 8),
      axis.text = element_text(size = 7)
    )
}

# Function to arrange plots in a grid
arrange_plots <- function(plots, main_title) {
  n_plots <- length(plots)
  n_cols <- min(3, n_plots)  
  n_rows <- ceiling(n_plots / n_cols)
  
  gridExtra::grid.arrange(
    grobs = plots,
    ncol = n_cols,
    nrow = n_rows,
    top = grid::textGrob(
      main_title,
      gp = grid::gpar(fontsize = 12)  
    )
  )
}

# Main display function
display_qq_plots_by_model_type <- function(models_list) {
  # Get all unique model types
  model_types <- models_list %>%
    map(names) %>%
    reduce(union)
  
  # Create plots for each model type
  walk(model_types, function(model_type) {
    # Extract all models of this type across exposures
    plots <- models_list %>%
      imap(function(exposure_models, exposure_name) {
        if (!model_type %in% names(exposure_models)) return(NULL)
        
        create_single_qq_plot(
          exposure_models[[model_type]], 
          title = paste0(exposure_name, "\n", model_type)
        )
      }) %>%
      compact()
    
    # Arrange and display plots for this model type
    arrange_plots(
      plots,
      main_title = paste("QQ Plots for", model_type)
    )
  })
}

```

Display QQ plots with confidence bands for models
```{r}
# Display QQ plots
display_qq_plots_by_model_type(all_models)
```


Create functions for comprehensive diagnostic plots (JH 10.26.24 - should convert this to purr map if keeping 11/2/24 - displaying tables instead of the plots)
```{r}
# Function to create comprehensive diagnostic plots for a specific model type
create_model_diagnostics_by_type <- function(models_list, model_type) {
  # List to store plots for each exposure
  diagnostic_plots <- list()
  
  for (exposure_name in names(models_list)) {
    exposure_models <- models_list[[exposure_name]]
    
    # Check if the model type exists for this exposure
    if (model_type %in% names(exposure_models)) {
      model <- exposure_models[[model_type]]
      
      # Skip if model is an error
      if (inherits(model, "try-error")) {
        next
      }
      
      # Extract residuals and fitted values
      resid_df <- data.frame(
        residuals = residuals(model, type = "deviance"),
        fitted = fitted(model),
        scale_residuals = residuals(model, type = "scaled.pearson"),
        response = model$y
      )
      
      # Create plot list for this exposure
      exposure_plots <- list()
      
      # 1. QQ plot
      exposure_plots[["qq"]] <- ggplot(resid_df, aes(sample = residuals)) +
        stat_qq() +
        stat_qq_line(color = "red") +
        labs(title = paste(exposure_name, "- Normal Q-Q Plot")) +
        theme_minimal() +
        theme(plot.title = element_text(size = 8))
      
      # 2. Residuals vs Fitted
      exposure_plots[["resid"]] <- ggplot(resid_df, aes(x = fitted, y = residuals)) +
        geom_point(alpha = 0.5) +
        geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
        geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
        labs(title = paste(exposure_name, "- Residuals vs Fitted"),
             x = "Fitted values",
             y = "Deviance residuals") +
        theme_minimal() +
        theme(plot.title = element_text(size = 8))
      
      # 3. Scale-location plot
      exposure_plots[["scale"]] <- ggplot(resid_df, 
                                        aes(x = fitted, y = sqrt(abs(scale_residuals)))) +
        geom_point(alpha = 0.5) +
        geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
        labs(title = paste(exposure_name, "- Scale-Location"),
             x = "Fitted values",
             y = "√|Standardized residuals|") +
        theme_minimal() +
        theme(plot.title = element_text(size = 8))
      
      # 4. Response vs Fitted
      exposure_plots[["resp"]] <- ggplot(resid_df, aes(x = fitted, y = response)) +
        geom_point(alpha = 0.5) +
        geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
        labs(title = paste(exposure_name, "- Response vs Fitted"),
             x = "Fitted values",
             y = "Response") +
        theme_minimal() +
        theme(plot.title = element_text(size = 8))
      
      # Store all plots for this exposure
      diagnostic_plots[[exposure_name]] <- exposure_plots
    }
  }
  
  # Calculate layout dimensions
  n_exposures <- length(diagnostic_plots)
  if (n_exposures == 0) {
    stop("No valid models found for the specified model type")
  }
  
  # Create a list of all plots in the desired order
  all_plots <- list()
  for (exposure_name in names(diagnostic_plots)) {
    all_plots <- c(all_plots, diagnostic_plots[[exposure_name]])
  }
  
  # Calculate grid dimensions based on number of exposures
  n_cols <- 4  # One column for each type of diagnostic plot
  n_rows <- n_exposures  # One row for each exposure
  
  # Combine all plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = all_plots,
    ncol = n_cols,
    nrow = n_rows,
    top = grid::textGrob(
      paste("Diagnostic Plots for Model Type:", model_type),
      gp = grid::gpar(fontsize = 12, fontface = "bold")
    )
  )
  
  return(combined_plot)
}

# Function to display diagnostics for all model types
display_all_model_diagnostics <- function(models_list) {
  # Get unique model types
  model_types <- unique(unlist(lapply(models_list, names)))
  
  # Create plots for each model type
  for (model_type in model_types) {
    tryCatch({
      plot <- create_model_diagnostics_by_type(models_list, model_type)
      print(plot)
      # Add a pause between plots if running interactively
      if (interactive()) {
        readline(prompt = paste("Press [Enter] to see next plot for model type:", 
                              model_type))
      }
    }, error = function(e) {
      message(paste("Error creating plots for model type:", model_type))
      message(e)
    })
  }
}
```

Display all diagnostic plots for models
```{r}
# List all model types
model_types <- list_model_types(all_models)

# Display all diagnostics 
display_all_model_diagnostics(all_models)

```

GET SUMMARY STATS PER MODEL TYPE 
```{r}
#Function to generate summary statistics by model type for every exposure
get_model_type_summary <- function(models_list, model_type) {
  summary_stats <- data.frame(
    exposure = character(),
    deviance_explained = numeric(),
    aic = numeric(),
    edf = numeric(),
    ref_df = numeric(),
    chi_sq = numeric(),
    p_value = numeric(),
    stringsAsFactors = FALSE
  )
  
  for (exposure_name in names(models_list)) {
    exposure_models <- models_list[[exposure_name]]
    
    if (model_type %in% names(exposure_models)) {
      model <- exposure_models[[model_type]]
      
      if (!inherits(model, "try-error")) {
        model_summary <- summary(model)
        
        # Get the smooth terms table
        smooth_table <- model_summary$s.table
        
        if (!is.null(smooth_table)) {
          summary_stats <- rbind(summary_stats, data.frame(
            exposure = exposure_name,
            deviance_explained = model_summary$dev.expl * 100,
            aic = AIC(model),
            edf = smooth_table[1, "edf"],
            ref_df = smooth_table[1, "Ref.df"],
            chi_sq = smooth_table[1, "Chi.sq"],  
            p_value = smooth_table[1, "p-value"],
            stringsAsFactors = FALSE
          ))
        }
      }
    }
  }
  
  return(summary_stats)
}

# Generate summary stats for mrf_income model for every exposure
summary_stats <- get_model_type_summary(all_models, "mrf_income")
print(summary_stats)

```

Create effects summary data frame and forest plot
```{r}
# Function to extract effect estimates for score_0.5
get_score_effects <- function(models_list, model_type = "mrf_income") {
  effects_list <- list()
  
  for (exposure_name in names(models_list)) {
    model <- models_list[[exposure_name]][[model_type]]
    
    if (!inherits(model, "try-error")) {
      # Get predicted values across range of scores
      score_seq <- seq(min(model$model$score_0.5), max(model$model$score_0.5), length.out = 100)
      pred_data <- data.frame(
        score_0.5 = score_seq,
        med_income_2016 = mean(model$model$med_income_2016),
        COUNTYFP10 = levels(model$model$COUNTYFP10)[1],
        GEOID = levels(model$model$GEOID)[1]
      )
      
      # Get predictions with standard errors
      preds <- predict(model, pred_data, type = "terms", se.fit = TRUE)
      
      # Extract effect for score_0.5 smooth term
      score_effect <- data.frame(
        exposure = exposure_name,
        estimate = diff(range(preds$fit[,"s(score_0.5)"])),
        se = diff(range(preds$se.fit[,"s(score_0.5)"])),
        p_value = summary(model)$s.table["s(score_0.5)", "p-value"]
      )
      
      effects_list[[exposure_name]] <- score_effect
    }
  }
  
  # Combine all effects into one dataframe
  effects_df <- do.call(rbind, effects_list)
  
  # Calculate confidence intervals
  effects_df$lower <- effects_df$estimate - 1.96 * effects_df$se
  effects_df$upper <- effects_df$estimate + 1.96 * effects_df$se
  
  return(effects_df)
}

# Get effects
effects_df <- get_score_effects(all_models)

# Create forest plot
library(ggplot2)

# Create exposure labels with significance indicators
effects_df$exposure_label <- paste0(
  toupper(effects_df$exposure), 
  ifelse(effects_df$p_value < 0.001, "***",
         ifelse(effects_df$p_value < 0.01, "**",
                ifelse(effects_df$p_value < 0.05, "*", "")))
)

# Create the forest plot
ggplot(effects_df, aes(y = exposure_label)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(x = estimate), size = 3) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  labs(
    title = "Estimated Effects of Gentrification Score on Environmental Exposures",
    subtitle = "MRF Income Model Results",
    x = "Effect Size (Range of Smooth Term)",
    y = "Exposure Type",
    caption = "* p < 0.05, ** p < 0.01, *** p < 0.001"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    plot.caption = element_text(size = 8, hjust = 0),
    panel.grid.minor = element_blank()
  )

# Print numerical results
print(effects_df[, c("exposure_label", "estimate", "se", "p_value", "lower", "upper")])
```


Scale mrf_income models by IQR and create scaled forest plot
```{r}
# Function to extract IQR-scaled effect estimates for score_0.5
get_score_effects_scaled <- function(models_list, model_type = "mrf_income") {
  effects_list <- list()
  
  for (exposure_name in names(models_list)) {
    model <- models_list[[exposure_name]][[model_type]]
    
    if (!inherits(model, "try-error")) {
      # Get model data
      model_data <- model$model
      
      # Calculate IQRs
      score_iqr <- IQR(model_data$score_0.5)
      exposure_var <- exposure_list[[exposure_name]]$change
      exposure_iqr <- IQR(model_data[[exposure_var]])
      
      # Get predicted values across range of scores
      score_seq <- seq(min(model_data$score_0.5), max(model_data$score_0.5), length.out = 100)
      pred_data <- data.frame(
        score_0.5 = score_seq,
        med_income_2016 = mean(model_data$med_income_2016),
        COUNTYFP10 = levels(model_data$COUNTYFP10)[1],
        GEOID = levels(model_data$GEOID)[1]
      )
      
      # Get predictions with standard errors
      preds <- predict(model, pred_data, type = "terms", se.fit = TRUE)
      
      # Extract and scale effect for score_0.5 smooth term
      raw_effect <- diff(range(preds$fit[,"s(score_0.5)"]))
      raw_se <- diff(range(preds$se.fit[,"s(score_0.5)"]))
      
      # Scale effects by IQRs
      scaled_effect <- (raw_effect * score_iqr) / exposure_iqr
      scaled_se <- (raw_se * score_iqr) / exposure_iqr
      
      score_effect <- data.frame(
        exposure = exposure_name,
        estimate = scaled_effect,
        se = scaled_se,
        p_value = summary(model)$s.table["s(score_0.5)", "p-value"],
        exposure_iqr = exposure_iqr,
        score_iqr = score_iqr
      )
      
      effects_list[[exposure_name]] <- score_effect
    }
  }
  
  # Combine all effects into one dataframe
  effects_df <- do.call(rbind, effects_list)
  
  # Calculate confidence intervals
  effects_df$lower <- effects_df$estimate - 1.96 * effects_df$se
  effects_df$upper <- effects_df$estimate + 1.96 * effects_df$se
  
  return(effects_df)
}

# Get scaled effects
effects_df_scaled <- get_score_effects_scaled(all_models)

# Create forest plot with scaled effects
effects_df_scaled$exposure_label <- paste0(
  toupper(effects_df_scaled$exposure), 
  ifelse(effects_df_scaled$p_value < 0.001, "***",
         ifelse(effects_df_scaled$p_value < 0.01, "**",
                ifelse(effects_df_scaled$p_value < 0.05, "*", "")))
)

ggplot(effects_df_scaled, aes(y = exposure_label)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(x = estimate), size = 3) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  labs(
    title = "IQR-Scaled Effects of Gentrification Score on Environmental Exposures",
    subtitle = "MRF Income Model Results",
    x = "Effect Size (Change in Exposure IQR per Score IQR)",
    y = "Exposure Type",
    caption = "* p < 0.05, ** p < 0.01, *** p < 0.001"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    plot.caption = element_text(size = 8, hjust = 0),
    panel.grid.minor = element_blank()
  )

```

Understanding direction and magnitude of exposure changes 
```{r}
# Function to analyze changes in exposures
analyze_exposure_changes <- function(data, exposure_list) {
  results <- list()
  
  for (exp_name in names(exposure_list)) {
    exp_info <- exposure_list[[exp_name]]
    change_var <- exp_info$change
    
    # Skip temperature which doesn't have baseline/change structure
    if (exp_name != "temp") {
      baseline_var <- exp_info$baseline
      
      # Calculate summary statistics
      change_stats <- data %>%
        summarise(
          n = nrow(data),
          mean_change = mean(.data[[change_var]], na.rm = TRUE),
          median_change = median(.data[[change_var]], na.rm = TRUE),
          sd_change = sd(.data[[change_var]], na.rm = TRUE),
          pct_positive = mean(.data[[change_var]] > 0, na.rm = TRUE) * 100,
          q25 = quantile(.data[[change_var]], 0.25, na.rm = TRUE),
          q75 = quantile(.data[[change_var]], 0.75, na.rm = TRUE)
        )
      
      # Create histogram
      hist_plot <- ggplot(data, aes(x = .data[[change_var]])) +
        geom_histogram(bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
        geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
        labs(
          title = paste("Distribution of Changes in", toupper(exp_name)),
          x = "Change",
          y = "Count"
        ) +
        theme_minimal()
      
      # Store results
      results[[exp_name]] <- list(
        stats = change_stats,
        plot = hist_plot
      )
    }
  }
  
  return(results)
}

# Run the analysis
change_analysis <- analyze_exposure_changes(data_prepared, exposure_list)

# Print summary statistics 
print(change_analysis)

# Create a combined plot of all histograms
plot_list <- lapply(names(change_analysis), function(exp_name) {
  change_analysis[[exp_name]]$plot
})

# Arrange all plots in a grid
library(gridExtra)
grid.arrange(grobs = plot_list, ncol = 2)

# Create boxplots for another view of the distributions
combined_changes <- data.frame(
  Exposure = character(),
  Change = numeric()
)

for (exp_name in names(exposure_list)) {
  if (exp_name != "temp") {
    change_var <- exposure_list[[exp_name]]$change
    tmp_data <- data.frame(
      Exposure = toupper(exp_name),
      Change = data_prepared[[change_var]]
    )
    combined_changes <- rbind(combined_changes, tmp_data)
  }
}

# Create boxplot
ggplot(combined_changes, aes(x = Exposure, y = Change)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of Changes Across All Exposures",
    x = "Exposure Type",
    y = "Change"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
