---
title: "gent_clean_1.21.25"
author: "Joyce Hu"
date: "2025-1-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages (JH - check if any are unused after code cleaned)
```{r}
library(tidyverse)
library(tigris)
library(arrow)
library(here)
library(mgcv)
library(fixest)
library(lmtest)
library(terra)
library(exactextractr)
library(tidyterra)
library(sf)
library(mgcViz)

library(DHARMa)
library(qqplotr)
library(readxl) # for read_excel()
library(spdep) # for moran.test()
library(tidycensus) # for median income
library(distr) # for qqbounds()
library(spatialreg) # for spatial lag and error models
library(gridExtra) # for viewing plots side-by-side
library(qgam) 
library(dplyr)
library(INLA)
library(ggregplot) #INLA compatible plotting library (efxplot)
library(ggfields) # for plotting spatial field
library(lattice) # for levelplot
library(inlatools) # for INLA diagnostics 
```

Load and clean data
```{r}
# Check if 'data' directory exists; if not, create it
if (!dir.exists(here("data"))) {
  dir.create(here("data"))
}

# Load gentrification score by tract
gentrification <- read.csv(here("data", "Data_NYC_Gentrification_2000_16.csv")) %>% 
  mutate(tractid = as.character(tractid)) %>% 
  dplyr::select(tractid, score_0.5)

# Load gentrification score if using xlsx file:
# gentrification <- read_xlsx(here("Gentrification","Data_NYC_Gentrification_2000_16.xlsx")) %>%
#   mutate(tractid = as.character(tractid)) %>%
#   dplyr::select(tractid, score_0.5)

# pull NYC tract IDs/fips from gentrification index dataset for filtering
nyc_fips <- gentrification$tractid

# Load NYC tracts shapefile
nyc_tracts <- tracts(state = "NY", county = c("Bronx", "Kings", "New York", "Queens", "Richmond"), year = 2010) %>%
  # clean lat and long for Conley SEs
  mutate(lat = as.numeric(INTPTLAT10), long = as.numeric(INTPTLON10), GEOID = GEOID10) %>% 
  # filter for FIPS codes included in gentrification index dataset
  filter(GEOID %in% nyc_fips)


# Load NY tracts populatuion weighted centroid points from Census data
nyc_weighted_centroids <- read.table("https://www2.census.gov/geo/docs/reference/cenpop2010/tract/CenPop2010_Mean_TR36.txt", 
                         header = TRUE, sep = ",", stringsAsFactors = FALSE, 
                         colClasses = c("character", "character", "character", "numeric", "numeric", "numeric")) %>%
  mutate(FIPS = paste(STATEFP, COUNTYFP, TRACTCE, sep = "")) %>% 
  ## filter for only NYC tracts
  filter(FIPS %in% nyc_fips) %>%  
  rename(Longitude_weighted = LONGITUDE, 
         Latitude_weighted = LATITUDE)

# JH - 10/12/24 warning persists 10/13, 10/14 Warning: R metadata may have unsafe or invalid elements ℹ Type: "externalptr" options(arrow.unsafe_metadata = TRUE) #

# Temperature data
if(!file.exists(here("data","summarized_daily_temp_preds.parquet"))){download.file("https://zenodo.org/records/10557980/files/summarized_daily_temp_preds.parquet?download=1", 
                                                                                   destfile = here("data","summarized_daily_temp_preds.parquet"))}

# Temperature by tract
temp <- read_parquet(here("data","summarized_daily_temp_preds.parquet")) %>%
  filter(str_detect(date, "2016")) %>%
  mutate(temp_c = mean_temp_daily_cK/100-273.15) %>% 
  mutate(temp_f = temp_c*9/5+32) %>% 
  group_by(GEOID, date) %>%
  ungroup() %>% 
  group_by(GEOID) %>% 
  summarize(mean_temp = mean(mean_temp_daily_cK/100))

# NDVI by tract (need to have it downloaded and stored locally)
ndvi_2016 <- rast(here("data", "ndvi_c2_2016_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

ndvi_2000 <- rast(here("data", "ndvi_c2_2000_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

# PM2.5 data
# Download data from URL
download_data <- function() {
  # set longer timeout
  options(timeout = 7200)
  if(!file.exists(here("data", "2016_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_pm25_daily_average.txt.gz", destfile = here("data", "2016_pm25_daily_average.txt.gz"))} #2016
  if(!file.exists(here("data", "2002_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_pm25_daily_average.txt.gz", destfile = here("data", "2002_pm25_daily_average.txt.gz"))}} #2002
download_data()

# Read in PM2.5 daily average by census tract from EPA FAQSD - 2002
pm_tract_2002 <- read.delim(gzfile(here("data", "2002_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## Include only NYC tracts
  filter(FIPS %in% nyc_fips) %>%
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## Calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(pm_mean_2002 = mean(pm25_daily_average.ug.m3.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# Read in PM2.5 daily average by census tract from EPA FAQSD - 2016
pm_tract_2016 <- read.delim(gzfile(here("data", "2016_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(pm_mean_2016 = mean(Prediction), 
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))
  
# Combine 2002 and 2016 pm data for nyc_tracts
pm_tract <- pm_tract_2002 %>%
  inner_join(pm_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# NO2 CACES
no2_2016_caces <- read.csv(here("data", "2016_no2_caces.csv")) %>% 
  mutate(no2_2016 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_2000_caces <- read.csv(here("data", "2000_no2_caces.csv")) %>% 
  mutate(no2_2000 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_caces <- left_join(no2_2000_caces, no2_2016_caces, by = "fips") %>% 
  select(no2_2000, no2_2016, fips)

#O3 data download
options(timeout = 7200)
if(!file.exists(here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))} #2016
if(!file.exists(here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))} #2002

# Read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2002
o3_tract_2002 <- read.delim(gzfile(here("data", "2002_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>% 
  ## include only NYC tracts
  filter(FIPS %in% nyc_fips) %>% 
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(o3_mean_2002 = mean(ozone_daily_8hour_maximum.ppb.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# Read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2016
o3_tract_2016 <- read.delim(gzfile(here("data", "2016_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>% 
  ## calculate annual average by tract using  population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(o3_mean_2016 = mean(Prediction),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

o3_tract <- o3_tract_2002 %>%
  left_join(o3_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# Merge all exposure data
all_data <- nyc_tracts %>% 
  # join gentrification index by tract
  right_join(gentrification, by = c("GEOID10" = "tractid")) %>%
  left_join(pm_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(pm_2016 = pm_mean_2016, pm_2002 = pm_mean_2002, pm_change = pm_2016-pm_2002) %>% 
  left_join(o3_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(o3_2016 = o3_mean_2016, o3_2002 = o3_mean_2002, o3_change = o3_2016-o3_2002) %>% 
  left_join(no2_caces, by = c("GEOID" = "fips")) %>% 
  mutate(no2_change = no2_2016-no2_2000) %>% 
  mutate(ndvi_2016 = ndvi_2016, ndvi_2000 = ndvi_2000, ndvi_change = ndvi_2016-ndvi_2000) %>% 
  left_join(temp, by = "GEOID")

############ Median income ###############################
# Find median household income variable in 2000 decennial census
all_vars_dec <- 
  load_variables(year = 2000, dataset = "sf3")

# Models adjusting for median income in 2000 (note: 2000 decennial census is 1999 data)
income_2000 <- get_decennial(variables = "P053001", year = 2000,  sumfile = "sf3", state = "NY", 
                  county = c("Queens", "King", "New York", "Richmond", "Bronx"), geography = "tract") %>% 
  rename(med_income_2000 = value)
#results in 226 missing tracts in terms of 2010 tracts.

# Find median household income variable in 2010 ACS
income_2010 <- get_acs(variables = "B06011_001", year = 2010, state = "NY", 
                  county = c("Queens", "King", "New York", "Richmond", "Bronx"), geography = "tract") %>% 
  rename(med_income_2010 = estimate)

# Joining both income sets with data and imputing 2010 data where 2000 data is missing
all_data_inc <- all_data %>%
  left_join(income_2000, by = "GEOID") %>%
  left_join(income_2010, by = "GEOID") %>%
  mutate(med_income_2000_impute = coalesce(med_income_2000, med_income_2010)) %>% 
  # coerce GEOID to numeric then factor class to avoid error in MRF smooth
  mutate(GEOID = as.numeric(GEOID)) %>% 
  mutate(GEOID = factor(GEOID)) %>% 
 # coerce COUNTYFP10 to numeric then factor class to avoid error ?
  mutate(COUNTYFP10 = as.numeric(COUNTYFP10)) %>% 
  mutate(COUNTYFP10 = factor(COUNTYFP10))

# Ranking median income for use in GAM models
all_data_inc$med_income_ranked <- rank(all_data_inc$med_income_2000_impute)


# Converting median income to quantiles and scaling
# Create deciles
income_quantiles <- quantile(all_data_inc$med_income_2000_impute, probs = seq(0, 1, by = 0.1))
all_data_inc$med_income_decile <- findInterval(all_data_inc$med_income_2000_impute, 
                                             income_quantiles,
                                             rightmost.closed = TRUE)
```

INLA set up
```{r}
# # Set INLA repository
# options(repos = c(getOption("repos"),
#    INLA="https://inla.r-inla-download.org/R/stable"))
# 
# # Install INLA and dependencies (from CRAN)
# install.packages("INLA", dep = TRUE)

#need older version of INLA to be compatible with R version
 # remotes::install_version("INLA", version="22.05.07",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/testing"), dep=TRUE)

# Install an even older version of INLA for GLIBC compatibility

# remotes::install_version("INLA", 
#                          version="21.02.23", 
#                          repos=c(getOption("repos"),
#                                  INLA="https://inla.r-inla-download.org/R/stable"), 
#                          dep=TRUE)

```

Neighbor relationships for MRF smooth model (to facilitate running MRF models for setting priors)
```{r}
# Establish neighbor relationships for MRF model
nb <- poly2nb(all_data_inc, row.names = all_data_inc$GEOID)
names(nb) <- attr(nb, "region.id") 

# To facilitate MRF smooth model for NDVI
# Create clean dataset without 8 NA values (follow up)
all_data_inc_clean <- all_data_inc[is.finite(all_data_inc$ndvi_change), ]

# Create new neighborhood structure from cleaned data
nb_clean <- poly2nb(all_data_inc_clean)
names(nb_clean) <- all_data_inc_clean$GEOID

```

INLA - Mesh creation
```{r}
graphics.off()
#### Create mesh using tract centroids as locations

# Create matrix of locations
Locations <- as.matrix(cbind(
  as.numeric(nyc_weighted_centroids$Longitude_weighted),
  as.numeric(nyc_weighted_centroids$Latitude_weighted)
))

# Create mesh options
# Various mesh option specifications 
MeshA <- inla.mesh.2d(jitter(Locations), max.edge = c(20, 40))
MeshB <- inla.mesh.2d(Locations, max.edge = c(20, 40))
MeshC <- inla.mesh.2d(Locations, max.edge = c(50, 100))
mesh_simple <- inla.mesh.create(Locations)

# Visualize mesh options
plot(MeshA)
points(Locations, col = "red", pch = 2)

plot(MeshB)
points(Locations, col = "red", pch = 2)

plot(MeshC)
points(Locations, col = "red", pch = 2)


# "Generally, the mesh function automatically creates a mesh like mesh A, where closer-together sampling locations produce smaller triangles. When exploring/setting up preliminary analyses, use a mesh like mesh B. For analyses to be reported in a paper, use a mesh like mesh C. Be careful of edges, and try to allow some space around your sampling area for INLA to estimate. The edge triangles can be bigger to reduce computing power."

# Set chosen mesh to use for analysis
Mesh <- MeshB
```

INLA - Creating the A Matrix and SPDE
```{r}
# Create A matrix
A <- inla.spde.make.A(Mesh, loc = Locations) 
# "the A matrix maps the Gaussian Markov Random Field (GMRF) from the mesh nodes to the n observation location. This is represented by a matrix with the number of observations as column and the number of nodes as rows."


# Create spatial structure (SPDE object)
spde <- inla.spde2.matern(Mesh, alpha=2) #they fix alpha as 2

# Create all the required indexes for the SPDE model, naming the spatial field as “spatial.field”, which facilitates the extraction of the estimated values of the spatial field after the fitting process.
iset <- inla.spde.make.index(name = "spatial.field", spde$n.spde)
```
Function for setting priors from GAM models for later use
```{r}
# Function to extract priors from GAM 
simple_extract <- function(gam_model) {
  coef_summary <- summary(gam_model)$p.table
  list(
    mean = setNames(coef_summary[,"Estimate"], rownames(coef_summary)),
    prec = setNames(1/coef_summary[,"Std. Error"]^2, rownames(coef_summary))
  )
}
```


INLA stack (PM2.5)
```{r}
# Create stack
stk_pm <- inla.stack(
  data=list(y=all_data_inc$pm_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10)))


#"Two projector matrices are needed A = list(A,1): the projector matrix for the latent field and a matrix that is a one-to-one map of the covariate. The latter matrix can simply be a constant rather than a diagonal matrix."
```


INLA - model fitting (no priors) for PM2.5
```{r}
# Specify formula
formula1_pm <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + #not sure if this needs to have a particular specification?
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 

# Specify model
model1_pm<-inla(formula1_pm, 
             data = inla.stack.data(stk_pm, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_pm),compute = TRUE), 
             control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE))

```

INLA model results for PM2.5
```{r}
###  Model results

# Posterior distribution summary of the fixed effects
model1_pm$summary.fix


# Posterior distribution summary of the hyper parameters (spatial field parameters)
model1_pm$summary.hyperpar

# Rounded summaries
round(model1_pm$summary.fixed,3) 
round(model1_pm$summary.hyperpar,3)

# "The posterior distribution of the hyperparameters θ1 and θ2 are difficult to interpret. We can transform them into more interpretable quantities such as the range, and the variance. The range is of particular interest. It provides the distance value (in the unit of the point coordinates) above which spatial dependencies become negligible."
model1_pm.res<-inla.spde2.result(model1_pm, 'spatial.field', spde, do.transf=TRUE) #do.tranf = TRUE makes sure that marginals are calculated in the same scale as the data.

model1_pm.res$summary.log.range.nominal
model1_pm.res$summary.log.variance.nominal

# Plot effect estimate and credible intervals
graphics.off()
Efxplot(list(model1_pm))

# Plot for fixed parameters for insight into the effect of covariates
par(mfrow=c(2,2))
plot(model1_pm$marginals.fixed[[1]], ty = "l", xlab = expression(beta[0]), ylab = "Density")
plot(model1_pm$marginals.fixed[[2]], ty = "l", xlab = expression(beta[score_0.5]), ylab = "Density") 
plot(model1_pm$marginals.fixed[[3]], ty = "l", xlab = expression(beta[med_income_decile]), ylab = "Density") 

# Plots for hyper parameters for insight into the effect of the structure of the spatial field
par(mfrow=c(1,3))
plot(model1_pm.res$marginals.var[[1]], ty = "l", xlab = expression(sigma[randomfield]^2), ylab = "Density") 
plot(model1_pm.res$marginals.kap[[1]], type = "l", xlab = expression(kappa), ylab = "Density")
plot(model1_pm.res$marginals.range[[1]], type = "l", xlab = "range nominal", ylab = "Density")

graphics.off()

# Plot spatial field 
gproj <- inla.mesh.projector(Mesh,  dims = c(300, 300))
g.mean <- inla.mesh.project(gproj, model1_pm$summary.random$spatial.field$mean)
g.sd <- inla.mesh.project(gproj, model1_pm$summary.random$spatial.field$sd)

grid.arrange(levelplot(g.mean, scales=list(draw=F), xlab='', ylab='', main='mean',col.regions = heat.colors(16)),
             levelplot(g.sd, scal=list(draw=F), xla='', yla='', main='sd' ,col.regions = heat.colors(16)),
             nrow=1)
# "take a look at the spatial field to understand the spatial structure. It can inform on potential omitted variable - how much of the variance in the dependent variable is not explained by the covariates." 

# Other model plots
plot(model1_pm, single = TRUE) 

# Model selection (not relevant right now with just one model)
model1_pm$dic  # Deviance Information Criterion
model1_pm$waic # Watanabe-Akaike Information Criterion

# Conditional Predictive Ordinate (CPO) diagnostics
summary(model1_pm$cpo)  # Predictive performance 
1 - mean(model1_pm$cpo$failure)  # Model fit success rate


# Posterior predictive checks
# posterior_samples <- inla.posterior.sample(n = 1000, result = model1_pm)

# Residual Analysis
par(mfrow=c(2,2))
plot(model1_pm$summary.random$spatial.field$mean, type="l")
qqnorm(model1_pm$summary.random$spatial.field$mean)
hist(model1_pm$summary.random$spatial.field$mean)

# QQ Plot to check normality
residuals <- model1_pm$summary.random$spatial.field$mean
qqnorm(residuals)

# Shapiro-Wilk Test for Normality
shapiro.test(residuals)

# Kolmogorov-Smirnov test against normal distribution
ks.test(residuals, "pnorm", mean=mean(residuals), sd=sd(residuals))

# Check for outliers
plot(model1_pm$summary.random$spatial.field$mean)
boxplot(model1_pm$summary.random$spatial.field$mean)

```
Additional diagnostics? 
```{r}
# inlatools setting a prior https://inlatools.netlify.app/articles/prior - not sure how to apply this
str(x <- simulate_iid(sigma = 1))
plot(x)
plot(x, link = "log")
plot(x, link = "logit")

plot(model1_pm)

# https://www.paulamoraga.com/book-geospatial/sec-inla.html#control-variables-to-compute-approximations
alpha <- model1_pm$marginals.fixed[[1]]
ggplot(data.frame(inla.smarginal(alpha)), aes(x, y)) +
  geom_line() +
  theme_bw()

quant <- inla.qmarginal(0.05, alpha)
quant

inla.pmarginal(quant, alpha)

ggplot(data.frame(inla.smarginal(alpha)), aes(x, y)) +
  geom_line() +
  geom_area(data = subset(data.frame(inla.smarginal(alpha)),
                                     x < quant),
            fill = "black") +
  theme_bw()

inla.dmarginal(-6.3, alpha)

ggplot(data.frame(inla.smarginal(alpha)), aes(x, y)) +
  geom_line() +
  geom_vline(xintercept = -6.3, linetype = "dashed") +
  theme_bw()


marg.variance <- inla.tmarginal(function(x) 1/x, 
model1_pm$marginals.hyperpar[[1]])

ggplot(data.frame(inla.smarginal(marg.variance)), aes(x, y)) +
  geom_line() +
  theme_bw()

m <- inla.emarginal(function(x) x, marg.variance)
m

mm <- inla.emarginal(function(x) x^2, marg.variance)
sqrt(mm - m^2)

inla.qmarginal(c(0.025, 0.5, 0.975), marg.variance)

inla.zmarginal(marg.variance)

# Extract fitted values marginals
list_marginals <- model1_pm$marginals.fitted.values

# Convert to dataframe
marginals <- data.frame(do.call(rbind, list_marginals))

# Add identifier column (modify as needed)
marginals$location <- rep(names(list_marginals), 
                          times = sapply(list_marginals, nrow))

```



INLA - setting priors in the model for PM2.5
```{r}
# MRF smooth GAM model for pm 2.5
mrf_income_pm <- bam(pm_change ~ score_0.5 + med_income_decile + s(GEOID, bs='mrf', k=200, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 

summary(mrf_income_pm)

## trying to set priors using GAM model results
formula2_pm <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + #not sure if this needs to have a particular specification?
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect

# Using the estimates and standard errors to inform priors
# PM model
priors_pm <- simple_extract(mrf_income_pm)
model2_pm <- inla(
  formula2_pm, 
  data = inla.stack.data(stk_pm, spde=spde), 
  family = 'gaussian',
  control.fixed = priors_pm,
  control.predictor = list(A=inla.stack.A(stk_pm), compute=TRUE),
  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE),
  verbose = FALSE
)


# Comparing results for models with and without setting priors from MRF model
# Plot effect estimates and credible intervals
graphics.off()
Efxplot(list(model2_pm)) # with priors
Efxplot(list(model1_pm)) # without priors


plot(model2_pm, single = TRUE) #don't know what this plot is and don't know why output is different from model 1

# Compare model 1 (no priors) and model 2 (priors) results
model1_pm$summary.fix
model1_pm$summary.hyperpar

model2_pm$summary.fix
model2_pm$summary.hyperpar

# Residual Analysis
par(mfrow=c(2,2))
plot(model2_pm$summary.random$spatial.field$mean, type="l")
qqnorm(model2_pm$summary.random$spatial.field$mean)
hist(model2_pm$summary.random$spatial.field$mean)

# QQ Plot to check normality
residuals <- model2_pm$summary.random$spatial.field$mean
qqnorm(residuals)

# Shapiro-Wilk Test for Normality
shapiro.test(residuals)

# Kolmogorov-Smirnov test against normal distribution
ks.test(residuals, "pnorm", mean=mean(residuals), sd=sd(residuals))

# Check for outliers
plot(model2_pm$summary.random$spatial.field$mean)
boxplot(model2_pm$summary.random$spatial.field$mean)


```

INLA stack (NDVI)
```{r}
# Create stack
stk_ndvi <- inla.stack(
  data=list(y=all_data_inc$ndvi_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10)))


#"Two projector matrices are needed A = list(A,1): the projector matrix for the latent field and a matrix that is a one-to-one map of the covariate. The latter matrix can simply be a constant rather than a diagonal matrix."
```

INLA - model fitting (no priors) for ndvi
```{r}
# Specify formula
formula1_ndvi <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model
model1_ndvi<-inla(formula1_ndvi, 
             data = inla.stack.data(stk_ndvi, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_ndvi),compute = TRUE), 
             control.compute = list(cpo = TRUE))

```

INLA model results for NDVI
```{r}
###  Model results
# Posterior distribution summary of the fixed effects
model1_ndvi$summary.fix

# Posterior distribution summary of the hyper parameters (spatial field parameters)
model1_ndvi$summary.hyperpar

# Rounded summaries
round(model1_ndvi$summary.fixed,3) 
round(model1_ndvi$summary.hyperpar,3)

# "The posterior distribution of the hyperparameters θ1 and θ2 are difficult to interpret. We can transform them into more interpretable quantities such as the range, and the variance. The range is of particular interest. It provides the distance value (in the unit of the point coordinates) above which spatial dependencies become negligible."
model1_ndvi.res<-inla.spde2.result(model1_ndvi, 'spatial.field', spde, do.transf=TRUE) #do.tranf = TRUE makes sure that marginals are calculated in the same scale as the data.

model1_ndvi.res$summary.log.range.nominal
model1_ndvi.res$summary.log.variance.nominal

# Plot effect estimate and credible intervals
graphics.off()
Efxplot(list(model1_ndvi))

# Plot for fixed parameters for insight into the effect of covariates
par(mfrow=c(2,2))
plot(model1_ndvi$marginals.fixed[[1]], ty = "l", xlab = expression(beta[0]), ylab = "Density")
plot(model1_ndvi$marginals.fixed[[2]], ty = "l", xlab = expression(beta[score_0.5]), ylab = "Density") 
plot(model1_ndvi$marginals.fixed[[3]], ty = "l", xlab = expression(beta[med_income_ranked]), ylab = "Density") 

# Plots for hyper parameters for insight into the effect of the structure of the spatial field
par(mfrow=c(1,3))
plot(model1_ndvi.res$marginals.var[[1]], ty = "l", xlab = expression(sigma[randomfield]^2), ylab = "Density") 
plot(model1_ndvi.res$marginals.kap[[1]], type = "l", xlab = expression(kappa), ylab = "Density")
plot(model1_ndvi.res$marginals.range[[1]], type = "l", xlab = "range nominal", ylab = "Density")


# Plot spatial field 
gproj <- inla.mesh.projector(Mesh,  dims = c(300, 300))
g.mean <- inla.mesh.project(gproj, model1_ndvi$summary.random$spatial.field$mean)
g.sd <- inla.mesh.project(gproj, model1_ndvi$summary.random$spatial.field$sd)

grid.arrange(levelplot(g.mean, scales=list(draw=F), xlab='', ylab='', main='mean',col.regions = heat.colors(16)),
             levelplot(g.sd, scal=list(draw=F), xla='', yla='', main='sd' ,col.regions = heat.colors(16)),
             nrow=1)
# "take a look at the spatial field to understand the spatial structure. It can inform on potential omitted variable - how much of the variance in the dependent variable is not explained by the covariates." 

# Other model plots
plot(model1_ndvi, single = TRUE) 

# Residual Analysis
par(mfrow=c(2,2))
plot(model1_ndvi$summary.random$spatial.field$mean, type="l")
qqnorm(model1_ndvi$summary.random$spatial.field$mean)
hist(model1_ndvi$summary.random$spatial.field$mean)

# QQ Plot to check normality
residuals <- model1_ndvi$summary.random$spatial.field$mean
qqnorm(residuals)

# Shapiro-Wilk Test for Normality
shapiro.test(residuals)

# Kolmogorov-Smirnov test against normal distribution
ks.test(residuals, "pnorm", mean=mean(residuals), sd=sd(residuals))

# Check for outliers
plot(model1_ndvi$summary.random$spatial.field$mean)
boxplot(model1_ndvi$summary.random$spatial.field$mean)



```

INLA - setting priors in the model for NDVI
```{r}
# MRF smooth GAM model for NDVI
mrf_income_ndvi <- bam(ndvi_change ~ score_0.5 + med_income_decile + 
                      s(GEOID, bs='mrf', k=200, xt=list(nb=nb_clean)), 
                      family = scat, data = all_data_inc_clean, method = "REML")

summary(mrf_income_ndvi)

## trying to set priors using GAM model results
formula2_ndvi <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 

# Using the estimates and standard errors to inform priors
priors_ndvi <- simple_extract(mrf_income_ndvi)
model2_ndvi <- inla(
  formula2_ndvi,
  data = inla.stack.data(stk_ndvi, spde=spde),
  family = 'gaussian',
  control.fixed = priors_ndvi,
  control.predictor = list(A=inla.stack.A(stk_ndvi), compute=TRUE),
  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE),
  verbose = FALSE
)

# Plot effect estimates and credible intervals
graphics.off()
Efxplot(list(model2_ndvi)) 
Efxplot(list(model1_ndvi))
# SETTING PRIORS CHANGES DIRECTION OF SIGNIFICANCE.......

plot(model2_ndvi, single = TRUE) #don't know what this plot is and don't know why output is different from model 1

# Compare model 1 (no priors) and model 2 (priors) results
model1_ndvi$summary.fix
model1_ndvi$summary.hyperpar

model2_ndvi$summary.fix
model2_ndvi$summary.hyperpar

# Residual Analysis
graphics.off()
par(mfrow=c(2,2))
plot(model2_ndvi$summary.random$spatial.field$mean, type="l")
qqnorm(model2_ndvi$summary.random$spatial.field$mean)
hist(model2_ndvi$summary.random$spatial.field$mean)

# QQ Plot to check normality
residuals <- model2_ndvi$summary.random$spatial.field$mean
qqnorm(residuals)

# Shapiro-Wilk Test for Normality
shapiro.test(residuals)

# Kolmogorov-Smirnov test against normal distribution
ks.test(residuals, "pnorm", mean=mean(residuals), sd=sd(residuals))

# Check for outliers
plot(model2_ndvi$summary.random$spatial.field$mean)
boxplot(model2_ndvi$summary.random$spatial.field$mean)


```


INLA stack (O3)
```{r}
# Create stack
stk_o3 <- inla.stack(
  data=list(y=all_data_inc$o3_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10)))


#"Two projector matrices are needed A = list(A,1): the projector matrix for the latent field and a matrix that is a one-to-one map of the covariate. The latter matrix can simply be a constant rather than a diagonal matrix."

```

INLA - model fitting (no priors) for O3
```{r}
# Specify formula
formula1_o3 <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model
model1_o3<-inla(formula1_o3, 
             data = inla.stack.data(stk_o3, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_o3),compute = TRUE), 
             control.compute = list(cpo = TRUE))

```

INLA model results for O3
```{r}
###  Model results
# Posterior distribution summary of the fixed effects
model1_o3$summary.fix

# Posterior distribution summary of the hyper parameters (spatial field parameters)
model1_o3$summary.hyperpar

# Rounded summaries
round(model1_o3$summary.fixed,3) 
round(model1_o3$summary.hyperpar,3)

# "The posterior distribution of the hyperparameters θ1 and θ2 are difficult to interpret. We can transform them into more interpretable quantities such as the range, and the variance. The range is of particular interest. It provides the distance value (in the unit of the point coordinates) above which spatial dependencies become negligible."
model1_o3.res<-inla.spde2.result(model1_ndvi, 'spatial.field', spde, do.transf=TRUE) #do.tranf = TRUE makes sure that marginals are calculated in the same scale as the data.

model1_o3.res$summary.log.range.nominal
model1_o3.res$summary.log.variance.nominal

# Plot effect estimate and credible intervals
graphics.off()
Efxplot(list(model1_o3))

# Plot for fixed parameters for insight into the effect of covariates
par(mfrow=c(2,2))
plot(model1_o3$marginals.fixed[[1]], ty = "l", xlab = expression(beta[0]), ylab = "Density")
plot(model1_o3$marginals.fixed[[2]], ty = "l", xlab = expression(beta[score_0.5]), ylab = "Density") 
plot(model1_o3$marginals.fixed[[3]], ty = "l", xlab = expression(beta[med_income_decile]), ylab = "Density") 

# Plots for hyper parameters for insight into the effect of the structure of the spatial field
par(mfrow=c(1,3))
plot(model1_o3.res$marginals.var[[1]], ty = "l", xlab = expression(sigma[randomfield]^2), ylab = "Density") 
plot(model1_o3.res$marginals.kap[[1]], type = "l", xlab = expression(kappa), ylab = "Density")
plot(model1_o3.res$marginals.range[[1]], type = "l", xlab = "range nominal", ylab = "Density")


# Plot spatial field 
gproj <- inla.mesh.projector(Mesh,  dims = c(300, 300))
g.mean <- inla.mesh.project(gproj, model1_o3$summary.random$spatial.field$mean)
g.sd <- inla.mesh.project(gproj, model1_o3$summary.random$spatial.field$sd)

grid.arrange(levelplot(g.mean, scales=list(draw=F), xlab='', ylab='', main='mean',col.regions = heat.colors(16)),
             levelplot(g.sd, scal=list(draw=F), xla='', yla='', main='sd' ,col.regions = heat.colors(16)),
             nrow=1)
# "take a look at the spatial field to understand the spatial structure. It can inform on potential omitted variable - how much of the variance in the dependent variable is not explained by the covariates." 

# Other model plots
plot(model1_o3, single = TRUE) 

# Residual Analysis
par(mfrow=c(2,2))
plot(model1_o3$summary.random$spatial.field$mean, type="l")
qqnorm(model1_o3$summary.random$spatial.field$mean)
hist(model1_o3$summary.random$spatial.field$mean)

# QQ Plot to check normality
residuals <- model1_o3$summary.random$spatial.field$mean
qqnorm(residuals)

# Shapiro-Wilk Test for Normality
shapiro.test(residuals)

# Kolmogorov-Smirnov test against normal distribution
ks.test(residuals, "pnorm", mean=mean(residuals), sd=sd(residuals))

# Check for outliers
plot(model1_o3$summary.random$spatial.field$mean)
boxplot(model1_o3$summary.random$spatial.field$mean)
```

INLA - setting priors for O3
```{r}
# MRF smooth model for o3
mrf_income_o3 <- bam(o3_change ~ score_0.5 + med_income_decile + s(GEOID, bs='mrf', k=200, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 

summary(mrf_income_o3)

## trying to set priors using GAM model results
formula2_o3 <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect

# O3 model
priors_o3 <- simple_extract(mrf_income_o3)
model2_o3 <- inla(
  formula2_o3,
  data = inla.stack.data(stk_o3, spde=spde),
  family = 'gaussian',
  control.fixed = priors_o3,
  control.predictor = list(A=inla.stack.A(stk_o3), compute=TRUE),
  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE),
  verbose = FALSE
)

# Plot effect estimates and credible intervals
graphics.off()
Efxplot(list(model2_o3)) 
Efxplot(list(model1_o3))

plot(model2_o3, single = TRUE) #don't know what this plot is and don't know why output is different from model 1

# Compare model 1 (no priors) and model 2 (priors) results
model1_o3$summary.fix
model1_o3$summary.hyperpar

model2_o3$summary.fix
model2_o3$summary.hyperpar

# Residual Analysis
graphics.off()
par(mfrow=c(2,2))
plot(model2_o3$summary.random$spatial.field$mean, type="l")
qqnorm(model2_o3$summary.random$spatial.field$mean)
hist(model2_o3$summary.random$spatial.field$mean)

# QQ Plot to check normality
residuals <- model2_o3$summary.random$spatial.field$mean
qqnorm(residuals)

# Shapiro-Wilk Test for Normality
shapiro.test(residuals)

# Kolmogorov-Smirnov test against normal distribution
ks.test(residuals, "pnorm", mean=mean(residuals), sd=sd(residuals))

# Check for outliers
plot(model2_o3$summary.random$spatial.field$mean)
boxplot(model2_o3$summary.random$spatial.field$mean)

```


INLA stack (NO2)
```{r}
# Create stack
stk_no2 <- inla.stack(
  data=list(y=all_data_inc$no2_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10)))


#"Two projector matrices are needed A = list(A,1): the projector matrix for the latent field and a matrix that is a one-to-one map of the covariate. The latter matrix can simply be a constant rather than a diagonal matrix."

```
INLA - model fitting (no priors) for NO2
```{r}
# Specify formula
formula1_no2 <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model
model1_no2<-inla(formula1_no2, 
             data = inla.stack.data(stk_no2, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_no2),compute = TRUE), 
             control.compute = list(cpo = TRUE))

# Additional warnings:
# *** WARNING *** Eigenvalue 1 of the Hessian is -2.1127 < 0
# 	*** WARNING *** This have consequence for the accurancy of the hyperpar
# 	*** WARNING *** Continue with a diagonal Hessian.


```

INLA model results for NO2
```{r}
###  Model results
# Posterior distribution summary of the fixed effects
model1_no2$summary.fix

# Posterior distribution summary of the hyper parameters (spatial field parameters)
model1_no2$summary.hyperpar

# Rounded summaries
round(model1_no2$summary.fixed,3) 
round(model1_no2$summary.hyperpar,3)

# "The posterior distribution of the hyperparameters θ1 and θ2 are difficult to interpret. We can transform them into more interpretable quantities such as the range, and the variance. The range is of particular interest. It provides the distance value (in the unit of the point coordinates) above which spatial dependencies become negligible."
model1_no2.res<-inla.spde2.result(model1_ndvi, 'spatial.field', spde, do.transf=TRUE) #do.tranf = TRUE makes sure that marginals are calculated in the same scale as the data.

model1_no2.res$summary.log.range.nominal
model1_no2.res$summary.log.variance.nominal

# Plot effect estimate and credible intervals
graphics.off()
Efxplot(list(model1_no2))

# Plot for fixed parameters for insight into the effect of covariates
par(mfrow=c(2,2))
plot(model1_no2$marginals.fixed[[1]], ty = "l", xlab = expression(beta[0]), ylab = "Density")
plot(model1_no2$marginals.fixed[[2]], ty = "l", xlab = expression(beta[score_0.5]), ylab = "Density") 
plot(model1_no2$marginals.fixed[[3]], ty = "l", xlab = expression(beta[med_income_decile]), ylab = "Density") 

# Plots for hyper parameters for insight into the effect of the structure of the spatial field
par(mfrow=c(1,3))
plot(model1_no2.res$marginals.var[[1]], ty = "l", xlab = expression(sigma[randomfield]^2), ylab = "Density") 
plot(model1_no2.res$marginals.kap[[1]], type = "l", xlab = expression(kappa), ylab = "Density")
plot(model1_no2.res$marginals.range[[1]], type = "l", xlab = "range nominal", ylab = "Density")


# Plot spatial field 
gproj <- inla.mesh.projector(Mesh,  dims = c(300, 300))
g.mean <- inla.mesh.project(gproj, model1_no2$summary.random$spatial.field$mean)
g.sd <- inla.mesh.project(gproj, model1_no2$summary.random$spatial.field$sd)

grid.arrange(levelplot(g.mean, scales=list(draw=F), xlab='', ylab='', main='mean',col.regions = heat.colors(16)),
             levelplot(g.sd, scal=list(draw=F), xla='', yla='', main='sd' ,col.regions = heat.colors(16)),
             nrow=1)
# "take a look at the spatial field to understand the spatial structure. It can inform on potential omitted variable - how much of the variance in the dependent variable is not explained by the covariates." 

# Other model plots
plot(model1_no2, single = TRUE)

# Residual Analysis
par(mfrow=c(2,2))
plot(model1_no2$summary.random$spatial.field$mean, type="l")
qqnorm(model1_no2$summary.random$spatial.field$mean)
hist(model1_no2$summary.random$spatial.field$mean)

# QQ Plot to check normality
residuals <- model1_no2$summary.random$spatial.field$mean
qqnorm(residuals)

# Shapiro-Wilk Test for Normality
shapiro.test(residuals)

# Kolmogorov-Smirnov test against normal distribution
ks.test(residuals, "pnorm", mean=mean(residuals), sd=sd(residuals))

# Check for outliers
plot(model1_no2$summary.random$spatial.field$mean)
boxplot(model1_no2$summary.random$spatial.field$mean)
```

INLA - Setting priors in the model for NO2
```{r}
# MRF smooth model for no2
mrf_income_no2 <- bam(no2_change ~ score_0.5 + med_income_decile + s(GEOID, bs='mrf', k=200, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML") 

formula2_no2 <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect

priors_no2 <- simple_extract(mrf_income_no2)
model2_no2 <- inla(
  formula2_no2,
  data = inla.stack.data(stk_no2, spde=spde),
  family = 'gaussian',
  control.fixed = priors_no2,
  control.predictor = list(A=inla.stack.A(stk_no2), compute=TRUE),
  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE),
  verbose = FALSE
)

# Plot effect estimates and credible intervals
graphics.off()
Efxplot(list(model2_no2)) 
Efxplot(list(model1_no2))

plot(model2_no2, single = TRUE) #don't know what this plot is and don't know why output is different from model 1

# Compare model 1 (no priors) and model 2 (priors) results
model1_no2$summary.fix
model1_no2$summary.hyperpar

model2_no2$summary.fix
model2_no2$summary.hyperpar

# Residual Analysis
graphics.off()
par(mfrow=c(2,2))
plot(model2_no2$summary.random$spatial.field$mean, type="l")
qqnorm(model2_no2$summary.random$spatial.field$mean)
hist(model2_no2$summary.random$spatial.field$mean)

# QQ Plot to check normality
residuals <- model2_no2$summary.random$spatial.field$mean
qqnorm(residuals)

# Shapiro-Wilk Test for Normality
shapiro.test(residuals)

# Kolmogorov-Smirnov test against normal distribution
ks.test(residuals, "pnorm", mean=mean(residuals), sd=sd(residuals))

# Check for outliers
plot(model2_no2$summary.random$spatial.field$mean)
boxplot(model2_no2$summary.random$spatial.field$mean)

```


INLA stack (temp)
```{r}
# Create stack
stk_temp <- inla.stack(
  data=list(y=all_data_inc$mean_temp), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10)))


#"Two projector matrices are needed A = list(A,1): the projector matrix for the latent field and a matrix that is a one-to-one map of the covariate. The latter matrix can simply be a constant rather than a diagonal matrix."

```

INLA - model fitting (no priors) for temp
```{r}
# Specify formula
formula1_temp <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model
model1_temp <-inla(formula1_temp, 
             data = inla.stack.data(stk_temp, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_temp),compute = TRUE), 
             control.compute = list(cpo = TRUE))

```

INLA model results for mean temp
```{r}
###  Model results
# Posterior distribution summary of the fixed effects
model1_temp$summary.fix

# Posterior distribution summary of the hyper parameters (spatial field parameters)
model1_temp$summary.hyperpar

# Rounded summaries
round(model1_temp$summary.fixed,3) 
round(model1_temp$summary.hyperpar,3)

# "The posterior distribution of the hyperparameters θ1 and θ2 are difficult to interpret. We can transform them into more interpretable quantities such as the range, and the variance. The range is of particular interest. It provides the distance value (in the unit of the point coordinates) above which spatial dependencies become negligible."
model1_temp.res<-inla.spde2.result(model1_ndvi, 'spatial.field', spde, do.transf=TRUE) #do.tranf = TRUE makes sure that marginals are calculated in the same scale as the data.

model1_temp.res$summary.log.range.nominal
model1_temp.res$summary.log.variance.nominal

# Plot effect estimate and credible intervals
graphics.off()
Efxplot(list(model1_temp))

# Plot for fixed parameters for insight into the effect of covariates
par(mfrow=c(2,2))
plot(model1_temp$marginals.fixed[[1]], ty = "l", xlab = expression(beta[0]), ylab = "Density")
plot(model1_temp$marginals.fixed[[2]], ty = "l", xlab = expression(beta[score_0.5]), ylab = "Density") 
plot(model1_temp$marginals.fixed[[3]], ty = "l", xlab = expression(beta[med_income_decile]), ylab = "Density") 

# Plots for hyper parameters for insight into the effect of the structure of the spatial field
par(mfrow=c(1,3))
plot(model1_temp.res$marginals.var[[1]], ty = "l", xlab = expression(sigma[randomfield]^2), ylab = "Density") 
plot(model1_temp.res$marginals.kap[[1]], type = "l", xlab = expression(kappa), ylab = "Density")
plot(model1_temp.res$marginals.range[[1]], type = "l", xlab = "range nominal", ylab = "Density")


# Plot spatial field 
gproj <- inla.mesh.projector(Mesh,  dims = c(300, 300))
g.mean <- inla.mesh.project(gproj, model1_temp$summary.random$spatial.field$mean)
g.sd <- inla.mesh.project(gproj, model1_temp$summary.random$spatial.field$sd)

grid.arrange(levelplot(g.mean, scales=list(draw=F), xlab='', ylab='', main='mean',col.regions = heat.colors(16)),
             levelplot(g.sd, scal=list(draw=F), xla='', yla='', main='sd' ,col.regions = heat.colors(16)),
             nrow=1)
# "take a look at the spatial field to understand the spatial structure. It can inform on potential omitted variable - how much of the variance in the dependent variable is not explained by the covariates." 

# Other model plots
plot(model1_temp, single = TRUE) 

# Residual Analysis
par(mfrow=c(2,2))
plot(model1_temp$summary.random$spatial.field$mean, type="l")
qqnorm(model1_temp$summary.random$spatial.field$mean)
hist(model1_temp$summary.random$spatial.field$mean)

# QQ Plot to check normality
residuals <- model1_temp$summary.random$spatial.field$mean
qqnorm(residuals)

# Shapiro-Wilk Test for Normality
shapiro.test(residuals)

# Kolmogorov-Smirnov test against normal distribution
ks.test(residuals, "pnorm", mean=mean(residuals), sd=sd(residuals))

# Check for outliers
plot(model1_temp$summary.random$spatial.field$mean)
boxplot(model1_temp$summary.random$spatial.field$mean)
```
INLA - setting priors for temp model
```{r}
#MRF smooth model for 2016 mean temp
mrf_income_temp <- bam(mean_temp ~ score_0.5 + med_income_decile + s(GEOID, bs='mrf', k=200, xt=list(nb=nb)), family = scat, data = all_data_inc, method = "REML")  

summary(mrf_income_temp)

formula2_temp <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect

priors_temp <- simple_extract(mrf_income_temp)
model2_temp <- inla(
  formula2_temp,
  data = inla.stack.data(stk_temp, spde=spde),
  family = 'gaussian',
  control.fixed = priors_temp,
  control.predictor = list(A=inla.stack.A(stk_temp), compute=TRUE),
  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE),
  verbose = FALSE
)

# Plot effect estimates and credible intervals
graphics.off()
Efxplot(list(model2_temp)) 
Efxplot(list(model1_temp))

plot(model2_temp, single = TRUE) #don't know what this plot is and don't know why output is different from model 1

# Compare model 1 (no priors) and model 2 (priors) results
model1_temp$summary.fix
model1_temp$summary.hyperpar

model2_temp$summary.fix
model2_temp$summary.hyperpar

# Residual Analysis
graphics.off()
par(mfrow=c(2,2))
plot(model2_temp$summary.random$spatial.field$mean, type="l")
qqnorm(model2_temp$summary.random$spatial.field$mean)
hist(model2_temp$summary.random$spatial.field$mean)

# QQ Plot to check normality
residuals <- model2_temp$summary.random$spatial.field$mean
qqnorm(residuals)

# Shapiro-Wilk Test for Normality
shapiro.test(residuals)

# Kolmogorov-Smirnov test against normal distribution
ks.test(residuals, "pnorm", mean=mean(residuals), sd=sd(residuals))

# Check for outliers
plot(model2_temp$summary.random$spatial.field$mean)
boxplot(model2_temp$summary.random$spatial.field$mean)

```



