---
title: "gent_3.1.25"
author: "Joyce Hu"
date: "2025-1-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages (JH - check if any are unused after code cleaned)
```{r}
library(tidyverse)
library(tigris)
library(arrow)
library(here)
library(mgcv)
library(fixest)
library(lmtest)
library(terra)
library(exactextractr)
library(tidyterra)
library(sf)
library(mgcViz)

library(DHARMa)
library(qqplotr)
library(readxl) # for read_excel()
library(spdep) # for moran.test()
library(tidycensus) # for median income
library(distr) # for qqbounds()
library(spatialreg) # for spatial lag and error models
library(gridExtra) # for viewing plots side-by-side
library(qgam) 
library(dplyr)
library(INLA)
library(ggregplot) #INLA compatible plotting library (efxplot)
library(ggfields) # for plotting spatial field
library(lattice) # for levelplot
library(inlatools) # for INLA diagnostics 
library(kableExtra)
library(gridExtra)
library(lattice)
```

Load and clean data (note: takes approximately 12 mins)
```{r}
# Check if 'data' directory exists; if not, create it
if (!dir.exists(here("data"))) {
  dir.create(here("data"))
}

# Load gentrification score by tract
gentrification <- read.csv(here("data", "Data_NYC_Gentrification_2000_16.csv")) %>% 
  mutate(tractid = as.character(tractid)) %>% 
  dplyr::select(tractid, score_0.5)

# Load gentrification score if using xlsx file:
# gentrification <- read_xlsx(here("Gentrification","Data_NYC_Gentrification_2000_16.xlsx")) %>%
#   mutate(tractid = as.character(tractid)) %>%
#   dplyr::select(tractid, score_0.5)

# Pull NYC tract IDs/fips from gentrification index dataset for filtering
nyc_fips <- gentrification$tractid

# Load NYC tracts shapefile
nyc_tracts <- tracts(state = "NY", county = c("Bronx", "Kings", "New York", "Queens", "Richmond"), year = 2010) %>%
  # clean lat and long for Conley SEs
  mutate(lat = as.numeric(INTPTLAT10), long = as.numeric(INTPTLON10), GEOID = GEOID10) %>% 
  # filter for FIPS codes included in gentrification index dataset
  filter(GEOID %in% nyc_fips)


# Load NY tracts populatuion weighted centroid points from Census data
nyc_weighted_centroids <- read.table("https://www2.census.gov/geo/docs/reference/cenpop2010/tract/CenPop2010_Mean_TR36.txt", 
                         header = TRUE, sep = ",", stringsAsFactors = FALSE, 
                         colClasses = c("character", "character", "character", "numeric", "numeric", "numeric")) %>%
  mutate(FIPS = paste(STATEFP, COUNTYFP, TRACTCE, sep = "")) %>% 
  ## filter for only NYC tracts
  filter(FIPS %in% nyc_fips) %>%  
  rename(Longitude_weighted = LONGITUDE, 
         Latitude_weighted = LATITUDE)

# JH - 10/12/24 warning persists 10/13, 10/14 Warning: R metadata may have unsafe or invalid elements ℹ Type: "externalptr" options(arrow.unsafe_metadata = TRUE) #

# Temperature data
if(!file.exists(here("data","summarized_daily_temp_preds.parquet"))){download.file("https://zenodo.org/records/10557980/files/summarized_daily_temp_preds.parquet?download=1", 
                                                                                   destfile = here("data","summarized_daily_temp_preds.parquet"))}

# Temperature by tract
temp <- read_parquet(here("data","summarized_daily_temp_preds.parquet")) %>%
  filter(str_detect(date, "2016")) %>%
  mutate(temp_c = mean_temp_daily_cK/100-273.15) %>% 
  mutate(temp_f = temp_c*9/5+32) %>% 
  group_by(GEOID, date) %>%
  ungroup() %>% 
  group_by(GEOID) %>% 
  summarize(mean_temp = mean(mean_temp_daily_cK/100))

# NDVI by tract (need to have it downloaded and stored locally)
ndvi_mean_2016 <- rast(here("data", "ndvi_c2_2016_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

ndvi_mean_2000 <- rast(here("data", "ndvi_c2_2000_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

ndvi_sd_2016 <- rast(here("data", "ndvi_c2_2000_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "stdev")

ndvi_sd_2000 <- rast(here("data", "ndvi_c2_2016_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "stdev")

# PM2.5 data
# Download data from URL
download_data <- function() {
  # set longer timeout
  options(timeout = 7200)
  if(!file.exists(here("data", "2016_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_pm25_daily_average.txt.gz", destfile = here("data", "2016_pm25_daily_average.txt.gz"))} #2016
  if(!file.exists(here("data", "2002_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_pm25_daily_average.txt.gz", destfile = here("data", "2002_pm25_daily_average.txt.gz"))}} #2002
download_data()

# Read in PM2.5 daily average by census tract from EPA FAQSD - 2002
pm_tract_2002 <- read.delim(gzfile(here("data", "2002_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## Include only NYC tracts
  filter(FIPS %in% nyc_fips) %>%
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## Calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(pm_mean_2002 = mean(pm25_daily_average.ug.m3.),
            pm_sd_2002 = sd(pm25_daily_average.ug.m3.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# Read in PM2.5 daily average by census tract from EPA FAQSD - 2016
pm_tract_2016 <- read.delim(gzfile(here("data", "2016_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(pm_mean_2016 = mean(Prediction),
            pm_sd_2016 = sd(Prediction),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))
  
# Combine 2002 and 2016 pm data for nyc_tracts
pm_tract <- pm_tract_2002 %>%
  inner_join(pm_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# NO2 CACES
no2_2016_caces <- read.csv(here("data", "2016_no2_caces.csv")) %>% 
  mutate(no2_2016 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_2000_caces <- read.csv(here("data", "2000_no2_caces.csv")) %>% 
  mutate(no2_2000 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_caces <- left_join(no2_2000_caces, no2_2016_caces, by = "fips") %>% 
  select(no2_2000, no2_2016, fips)

#O3 data download
options(timeout = 7200)
if(!file.exists(here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))} #2016
if(!file.exists(here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))} #2002

# Read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2002
o3_tract_2002 <- read.delim(gzfile(here("data", "2002_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>% 
  ## include only NYC tracts
  filter(FIPS %in% nyc_fips) %>% 
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(o3_mean_2002 = mean(ozone_daily_8hour_maximum.ppb.),
            o3_sd_2002 = sd(ozone_daily_8hour_maximum.ppb.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# Read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2016
o3_tract_2016 <- read.delim(gzfile(here("data", "2016_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>% 
  ## calculate annual average by tract using  population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(o3_mean_2016 = mean(Prediction),
            o3_sd_2016 = sd(Prediction),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

o3_tract <- o3_tract_2002 %>%
  left_join(o3_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# Merge all exposure data
all_data <- nyc_tracts %>% 
  # join gentrification index by tract
  right_join(gentrification, by = c("GEOID10" = "tractid")) %>%
  left_join(pm_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(pm_2016 = pm_mean_2016, pm_2002 = pm_mean_2002, pm_change = pm_2016-pm_2002) %>% 
  left_join(o3_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(o3_2016 = o3_mean_2016, o3_2002 = o3_mean_2002, o3_change = o3_2016-o3_2002, o3_sd_2016 = o3_sd_2016, o3_sd_2002 = o3_sd_2002) %>% 
  left_join(no2_caces, by = c("GEOID" = "fips")) %>% 
  mutate(no2_change = no2_2016-no2_2000) %>% 
  mutate(ndvi_mean_2016 = ndvi_mean_2016, ndvi_mean_2000 = ndvi_mean_2000, ndvi_change = ndvi_mean_2016-ndvi_mean_2000) %>%
  mutate(ndvi_sd_2016 = ndvi_sd_2016, ndvi_sd_2000 = ndvi_sd_2000) %>%
  left_join(temp, by = "GEOID")


############ Median income ###############################
# Find median household income variable in 2000 decennial census
all_vars_dec <- 
  load_variables(year = 2000, dataset = "sf3")

# Models adjusting for median income in 2000 (note: 2000 decennial census is 1999 data)
income_2000 <- get_decennial(variables = "P053001", year = 2000,  sumfile = "sf3", state = "NY", 
                  county = c("Queens", "King", "New York", "Richmond", "Bronx"), geography = "tract") %>% 
  rename(med_income_2000 = value)
#results in 226 missing tracts in terms of 2010 tracts.

# Find median household income variable in 2010 ACS
income_2010 <- get_acs(variables = "B06011_001", year = 2010, state = "NY", 
                  county = c("Queens", "King", "New York", "Richmond", "Bronx"), geography = "tract") %>% 
  rename(med_income_2010 = estimate)

# Joining both income sets with data and imputing 2010 data where 2000 data is missing
all_data_inc <- all_data %>%
  left_join(income_2000, by = "GEOID") %>%
  left_join(income_2010, by = "GEOID") %>%
  mutate(med_income_2000_impute = coalesce(med_income_2000, med_income_2010)) %>% 
  # coerce GEOID to numeric then factor class to avoid error in MRF smooth
  mutate(GEOID = as.numeric(GEOID)) %>% 
  mutate(GEOID = factor(GEOID)) %>% 
 # coerce COUNTYFP10 to numeric then factor class to avoid error ?
  mutate(COUNTYFP10 = as.numeric(COUNTYFP10)) %>% 
  mutate(COUNTYFP10 = factor(COUNTYFP10))

# Ranking median income for use in GAM models
all_data_inc$med_income_ranked <- rank(all_data_inc$med_income_2000_impute)


# Converting median income to quantiles and scaling
# Create deciles
income_quantiles <- quantile(all_data_inc$med_income_2000_impute, probs = seq(0, 1, by = 0.1))
all_data_inc$med_income_decile <- findInterval(all_data_inc$med_income_2000_impute, 
                                             income_quantiles,
                                             rightmost.closed = TRUE)
```
Backing up data (writing data frames as csv files)
```{r}
###################### BACKING UP NYC TRACTS INFO ######################
# # Convert geometry to WKT format
# nyc_tracts$geometry <- st_as_text(nyc_tracts$geometry)
# # Save to CSV (now including geometry as text)
# write.csv(nyc_tracts, "nyc_tracts_with_geometry.csv", row.names = FALSE)
# nyc_tracts_with_geometry <- read.csv(here("nyc_tracts_with_geometry.csv"))
# nyc_tracts <- st_as_sf(nyc_tracts_with_geometry, wkt = "geometry") %>%
#   mutate(
#     lat = as.numeric(INTPTLAT10),
#     long = as.numeric(INTPTLON10),
#     tractid = GEOID10,
#     tractid = as.character(tractid)  
#   ) %>%
#   filter(tractid %in% nyc_fips)
# 
# ############## BACKING UP NYC WEIGHTED CENTROIDS ######################
# write.csv(nyc_weighted_centroids, "nyc_weighted_centroids.csv",row.names=FALSE)
# nyc_weighted_centroids_copy <- read_csv(here("nyc_weighted_centroids.csv"))
# 
# #################### BACKING UP all_data_inc as csv ######################
# # Convert geometry to WKT format and save as CSV
# all_data_inc_csv <- all_data_inc %>%
#   mutate(geometry = st_as_text(geometry))
# 
# write.csv(all_data_inc_csv, "all_data_inc_csv.csv", row.names = FALSE)
# 
# #################### BACKING UP all_data_inc_clean as csv (for NDVI MRF smooth) ######################
# this runs only when #323 is already run to then save the output
# all_data_inc_clean_csv <- all_data_inc %>%
#   mutate(geometry = st_as_text(geometry))
# 
# write.csv(all_data_inc_clean_csv, "all_data_inc_clean_csv.csv", row.names = FALSE)

```

Read in backed up data (alternative to code chunk 2 - lines 45-247)
```{r}
# # read in as data frame 
# all_data_inc <- read.csv("here"("all_data_inc_csv.csv"))
# 
# all_data_inc_clean <- read.csv("here"("all_data_inc_clean_csv.csv"))

```

INLA set up
```{r}
# # Set INLA repository
# options(repos = c(getOption("repos"),
#    INLA="https://inla.r-inla-download.org/R/stable"))
# 
# # Install INLA and dependencies (from CRAN)
# install.packages("INLA", dep = TRUE)

#need older version of INLA to be compatible with R version
 # remotes::install_version("INLA", version="22.05.07",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/testing"), dep=TRUE)

# Install an even older version of INLA for GLIBC compatibility

# remotes::install_version("INLA", 
#                          version="21.02.23", 
#                          repos=c(getOption("repos"),
#                                  INLA="https://inla.r-inla-download.org/R/stable"), 
#                          dep=TRUE)

```

INLA - Mesh creation
```{r}
graphics.off()
#### Create mesh using tract centroids as locations

# Create matrix of locations
Locations <- as.matrix(cbind(
  as.numeric(nyc_weighted_centroids$Longitude_weighted),
  as.numeric(nyc_weighted_centroids$Latitude_weighted)
))

# Create mesh options
# Various mesh option specifications 
MeshA <- inla.mesh.2d(jitter(Locations), max.edge = c(20, 40))
MeshB <- inla.mesh.2d(Locations, max.edge = c(20, 40))
MeshC <- inla.mesh.2d(Locations, max.edge = c(50, 100))
mesh_simple <- inla.mesh.create(Locations)

# Visualize mesh options
plot(MeshA)
points(Locations, col = "red", pch = 2)

plot(MeshB)
points(Locations, col = "red", pch = 2)

plot(MeshC)
points(Locations, col = "red", pch = 2)


# "Generally, the mesh function automatically creates a mesh like mesh A, where closer-together sampling locations produce smaller triangles. When exploring/setting up preliminary analyses, use a mesh like mesh B. For analyses to be reported in a paper, use a mesh like mesh C. Be careful of edges, and try to allow some space around your sampling area for INLA to estimate. The edge triangles can be bigger to reduce computing power."

# Set chosen mesh to use for analysis
Mesh <- MeshB
```

INLA - Creating the A Matrix and SPDE
```{r}
# Create A matrix
A <- inla.spde.make.A(Mesh, loc = Locations) 
# "the A matrix maps the Gaussian Markov Random Field (GMRF) from the mesh nodes to the n observation location. This is represented by a matrix with the number of observations as column and the number of nodes as rows."


# Create spatial structure (SPDE object)
spde <- inla.spde2.matern(Mesh, alpha=2) #they fix alpha as 2

# Create all the required indexes for the SPDE model, naming the spatial field as “spatial.field”, which facilitates the extraction of the estimated values of the spatial field after the fitting process.
iset <- inla.spde.make.index(name = "spatial.field", spde$n.spde)
```

INLA stack (PM2.5)
```{r}
# Create stack 
stk_pm <- inla.stack(
  data=list(y=all_data_inc$pm_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10)))

# Create stack with standard deviations (both years)
stk_pm_sd <- inla.stack(
  data=list(y=all_data_inc$pm_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 pm_sd_2002 = all_data_inc$pm_sd_2002,
                                 pm_sd_2016 = all_data_inc$pm_sd_2016)))

# Create stack with standard deviation (2002 only)
stk_pm_sd_2002 <- inla.stack(
  data=list(y=all_data_inc$pm_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 pm_sd_2002 = all_data_inc$pm_sd_2002)))

# Create stack with baseline
stk_pm_baseline <- inla.stack(
  data=list(y=all_data_inc$pm_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 pm_mean_2002 = all_data_inc$pm_mean_2002)))


#"Two projector matrices are needed A = list(A,1): the projector matrix for the latent field and a matrix that is a one-to-one map of the covariate. The latter matrix can simply be a constant rather than a diagonal matrix."
```

INLA - model fitting (no priors) for PM2.5
```{r}
# Specify formula
formula1_pm <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + #not sure if this needs to have a particular specification?
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 

# Specify model
model1_pm<-inla(formula1_pm, 
             data = inla.stack.data(stk_pm, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_pm),compute = TRUE), 
             control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE))

```

INLA - model fitting (no priors) for PM2.5 with std dev. or baseline as covariates
```{r}
# Specify formula - using both years' sd
formula_pm_sd <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  pm_sd_2002 + 
  pm_sd_2016 +
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model - using both years' sd
model_pm_sd <-inla(formula_pm_sd, 
             data = inla.stack.data(stk_pm_sd, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_pm_sd),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))

# Specify formula - using 2002 sd
formula_pm_sd_2002 <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  pm_sd_2002 + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model - using 2002 sd
model_pm_sd_2002 <-inla(formula_pm_sd_2002, 
             data = inla.stack.data(stk_pm_sd_2002, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_pm_sd_2002),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))

# Specify formula - using baseline
formula_pm_baseline <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  pm_mean_2002 + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model - using baseline
model_pm_baseline <-inla(formula_pm_baseline, 
             data = inla.stack.data(stk_pm_baseline, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_pm_baseline),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))

```

INLA stack (NDVI)
```{r}
# Create stack
stk_ndvi <- inla.stack(
  data=list(y=all_data_inc$ndvi_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10)))

# Create stack with s.d. - both years
stk_ndvi_sd <- inla.stack(
  data=list(y=all_data_inc$ndvi_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 ndvi_sd_2000 = all_data_inc$ndvi_sd_2000,
                                 ndvi_sd_2016 = all_data_inc$ndvi_sd_2016)))

# Create stack with s.d. - 2000
stk_ndvi_sd_2000 <- inla.stack(
  data=list(y=all_data_inc$ndvi_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 ndvi_sd_2000 = all_data_inc$ndvi_sd_2000)))

# Create stack with baseline
stk_ndvi_baseline <- inla.stack(
  data=list(y=all_data_inc$ndvi_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 ndvi_mean_2000 = ndvi_mean_2000)))


#"Two projector matrices are needed A = list(A,1): the projector matrix for the latent field and a matrix that is a one-to-one map of the covariate. The latter matrix can simply be a constant rather than a diagonal matrix."
```

INLA - model fitting (no priors) for NDVI
```{r}
# Specify formula
formula1_ndvi <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model
model1_ndvi<-inla(formula1_ndvi, 
             data = inla.stack.data(stk_ndvi, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_ndvi),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))

# Specify formula - sd for both years
formula_ndvi_sd <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  ndvi_sd_2000 +
  ndvi_sd_2016 +
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model
model_ndvi_sd <-inla(formula_ndvi_sd, 
             data = inla.stack.data(stk_ndvi_sd, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_ndvi_sd),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))

# Specify formula - sd for 2000
formula_ndvi_sd_2000 <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  ndvi_sd_2000 +
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model
model_ndvi_sd_2000 <-inla(formula_ndvi_sd_2000, 
             data = inla.stack.data(stk_ndvi_sd_2000, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_ndvi_sd_2000),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))


# Specify formula with baseline
formula_ndvi_baseline <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  ndvi_mean_2000 +
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model
model_ndvi_baseline <-inla(formula_ndvi_baseline, 
             data = inla.stack.data(stk_ndvi_baseline, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_ndvi_baseline),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))
```

INLA stack (O3)
```{r}
# Create stack
stk_o3 <- inla.stack(
  data=list(y=all_data_inc$o3_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10)))

# Create stack with std dev. for both years
stk_o3_sd <- inla.stack(
  data=list(y=all_data_inc$o3_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 o3_sd_2002 = all_data_inc$o3_sd_2002,
                                 o3_sd_2016 = all_data_inc$o3_sd_2016)))

# Create stack with std dev. for 2002
stk_o3_sd_2002 <- inla.stack(
  data=list(y=all_data_inc$o3_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 o3_sd_2002 = all_data_inc$o3_sd_2002)))


# Create stack with baseline
stk_o3_baseline <- inla.stack(
  data=list(y=all_data_inc$o3_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 o3_mean_2002 = all_data_inc$o3_mean_2002)))
```

INLA - model fitting (no priors) for O3
```{r}
# Specify formula
formula1_o3 <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model
model1_o3<-inla(formula1_o3, 
             data = inla.stack.data(stk_o3, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_o3),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))

```

INLA - model fitting (no priors) for O3 with std. dev as covariates
```{r}
# Specify formula - using both years' sd
formula_o3_sd <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  o3_sd_2002 + 
  o3_sd_2016 +
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model - using both years' sd
model_o3_sd <-inla(formula_o3_sd, 
             data = inla.stack.data(stk_o3_sd, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_o3_sd),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))

# Specify formula - using 2002 sd
formula_o3_sd_2002 <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  o3_sd_2002 + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model - using 2002 sd
model_o3_sd_2002 <-inla(formula_o3_sd_2002, 
             data = inla.stack.data(stk_o3_sd_2002, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_o3_sd_2002),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))

# Specify formula - using baseline
formula_o3_baseline <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  o3_mean_2002 + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model - using baseline
model_o3_baseline <-inla(formula_o3_baseline, 
             data = inla.stack.data(stk_o3_baseline, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_o3_baseline),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))


```

INLA stack (NO2)
```{r}
# Create stack
stk_no2 <- inla.stack(
  data=list(y=all_data_inc$no2_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10)))

# Create stack - no no2 standard deviations, try to add in o3 std dev and see how that goes since they often have an inverse r/s?
stk_no2_sd <- inla.stack(
  data=list(y=all_data_inc$no2_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 o3_sd_2002 = all_data_inc$o3_sd_2002,
                                 o3_sd_2016 = all_data_inc$o3_sd_2016)))


# Create stack with baseline
stk_no2_baseline <- inla.stack(
  data=list(y=all_data_inc$no2_change), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 no2_2000 = all_data_inc$no2_2000)))

```

INLA - model fitting (no priors) for NO2
```{r}
# Specify formula
formula1_no2 <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model
model1_no2<-inla(formula1_no2, 
             data = inla.stack.data(stk_no2, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_no2),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))

# Additional warnings:
# *** WARNING *** Eigenvalue 1 of the Hessian is -2.1127 < 0
# 	*** WARNING *** This have consequence for the accurancy of the hyperpar
# 	*** WARNING *** Continue with a diagonal Hessian.

# Specify formula using o3 std dev
formula_no2_sd <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  o3_sd_2002 +
  o3_sd_2016 +
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model with o3 std dev
model_no2_sd<-inla(formula_no2_sd, 
             data = inla.stack.data(stk_no2_sd, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_no2_sd),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))

# Specify formula using no2 baseline
formula_no2_baseline <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  no2_2000 +
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model with o3 std dev
model_no2_baseline <-inla(formula_no2_baseline, 
             data = inla.stack.data(stk_no2_baseline, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_no2_baseline),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))
```

INLA stack (temp)
```{r}
# Create stack
stk_temp <- inla.stack(
  data=list(y=all_data_inc$mean_temp), # the response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  #the spatial index
                               #the covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10)))
```

INLA - model fitting (no priors) for temp
```{r}
# Specify formula
formula1_temp <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model
model1_temp <-inla(formula1_temp, 
             data = inla.stack.data(stk_temp, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_temp),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))
```


######################################################################

List all models
```{r}
pm_models <- list(model1_pm, model_pm_sd, model_pm_sd_2002, model_pm_baseline)
pm_names <- c("model1_pm", "model_pm_sd", "model_pm_sd_2002", "model_pm_baseline")

o3_models <- list(model1_o3, model_o3_sd, model_o3_sd_2002, model_o3_baseline)
o3_names <- c("model1_o3", "model_o3_sd", "model_o3_sd_2002", "model_o3_baseline")

no2_models <- list(model1_no2, model_no2_sd, model_no2_baseline)
no2_names <- c("model1_no2", "model_no2_sd", "model_no2_baseline")

ndvi_models <- list(model1_ndvi, model_ndvi_baseline)
ndvi_names <- c("model1_ndvi", "model_ndvi_sd", "model_ndvi_sd_2000","model_ndvi_baseline")

temp_models <- list(model1_temp)
temp_names <- c("model1_temp")
```


Function to calculate Moran's I 
```{r}
# Function to calculate and visualize Moran's I for multiple models
analyze_spatial_autocorrelation <- function(models, model_names, observed_values, coords, 
                                           k = 5, output_dir) {
  # Create output directory if it doesn't exist
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # Initialize results dataframe
  results <- data.frame(
    model = character(),
    morans_i = numeric(),
    p_value = numeric(),
    z_score = numeric(),
    significance = character(),
    stringsAsFactors = FALSE
  )
  
  # Process each model
  for (i in 1:length(models)) {
    model <- models[[i]]
    model_name <- model_names[i]
    
    # Extract fitted values
    fitted_values <- model$summary.fitted.values$mean
    
    # Calculate residuals
    residuals <- observed_values - fitted_values[1:length(observed_values)]
    
    # Create neighbors list
    knn <- knearneigh(coords, k = k)
    nb <- knn2nb(knn)
    lw <- nb2listw(nb, style = "W")
    
    # Calculate Moran's I 
    mi_test <- moran.test(residuals, lw, zero.policy = TRUE)
    moran_i <- round(mi_test$estimate[1], 4)
    p_value <- mi_test$p.value
    z_score <- round(mi_test$statistic, 4)
    
    # Add significance 
    sig_indicator <- ifelse(p_value < 0.001, "***", 
                     ifelse(p_value < 0.01, "**",
                     ifelse(p_value < 0.05, "*", "ns")))
    
    # Create Moran scatter plot
    png(file.path(output_dir, paste0("moran_plot_", gsub(" ", "_", model_name), ".png")), 
        width = 800, height = 600)
    mp <- moran.plot(residuals, lw, main = paste("Moran's I Scatter Plot -", model_name),
                    xlab = "Residuals", ylab = "Spatially Lagged Residuals")
    
    # Add regression line with the Moran's I slope
    abline(lm(mp$wx ~ mp$x), col = "red", lwd = 2)
    
    # Add stats to plot
    legend_text <- c(
      paste("Moran's I:", moran_i),
      paste("p-value:", format.pval(p_value, digits = 4), sig_indicator),
      paste("z-score:", z_score)
    )
    
    legend("topright", legend = legend_text, bty = "n", cex = 0.9)
    
    dev.off()
    
    # Add to results dataframe
    results <- rbind(results, data.frame(
      model = model_name,
      morans_i = moran_i,
      p_value = p_value,
      z_score = z_score,
      significance = sig_indicator,
      stringsAsFactors = FALSE
    ))
  }
  
  # Save results to CSV
  write.csv(results, file.path(output_dir, "morans_i_results.csv"), row.names = FALSE)
  
  return(results)
}

# Run Moran's I analysis for all model sets
# PM models
pm_morans <- analyze_spatial_autocorrelation(
  pm_models, 
  pm_names, 
  all_data_inc$pm_change, 
  Locations, 
  k = 5, 
  output_dir = "inla_results_pm"
)

# O3 models
o3_morans <- analyze_spatial_autocorrelation(
  o3_models, 
  o3_names, 
  all_data_inc$o3_change, 
  Locations, 
  k = 5, 
  output_dir = "inla_results_o3"
)

# NO2 models
no2_morans <- analyze_spatial_autocorrelation(
  no2_models, 
  no2_names, 
  all_data_inc$no2_change, 
  Locations, 
  k = 5, 
  output_dir = "inla_results_no2"
)

# NDVI models
ndvi_morans <- analyze_spatial_autocorrelation(
  ndvi_models, 
  ndvi_names, 
  all_data_inc$ndvi_change, 
  Locations, 
  k = 5, 
  output_dir = "inla_results_ndvi"
)

# Temperature models
temp_morans <- analyze_spatial_autocorrelation(
  temp_models, 
  temp_names, 
  all_data_inc$mean_temp, 
  Locations, 
  k = 5, 
  output_dir = "inla_results_temp"
)

# Function to create summary table of all Moran's I results
create_morans_summary <- function(results_list, model_type_names) {
  combined_results <- data.frame()
  
  for (i in 1:length(results_list)) {
    results <- results_list[[i]]
    results$model_type <- model_type_names[i]
    combined_results <- rbind(combined_results, results)
  }
  
  # Reorder columns to put model_type first
  combined_results <- combined_results[, c("model_type", "model", "morans_i", 
                                          "p_value", "z_score", "significance")]
  
  # Save combined results
  write.csv(combined_results, "all_morans_i_results.csv", row.names = FALSE)
  
  return(combined_results)
}

# Create summary of all results
all_morans_summary <- create_morans_summary(
  list(pm_morans, o3_morans, no2_morans, ndvi_morans, temp_morans),
  c("PM", "O3", "NO2", "NDVI", "Temperature")
)
```
Function to analyze INLA models - summary stats + plots
(JH to do 3.3.25 - troubleshoot qqplots - right now they don't load (and figure out residuals vs spatial mean as proxy(?) for residuals)
```{r}
# Function to analyze INLA models 
analyze_inla_models <- function(models, model_names = NULL, plot_results = TRUE, 
                               include_spatial = TRUE, include_residuals = TRUE,  
                               output_dir = "inla_model_results") {
  
  # Create output directory structure
  dir.create(output_dir, showWarnings = FALSE)
  plots_dir <- file.path(output_dir, "plots")
  tables_dir <- file.path(output_dir, "tables")
  dir.create(plots_dir, showWarnings = FALSE)
  dir.create(tables_dir, showWarnings = FALSE)
  
  # Make sure model_names are available
  if (is.null(model_names)) {
    model_names <- paste0("Model", 1:length(models))
  }
  
  if (length(model_names) != length(models)) {
    warning("Number of model names doesn't match number of models. Using default names.")
    model_names <- paste0("Model", 1:length(models))
  }
  
  # Create result lists
  results <- list()
  
  # Loop through each model
  for (i in seq_along(models)) {
    model <- models[[i]]
    name <- model_names[i]
    model_results <- list()
    
    # Handle NULL models
    if (is.null(model)) {
      results[[name]] <- list(
        fixed_summary = NULL,
        hyper_summary = NULL,
        dic = NA
      )
      next
    }
    
    # Fixed effects summary
    model_results$fixed_summary <- round(model$summary.fixed, 3)
    
    # Hyperparameter summary
    model_results$hyper_summary <- round(model$summary.hyperpar, 3)
    
    # DIC 
    model_results$dic <- model$dic$dic
    
    
    # Extract spatial field info
    if (include_spatial && !is.null(model$summary.random$spatial.field)) {
      # Check if we have a spatial model and can apply inla.spde2.result
      if ("spde" %in% ls(.GlobalEnv) && !is.null(model$summary.random$spatial.field)) {
        tryCatch({
          spde_obj <- get("spde", envir = .GlobalEnv)
          model.res <- inla.spde2.result(model, 'spatial.field', spde_obj, do.transf = TRUE)
          model_results$spatial_range <- model.res$summary.log.range.nominal
          model_results$spatial_variance <- model.res$summary.log.variance.nominal
          model_results$marginals_var <- model.res$marginals.var
          model_results$marginals_kap <- model.res$marginals.kap
          model_results$marginals_range <- model.res$marginals.range
        }, error = function(e) {
          message(paste("Could not calculate spatial field parameters for model", name, "- Error:", e$message))
        })
      }
      
      # Extract spatial field mean and sd for residual analysis
      model_results$spatial_field_mean <- model$summary.random$spatial.field$mean
      model_results$spatial_field_sd <- model$summary.random$spatial.field$sd
    }
    
    # Store results
    results[[name]] <- model_results
    
    # Create plots
    if (plot_results) {
      # Create model-specific directory
      model_plots_dir <- file.path(plots_dir, name)
      dir.create(model_plots_dir, showWarnings = FALSE)
      
      # 1. Fixed effects parameter plots
      pdf(file.path(model_plots_dir, "fixed_effects_parameters.pdf"), width = 10, height = 8)
      par(mfrow = c(2, 2))
      for (j in 1:min(nrow(model$summary.fixed), 4)) {
        tryCatch({
          plot(model$marginals.fixed[[j]], ty = "l", 
               xlab = bquote(beta[.(rownames(model$summary.fixed)[j])]), 
               ylab = "Density")
        }, error = function(e) {
          message(paste("Error plotting fixed effect", j, "for model", name))
        })
      }
      dev.off()
      
      # Spatial field plots 
      if ("Mesh" %in% ls(.GlobalEnv) && !is.null(model_results$spatial_field_mean)) {
        tryCatch({
          # Spatial field mean and SD plots
          pdf(file.path(model_plots_dir, "spatial_field.pdf"), width = 10, height = 5)
          mesh_obj <- get("Mesh", envir = .GlobalEnv)
          gproj <- inla.mesh.projector(mesh_obj, dims = c(300, 300))
          g.mean <- inla.mesh.project(gproj, model_results$spatial_field_mean)
          g.sd <- inla.mesh.project(gproj, model_results$spatial_field_sd)
          
          grid.arrange(
            levelplot(g.mean, scales = list(draw = FALSE), 
                     xlab = '', ylab = '', main = 'mean', 
                     col.regions = heat.colors(16)),
            levelplot(g.sd, scales = list(draw = FALSE), 
                     xlab = '', ylab = '', main = 'sd', 
                     col.regions = heat.colors(16)),
            nrow = 1
          )
          dev.off()
        }, error = function(e) {
          message(paste("Could not create spatial field levelplot for model", name, "- Error:", e$message))
        })
      }
      
      # Hyperparameter plots
      if (!is.null(model_results$marginals_var)) {
        pdf(file.path(model_plots_dir, "hyperparameters.pdf"), width = 10, height = 4)
        par(mfrow = c(1, 3))
        tryCatch({
          # Plot variance, kappa, and range distributions
          plot(model_results$marginals_var[[1]], ty = "l", 
               xlab = expression(sigma[randomfield]^2), ylab = "Density") 
          plot(model_results$marginals_kap[[1]], type = "l", 
               xlab = expression(kappa), ylab = "Density")
          plot(model_results$marginals_range[[1]], type = "l", 
               xlab = "range nominal", ylab = "Density")
        }, error = function(e) {
          message(paste("Error creating hyperparameter plots for model", name, "- Error:", e$message))
        })
        dev.off()
      }
      
      # Residual analysis plots
      if (!is.null(model_results$spatial_field_mean)) {
        pdf(file.path(model_plots_dir, "residual_analysis.pdf"), width = 10, height = 8)
        par(mfrow = c(2, 2))
        
        # Plot time series of residuals
        plot(model_results$spatial_field_mean, type = "l", 
             main = "Spatial Field Mean", xlab = "Index", ylab = "Value")
        
        # QQ plot (not working rn)
        qqnorm(model_results$spatial_field_mean, main = "Normal Q-Q Plot")
        qqline(model_results$spatial_field_mean)
        
        # Histogram
        hist(model_results$spatial_field_mean, 
             main = "Histogram of Spatial Field Mean", 
             xlab = "Value")
        
        # Boxplot
        boxplot(model_results$spatial_field_mean, 
                main = "Boxplot of Spatial Field Mean")
        
        dev.off()
      }
    }
  }
  
  # Create comparison tables
  
  # Fixed effects table
  fixed_effects_list <- lapply(results, function(x) x$fixed_summary)
  fixed_effects_table <- create_comparison_table(fixed_effects_list, model_names, "Fixed Effects")
  
  # Hyperparameters table
  hyper_list <- lapply(results, function(x) x$hyper_summary)
  hyper_table <- create_comparison_table(hyper_list, model_names, "Hyperparameters")
  
  # Spatial parameters table
  spatial_range_list <- lapply(results, function(x) x$spatial_range)
  spatial_var_list <- lapply(results, function(x) x$spatial_variance)
  
  # Model comparison metrics table
  model_metrics <- data.frame(
    Model = model_names,
    DIC = sapply(results, function(x) {
      if (is.null(x) || is.null(x$dic) || is.na(x$dic)) {
        return("N/A")
      } else {
        return(round(x$dic, 3))
      }
    }),
    stringsAsFactors = FALSE
  )
  
  # Create spatial parameters table 
  if (any(sapply(spatial_range_list, function(x) !is.null(x)))) {
    spatial_params <- data.frame(
      Model = model_names,
      LogRange = sapply(spatial_range_list, function(x) 
        ifelse(is.null(x), "N/A", paste0(round(x[,"mean"], 3), " (", round(x[,"sd"], 3), ")"))),
      LogVariance = sapply(spatial_var_list, function(x) 
        ifelse(is.null(x), "N/A", paste0(round(x[,"mean"], 3), " (", round(x[,"sd"], 3), ")")))
    )
    
    spatial_table <- kableExtra::kbl(spatial_params, caption = "Spatial Field Parameters") %>%
      kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
  } else {
    spatial_table <- NULL
  }
  
  metrics_table <- kableExtra::kbl(model_metrics, caption = "Model Comparison Metrics") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
  
  # Save tables to files
  writeLines(as.character(fixed_effects_table), file.path(tables_dir, "fixed_effects.html"))
  writeLines(as.character(hyper_table), file.path(tables_dir, "hyperparameters.html"))
  writeLines(as.character(metrics_table), file.path(tables_dir, "model_metrics.html"))
  if (!is.null(spatial_table)) {
    writeLines(as.character(spatial_table), file.path(tables_dir, "spatial_parameters.html"))
  }
  
  # Create a combined report
  report_file <- file.path(output_dir, "analysis_report.html")
  
  report_content <- c(
    "<!DOCTYPE html>",
    "<html>",
    "<head>",
    "<title>INLA Model Analysis Report</title>",
    "<style>",
    "body { font-family: Arial, sans-serif; margin: 20px; }",
    "h1, h2 { color: #2c3e50; }",
    "table { border-collapse: collapse; width: 100%; margin-bottom: 20px; }",
    "th, td { border: 1px solid #ddd; padding: 8px; text-align: left; }",
    "th { background-color: #f2f2f2; }",
    "tr:nth-child(even) { background-color: #f9f9f9; }",
    ".model-section { margin-bottom: 30px; padding-bottom: 20px; border-bottom: 1px solid #eee; }",
    "</style>",
    "</head>",
    "<body>",
    "<h1>INLA Model Analysis Report</h1>",
    paste0("<p>Generated on: ", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "</p>"),
    "<h2>Model Comparison Metrics</h2>",
    as.character(metrics_table),
    "<h2>Fixed Effects</h2>",
    as.character(fixed_effects_table),
    "<h2>Hyperparameters</h2>",
    as.character(hyper_table)
  )
  
  if (!is.null(spatial_table)) {
    report_content <- c(report_content,
                        "<h2>Spatial Field Parameters</h2>",
                        as.character(spatial_table))
  }
  
  # Add links to individual model plots
  report_content <- c(report_content, "<h2>Model Plots</h2>", "<ul>")
  
  for (name in model_names) {
    model_plots_dir <- file.path("plots", name)
    plot_files <- list.files(file.path(plots_dir, name), pattern = "\\.pdf$|\\.png$")
    
    if (length(plot_files) > 0) {
      report_content <- c(report_content, paste0("<li><strong>", name, "</strong><ul>"))
      
      for (plot_file in plot_files) {
        plot_path <- file.path(model_plots_dir, plot_file)
        report_content <- c(report_content, 
                           paste0("<li><a href='", plot_path, "'>", 
                                 gsub("\\.(pdf|png)$", "", plot_file), "</a></li>"))
      }
      
      report_content <- c(report_content, "</ul></li>")
    }
  }
  
  report_content <- c(report_content, "</ul>", "</body>", "</html>")
  
  writeLines(report_content, report_file)
  
  # Return all results and tables
  return(list(
    results = results,
    fixed_effects_table = fixed_effects_table,
    hyper_table = hyper_table,
    metrics_table = metrics_table,
    spatial_table = spatial_table,
    output_directory = output_dir,
    report_file = report_file
  ))
}

# Function to create comparison tables
create_comparison_table <- function(data_list, model_names, caption) {
  # Check if we have data
  if (length(data_list) == 0 || all(sapply(data_list, is.null))) {
    return(kableExtra::kbl(data.frame(Message = "No data available"), caption = caption))
  }
  
  # Get all row names from all models
  all_rows <- unique(unlist(lapply(data_list, function(x) if (!is.null(x)) rownames(x) else NULL)))
  
  # If no rows found, return empty table
  if (length(all_rows) == 0) {
    return(kableExtra::kbl(data.frame(Message = "No parameters found"), caption = caption))
  }
  
  # Create empty dataframe
  result <- data.frame(Parameter = all_rows, stringsAsFactors = FALSE)
  
  # Get columns to extract (mean, sd, etc.)
  first_valid_list <- NULL
  for (dl in data_list) {
    if (!is.null(dl) && nrow(dl) > 0) {
      first_valid_list <- dl
      break
    }
  }
  
  if (!is.null(first_valid_list)) {
    cols_to_extract <- colnames(first_valid_list)
  } else {
    cols_to_extract <- c("mean", "sd", "0.025quant", "0.5quant", "0.975quant")
  }
  
  # Add data for each model
  for (i in seq_along(data_list)) {
    model_data <- data_list[[i]]
    model_name <- model_names[i]
    
    if (is.null(model_data) || nrow(model_data) == 0) {
      # Skip this model if no data
      next
    }
    
    for (col in cols_to_extract) {
      col_name <- paste0(model_name, "_", col)
      result[[col_name]] <- NA
      
      for (row in rownames(model_data)) {
        row_idx <- which(result$Parameter == row)
        if (length(row_idx) > 0) {
          result[row_idx, col_name] <- model_data[row, col]
        }
      }
    }
  }
  
  # Create kable table
  table <- kableExtra::kbl(result, caption = caption) %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
  
  # Grouped headers
  tryCatch({
    groups <- sapply(model_names, function(name) {
      cols <- grep(paste0("^", name, "_"), colnames(result), value = TRUE)
      length(cols)
    })
    
    if (all(groups > 0)) {
      table <- table %>% 
        kableExtra::add_header_above(c(" " = 1, setNames(groups, model_names)))
    }
  }, error = function(e) {
    message("Could not add grouped headers to table: ", e$message)
  })
  
  return(table)
}

# Function to run the analysis with a list of models
compare_models <- function(model_list, names = NULL, plot = TRUE, 
                          include_spatial = TRUE, include_residuals = TRUE,  
                          output_dir = "inla_model_results") {
  
  # Filter out NULL models
  valid_indices <- which(!sapply(model_list, is.null))
  
  if (length(valid_indices) == 0) {
    message("No valid models to analyze!")
    return(NULL)
  }
  
  if (length(valid_indices) < length(model_list)) {
    message("Some models are NULL and will be analyzed as such.")
  }
  
  # Run the analysis
  results <- analyze_inla_models(
    models = model_list,
    model_names = names,
    plot_results = plot,
    include_spatial = include_spatial,
    include_residuals = include_residuals,
    output_dir = output_dir
  )
  
  # Print info about where results are stored
  cat("\nAnalysis complete!\n")
  cat(paste0("Results stored in directory: ", output_dir, "\n"))
  cat(paste0("HTML report available at: ", results$report_file, "\n\n"))
  
  # Return the complete results
  return(results)
}
```

Get all results and store them in respective directories
```{r}
pm_results <- compare_models(pm_models, pm_names, output_dir = "inla_results_pm")
o3_results <- compare_models(o3_models, o3_names, output_dir = "inla_results_o3")
no2_results <- compare_models(no2_models, no2_names, output_dir = "inla_results_no2")
ndvi_results <- compare_models(ndvi_models, ndvi_names, output_dir = "inla_results_ndvi")
temp_results <- compare_models(temp_models, temp_names, output_dir = "inla_results_temp")
```
