---
title: "gam_cleaning_9.26.24"
output: html_document
date: "2024-09-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages (JH - check if any are unused after code cleaned)
```{r}
library(tidyverse)
library(tigris)
library(arrow)
library(here)
library(mgcv)
library(fixest)
library(lmtest)
library(terra)
library(exactextractr)
library(tidyterra)
library(sf)
library(mgcViz)

library(DHARMa)
library(qqplotr)
library(readxl) # for read_excel()
library(spdep) # for moran.test()
library(tidycensus) # for median income
library(distr) # for qqbounds()
library(spatialreg) # for spatial lag and error models
library(gridExtra) # for viewing plots side-by-side
library(qgam) 
```

Load and clean data
```{r}
# Check if 'data' directory exists; if not, create it
if (!dir.exists(here("data"))) {
  dir.create(here("data"))
}

# load gentrification score by tract
gentrification <- read.csv(here("data", "Data_NYC_Gentrification_2000_16.csv")) %>% 
  mutate(tractid = as.character(tractid)) %>% 
  dplyr::select(tractid, score_0.5)

gentrification <- read_xlsx(here("Gentrification","Data_NYC_Gentrification_2000_16.xlsx")) %>%

  mutate(tractid = as.character(tractid)) %>%

  dplyr::select(tractid, score_0.5)

# pull NYC tract IDs/fips from gentrification index dataset for filtering
nyc_fips <- gentrification$tractid

# load NYC tracts shapefile
nyc_tracts <- tracts(state = "NY", county = c("Bronx", "Kings", "New York", "Queens", "Richmond"), year = 2010) %>%
  # clean lat and long for Conley SEs
  mutate(lat = as.numeric(INTPTLAT10), long = as.numeric(INTPTLON10), GEOID = GEOID10) %>% 
  # filter for FIPS codes included in gentrification index dataset
  filter(GEOID %in% nyc_fips)


#load NY tracts populatuion weighted centroid points from Census data
nyc_weighted_centroids <- read.table("https://www2.census.gov/geo/docs/reference/cenpop2010/tract/CenPop2010_Mean_TR36.txt", 
                         header = TRUE, sep = ",", stringsAsFactors = FALSE, 
                         colClasses = c("character", "character", "character", "numeric", "numeric", "numeric")) %>%
  mutate(FIPS = paste(STATEFP, COUNTYFP, TRACTCE, sep = "")) %>% 
  ## filter for only NYC tracts
  filter(FIPS %in% nyc_fips) %>%  
  rename(Longitude_weighted = LONGITUDE, 
         Latitude_weighted = LATITUDE)

# JH - 10/12/24, warning persists 10/13, 10/14 Warning: R metadata may have unsafe or invalid elements ℹ Type: "externalptr" options(arrow.unsafe_metadata = TRUE) #

# temperature data
if(!file.exists(here("data","summarized_daily_temp_preds.parquet"))){download.file("https://zenodo.org/records/10557980/files/summarized_daily_temp_preds.parquet?download=1", 
                                                                                   destfile = here("data","summarized_daily_temp_preds.parquet"))}

# temperature by tract
temp <- read_parquet(here("data","summarized_daily_temp_preds.parquet")) %>%
  filter(str_detect(date, "2016")) %>%
  mutate(temp_c = mean_temp_daily_cK/100-273.15) %>% 
  mutate(temp_f = temp_c*9/5+32) %>% 
  group_by(GEOID, date) %>%
  ungroup() %>% 
  group_by(GEOID) %>% 
  summarize(mean_temp = mean(mean_temp_daily_cK/100))

# ndvi by tract (need to have it downloaded and stored locally)
ndvi_2016 <- rast(here("data", "ndvi_c2_2016_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

ndvi_2000 <- rast(here("data", "ndvi_c2_2000_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

# PM2.5 data
# download data from URL
download_data <- function() {
  # set longer timeout
  options(timeout = 7200)
  if(!file.exists(here("data", "2016_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_pm25_daily_average.txt.gz", destfile = here("data", "2016_pm25_daily_average.txt.gz"))} #2016
  if(!file.exists(here("data", "2002_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_pm25_daily_average.txt.gz", destfile = here("data", "2002_pm25_daily_average.txt.gz"))}} #2002
download_data()

# read in PM2.5 daily average by census tract from EPA FAQSD - 2002
pm_tract_2002 <- read.delim(gzfile(here("data", "2002_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## Include only NYC tracts
  filter(FIPS %in% nyc_fips) %>%
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## Calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(pm_mean_2002 = mean(pm25_daily_average.ug.m3.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# read in PM2.5 daily average by census tract from EPA FAQSD - 2016
pm_tract_2016 <- read.delim(gzfile(here("data", "2016_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(pm_mean_2016 = mean(Prediction), 
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))
  

# combine 2002 and 2016 pm data for nyc_tracts
pm_tract <- pm_tract_2002 %>%
  inner_join(pm_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# NO2 CACES
no2_2016_caces <- read.csv(here("data", "2016_no2_caces.csv")) %>% 
  mutate(no2_2016 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_2000_caces <- read.csv(here("data", "2000_no2_caces.csv")) %>% 
  mutate(no2_2000 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_caces <- left_join(no2_2000_caces, no2_2016_caces, by = "fips") %>% 
  select(no2_2000, no2_2016, fips)

#O3 data download
options(timeout = 7200)
if(!file.exists(here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))} #2016
if(!file.exists(here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))} #2002

# read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2002
o3_tract_2002 <- read.delim(gzfile(here("data", "2002_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>% 
  ## include only NYC tracts
  filter(FIPS %in% nyc_fips) %>% 
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(o3_mean_2002 = mean(ozone_daily_8hour_maximum.ppb.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2016
o3_tract_2016 <- read.delim(gzfile(here("data", "2016_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>% 
  ## calculate annual average by tract using  population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(o3_mean_2016 = mean(Prediction),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))


o3_tract <- o3_tract_2002 %>%
  left_join(o3_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# merge all exposure data
all_data <- nyc_tracts %>% 
  # join gentrification index by tract
  right_join(gentrification, by = c("GEOID10" = "tractid")) %>%
  left_join(pm_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(pm_2016 = pm_mean_2016, pm_2002 = pm_mean_2002, pm_change = pm_2016-pm_2002) %>% 
  left_join(o3_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(o3_2016 = o3_mean_2016, o3_2002 = o3_mean_2002, o3_change = o3_2016-o3_2002) %>% 
  left_join(no2_caces, by = c("GEOID" = "fips")) %>% 
  mutate(no2_change = no2_2016-no2_2000) %>% 
  mutate(ndvi_2016 = ndvi_2016, ndvi_2000 = ndvi_2000, ndvi_change = ndvi_2016-ndvi_2000) %>% 
  left_join(temp, by = "GEOID")

# models adjusting for median income in 2016
income <- get_acs(variables = "B06011_001", year = 2016, state = "NY", 
                  county = c("Queens", "King", "New York", "Richmond", "Bronx"), geography = "tract") %>% 
  rename(med_income_2016 = estimate)

all_data_inc <- all_data %>% 
  left_join(income, by = "GEOID") %>% 
  # coerce GEOID to numeric then factor class to avoid error in MRF smooth
  mutate(GEOID = as.numeric(GEOID)) %>% 
  mutate(GEOID = factor(GEOID))

##################### models adjusting for median income in 2016 and distance to downtown ###############
# load nyc roads
downtown_st <- roads(state = "NY", county = "New York", year = 2011)
ggplot() + geom_sf(data = downtown_st)

# Define bounding box
bbox_coords <- matrix(c(
  -74.0187, 40.70158,  # Southwest
  -74.0187, 40.77279,  # Northwest
  -73.95953, 40.77279, # Northeast
  -73.95953, 40.70158, # Southeast
  -74.0187, 40.70158   # Closing the polygon
), ncol = 2, byrow = TRUE)

# convert bounding box to sf
bbox_polygon <- st_sfc(st_polygon(list(bbox_coords)), crs = 4326)

# Ensure roads are the same CRS and filter for streets within bbox
downtown_st <- st_transform(downtown_st, crs = st_crs(bbox_polygon))
filtered_streets <- downtown_st[st_within(downtown_st, bbox_polygon, sparse = FALSE), ]

# Plot filtered downtown streets and bounding box
ggplot() +
  geom_sf(data = filtered_streets) +
  geom_sf(data = bbox_polygon, fill = NA, color = 'red') +
  ggtitle("Downtown Streets within Bounding Box")

# Ensure census tracts are in the same CRS
nyc_tracts <- st_transform(nyc_tracts, crs = st_crs(bbox_polygon))

# Filter intersecting tracts within downtown
intersecting_tracts <- st_intersection(nyc_tracts, bbox_polygon)
combined_area <- st_union(intersecting_tracts)
#get downtown centroid
area_centroid <- st_centroid(combined_area)

ggplot() +
  geom_sf(data = intersecting_tracts, fill = 'lightblue', color = 'black') +
  geom_sf(data = area_centroid, color = 'red', size = 3) +
  ggtitle('Centroid of Downtown Census Tracts')

# Calculate distances and add to "all_data_inc" dataframe
distances <- st_distance(nyc_tracts, area_centroid)
all_data_inc$dt_dist <- as.numeric(distances)
```
Create model functions
```{r}
# Create exposure list with proper variable names
exposure_list <- list(
  pm = list(
    name = "pm",
    baseline = "pm_2002",
    change = "pm_change"
  ),
  o3 = list(
    name = "o3",
    baseline = "o3_2002",
    change = "o3_change"
  ),
  no2 = list(
    name = "no2",
    baseline = "no2_2000",
    change = "no2_change"
  ),
  ndvi = list(
    name = "ndvi",
    baseline = "ndvi_2000",
    change = "ndvi_change"
  ),
  temp = list(
    name = "mean_temp",
    baseline = NULL,
    change = "mean_temp"  # No change variable for temp
  )
)

# Function to run GAM models
run_gam_models <- function(data, exposure_info) {
  # Extract exposure information
  exp_var <- exposure_info$change
  baseline_exposure <- exposure_info$baseline
  
  # Skip if data is missing the exposure variable
  if (!exp_var %in% names(data)) {
    warning(paste("Variable", exp_var, "not found in dataset"))
    return(NULL)
  }
  
  # Base components that will be in all models
  base_formula <- paste(exp_var, "~ s(score_0.5) + s(COUNTYFP10, bs='re')")
  
  # Create model formulas
  formulas <- list(
    # Base models (score + county + spatial)
    tp_base = paste(base_formula, "+ te(long, lat, d = 2, bs = 'gp', m = 2)"),
    mrf_base = paste(base_formula, "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),
    
    # Models with income
    tp_income = paste(base_formula, "+ s(med_income_2016) + te(long, lat, d = 2, bs = 'gp', m = 2)"),
    mrf_income = paste(base_formula, "+ s(med_income_2016) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"),
    
    # Models with distance
    tp_dist = paste(base_formula, "+ s(dt_dist) + te(long, lat, d = 2, bs = 'gp', m = 2)"),
    mrf_dist = paste(base_formula, "+ s(dt_dist) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  )
  
  # Add baseline models if baseline exists
  if (!is.null(baseline_exposure) && baseline_exposure %in% names(data)) {
    formulas$tp_baseline <- paste(base_formula, "+", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_baseline <- paste(base_formula, "+", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    # Combinations with baseline
    formulas$tp_income_baseline <- paste(base_formula, "+ s(med_income_2016) +", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_income_baseline <- paste(base_formula, "+ s(med_income_2016) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    formulas$tp_dist_baseline <- paste(base_formula, "+ s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_dist_baseline <- paste(base_formula, "+ s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
    
    formulas$tp_full <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ te(long, lat, d = 2, bs = 'gp', m = 2)")
    formulas$mrf_full <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) +", paste0("s(", baseline_exposure, ")"), "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  }
  
  # Add income + distance combinations
  formulas$tp_income_dist <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) + te(long, lat, d = 2, bs = 'gp', m = 2)")
  formulas$mrf_income_dist <- paste(base_formula, "+ s(med_income_2016) + s(dt_dist) + s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))")
  
  # Convert formulas to formula objects
  formulas <- lapply(formulas, as.formula)
  
  # Run models
  models <- purrr::map(formulas, ~try(gam(., 
                                        family = scat, 
                                        data = data, 
                                        method = "REML"), 
                                    silent = TRUE))
  
  return(models)
}

# Main execution
# First drop geometry from spatial data while keeping coordinates
data_prepared <- all_data_inc %>%
  mutate(
    COUNTYFP10 = as.factor(COUNTYFP10),
    GEOID = as.factor(GEOID)
  ) %>%
  st_drop_geometry()

# Construct neighbors list for MRF models
nb <- poly2nb(all_data_inc, row.names = all_data_inc$GEOID)
names(nb) <- attr(nb, "region.id")

# Run models for each exposure
all_models <- purrr::map(exposure_list, ~run_gam_models(data_prepared, .))

# Function to check which models succeeded/failed
check_models <- function(models) {
  model_status <- map(models, ~!inherits(., "try-error"))
  successful_models <- names(models)[unlist(model_status)]
  failed_models <- names(models)[!unlist(model_status)]
  
  return(list(
    successful = successful_models,
    failed = failed_models
  ))
}

# Check status of all models
model_status <- map(all_models, check_models)
```
display_all_model_diagnostics(all_models)
# Print model names for each exposure
for(exposure in names(all_models)) {
  cat("\nExposure:", exposure, "\n")
  print(names(all_models[[exposure]]))
}

# Check for failed models
for(exposure in names(all_models)) {
  cat("\nExposure:", exposure, "\n")
  for(model_name in names(all_models[[exposure]])) {
    if(inherits(all_models[[exposure]][[model_name]], "try-error")) {
      cat("Failed model:", model_name, "\n")
      print(all_models[[exposure]][[model_name]])
    }
  }
}

Create qqplot functions
```{r}
# Function to create QQ plots for a specific exposure
create_exposure_qq_plots <- function(models_list, exposure_name) {
  exposure_models <- models_list[[exposure_name]]
  plots <- list()
  
  for (model_name in names(exposure_models)) {
    model <- exposure_models[[model_name]]
    
    # Skip if model is an error
    if (inherits(model, "try-error")) {
      next
    }
    
    # Create QQ plot
    resid_df <- data.frame(
      residuals = residuals(model, type = "deviance"),
      fitted = fitted(model)
    )
    
    plots[[model_name]] <- ggplot(resid_df, aes(sample = residuals)) +
      stat_qq() +
      stat_qq_line(color = "red") +
      labs(
        title = paste0(exposure_name, "\n", model_name),
        x = "Theoretical Quantiles",
        y = "Sample Quantiles"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 10, hjust = 0.5),
        axis.title = element_text(size = 8),
        axis.text = element_text(size = 7)
      )
  }
  
  # Calculate grid layout
  n_plots <- length(plots)
  n_cols <- min(3, n_plots)
  n_rows <- ceiling(n_plots / n_cols)
  
  # Combine plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = plots,
    ncol = n_cols,
    nrow = n_rows,
    top = grid::textGrob(paste("QQ Plots for", exposure_name), 
                        gp = grid::gpar(fontsize = 12))
  )
  
  return(combined_plot)
}

# Function to create QQ plots for a specific model type across exposures
create_model_type_qq_plots <- function(models_list, model_type) {
  plots <- list()
  
  for (exposure_name in names(models_list)) {
    exposure_models <- models_list[[exposure_name]]
    
    # Check if the model type exists for this exposure
    if (model_type %in% names(exposure_models)) {
      model <- exposure_models[[model_type]]
      
      # Skip if model is an error
      if (inherits(model, "try-error")) {
        next
      }
      
      # Create QQ plot
      resid_df <- data.frame(
        residuals = residuals(model, type = "deviance"),
        fitted = fitted(model)
      )
      
      plots[[exposure_name]] <- ggplot(resid_df, aes(sample = residuals)) +
        stat_qq() +
        stat_qq_line(color = "red") +
        labs(
          title = paste0(exposure_name, "\n", model_type),
          x = "Theoretical Quantiles",
          y = "Sample Quantiles"
        ) +
        theme_minimal() +
        theme(
          plot.title = element_text(size = 10, hjust = 0.5),
          axis.title = element_text(size = 8),
          axis.text = element_text(size = 7)
        )
    }
  }
  
  # Calculate grid layout
  n_plots <- length(plots)
  n_cols <- min(3, n_plots)
  n_rows <- ceiling(n_plots / n_cols)
  
  # Combine plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = plots,
    ncol = n_cols,
    nrow = n_rows,
    top = grid::textGrob(paste("QQ Plots for", model_type), 
                        gp = grid::gpar(fontsize = 12))
  )
  
  return(combined_plot)
}

# Function to list available model types
list_model_types <- function(models_list) {
  unique(unlist(lapply(models_list, names)))
}

# Function to list available exposures
list_exposures <- function(models_list) {
  names(models_list)
}

create_exposure_qq_plots(all_models, "pm")
create_model_type_qq_plots(all_models, "tp_base")
list_model_types(all_models)
```

create_model_type_qq_plots(all_models, "mrf_base")
create_model_type_qq_plots(all_models, "mrf_income")
create_model_type_qq_plots(all_models, "mrf_baseline")
create_model_type_qq_plots(all_models, "mrf_dist")
create_model_type_qq_plots(all_models, "mrf_dist_baseline")
create_model_type_qq_plots(all_models, "mrf_income_baseline")
create_model_type_qq_plots(all_models, "mrf_full")


create_model_type_qq_plots(all_models, "tp_base")
create_model_type_qq_plots(all_models, "tp_income")
create_model_type_qq_plots(all_models, "tp_baseline")
create_model_type_qq_plots(all_models, "tp_income_dist")
create_model_type_qq_plots(all_models, "tp_income_baseline")
create_model_type_qq_plots(all_models, "tp_dist_baseline")
create_model_type_qq_plots(all_models, "tp_full")



```{r}
# Function to create comprehensive diagnostic plots for a specific model type
create_model_diagnostics_by_type <- function(models_list, model_type) {
  # List to store plots for each exposure
  diagnostic_plots <- list()
  
  for (exposure_name in names(models_list)) {
    exposure_models <- models_list[[exposure_name]]
    
    # Check if the model type exists for this exposure
    if (model_type %in% names(exposure_models)) {
      model <- exposure_models[[model_type]]
      
      # Skip if model is an error
      if (inherits(model, "try-error")) {
        next
      }
      
      # Extract residuals and fitted values
      resid_df <- data.frame(
        residuals = residuals(model, type = "deviance"),
        fitted = fitted(model),
        scale_residuals = residuals(model, type = "scaled.pearson"),
        response = model$y
      )
      
      # Create plot list for this exposure
      exposure_plots <- list()
      
      # 1. QQ plot
      exposure_plots[["qq"]] <- ggplot(resid_df, aes(sample = residuals)) +
        stat_qq() +
        stat_qq_line(color = "red") +
        labs(title = paste(exposure_name, "- Normal Q-Q Plot")) +
        theme_minimal() +
        theme(plot.title = element_text(size = 8))
      
      # 2. Residuals vs Fitted
      exposure_plots[["resid"]] <- ggplot(resid_df, aes(x = fitted, y = residuals)) +
        geom_point(alpha = 0.5) +
        geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
        geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
        labs(title = paste(exposure_name, "- Residuals vs Fitted"),
             x = "Fitted values",
             y = "Deviance residuals") +
        theme_minimal() +
        theme(plot.title = element_text(size = 8))
      
      # 3. Scale-location plot
      exposure_plots[["scale"]] <- ggplot(resid_df, 
                                        aes(x = fitted, y = sqrt(abs(scale_residuals)))) +
        geom_point(alpha = 0.5) +
        geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
        labs(title = paste(exposure_name, "- Scale-Location"),
             x = "Fitted values",
             y = "√|Standardized residuals|") +
        theme_minimal() +
        theme(plot.title = element_text(size = 8))
      
      # 4. Response vs Fitted
      exposure_plots[["resp"]] <- ggplot(resid_df, aes(x = fitted, y = response)) +
        geom_point(alpha = 0.5) +
        geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
        labs(title = paste(exposure_name, "- Response vs Fitted"),
             x = "Fitted values",
             y = "Response") +
        theme_minimal() +
        theme(plot.title = element_text(size = 8))
      
      # Store all plots for this exposure
      diagnostic_plots[[exposure_name]] <- exposure_plots
    }
  }
  
  # Calculate layout dimensions
  n_exposures <- length(diagnostic_plots)
  if (n_exposures == 0) {
    stop("No valid models found for the specified model type")
  }
  
  # Create a list of all plots in the desired order
  all_plots <- list()
  for (exposure_name in names(diagnostic_plots)) {
    all_plots <- c(all_plots, diagnostic_plots[[exposure_name]])
  }
  
  # Calculate grid dimensions based on number of exposures
  n_cols <- 4  # One column for each type of diagnostic plot
  n_rows <- n_exposures  # One row for each exposure
  
  # Combine all plots
  combined_plot <- gridExtra::grid.arrange(
    grobs = all_plots,
    ncol = n_cols,
    nrow = n_rows,
    top = grid::textGrob(
      paste("Diagnostic Plots for Model Type:", model_type),
      gp = grid::gpar(fontsize = 12, fontface = "bold")
    )
  )
  
  return(combined_plot)
}

# Function to display diagnostics for all model types
display_all_model_diagnostics <- function(models_list) {
  # Get unique model types
  model_types <- unique(unlist(lapply(models_list, names)))
  
  # Create plots for each model type
  for (model_type in model_types) {
    tryCatch({
      plot <- create_model_diagnostics_by_type(models_list, model_type)
      print(plot)
      # Add a pause between plots if running interactively
      if (interactive()) {
        readline(prompt = paste("Press [Enter] to see next plot for model type:", 
                              model_type))
      }
    }, error = function(e) {
      message(paste("Error creating plots for model type:", model_type))
      message(e)
    })
  }
}

# Function to get summary statistics for a specific model type
get_model_type_summary <- function(models_list, model_type) {
  summary_stats <- data.frame(
    exposure = character(),
    r_squared = numeric(),
    deviance_explained = numeric(),
    aic = numeric(),
    stringsAsFactors = FALSE
  )
  
  for (exposure_name in names(models_list)) {
    exposure_models <- models_list[[exposure_name]]
    
    if (model_type %in% names(exposure_models)) {
      model <- exposure_models[[model_type]]
      
      if (!inherits(model, "try-error")) {
        summary_stats <- rbind(summary_stats, data.frame(
          exposure = exposure_name,
          r_squared = summary(model)$r.sq,
          deviance_explained = summary(model)$dev.expl * 100,
          aic = AIC(model),
          stringsAsFactors = FALSE
        ))
      }
    }
  }
  
  return(summary_stats)
}
```


# Function to scale exposures by IQR and fit MRF income models
scale_and_fit_mrf <- function(data, exposure_list) {
  # Function to scale a variable by its IQR
  scale_by_iqr <- function(x) {
    x / (IQR(x, na.rm = TRUE))
  }
  
  # Create scaled dataset
  scaled_data <- data %>%
    mutate(
      # Scale changes
      pm_change_scaled = scale_by_iqr(pm_change),
      o3_change_scaled = scale_by_iqr(o3_change),
      no2_change_scaled = scale_by_iqr(no2_change),
      ndvi_change_scaled = scale_by_iqr(ndvi_change),
      mean_temp_scaled = scale_by_iqr(mean_temp),
      
      # Scale baselines
      pm_2002_scaled = scale_by_iqr(pm_2002),
      o3_2002_scaled = scale_by_iqr(o3_2002),
      no2_2000_scaled = scale_by_iqr(no2_2000),
      ndvi_2000_scaled = scale_by_iqr(ndvi_2000)
    )
  
  # Create modified exposure list with scaled variables
  scaled_exposure_list <- list(
    pm = list(
      name = "pm",
      baseline = "pm_2002_scaled",
      change = "pm_change_scaled"
    ),
    o3 = list(
      name = "o3",
      baseline = "o3_2002_scaled",
      change = "o3_change_scaled"
    ),
    no2 = list(
      name = "no2",
      baseline = "no2_2000_scaled",
      change = "no2_change_scaled"
    ),
    ndvi = list(
      name = "ndvi",
      baseline = "ndvi_2000_scaled",
      change = "ndvi_change_scaled"
    ),
    temp = list(
      name = "mean_temp",
      baseline = NULL,
      change = "mean_temp_scaled"
    )
  )
  
  # Function to fit MRF income model for each exposure
  fit_mrf_income <- function(data, exposure_info) {
    exp_var <- exposure_info$change
    baseline_exposure <- exposure_info$baseline
    
    if (!exp_var %in% names(data)) {
      warning(paste("Variable", exp_var, "not found in dataset"))
      return(NULL)
    }
    
    # Base formula components
    base_formula <- paste(exp_var, "~ s(score_0.5) + s(COUNTYFP10, bs='re')")
    
    # Add baseline if it exists
    if (!is.null(baseline_exposure) && baseline_exposure %in% names(data)) {
      formula <- as.formula(paste(
        base_formula,
        "+ s(med_income_2016) +",
        paste0("s(", baseline_exposure, ")"),
        "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"
      ))
    } else {
      formula <- as.formula(paste(
        base_formula,
        "+ s(med_income_2016)",
        "+ s(GEOID, bs = 'mrf', k = 5, xt = list(nb = nb))"
      ))
    }
    
    # Fit model
    model <- try(gam(formula,
                    family = scat,
                    data = data,
                    method = "REML"),
                 silent = TRUE)
    
    return(model)
  }
  
  # Fit models for each exposure
  scaled_models <- purrr::map(scaled_exposure_list, ~fit_mrf_income(scaled_data, .))
  
  # Return both scaled data and models
  return(list(
    scaled_data = scaled_data,
    models = scaled_models
  ))
}

# Usage example:
scaled_results <- scale_and_fit_mrf(data_prepared, exposure_list)

# Function to extract and summarize model results
summarize_scaled_models <- function(scaled_results) {
  map_df(names(scaled_results$models), function(exposure) {
    model <- scaled_results$models[[exposure]]
    if(inherits(model, "try-error")) {
      return(NULL)
    }
    
    summary <- summary(model)
    data.frame(
      exposure = exposure,
      r_squared = summary$r.sq,
      deviance_explained = summary$dev.expl * 100,
      n = summary$n
    )
  })
}

# Get model summaries
model_summaries <- summarize_scaled_models(scaled_results)
model_summaries

######################NEW
library(tidyverse)
library(mgcv)

analyze_gentrification_effects <- function(scaled_results) {
  # First scale the gentrification score in the data
  scaled_results$scaled_data <- scaled_results$scaled_data %>%
    mutate(score_0.5_scaled = score_0.5 / IQR(score_0.5, na.rm = TRUE))
  
  # Update models to use scaled score
  models <- scaled_results$models
  
  # Function to process each model
  process_model <- function(model, exposure_name) {
    if(inherits(model, "try-error")) {
      return(NULL)
    }
    
    # Extract smooth term for gentrification score
    smooth_term <- summary(model)$s.table["s(score_0.5_scaled)", ]
    
    # Get effective degrees of freedom (complexity of relationship)
    edf <- smooth_term["edf"]
    
    # Get p-value
    p_value <- smooth_term["p-value"]
    
    # Calculate range of fitted values for scaled gentrification score
    score_seq <- seq(min(scaled_results$scaled_data$score_0.5_scaled, na.rm = TRUE),
                    max(scaled_results$scaled_data$score_0.5_scaled, na.rm = TRUE),
                    length.out = 100)
    
    # Create prediction data frame
    pred_data <- data.frame(
      score_0.5_scaled = score_seq,
      # Set other variables to their median/reference values
      med_income_2016 = median(scaled_results$scaled_data$med_income_2016, na.rm = TRUE),
      COUNTYFP10 = levels(scaled_results$scaled_data$COUNTYFP10)[1],
      GEOID = levels(scaled_results$scaled_data$GEOID)[1]
    )
    
    # Add baseline exposure if it exists in the model
    if(exposure_name != "temp") {
      baseline_col <- paste0(tolower(exposure_name), "_2002_scaled")
      if(exposure_name == "no2") baseline_col <- "no2_2000_scaled"
      if(exposure_name == "ndvi") baseline_col <- "ndvi_2000_scaled"
      pred_data[[baseline_col]] <- median(scaled_results$scaled_data[[baseline_col]], na.rm = TRUE)
    }
    
    # Get predictions with standard errors
    pred <- predict(model, newdata = pred_data, type = "terms", se.fit = TRUE)
    
    # Extract just the gentrification score component
    score_effect <- pred$fit[, "s(score_0.5_scaled)"]
    score_se <- pred$se.fit[, "s(score_0.5_scaled)"]
    
    data.frame(
      exposure = exposure_name,
      edf = edf,
      p_value = p_value,
      score = score_seq,
      effect = score_effect,
      se_lower = score_effect - 1.96 * score_se,
      se_upper = score_effect + 1.96 * score_se
    )
  }
  
  # Apply to all models
  results <- map_df(names(models), ~process_model(models[[.]], .))
  
  return(results)
}

# Create visualization with proper axis labels
ggplot(gent_analysis, aes(x = score, y = effect)) +
  geom_line() +
  geom_ribbon(aes(ymin = se_lower, ymax = se_upper), alpha = 0.2) +
  facet_wrap(~exposure, scales = "free_y") +
  labs(x = "Gentrification Score (IQR units)",
       y = "Standardized Effect",
       title = "Relationship between Gentrification Score and Environmental Exposures",
       subtitle = "Effects shown per IQR increase in exposure and gentrification score") +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12),
    axis.title = element_text(size = 11),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )

# Updated summary table
gent_summary <- gent_analysis %>%
  group_by(exposure) %>%
  summarize(
    edf = first(edf),
    p_value = first(p_value),
    max_effect = max(effect) - min(effect),
    direction = case_when(
      max_effect > 0 ~ "Positive",
      max_effect < 0 ~ "Negative",
      TRUE ~ "No effect"
    )
  ) %>%
  arrange(p_value)

# Print formatted summary with note about scaling
print(knitr::kable(gent_summary, 
                   digits = c(0, 2, 4, 2, 0),
                   col.names = c("Exposure", "EDF", "P-value", "Max Effect Size", "Direction"),
                   caption = "Results per IQR increase in both exposure and gentrification score"))

  ########################################

create_forest_plots <- function(scaled_results) {
  # Function to extract model coefficients and CIs
  extract_model_stats <- function(model, exposure_name) {
    if(inherits(model, "try-error")) {
      return(NULL)
    }
    
    # Extract summary statistics
    model_summary <- summary(model)
    
    # Get parametric coefficients (if any)
    param_coef <- coef(model)
    param_se <- sqrt(diag(vcov(model)))
    
    # Calculate confidence intervals
    ci_lower <- param_coef - (1.96 * param_se)
    ci_upper <- param_coef + (1.96 * param_se)
    
    # Create data frame with results
    data.frame(
      exposure = exposure_name,
      estimate = param_coef[1],
      ci_lower = ci_lower[1],
      ci_upper = ci_upper[1],
      p_value = summary(model)$s.table[1, 4],
      r_squared = summary(model)$r.sq,
      deviance_explained = summary(model)$dev.expl * 100
    )
  }
  
  # Extract results for selected models (excluding temperature)
  selected_exposures <- c("pm", "o3", "no2", "ndvi")
  
  results_df <- map_df(selected_exposures, function(exp_name) {
    extract_model_stats(scaled_results$models[[exp_name]], exp_name)
  })
  
  # Create exposure labels mapping
  exposure_labels <- c(
    "pm" = "PM2.5",
    "o3" = "O3",
    "no2" = "NO2",
    "ndvi" = "NDVI"
  )
  
  # Add formatted labels
  results_df$exposure_label <- exposure_labels[results_df$exposure]
  
  # Create forest plot
  forest_plot <- ggplot(results_df, aes(y = reorder(exposure_label, estimate))) +
    # Add vertical line at null effect
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    # Add points and confidence intervals
    geom_point(aes(x = estimate), size = 3) +
    geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper), height = 0.2) +
    # Customize appearance
    theme_bw() +
    theme(
      panel.grid.minor = element_blank(),
      axis.title.y = element_blank(),
      text = element_text(size = 12),
      plot.title = element_text(size = 14, face = "bold"),
      axis.text = element_text(size = 10)
    ) +
    # Add labels
    labs(
      title = "Gent and environmental exposure",
      subtitle = "Standardized effects per interquartile range increase",
      x = "Effect Size",
      caption = "Error bars represent 95% confidence intervals"
    )
  
  # Create summary table
  summary_table <- results_df %>%
    mutate(
      exposure = exposure_label,
      estimate = round(estimate, 3),
      ci = sprintf("(%0.3f, %0.3f)", ci_lower, ci_upper),
      p_value = sprintf("%0.3f", p_value),
      r_squared = sprintf("%0.1f%%", r_squared * 100),
      deviance_explained = sprintf("%0.1f%%", deviance_explained)
    ) %>%
    select(exposure, estimate, ci, p_value, r_squared, deviance_explained)
  
  return(list(
    plot = forest_plot,
    summary = summary_table
  ))
}

# Usage example:
results <- create_forest_plots(scaled_results)

# Display plot
print(results$plot)

# Display summary table
print(results$summary)
  
  ###################################
  
  # Function to create forest plot data
create_forest_plot_data <- function(gent_summary) {
  # Add confidence intervals (placeholder - you may want to extract these from your models)
  forest_data <- gent_summary %>%
    mutate(
      # Create labels that include EDF and p-value
      label = paste0(exposure, "\n(EDF = ", sprintf("%.2f", edf), 
                    ", p = ", sprintf("%.3f", p_value), ")"),
      # Order exposures by effect size
      label = factor(label, levels = label[order(max_effect)])
    )
  
  return(forest_data)
}

# Create basic forest plot
create_basic_forest_plot <- function(forest_data) {
  ggplot(forest_data, aes(y = label, x = max_effect)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    geom_point(size = 3, color = "darkblue") +
    theme_minimal() +
    theme(
      axis.title.y = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      text = element_text(size = 12)
    ) +
    labs(
      x = "Maximum Effect Size (IQR units)",
      title = "Environmental Exposure Changes Associated with Gentrification",
      subtitle = "Effect sizes shown in interquartile range (IQR) units"
    )
}

# Create enhanced forest plot with confidence intervals and EDF visualization
create_enhanced_forest_plot <- function(forest_data) {
  # Create base plot
  p1 <- ggplot(forest_data, aes(y = label, x = max_effect)) +
    # Add reference line
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    # Add points
    geom_point(aes(size = edf), color = "darkblue") +
    # Add color gradient based on p-value
    scale_size_continuous(name = "EDF\n(complexity)", 
                         range = c(3, 8)) +
    # Customize theme
    theme_minimal() +
    theme(
      axis.title.y = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      text = element_text(size = 12),
      legend.position = "right"
    ) +
    labs(
      x = "Maximum Effect Size (IQR units)",
      title = "Environmental Exposure Changes Associated with Gentrification",
      subtitle = "Effect sizes shown in interquartile range (IQR) units\nPoint size indicates relationship complexity (EDF)"
    )
  
  return(p1)
}

# Create plot showing standardized effects across gentrification scores
create_trajectory_plot <- function(gent_analysis) {
  ggplot(gent_analysis, aes(x = score, y = effect, color = exposure)) +
    geom_line(size = 1) +
    geom_ribbon(aes(ymin = se_lower, ymax = se_upper, fill = exposure), 
                alpha = 0.2, color = NA) +
    theme_minimal() +
    theme(
      legend.position = "right",
      panel.grid.minor = element_blank()
    ) +
    labs(
      x = "Gentrification Score",
      y = "Standardized Effect (IQR units)",
      title = "Trajectories of Environmental Exposures Across Gentrification Scores",
      subtitle = "Shaded areas represent 95% confidence intervals"
    )
}

# Generate plots
forest_data <- create_forest_plot_data(gent_summary)
basic_plot <- create_basic_forest_plot(forest_data)
enhanced_plot <- create_enhanced_forest_plot(forest_data)
trajectory_plot <- create_trajectory_plot(gent_analysis)

# Display plots
print(basic_plot)
print(enhanced_plot)
print(trajectory_plot)

# Save plots if needed
#ggsave("basic_forest_plot.png", basic_plot, width = 10, height = 6)
#ggsave("enhanced_forest_plot.png", enhanced_plot, width = 10, height = 6)
#ggsave("trajectory_plot.png", trajectory_plot, width = 12, height = 6)
