---
title: "Environmental Gentrification - INLA Analysis Overview"
author: "Joyce Hu"
date: "2025-4-18"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, progress = FALSE)
```

# Project overview
This project examines the relationship between gentrification and environmental changes in New York City between 2000 and 2016, investigating whether census tracts experiencing higher levels of gentrification are associated with patterns of environmental improvement or degradation. This analysis informs the basis of a second part of the project which evaluates the relationship between environmental gentrification (as a combined exposure) and the change in prevalence of health-relevant endpoints.

## Research questions
The first research question we seek to answer with this project is, “does environmental gentrification impact New York City?” More specifically, do areas undergoing gentrification experience different environmental changes compared to non-gentrifying areas?

## Data inputs
We use a census tract-level index of gentrification (Johnson et al., 2021) as the independent variable and examines its relationship with several environmental factors (dependent variables): Air quality indicators (PM2.5,O3, NO2), vegetation measured through NDVI - Normalized Difference Vegetation Index, mean temperature in 2016.

<details>
<summary><h3>Click to expand - background on environmental gentrification and hypothesis</h3></summary>
Although gentrification has been differentially conceptualized in the literature, it is broadly recognized as the process where wealthier populations enter a neighborhood of historical disinvestment, contributing to the displacement of existing residents via increasing prices of housing, goods, services. In analyses, the following characteristics are often linked to gentrification: capital reinvestment in previously disinvested neighborhoods, demographic shifts with influx of higher-income residents, physical landscape transformation, changes in commercial services and amenities.

Green or environmental gentrification is the idea that initiatives and policies seeking to improve environmental conditions (e.g., urban greening like parks and street trees, air pollution and brownfield site remediation) may contribute to, or in some cases even drive, gentrification e.g., driving up the housing prices in a neighborhood. This has implications for environmental and health justice. For example, urban green space is associated with improved mental health and increased physical activity but if greening initiatives also contribute to gentrification, existing residents may not experience these health benefits.

We hypothesize that NYC census tracts with higher gentrification levels between 2000-2016 will be associated with: decreased concentrations of PM2.5 and NO2, increased vegetation (higher NDVI values), smaller increases in ozone (O3) concentrations, lower mean temperatures
</details>

<details>
<summary><h3> Load packages </h3></summary>
```{r}
library(tidyverse)
library(here) 
library(arrow)          # Reading parquet files
library(readxl)         # Reading excel files
library(tidycensus)     # Census data access
library(tigris)         # Census tract boundaries
library(terra)          # Raster processing
library(exactextractr)  # Raster aerial interpolation
library(tidyterra)      # For terra objects in tidyverse
library(sf)             # Spatial data 
library(spdep)          # Spatial dependence tests
library(spatialreg)     # Spatial regression models
library(INLA)           # Integrated nested laplace approximation models

# Unused packages for this code (keeping for now)
# library(mgcv) library(fixest) library(lmtest) library(mgcViz) library(DHARMa) library(qqplotr)
# library(lattice)library(distr) #for qqbounds() 
# library(gridExtra) library(qgam)  library(ggfields) # for plotting spatial field
# library(ggregplot) #INLA compatible plotting library (efxplot)
# library(kableExtra)library(inlatools) # for INLA diagnostics
```
</details>


# Data acquisition and processing
This section loads and prepares all the necessary datasets for analysis.
We combine NYC census tract boundaries with:
1. Gentrification scores (2000-2016) sourced from [Johnson et al.'s (2021) gentrifiction index](https://doi.org/10.1080/13658816.2021.1931873) 2. Air pollutant measures (PM2.5, NO2, O3) 3. Normalized Difference Vegetation Index (NDVI) as the measure of vegetative greening  4. Temperature data (not used in final analysis) 5. Median income

<details>
<summary>Click to expand</summary>
```{r}
# Check if 'data' directory exists; if not, create it
if (!dir.exists(here("data"))) {
  dir.create(here("data"))
}

# Load gentrification score by tract
gentrification <- read.csv(here("data", "Data_NYC_Gentrification_2000_16.csv")) %>% 
  mutate(tractid = as.character(tractid)) %>% 
  dplyr::select(tractid, score_0.5)

# Load gentrification score if using xlsx file:
# gentrification <- read_xlsx(here("Gentrification","Data_NYC_Gentrification_2000_16.xlsx")) %>%
#   mutate(tractid = as.character(tractid)) %>%
#   dplyr::select(tractid, score_0.5)

# Pull NYC tract IDs/fips from gentrification index dataset for filtering
nyc_fips <- gentrification$tractid

# Load NYC tracts shapefile
nyc_tracts <- tracts(state = "NY", county = c("Bronx", "Kings", "New York", "Queens", "Richmond"), year = 2010) %>%
  # clean lat and long for Conley SEs
  mutate(lat = as.numeric(INTPTLAT10), long = as.numeric(INTPTLON10), GEOID = GEOID10) %>% 
  # filter for FIPS codes included in gentrification index dataset
  filter(GEOID %in% nyc_fips)


# Load NY tracts populatuion weighted centroid points from Census data
nyc_weighted_centroids <- read.table("https://www2.census.gov/geo/docs/reference/cenpop2010/tract/CenPop2010_Mean_TR36.txt", 
                         header = TRUE, sep = ",", stringsAsFactors = FALSE, 
                         colClasses = c("character", "character", "character", "numeric", "numeric", "numeric")) %>%
  mutate(FIPS = paste(STATEFP, COUNTYFP, TRACTCE, sep = "")) %>% 
  ## filter for only NYC tracts
  filter(FIPS %in% nyc_fips) %>%  
  rename(Longitude_weighted = LONGITUDE, 
         Latitude_weighted = LATITUDE)

# JH - 10/12/24 warning persists 10/13, 10/14 Warning: R metadata may have unsafe or invalid elements ℹ Type: "externalptr" options(arrow.unsafe_metadata = TRUE) #

# Temperature data
if(!file.exists(here("data","summarized_daily_temp_preds.parquet"))){download.file("https://zenodo.org/records/10557980/files/summarized_daily_temp_preds.parquet?download=1", 
                                                                                   destfile = here("data","summarized_daily_temp_preds.parquet"))}

# Temperature by tract
temp <- read_parquet(here("data","summarized_daily_temp_preds.parquet")) %>%
  filter(str_detect(date, "2016")) %>%
  mutate(temp_c = mean_temp_daily_cK/100-273.15) %>% 
  mutate(temp_f = temp_c*9/5+32) %>% 
  group_by(GEOID, date) %>%
  ungroup() %>% 
  group_by(GEOID) %>% 
  summarize(mean_temp = mean(mean_temp_daily_cK/100))

# NDVI by tract (need to have it downloaded and stored locally)
ndvi_mean_2016 <- rast(here("data", "ndvi_c2_2016_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

ndvi_mean_2000 <- rast(here("data", "ndvi_c2_2000_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "mean")

ndvi_sd_2016 <- rast(here("data", "ndvi_c2_2000_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "stdev")

ndvi_sd_2000 <- rast(here("data", "ndvi_c2_2016_2.tif")) %>% 
  exact_extract(nyc_tracts, fun = "stdev")

# PM2.5 data
# Download data from URL
download_data <- function() {
  # set longer timeout
  options(timeout = 7200)
  if(!file.exists(here("data", "2016_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_pm25_daily_average.txt.gz", destfile = here("data", "2016_pm25_daily_average.txt.gz"))} #2016
  if(!file.exists(here("data", "2002_pm25_daily_average.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_pm25_daily_average.txt.gz", destfile = here("data", "2002_pm25_daily_average.txt.gz"))}} #2002
download_data()

# Read in PM2.5 daily average by census tract from EPA FAQSD - 2002
pm_tract_2002 <- read.delim(gzfile(here("data", "2002_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## Include only NYC tracts
  filter(FIPS %in% nyc_fips) %>%
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## Calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(pm_mean_2002 = mean(pm25_daily_average.ug.m3.),
            pm_sd_2002 = sd(pm25_daily_average.ug.m3.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# Read in PM2.5 daily average by census tract from EPA FAQSD - 2016
pm_tract_2016 <- read.delim(gzfile(here("data", "2016_pm25_daily_average.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(pm_mean_2016 = mean(Prediction),
            pm_sd_2016 = sd(Prediction),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))
  
# Combine 2002 and 2016 pm data for nyc_tracts
pm_tract <- pm_tract_2002 %>%
  inner_join(pm_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# NO2 CACES
no2_2016_caces <- read.csv(here("data", "2016_no2_caces.csv")) %>% 
  mutate(no2_2016 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_2000_caces <- read.csv(here("data", "2000_no2_caces.csv")) %>% 
  mutate(no2_2000 = as.numeric(pred_wght)) %>%  
  mutate(fips = as.character(fips))

no2_caces <- left_join(no2_2000_caces, no2_2016_caces, by = "fips") %>% 
  select(no2_2000, no2_2016, fips)

#O3 data download
options(timeout = 7200)
if(!file.exists(here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2016_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2016_ozone_daily_8hour_maximum.txt.gz"))} #2016
if(!file.exists(here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))){download.file("https://ofmpub.epa.gov/rsig/rsigserver?data/FAQSD/outputs/2002_ozone_daily_8hour_maximum.txt.gz", destfile = here("data", "2002_ozone_daily_8hour_maximum.txt.gz"))} #2002

# Read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2002
o3_tract_2002 <- read.delim(gzfile(here("data", "2002_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>% 
  ## include only NYC tracts
  filter(FIPS %in% nyc_fips) %>% 
  ## Convert FIPS to character for compatibility
  mutate(FIPS = as.character(FIPS)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = "FIPS") %>%
  ## calculate annual average by tract using population weighted centroid
  group_by(FIPS) %>%
  summarize(o3_mean_2002 = mean(ozone_daily_8hour_maximum.ppb.),
            o3_sd_2002 = sd(ozone_daily_8hour_maximum.ppb.),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

# Read in ozone daily 8-hour maximum by census tract from EPA FAQSD (external file) - 2016
o3_tract_2016 <- read.delim(gzfile(here("data", "2016_ozone_daily_8hour_maximum.txt.gz")), sep = ",") %>%
  ## include only NYC tracts
  filter(Loc_Label1 %in% nyc_fips) %>% 
  ## Convert Loc_Label1 to character for compatibility
  mutate(Loc_Label1 = as.character(Loc_Label1)) %>%
  ## Join with population-weighted centroids
  left_join(nyc_weighted_centroids, by = c("Loc_Label1" = "FIPS")) %>% 
  ## calculate annual average by tract using  population weighted centroid
  group_by(Loc_Label1) %>%
  summarize(o3_mean_2016 = mean(Prediction),
            o3_sd_2016 = sd(Prediction),
            Longitude = first(Longitude_weighted),
            Latitude = first(Latitude_weighted))

o3_tract <- o3_tract_2002 %>%
  left_join(o3_tract_2016, by = c("FIPS" = "Loc_Label1", "Longitude", "Latitude")) %>% 
  mutate(FIPS = as.character(FIPS))

# Merge all exposure data
all_data <- nyc_tracts %>% 
  # join gentrification index by tract
  right_join(gentrification, by = c("GEOID10" = "tractid")) %>%
  left_join(pm_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(pm_2016 = pm_mean_2016, pm_2002 = pm_mean_2002, pm_change = pm_2016-pm_2002) %>% 
  left_join(o3_tract, by = c("GEOID" = "FIPS"))  %>% 
  mutate(o3_2016 = o3_mean_2016, o3_2002 = o3_mean_2002, o3_change = o3_2016-o3_2002, o3_sd_2016 = o3_sd_2016, o3_sd_2002 = o3_sd_2002) %>% 
  left_join(no2_caces, by = c("GEOID" = "fips")) %>% 
  mutate(no2_change = no2_2016-no2_2000) %>% 
  mutate(ndvi_mean_2016 = ndvi_mean_2016, ndvi_mean_2000 = ndvi_mean_2000, ndvi_change = ndvi_mean_2016-ndvi_mean_2000) %>%
  mutate(ndvi_sd_2016 = ndvi_sd_2016, ndvi_sd_2000 = ndvi_sd_2000) %>%
  left_join(temp, by = "GEOID")


############# Median income ###############################
# Find median household income variable in 2000 decennial census
all_vars_dec <- 
  load_variables(year = 2000, dataset = "sf3")

# Models adjusting for median income in 2000 (note: 2000 decennial census is 1999 data)
income_2000 <- get_decennial(variables = "P053001", year = 2000,  sumfile = "sf3", state = "NY", 
                  county = c("Queens", "King", "New York", "Richmond", "Bronx"), geography = "tract") %>% 
  rename(med_income_2000 = value)
#results in 226 missing tracts in terms of 2010 tracts.

# Find median household income variable in 2010 American Community Survey
income_2010 <- get_acs(variables = "B06011_001", year = 2010, state = "NY", 
                  county = c("Queens", "King", "New York", "Richmond", "Bronx"), geography = "tract") %>% 
  rename(med_income_2010 = estimate)

# Joining both income sets with data and imputing 2010 data where 2000 data is missing
all_data_inc <- all_data %>%
  left_join(income_2000, by = "GEOID") %>%
  left_join(income_2010, by = "GEOID") %>%
  mutate(med_income_2000_impute = coalesce(med_income_2000, med_income_2010)) %>% 
  # coerce GEOID to numeric then factor class to avoid error in MRF smooth
  mutate(GEOID = as.numeric(GEOID)) %>% 
  mutate(GEOID = factor(GEOID)) %>% 
 # coerce COUNTYFP10 to numeric then factor class to avoid error ?
  mutate(COUNTYFP10 = as.numeric(COUNTYFP10)) %>% 
  mutate(COUNTYFP10 = factor(COUNTYFP10))

# Ranking median income 
all_data_inc$med_income_ranked <- rank(all_data_inc$med_income_2000_impute)


# Converting median income to deciles and scaling
# Create deciles
income_quantiles <- quantile(all_data_inc$med_income_2000_impute, probs = seq(0, 1, by = 0.1))
all_data_inc$med_income_decile <- findInterval(all_data_inc$med_income_2000_impute, 
                                             income_quantiles,
                                             rightmost.closed = TRUE)
```
</details>


# Spatial Modeling with INLA
### We use Integrated Nested Laplace Approximation (INLA) to fit Bayesian spatial models that account for spatial autocorrelation in our variables.

## INLA set up for all variables
Currently, need to use INLA version 21.02.23 to be compatible with R version 4.2
<details>
<summary>Click to expand</summary>
```{r}
# Set INLA repository
options(repos = c(getOption("repos"),
   INLA="https://inla.r-inla-download.org/R/stable"))
# 
# # Install INLA and dependencies (from CRAN)
# install.packages("INLA", dep = TRUE)

# Need to download an older version of INLA to be compatible with HPC R version and GLIBC compatibility

# remotes::install_version("INLA", 
#                          version="21.02.23", 
#                          repos=c(getOption("repos"),
#                                  INLA="https://inla.r-inla-download.org/R/stable"), 
#                          dep=TRUE)

```
</details>


## Creating spatial mesh for INLA (to be used for all variables)
```{r}
graphics.off()

# Create matrix of population-weighted tract centroid locations
Locations <- as.matrix(cbind(
  as.numeric(nyc_weighted_centroids$Longitude_weighted),
  as.numeric(nyc_weighted_centroids$Latitude_weighted)
))

# Create mesh options for spatial modeling
# Various mesh option specifications 
MeshA <- inla.mesh.2d(jitter(Locations), max.edge = c(20, 40))
MeshB <- inla.mesh.2d(Locations, max.edge = c(20, 40))
MeshC <- inla.mesh.2d(Locations, max.edge = c(50, 100))
mesh_simple <- inla.mesh.create(Locations)

# Visualize mesh structure options
plot(MeshA)
points(Locations, col = "red", pch = 2)

plot(MeshB)
points(Locations, col = "red", pch = 2)

plot(MeshC)
points(Locations, col = "red", pch = 2)


# "Generally, the mesh function automatically creates a mesh like mesh A, where closer-together sampling locations produce smaller triangles. When exploring/setting up preliminary analyses, use a mesh like mesh B. For analyses to be reported in a paper, use a mesh like mesh C. Be careful of edges, and try to allow some space around your sampling area for INLA to estimate. The edge triangles can be bigger to reduce computing power."

# Set chosen mesh to use for analysis
Mesh <- MeshB
```

## INLA - Creating the A Matrix and Stochastic Partial Differential Equation (SPDE) object (to be used for all variables)
```{r}
# Create A matrix
A <- inla.spde.make.A(Mesh, loc = Locations) 
# "the A matrix maps the Gaussian Markov Random Field (GMRF) from the mesh nodes to the n observation location. This is represented by a matrix with the number of observations as column and the number of nodes as rows."


# Create spatial structure (SPDE object)
spde <- inla.spde2.matern(Mesh, alpha=2) # Tutorial fixed alpha as 2

# Create all the required indexes for the SPDE model, naming the spatial field as “spatial.field”, which facilitates the extraction of the estimated values of the spatial field after the fitting process.
iset <- inla.spde.make.index(name = "spatial.field", spde$n.spde)
```

## Create INLA stacks
INLA uses a concept called "stacks" to organize data for spatial modeling. Each stack contains the response variable, spatial projector matrices, and covariates i.e., a separate stack needs to be created for each of the environmental variables. 

Each stack is specific to the variables included in the model. The final model we have chosen incorporates the standard deviations for our environmental variables.

```{r}
# Create INLA stack for PM2.5
stk_pm_sd <- inla.stack(
  data=list(y=all_data_inc$pm_change), # PM2.5 change 2002-2016 as response
                  A=list(A,1),  #the A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), #the Intercept
                                 iset),  # The spatial index
                               # Covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 pm_sd_2002 = all_data_inc$pm_sd_2002, #2002 std devs
                                 pm_sd_2016 = all_data_inc$pm_sd_2016))) # 2016 std devs

# Create INLA stack for NDVI
stk_ndvi_sd <- inla.stack(
  data=list(y=all_data_inc$ndvi_change), 
                  A=list(A,1),  
                  effects=list(c(list(Intercept=1), 
                                 iset),  
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 ndvi_sd_2000 = all_data_inc$ndvi_sd_2000,
                                 ndvi_sd_2016 = all_data_inc$ndvi_sd_2016)))
# Create INLA stack for O3
stk_o3_sd <- inla.stack(
  data=list(y=all_data_inc$o3_change), 
                  A=list(A,1),  
                  effects=list(c(list(Intercept=1), 
                                 iset),  
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 o3_sd_2002 = all_data_inc$o3_sd_2002, # 2002 O3 std dev
                                 o3_sd_2016 = all_data_inc$o3_sd_2016))) # 2016 O3 std dev

# The NO2 data did not come with standard deviations. Here, we try using O3 standard deviations as a proxy as O3 and NO2 often have an inverse relationship. 
stk_no2_sd <- inla.stack(
  data=list(y=all_data_inc$no2_change), 
                  A=list(A,1),  
                  effects=list(c(list(Intercept=1), 
                                 iset),  
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10,
                                 o3_sd_2002 = all_data_inc$o3_sd_2002, #2002 O3 std dev
                                 o3_sd_2016 = all_data_inc$o3_sd_2016))) #2016 O3 std dev

# Create stack for temp - no standard deviation
stk_temp <- inla.stack(
  data=list(y=all_data_inc$mean_temp), # The response - 2016 mean temperatures
                  A=list(A,1),  # The A matrix; the 1 is included to make the list (covariates)
                  effects=list(c(list(Intercept=1), # The intercept
                                 iset),  # The spatial index
                               #Covariates
                               list(
                                 score_0.5 = all_data_inc$score_0.5,
                                 med_income_decile = all_data_inc$med_income_decile,
                                 COUNTYFP10 = all_data_inc$COUNTYFP10)))

# Note: The A matrix structure consists of two components:
# 1. The projector matrix for the spatial random field
# 2. A one-to-one mapping for fixed effect covariates (indicated by the '1')

```
 

## INLA Model Fitting

### Model fitting for PM2.5
```{r}
# Specify formula
formula_pm_sd <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  pm_sd_2002 +                          # Standard deviation 2002 
  pm_sd_2016 +                          # Standard deviation 2016
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid")

#Specify model 
model_pm_sd <- inla(formula_pm_sd, 
             data = inla.stack.data(stk_pm_sd, spde=spde),
             family = 'gaussian', 
             control.predictor = list(A = inla.stack.A(stk_pm_sd), compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))
```

### Model fitting for NDVI
<details>
<summary>Click to expand</summary>
```{r}
# Specify formula
formula_ndvi_sd <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  ndvi_sd_2000 + # Standard deviation 2000
  ndvi_sd_2016 + # Standard deviation 2016
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") 

#Specify model 
model_ndvi_sd <-inla(formula_ndvi_sd, 
             data = inla.stack.data(stk_ndvi_sd, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_ndvi_sd),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))
```
</details>

### Model fitting for O3
<details>
<summary>Click to expand</summary>
```{r}
# Specify formula
formula_o3_sd <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  o3_sd_2002 + # Standard deviation 2002
  o3_sd_2016 + # Standard deviation 2016
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") #random effect
 
# Specify model
model_o3_sd <-inla(formula_o3_sd, 
             data = inla.stack.data(stk_o3_sd, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_o3_sd),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))

```
</details>

### Model fitting for NO2
<details>
<summary>Click to expand</summary>
```{r}
# Specify formula
formula_no2_sd <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  o3_sd_2002 +
  o3_sd_2016 +
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") 
 
# Specify model 
model_no2_sd<-inla(formula_no2_sd, 
             data = inla.stack.data(stk_no2_sd, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_no2_sd),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))

```
</details>

### Model fitting for mean temperature (2016)
<details>
<summary>Click to expand</summary>
```{r}
formula1_temp <- y ~ -1 + Intercept + score_0.5 +
  med_income_decile + 
  f(spatial.field, model = spde) +
  f(COUNTYFP10, model = "iid") 
 
model1_temp <-inla(formula1_temp, 
             data = inla.stack.data(stk_temp, spde=spde),
             family= 'gaussian', 
             control.predictor=list(A = inla.stack.A(stk_temp),compute = TRUE), 
             control.compute = list(dic = TRUE, cpo = TRUE))
```
</details>

## INLA model results and diagnostics
This code defines functions to analyze spatial autocorrelation using Moran's I statistic to evaluate residual spatial autocorrelation. Next, it creates functions to analyze the INLA models, extracting fixed effects, hyperparameters, and spatial field characteristics, generating diagnostic plots and tables for model evaluation. 
<details>
<summary>Click to expand</summary>
```{r}
# List the available models (one per variable)
pm_models <- list(model_pm_sd)
pm_names <- c("model_pm_sd")

o3_models <- list(model_o3_sd)
o3_names <- c("model_o3_sd")

no2_models <- list(model_no2_sd)
no2_names <- c("model_no2_sd")

ndvi_models <- list(model_ndvi_sd)
ndvi_names <- c("model_ndvi_sd")

temp_models <- list(model1_temp)
temp_names <- c("model1_temp")

# Function to calculate Moran's I 
analyze_spatial_autocorrelation <- function(models, model_names, observed_values, coords, k = 5) {
  # Initialize results dataframe
  results <- data.frame(
    model = character(),
    morans_i = numeric(),
    p_value = numeric(),
    z_score = numeric(),
    significance = character(),
    stringsAsFactors = FALSE
  )
  
  # Process each model
  for (i in 1:length(models)) {
    model <- models[[i]]
    model_name <- model_names[i]
    
    # Skip NULL models
    if (is.null(model)) {
      next
    }
    
    # Extract fitted values
    fitted_values <- model$summary.fitted.values$mean
    
    # Calculate residuals
    residuals <- observed_values - fitted_values[1:length(observed_values)]
    
    # Create neighbors list
    knn <- knearneigh(coords, k = k)
    nb <- knn2nb(knn)
    lw <- nb2listw(nb, style = "W")
    
    # Calculate Moran's I 
    mi_test <- moran.test(residuals, lw, zero.policy = TRUE)
    moran_i <- round(mi_test$estimate[1], 4)
    p_value <- mi_test$p.value
    z_score <- round(mi_test$statistic, 4)
    
    # Add significance 
    sig_indicator <- ifelse(p_value < 0.001, "***", 
                     ifelse(p_value < 0.01, "**",
                     ifelse(p_value < 0.05, "*", "ns")))
    
    # Create Moran scatter plot 
    plot_title <- paste("Moran's I Scatter Plot -", model_name)
    mp <- moran.plot(residuals, lw, main = plot_title,
                    xlab = "Residuals", ylab = "Spatially Lagged Residuals")
    
    # Add regression line with the Moran's I slope
    abline(lm(mp$wx ~ mp$x), col = "red", lwd = 2)
    
    # Add stats to plot
    legend_text <- c(
      paste("Moran's I:", moran_i),
      paste("p-value:", format.pval(p_value, digits = 4), sig_indicator),
      paste("z-score:", z_score)
    )
    
    legend("topright", legend = legend_text, bty = "n", cex = 0.9)
    
    # Add to results dataframe
    results <- rbind(results, data.frame(
      model = model_name,
      morans_i = moran_i,
      p_value = p_value,
      z_score = z_score,
      significance = sig_indicator,
      stringsAsFactors = FALSE
    ))
  }
  
  return(results)
}

# Function to analyze INLA models - simplified for single model analysis
analyze_inla_models <- function(models, model_names = NULL, include_spatial = TRUE) {
  # Make sure model_names are available
  if (is.null(model_names)) {
    model_names <- paste0("Model", 1:length(models))
  }
  
  if (length(model_names) != length(models)) {
    warning("Number of model names doesn't match number of models. Using default names.")
    model_names <- paste0("Model", 1:length(models))
  }
  
  # Create result lists
  results <- list()
  
  # Loop through each model
  for (i in seq_along(models)) {
    model <- models[[i]]
    name <- model_names[i]
    model_results <- list()
    
    # Handle NULL models
    if (is.null(model)) {
      results[[name]] <- list(
        fixed_summary = NULL,
        hyper_summary = NULL,
        dic = NA
      )
      next
    }
    
    # Fixed effects summary
    model_results$fixed_summary <- round(model$summary.fixed, 3)
    
    # Hyperparameter summary
    model_results$hyper_summary <- round(model$summary.hyperpar, 3)
    
    # DIC 
    model_results$dic <- model$dic$dic
    
    # Extract spatial field info
    if (include_spatial && !is.null(model$summary.random$spatial.field)) {
      if ("spde" %in% ls(.GlobalEnv) && !is.null(model$summary.random$spatial.field)) {
        tryCatch({
          spde_obj <- get("spde", envir = .GlobalEnv)
          model.res <- inla.spde2.result(model, 'spatial.field', spde_obj, do.transf = TRUE)
          model_results$spatial_range <- model.res$summary.log.range.nominal
          model_results$spatial_variance <- model.res$summary.log.variance.nominal
          model_results$marginals_var <- model.res$marginals.var
          model_results$marginals_kap <- model.res$marginals.kap
          model_results$marginals_range <- model.res$marginals.range
        }, error = function(e) {
          message(paste("Could not calculate spatial field parameters for model", name, "- Error:", e$message))
        })
      }
      
      # Extract spatial field mean and sd for residual analysis
      model_results$spatial_field_mean <- model$summary.random$spatial.field$mean
      model_results$spatial_field_sd <- model$summary.random$spatial.field$sd
    }
    
    # Store results
    results[[name]] <- model_results
  }
  
  # Return the results
  return(results)
}

# Function to plot model diagnostics 
plot_model_diagnostics <- function(model, model_name) {
  if (is.null(model)) {
    cat("Model is NULL, cannot generate diagnostics\n")
    return(invisible(NULL))
  }
  
  if (!is.null(model$summary.random$spatial.field)) {
    par(mfrow = c(2, 2))
    
    # Extract spatial field mean as proxy for residuals
    residuals <- model$summary.random$spatial.field$mean
    
    # Time series plot of residuals
    plot(residuals, type = "p", main = paste("Spatial Field Values -", model_name), 
         xlab = "Index", ylab = "Value", pch = 19, cex = 0.5)
    abline(h = 0, col = "red", lty = 2)
    
    # QQplot
    qqnorm(residuals, main = paste("Normal Q-Q Plot -", model_name))
    qqline(residuals, col = "red")
    
    # Histogram
    hist(residuals, main = paste("Histogram -", model_name), 
         xlab = "Spatial Field Value", breaks = 20, 
         col = "lightblue", border = "white")
    
    # Boxplot
    boxplot(residuals, main = paste("Boxplot -", model_name), 
            ylab = "Spatial Field Value", col = "lightblue")
    
    par(mfrow = c(1, 1))
    return(invisible(residuals))
  } else {
    cat("No spatial field available for this model\n")
    return(invisible(NULL))
  }
}

# Function to create a summary table of all Moran's I results
create_morans_summary <- function(results_list, model_type_names) {
  combined_results <- data.frame()
  
  for (i in 1:length(results_list)) {
    results <- results_list[[i]]
    results$model_type <- model_type_names[i]
    combined_results <- rbind(combined_results, results)
  }
  
  # Reorder columns to put model_type first
  combined_results <- combined_results[, c("model_type", "model", "morans_i", 
                                          "p_value", "z_score", "significance")]
  
  return(combined_results)
}

# Run Moran's I analysis for all models
# PM model
cat("## Moran's I Analysis for PM Model\n")
pm_morans <- analyze_spatial_autocorrelation(
  pm_models, 
  pm_names, 
  all_data_inc$pm_change, 
  Locations, 
  k = 5
)
knitr::kable(pm_morans, caption = "Moran's I Results for PM Model")

# O3 model
cat("## Moran's I Analysis for O3 Model\n")
o3_morans <- analyze_spatial_autocorrelation(
  o3_models, 
  o3_names, 
  all_data_inc$o3_change, 
  Locations, 
  k = 5
)
knitr::kable(o3_morans, caption = "Moran's I Results for O3 Model")

# NO2 model
cat("## Moran's I Analysis for NO2 Model\n")
no2_morans <- analyze_spatial_autocorrelation(
  no2_models, 
  no2_names, 
  all_data_inc$no2_change, 
  Locations, 
  k = 5
)
knitr::kable(no2_morans, caption = "Moran's I Results for NO2 Model")

# NDVI model
cat("## Moran's I Analysis for NDVI Model\n")
ndvi_morans <- analyze_spatial_autocorrelation(
  ndvi_models, 
  ndvi_names, 
  all_data_inc$ndvi_change, 
  Locations, 
  k = 5
)
knitr::kable(ndvi_morans, caption = "Moran's I Results for NDVI Model")

# Temperature model
cat("## Moran's I Analysis for Temperature Model\n")
temp_morans <- analyze_spatial_autocorrelation(
  temp_models, 
  temp_names, 
  all_data_inc$mean_temp, 
  Locations, 
  k = 5
)
knitr::kable(temp_morans, caption = "Moran's I Results for Temperature Model")

# Create summary of all results
cat("## Combined Moran's I Results for All Models\n")
all_morans_summary <- create_morans_summary(
  list(pm_morans, o3_morans, no2_morans, ndvi_morans, temp_morans),
  c("PM", "O3", "NO2", "NDVI", "Temperature")
)
knitr::kable(all_morans_summary, caption = "Combined Moran's I Results")

# Get model results
cat("## Model Diagnostics\n")

# PM Model
cat("### PM Model Diagnostics\n")
plot_model_diagnostics(pm_models[[1]], pm_names[1])

# O3 Model
cat("### O3 Model Diagnostics\n")
plot_model_diagnostics(o3_models[[1]], o3_names[1])

# NO2 Model
cat("### NO2 Model Diagnostics\n")
plot_model_diagnostics(no2_models[[1]], no2_names[1])

# NDVI Model
cat("### NDVI Model Diagnostics\n")
plot_model_diagnostics(ndvi_models[[1]], ndvi_names[1])

# Temperature Model
cat("### Temperature Model Diagnostics\n")
plot_model_diagnostics(temp_models[[1]], temp_names[1])

# Create DIC comparison table
cat("## Model Comparison using DIC\n")
model_comparison <- data.frame(
  Model_Type = character(),
  Model = character(),
  DIC = numeric(),
  stringsAsFactors = FALSE
)

# Add PM model
if (!is.null(pm_models[[1]]) && !is.null(pm_models[[1]]$dic)) {
  model_comparison <- rbind(model_comparison, 
                           data.frame(
                             Model_Type = "PM",
                             Model = pm_names[1],
                             DIC = pm_models[[1]]$dic$dic,
                             stringsAsFactors = FALSE
                           ))
}

# Add O3 model
if (!is.null(o3_models[[1]]) && !is.null(o3_models[[1]]$dic)) {
  model_comparison <- rbind(model_comparison, 
                           data.frame(
                             Model_Type = "O3",
                             Model = o3_names[1],
                             DIC = o3_models[[1]]$dic$dic,
                             stringsAsFactors = FALSE
                           ))
}

# Add NO2 model
if (!is.null(no2_models[[1]]) && !is.null(no2_models[[1]]$dic)) {
  model_comparison <- rbind(model_comparison, 
                           data.frame(
                             Model_Type = "NO2",
                             Model = no2_names[1],
                             DIC = no2_models[[1]]$dic$dic,
                             stringsAsFactors = FALSE
                           ))
}

# Add NDVI model
if (!is.null(ndvi_models[[1]]) && !is.null(ndvi_models[[1]]$dic)) {
  model_comparison <- rbind(model_comparison, 
                           data.frame(
                             Model_Type = "NDVI",
                             Model = ndvi_names[1],
                             DIC = ndvi_models[[1]]$dic$dic,
                             stringsAsFactors = FALSE
                           ))
}

# Add Temperature model
if (!is.null(temp_models[[1]]) && !is.null(temp_models[[1]]$dic)) {
  model_comparison <- rbind(model_comparison, 
                           data.frame(
                             Model_Type = "Temperature",
                             Model = temp_names[1],
                             DIC = temp_models[[1]]$dic$dic,
                             stringsAsFactors = FALSE
                           ))
}

# Display model comparison table
knitr::kable(model_comparison, caption = "Model Comparison by DIC")
```
</details>

```{r}
```

